---
title: "UWMad 2024 Gutierrez B3K Genotyping — Quality Control"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 9
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(data.table)
library(ggplot2)

theme_set(theme_bw(base_size = 12))
```

---

## Configuration

```{r config}
# ── QC thresholds ─────────────────────────────────────────────────────────────
MISS_MK  <- 0.10   # max marker missing rate
MAF_MIN  <- 0.05   # min minor allele frequency
CR_SAMP  <- 0.80   # min sample call rate
HET_SAMP <- 0.15   # max sample observed heterozygosity (advanced RIL / near-inbred)

# ── File paths ─────────────────────────────────────────────────────────────────
orei_root <- if (dir.exists("OREI_Genotyping_X")) {
  "."
} else if (dir.exists("OREI Genotyping/OREI_Genotyping_X")) {
  "OREI Genotyping"
} else {
  stop("Cannot locate OREI_Genotyping_X. Run from 'OREI Genotyping/' or the project root.")
}

x_dir  <- file.path(orei_root, "OREI_Genotyping_X")
prefix <- "UWMad2024Gutierrez-B3K_01-06"
path   <- function(suffix) file.path(x_dir, paste0(prefix, suffix))

cat("Directory:", x_dir, "\n")
```

---

## 1  Data Import

### 1.1  Sample metadata

```{r import_samples}
samples_meta <- fread(path("-Samples.txt"),     sep = "\t", check.names = FALSE)
pass_meta    <- fread(path("-VCF-Samples.txt"), sep = "\t", check.names = FALSE)

# poorQC file is absent when all samples pass array QC
poorqc_file <- path("-poorQC_VCF-Samples.txt")
fail_meta <- if (file.exists(poorqc_file)) {
  dt <- fread(poorqc_file, sep = "\t", check.names = FALSE)
  dt[!is.na(`Sample ID`) & nzchar(`Sample ID`)]
} else {
  data.table(`Sample ID` = character())
}

samples_meta[, qc_pass := `Sample ID` %in% pass_meta[["Sample ID"]] &
                          !(`Sample ID` %in% fail_meta[["Sample ID"]])]
```

### 1.2  AB genotype calls → numeric matrix

```{r import_ab}
ab <- fread(path("-AB.txt"), sep = "\t", header = TRUE)
setnames(ab, names(ab)[1], "marker_id")

marker_map <- ab[, .(
  marker_id,
  chrom = sub("^chr", "", CHROM),   # normalise: chr1H → 1H
  pos   = as.integer(POS)
)]

sample_ids <- setdiff(names(ab), c("marker_id", "CHROM", "POS"))
char_mat   <- as.matrix(ab[, ..sample_ids])
rownames(char_mat) <- ab$marker_id

ab_to_int <- function(m) {
  out <- matrix(NA_integer_, nrow(m), ncol(m), dimnames = dimnames(m))
  out[which(m == "AA")] <- 0L
  out[which(m == "AB")] <- 1L
  out[which(m == "BB")] <- 2L
  out
}

geno_mat <- ab_to_int(char_mat)
```

### 1.3  Extract family metadata

Sample IDs follow the pattern `ParentA/ParentB-plantNumber`.
The `Sample Name` column in the metadata contains the cross name directly.

```{r family_metadata}
meta_idx   <- match(sample_ids, samples_meta[["Sample ID"]])
cross_name <- samples_meta[["Sample Name"]][meta_idx]

# Fallback: parse family from sample ID if Sample Name is missing
cross_name <- ifelse(
  !is.na(cross_name) & nzchar(cross_name),
  cross_name,
  sub("-[0-9]+$", "", sample_ids)
)

miss_sample    <- colMeans(is.na(geno_mat))
obs_het_sample <- colMeans(geno_mat == 1L, na.rm = TRUE)

sample_qc <- data.table(
  sample_id       = sample_ids,
  cross           = cross_name,
  plate           = samples_meta[["Sample Plate"]][meta_idx],
  call_rate_array = samples_meta[["Call Rate"]][meta_idx],
  qc_pass_array   = samples_meta[["qc_pass"]][meta_idx],
  call_rate_geno  = 1 - miss_sample,
  obs_het         = obs_het_sample
)

cat(sprintf("Samples: %d across %d crosses\n",
            nrow(sample_qc), length(unique(sample_qc$cross))))
```

### 1.4  Per-marker QC statistics

```{r compute_marker_qc}
miss_marker    <- rowMeans(is.na(geno_mat))
allele_freq    <- rowMeans(geno_mat, na.rm = TRUE) / 2
maf            <- pmin(allele_freq, 1 - allele_freq)
obs_het_marker <- rowMeans(geno_mat == 1L, na.rm = TRUE)

marker_qc <- data.table(
  marker_id = marker_map$marker_id,
  chrom     = marker_map$chrom,
  pos       = marker_map$pos,
  miss_rate = miss_marker,
  maf       = maf,
  obs_het   = obs_het_marker
)
```

---

## 2  Data Overview

```{r overview}
overview <- data.frame(
  Item  = c("Markers (raw)", "Chromosomes", "Samples (total)",
            "Crosses / families", "Samples passing array QC"),
  Count = c(
    nrow(marker_qc),
    length(unique(marker_map$chrom)),
    nrow(sample_qc),
    length(unique(sample_qc$cross)),
    sum(sample_qc$qc_pass_array, na.rm = TRUE)
  )
)
knitr::kable(overview, align = "lr")
```

Samples per cross:

```{r cross_counts}
cross_counts <- sample_qc[, .N, by = cross][order(-N)]
knitr::kable(cross_counts, col.names = c("Cross", "Samples"), align = "lr")
```

Markers per chromosome:

```{r chrom_counts}
chrom_counts <- marker_qc[, .N, by = chrom][order(chrom)]
knitr::kable(chrom_counts, col.names = c("Chromosome", "Markers"), align = "lr")
```

---

## 3  Sample QC

### 3.1  Call-rate distribution

```{r sample_callrate, fig.height=4}
ggplot(sample_qc, aes(call_rate_geno)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white", linewidth = 0.2) +
  geom_vline(xintercept = CR_SAMP, colour = "firebrick", linetype = "dashed") +
  annotate("text", x = CR_SAMP - 0.005, y = Inf,
           label = sprintf("threshold = %.2f", CR_SAMP),
           colour = "firebrick", hjust = 1, vjust = 1.5, size = 3.5) +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(x = "Call rate", y = "Samples", title = "Sample call-rate distribution")

cat(sprintf("Samples below %.0f%% call rate: %d / %d\n",
            CR_SAMP * 100, sum(sample_qc$call_rate_geno < CR_SAMP), nrow(sample_qc)))
```

### 3.2  Heterozygosity

```{r sample_het, fig.height=4}
ggplot(sample_qc, aes(obs_het)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white", linewidth = 0.2) +
  geom_vline(xintercept = HET_SAMP, colour = "firebrick", linetype = "dashed") +
  annotate("text", x = HET_SAMP + 0.002, y = Inf,
           label = sprintf("threshold = %.2f", HET_SAMP),
           colour = "firebrick", hjust = 0, vjust = 1.5, size = 3.5) +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(x = "Observed heterozygosity", y = "Samples",
       title = "Sample heterozygosity distribution")
```

### 3.3  Call rate vs heterozygosity – overall

```{r cr_vs_het}
sample_qc[, flag := fcase(
  call_rate_geno < CR_SAMP & obs_het > HET_SAMP, "fail: low CR + high het",
  call_rate_geno < CR_SAMP,                       "fail: low call rate",
  obs_het        > HET_SAMP,                      "fail: high heterozygosity",
  default = "pass"
)]

ggplot(sample_qc, aes(call_rate_geno, obs_het, colour = flag)) +
  geom_point(alpha = 0.55, size = 1.4) +
  scale_colour_manual(values = c(
    "pass"                      = "grey60",
    "fail: low call rate"       = "#E69F00",
    "fail: high heterozygosity" = "#56B4E9",
    "fail: low CR + high het"   = "firebrick"
  )) +
  scale_x_continuous(labels = scales::percent_format()) +
  scale_y_continuous(labels = scales::percent_format()) +
  geom_vline(xintercept = CR_SAMP,  linetype = "dashed", colour = "grey40") +
  geom_hline(yintercept = HET_SAMP, linetype = "dashed", colour = "grey40") +
  labs(x = "Call rate", y = "Observed heterozygosity",
       colour = "QC status", title = "Sample QC: call rate vs heterozygosity")
```

### 3.4  Per-family call rate

```{r family_callrate, fig.height=max(4, length(unique(sample_qc$cross)) * 0.35)}
family_qc <- sample_qc[, .(
  n_samples      = .N,
  mean_cr        = mean(call_rate_geno, na.rm = TRUE),
  mean_het       = mean(obs_het,        na.rm = TRUE),
  n_fail_cr      = sum(call_rate_geno < CR_SAMP),
  n_fail_het     = sum(obs_het > HET_SAMP)
), by = cross]

ggplot(sample_qc, aes(y = reorder(cross, call_rate_geno, median),
                       x = call_rate_geno)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6, outlier.size = 1) +
  geom_vline(xintercept = CR_SAMP, colour = "firebrick", linetype = "dashed") +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(x = "Call rate", y = "Cross", title = "Call rate by cross (sorted by median)")
```

### 3.5  Per-family heterozygosity

```{r family_het, fig.height=max(4, length(unique(sample_qc$cross)) * 0.35)}
ggplot(sample_qc, aes(y = reorder(cross, obs_het, median), x = obs_het)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6, outlier.size = 1) +
  geom_vline(xintercept = HET_SAMP, colour = "firebrick", linetype = "dashed") +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(x = "Observed heterozygosity", y = "Cross",
       title = "Heterozygosity by cross (sorted by median)")
```

### 3.6  Per-family summary table

```{r family_table}
family_qc[, mean_cr  := round(mean_cr,  4)]
family_qc[, mean_het := round(mean_het, 4)]
setorder(family_qc, cross)
knitr::kable(family_qc,
             col.names = c("Cross", "N", "Mean call rate",
                           "Mean het", "Fail CR", "Fail het"),
             align = "lrrrll")
```

### 3.7  Flagged samples

```{r flagged_samples}
flagged <- sample_qc[flag != "pass",
                     .(sample_id, cross, plate, call_rate_geno, obs_het, flag)]
flagged[, call_rate_geno := round(call_rate_geno, 4)]
flagged[, obs_het        := round(obs_het,        4)]
setorder(flagged, cross, flag)

cat(sprintf("Total flagged samples: %d / %d\n", nrow(flagged), nrow(sample_qc)))
if (nrow(flagged) > 0) {
  knitr::kable(flagged,
               col.names = c("Sample ID", "Cross", "Plate", "Call rate",
                             "Obs. het", "Flag"),
               align = "lllrrl")
}
```

---

## 4  Marker QC

### 4.1  Missing rate

```{r marker_miss, fig.height=4}
ggplot(marker_qc, aes(miss_rate)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white", linewidth = 0.2) +
  geom_vline(xintercept = MISS_MK, colour = "firebrick", linetype = "dashed") +
  annotate("text", x = MISS_MK + 0.003, y = Inf,
           label = sprintf("threshold = %.2f", MISS_MK),
           colour = "firebrick", hjust = 0, vjust = 1.5, size = 3.5) +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(x = "Missing rate", y = "Markers", title = "Marker missing-rate distribution")

cat(sprintf("Markers above %.0f%% missing: %d / %d\n",
            MISS_MK * 100, sum(marker_qc$miss_rate > MISS_MK), nrow(marker_qc)))
```

### 4.2  Minor allele frequency

```{r marker_maf, fig.height=4}
ggplot(marker_qc, aes(maf)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white", linewidth = 0.2) +
  geom_vline(xintercept = MAF_MIN, colour = "firebrick", linetype = "dashed") +
  annotate("text", x = MAF_MIN + 0.003, y = Inf,
           label = sprintf("threshold = %.2f", MAF_MIN),
           colour = "firebrick", hjust = 0, vjust = 1.5, size = 3.5) +
  labs(x = "Minor allele frequency", y = "Markers", title = "MAF distribution")

cat(sprintf("Monomorphic (MAF = 0): %d\n", sum(marker_qc$maf == 0, na.rm = TRUE)))
cat(sprintf("Markers MAF < %.2f  : %d / %d\n",
            MAF_MIN, sum(marker_qc$maf < MAF_MIN, na.rm = TRUE), nrow(marker_qc)))
```

### 4.3  Marker heterozygosity

```{r marker_het, fig.height=4}
ggplot(marker_qc, aes(obs_het)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white", linewidth = 0.2) +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(x = "Observed heterozygosity", y = "Markers",
       title = "Marker heterozygosity distribution")
```

### 4.4  Chromosomal distribution

```{r chrom_dist, fig.height=4}
ggplot(marker_qc[!is.na(pos)], aes(pos / 1e6, fill = chrom)) +
  geom_histogram(bins = 50, colour = "white", linewidth = 0.1) +
  facet_wrap(~chrom, nrow = 1) +
  scale_fill_brewer(palette = "Set1", guide = "none") +
  labs(x = "Physical position (Mb)", y = "Markers",
       title = "Marker distribution across chromosomes (Morex V3)")
```

---

## 5  QC Filtering

```{r apply_filters}
pass_samples <- sample_qc[call_rate_geno >= CR_SAMP & obs_het <= HET_SAMP, sample_id]
pass_markers <- marker_qc[miss_rate <= MISS_MK & maf >= MAF_MIN,           marker_id]

geno_filt <- geno_mat[pass_markers, pass_samples]

filter_summary <- data.frame(
  Step  = c("Input markers", "After miss-rate filter", "After MAF filter",
            "Final markers", "Input samples", "After call-rate filter",
            "After het filter", "Final samples"),
  Count = c(
    nrow(marker_qc),
    sum(marker_qc$miss_rate <= MISS_MK),
    sum(marker_qc$miss_rate <= MISS_MK & marker_qc$maf >= MAF_MIN),
    length(pass_markers),
    nrow(sample_qc),
    sum(sample_qc$call_rate_geno >= CR_SAMP),
    sum(sample_qc$call_rate_geno >= CR_SAMP & sample_qc$obs_het <= HET_SAMP),
    length(pass_samples)
  )
)
knitr::kable(filter_summary, align = "lr")
```

Samples retained per cross after filtering:

```{r filter_by_cross}
retained_cross <- sample_qc[sample_id %in% pass_samples, .N, by = cross][order(-N)]
knitr::kable(retained_cross, col.names = c("Cross", "Samples retained"), align = "lr")
```

---

## 6  PCA

```{r pca_compute}
impute_mean <- function(mat) {
  row_means <- rowMeans(mat, na.rm = TRUE)
  idx       <- which(is.na(mat), arr.ind = TRUE)
  mat[idx]  <- row_means[idx[, 1]]
  mat
}

geno_imp <- impute_mean(geno_filt)
pca_res  <- prcomp(t(geno_imp), center = TRUE, scale. = FALSE)
pve      <- pca_res$sdev^2 / sum(pca_res$sdev^2) * 100

pca_dt <- as.data.table(pca_res$x[, 1:10])
pca_dt[, sample_id := rownames(pca_res$x)]
pca_dt <- merge(pca_dt, sample_qc[, .(sample_id, cross, call_rate_geno)],
                by = "sample_id")

cross_names   <- sort(unique(pca_dt$cross))
n_crosses     <- length(cross_names)
cross_palette <- setNames(scales::hue_pal()(n_crosses), cross_names)
```

### 6.1  Scree plot

```{r scree, fig.height=4, fig.width=6}
data.table(PC = 1:20, PVE = pve[1:20]) |>
  ggplot(aes(PC, PVE)) +
  geom_col(fill = "steelblue", width = 0.7) +
  geom_line(aes(group = 1), colour = "grey30") +
  geom_point(colour = "grey30") +
  labs(x = "Principal component", y = "Variance explained (%)",
       title = "Scree plot – top 20 PCs")
```

### 6.2  PC1 vs PC2 – by cross

```{r pca_pc12}
ggplot(pca_dt, aes(PC1, PC2, colour = cross)) +
  geom_point(alpha = 0.65, size = 1.6) +
  scale_colour_manual(values = cross_palette) +
  labs(x = sprintf("PC1 (%.1f%%)", pve[1]),
       y = sprintf("PC2 (%.1f%%)", pve[2]),
       colour = "Cross", title = "PCA – PC1 vs PC2 coloured by cross") +
  theme(legend.position = "right", legend.text = element_text(size = 7))
```

### 6.3  PC1 vs PC3 – by cross

```{r pca_pc13}
ggplot(pca_dt, aes(PC1, PC3, colour = cross)) +
  geom_point(alpha = 0.65, size = 1.6) +
  scale_colour_manual(values = cross_palette) +
  labs(x = sprintf("PC1 (%.1f%%)", pve[1]),
       y = sprintf("PC3 (%.1f%%)", pve[3]),
       colour = "Cross", title = "PCA – PC1 vs PC3") +
  theme(legend.position = "right", legend.text = element_text(size = 7))
```

### 6.4  PC1 vs PC2 – by call rate

```{r pca_callrate}
ggplot(pca_dt, aes(PC1, PC2, colour = call_rate_geno)) +
  geom_point(alpha = 0.65, size = 1.6) +
  scale_colour_viridis_c(option = "plasma", labels = scales::percent_format(),
                         name = "Call rate") +
  labs(x = sprintf("PC1 (%.1f%%)", pve[1]),
       y = sprintf("PC2 (%.1f%%)", pve[2]),
       title = "PCA – call rate overlay")
```

---

## 7  Linkage Disequilibrium

Pairwise r² within each chromosome using QC-filtered markers and samples.

```{r ld_compute, cache=TRUE}
marker_filt <- marker_qc[marker_id %in% pass_markers & !is.na(pos)]

ld_all <- rbindlist(lapply(sort(unique(marker_filt$chrom)), function(chr) {
  mk_chr <- marker_filt[chrom == chr]
  if (nrow(mk_chr) < 2) return(NULL)

  m   <- geno_filt[mk_chr$marker_id, , drop = FALSE]
  pos <- mk_chr$pos

  r2_mat <- cor(t(m), use = "pairwise.complete.obs")^2
  upper  <- upper.tri(r2_mat)
  idx    <- which(upper, arr.ind = TRUE)

  data.table(
    chrom   = chr,
    dist_bp = abs(pos[idx[, 2]] - pos[idx[, 1]]),
    r2      = r2_mat[upper]
  )
}))

ld_all[, dist_bin := floor(dist_bp / 2e6) * 2]
ld_bin <- ld_all[, .(
  med_r2  = median(r2, na.rm = TRUE),
  mean_r2 = mean(r2,   na.rm = TRUE),
  n_pairs = .N
), by = .(chrom, dist_bin)]
```

### 7.1  LD decay – all chromosomes

```{r ld_decay_all, fig.height=5}
ggplot(ld_bin[dist_bin <= 200], aes(dist_bin, med_r2, colour = chrom, group = chrom)) +
  geom_line(linewidth = 0.8, alpha = 0.85) +
  geom_point(size = 1.2, alpha = 0.7) +
  scale_colour_brewer(palette = "Set1", name = "Chromosome") +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Physical distance (Mb)", y = expression(paste("Median ", r^2)),
       title = "LD decay by chromosome (2 Mb bins)")
```

### 7.2  LD decay – faceted by chromosome

```{r ld_decay_facet, fig.height=6}
ggplot(ld_bin[dist_bin <= 200], aes(dist_bin, med_r2)) +
  geom_line(colour = "steelblue", linewidth = 0.8) +
  geom_point(colour = "steelblue", size = 1.0) +
  facet_wrap(~chrom, nrow = 2) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Physical distance (Mb)", y = expression(paste("Median ", r^2)),
       title = "LD decay per chromosome")
```

### 7.3  Distance at r² = 0.2

```{r ld_decay_point}
r2_thr <- 0.2

ld_decay_pt <- ld_bin[, {
  above <- .SD[med_r2 >= r2_thr]
  .(decay_dist_mb = if (nrow(above) == 0) NA_real_ else max(above$dist_bin) + 2)
}, by = chrom]

setorder(ld_decay_pt, chrom)
knitr::kable(ld_decay_pt,
             col.names = c("Chromosome", sprintf("Distance r² ≥ %.1f (Mb)", r2_thr)),
             align = "lr")
```

---

## 8  Save Outputs

```{r save_outputs}
marker_map_filt <- marker_map[marker_id %in% pass_markers]
sample_qc_filt  <- sample_qc[sample_id %in% pass_samples]

# ── .Rdata: numeric matrix + map + QC tables ──────────────────────────────────
rdata_path <- file.path(orei_root, "OREI_X_geno.Rdata")
save(geno_filt, marker_map_filt, marker_qc, sample_qc, file = rdata_path)
cat("Saved:", rdata_path, "\n")

# ── Filtered HapMap (A/B notation) ────────────────────────────────────────────
calls_char <- matrix("NN", nrow(geno_filt), ncol(geno_filt),
                     dimnames = dimnames(geno_filt))
calls_char[which(geno_filt == 0L)] <- "AA"
calls_char[which(geno_filt == 1L)] <- "AB"
calls_char[which(geno_filt == 2L)] <- "BB"

hmp_out <- data.table(
  `rs#`       = marker_map_filt$marker_id,
  alleles     = "A/B",
  chrom       = marker_map_filt$chrom,
  pos         = marker_map_filt$pos,
  strand      = "+",
  `assembly#` = "MorexV3",
  center      = "NA",
  protLSID    = "NA",
  assayLSID   = "NA",
  panel       = "NA",
  QCcode      = "NA"
)
hmp_out <- cbind(hmp_out, as.data.table(calls_char))

hmp_path <- file.path(orei_root, "OREI_X_hapmap_filtered.txt")
fwrite(hmp_out, hmp_path, sep = "\t", quote = FALSE)
cat("Saved:", hmp_path, "\n")

cat(sprintf("\nFinal dataset: %d markers × %d samples\n",
            nrow(geno_filt), ncol(geno_filt)))
cross_tbl <- sample_qc_filt[, .N, by = cross][order(-N)]
cat(sprintf("Crosses retained: %d\n", nrow(cross_tbl)))
print(cross_tbl)
```

---

## 9  Session Info

```{r session_info}
sessionInfo()
```
