---
title: "OREI MNS2025 50K Genotyping — Quality Control"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 9
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(data.table)
library(ggplot2)

theme_set(theme_bw(base_size = 12))
```

---

## Configuration

```{r config}
# ── QC thresholds ─────────────────────────────────────────────────────────────
MISS_MK  <- 0.10   # max marker missing rate
MAF_MIN  <- 0.05   # min minor allele frequency
CR_SAMP  <- 0.80   # min sample call rate
HET_SAMP <- 0.15   # max sample observed heterozygosity

# Valid barley chromosomes (Morex V3 naming, no prefix)
VALID_CHROMS <- paste0(1:7, "H")

# Max markers per chromosome used for LD computation (subsampled if exceeded)
MAX_LD_MARKERS_PER_CHR <- 1000

# ── File paths ─────────────────────────────────────────────────────────────────
orei_root <- if (dir.exists("OREI_Genotyping_50K")) {
  "."
} else if (dir.exists("OREI Genotyping/OREI_Genotyping_50K")) {
  "OREI Genotyping"
} else {
  stop("Cannot locate OREI_Genotyping_50K. Run from 'OREI Genotyping/' or the project root.")
}

hmp_file <- file.path(orei_root, "OREI_Genotyping_50K",
                      "orei_mns2025_50K_hapmap_050726.txt")
stopifnot(file.exists(hmp_file))
cat("HapMap file:", hmp_file, "\n")
```

---

## 1  Data Import

### 1.1  Read HapMap

The 50K array exports genotype calls in HapMap format with nucleotide-coded calls
(e.g. `AA`, `CC`, `AC`). Columns 1–11 are marker metadata; columns 12+ are samples.

```{r import_hapmap}
hmp <- fread(hmp_file, sep = "\t", header = TRUE, check.names = FALSE)

META_COLS <- c("rs#", "alleles", "chrom", "pos", "strand",
               "assembly#", "center", "protLSID", "assayLSID", "panel", "QCcode")

# Identify sample columns (everything after the 11 metadata columns)
sample_ids <- setdiff(names(hmp), META_COLS)
cat(sprintf("Markers (raw): %d\n", nrow(hmp)))
cat(sprintf("Samples       : %d\n", length(sample_ids)))
```

### 1.2  Marker map

```{r marker_map}
marker_map <- hmp[, .(
  marker_id = `rs#`,
  alleles,
  chrom     = chrom,          # already "1H"–"7H" from TASSEL export
  pos       = as.integer(pos)
)]

# Drop markers on unassigned/unknown chromosomes
n_unmapped <- sum(!marker_map$chrom %in% VALID_CHROMS)
if (n_unmapped > 0)
  cat(sprintf("Removing %d markers on non-canonical chromosomes.\n", n_unmapped))

marker_map <- marker_map[chrom %in% VALID_CHROMS]
hmp        <- hmp[`rs#` %in% marker_map$marker_id]
```

### 1.3  Encode nucleotide calls → 0/1/2/NA

The `alleles` column specifies the two variant alleles (e.g., `A/C`).
Encoding: homozygous allele-1 = 0, heterozygous = 1, homozygous allele-2 = 2, `NC`/`NN` = NA.

```{r encode_numeric}
call_mat <- as.matrix(hmp[, ..sample_ids])
rownames(call_mat) <- hmp[["rs#"]]

allele1 <- sub("/.*", "", marker_map$alleles)
allele2 <- sub(".*/", "", marker_map$alleles)

hom1    <- paste0(allele1, allele1)
hom2    <- paste0(allele2, allele2)
het_ab  <- paste0(allele1, allele2)
het_ba  <- paste0(allele2, allele1)

geno_mat <- matrix(NA_integer_, nrow(call_mat), ncol(call_mat),
                   dimnames = dimnames(call_mat))

for (i in seq_len(nrow(call_mat))) {
  x <- call_mat[i, ]
  geno_mat[i, x == hom1[i]]                     <- 0L
  geno_mat[i, x %in% c(het_ab[i], het_ba[i])]   <- 1L
  geno_mat[i, x == hom2[i]]                     <- 2L
  # NC / NN / anything else remains NA
}
```

### 1.4  Per-marker and per-sample QC statistics

```{r compute_qc}
miss_marker    <- rowMeans(is.na(geno_mat))
allele_freq    <- rowMeans(geno_mat, na.rm = TRUE) / 2
maf            <- pmin(allele_freq, 1 - allele_freq)
obs_het_marker <- rowMeans(geno_mat == 1L, na.rm = TRUE)

marker_qc <- data.table(
  marker_id = marker_map$marker_id,
  chrom     = marker_map$chrom,
  pos       = marker_map$pos,
  miss_rate = miss_marker,
  maf       = maf,
  obs_het   = obs_het_marker
)

miss_sample    <- colMeans(is.na(geno_mat))
obs_het_sample <- colMeans(geno_mat == 1L, na.rm = TRUE)

# Label samples as parents/checks or NAM lines
sample_type <- ifelse(grepl("^25BS-NAM-", sample_ids), "NAM line", "Parent / check")

sample_qc <- data.table(
  sample_id      = sample_ids,
  sample_type    = sample_type,
  call_rate_geno = 1 - miss_sample,
  obs_het        = obs_het_sample
)
```

---

## 2  Data Overview

```{r overview}
overview <- data.frame(
  Item  = c("Markers (raw)", "Chromosomes", "Samples (total)",
            "  NAM lines", "  Parents / checks"),
  Count = c(
    nrow(hmp),
    length(unique(marker_map$chrom)),
    nrow(sample_qc),
    sum(sample_qc$sample_type == "NAM line"),
    sum(sample_qc$sample_type == "Parent / check")
  )
)
knitr::kable(overview, align = "lr")
```

Markers per chromosome:

```{r chrom_counts}
chrom_counts <- marker_qc[, .N, by = chrom][order(chrom)]
knitr::kable(chrom_counts, col.names = c("Chromosome", "Markers"), align = "lr")
```

---

## 3  Sample QC

### 3.1  Call-rate distribution

```{r sample_callrate, fig.height=4}
ggplot(sample_qc, aes(call_rate_geno, fill = sample_type)) +
  geom_histogram(bins = 60, colour = "white", linewidth = 0.2, alpha = 0.8,
                 position = "stack") +
  geom_vline(xintercept = CR_SAMP, colour = "firebrick", linetype = "dashed") +
  annotate("text", x = CR_SAMP - 0.005, y = Inf,
           label = sprintf("threshold = %.2f", CR_SAMP),
           colour = "firebrick", hjust = 1, vjust = 1.5, size = 3.5) +
  scale_x_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c("NAM line" = "steelblue",
                                "Parent / check" = "#E69F00")) +
  labs(x = "Call rate", y = "Samples", fill = NULL,
       title = "Sample call-rate distribution")

cat(sprintf("Samples below %.0f%% call rate: %d / %d\n",
            CR_SAMP * 100, sum(sample_qc$call_rate_geno < CR_SAMP), nrow(sample_qc)))
```

### 3.2  Heterozygosity

```{r sample_het, fig.height=4}
ggplot(sample_qc, aes(obs_het, fill = sample_type)) +
  geom_histogram(bins = 60, colour = "white", linewidth = 0.2, alpha = 0.8,
                 position = "stack") +
  geom_vline(xintercept = HET_SAMP, colour = "firebrick", linetype = "dashed") +
  annotate("text", x = HET_SAMP + 0.002, y = Inf,
           label = sprintf("threshold = %.2f", HET_SAMP),
           colour = "firebrick", hjust = 0, vjust = 1.5, size = 3.5) +
  scale_x_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c("NAM line" = "steelblue",
                                "Parent / check" = "#E69F00")) +
  labs(x = "Observed heterozygosity", y = "Samples", fill = NULL,
       title = "Sample heterozygosity distribution")
```

### 3.3  Call rate vs heterozygosity

```{r cr_vs_het}
sample_qc[, flag := fcase(
  call_rate_geno < CR_SAMP & obs_het > HET_SAMP, "fail: low CR + high het",
  call_rate_geno < CR_SAMP,                       "fail: low call rate",
  obs_het        > HET_SAMP,                      "fail: high heterozygosity",
  default = "pass"
)]

ggplot(sample_qc, aes(call_rate_geno, obs_het, colour = flag, shape = sample_type)) +
  geom_point(alpha = 0.6, size = 2) +
  scale_colour_manual(values = c(
    "pass"                      = "grey60",
    "fail: low call rate"       = "#E69F00",
    "fail: high heterozygosity" = "#56B4E9",
    "fail: low CR + high het"   = "firebrick"
  )) +
  scale_shape_manual(values = c("NAM line" = 16, "Parent / check" = 17)) +
  scale_x_continuous(labels = scales::percent_format()) +
  scale_y_continuous(labels = scales::percent_format()) +
  geom_vline(xintercept = CR_SAMP,  linetype = "dashed", colour = "grey40") +
  geom_hline(yintercept = HET_SAMP, linetype = "dashed", colour = "grey40") +
  labs(x = "Call rate", y = "Observed heterozygosity",
       colour = "QC status", shape = "Sample type",
       title = "Sample QC: call rate vs heterozygosity")
```

### 3.4  Flagged samples

```{r flagged_samples}
flagged <- sample_qc[flag != "pass",
                     .(sample_id, sample_type, call_rate_geno, obs_het, flag)]
flagged[, call_rate_geno := round(call_rate_geno, 4)]
flagged[, obs_het        := round(obs_het,        4)]
setorder(flagged, flag, call_rate_geno)

cat(sprintf("Total flagged samples: %d / %d\n", nrow(flagged), nrow(sample_qc)))
if (nrow(flagged) > 0) {
  knitr::kable(flagged, col.names = c("Sample ID", "Type", "Call rate",
                                      "Obs. het", "Flag"), align = "lllrrl")
}
```

---

## 4  Marker QC

### 4.1  Missing rate

```{r marker_miss, fig.height=4}
ggplot(marker_qc, aes(miss_rate)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white", linewidth = 0.2) +
  geom_vline(xintercept = MISS_MK, colour = "firebrick", linetype = "dashed") +
  annotate("text", x = MISS_MK + 0.003, y = Inf,
           label = sprintf("threshold = %.2f", MISS_MK),
           colour = "firebrick", hjust = 0, vjust = 1.5, size = 3.5) +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(x = "Missing rate", y = "Markers", title = "Marker missing-rate distribution")

cat(sprintf("Markers above %.0f%% missing: %d / %d\n",
            MISS_MK * 100, sum(marker_qc$miss_rate > MISS_MK), nrow(marker_qc)))
```

### 4.2  Minor allele frequency

```{r marker_maf, fig.height=4}
ggplot(marker_qc, aes(maf)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white", linewidth = 0.2) +
  geom_vline(xintercept = MAF_MIN, colour = "firebrick", linetype = "dashed") +
  annotate("text", x = MAF_MIN + 0.003, y = Inf,
           label = sprintf("threshold = %.2f", MAF_MIN),
           colour = "firebrick", hjust = 0, vjust = 1.5, size = 3.5) +
  labs(x = "Minor allele frequency", y = "Markers", title = "MAF distribution")

cat(sprintf("Monomorphic (MAF = 0): %d\n", sum(marker_qc$maf == 0, na.rm = TRUE)))
cat(sprintf("Markers MAF < %.2f  : %d / %d\n",
            MAF_MIN, sum(marker_qc$maf < MAF_MIN, na.rm = TRUE), nrow(marker_qc)))
```

### 4.3  Marker heterozygosity

```{r marker_het, fig.height=4}
ggplot(marker_qc, aes(obs_het)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white", linewidth = 0.2) +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(x = "Observed heterozygosity", y = "Markers",
       title = "Marker heterozygosity distribution")
```

### 4.4  Chromosomal distribution

```{r chrom_dist, fig.height=4}
ggplot(marker_qc[!is.na(pos)], aes(pos / 1e6, fill = chrom)) +
  geom_histogram(bins = 50, colour = "white", linewidth = 0.1) +
  facet_wrap(~chrom, nrow = 1) +
  scale_fill_brewer(palette = "Set1", guide = "none") +
  labs(x = "Physical position (Mb)", y = "Markers",
       title = "Marker distribution across chromosomes (Morex V3)")
```

---

## 5  QC Filtering

```{r apply_filters}
pass_samples <- sample_qc[call_rate_geno >= CR_SAMP & obs_het <= HET_SAMP, sample_id]
pass_markers <- marker_qc[miss_rate <= MISS_MK & maf >= MAF_MIN,           marker_id]

geno_filt <- geno_mat[pass_markers, pass_samples]

filter_summary <- data.frame(
  Step  = c("Input markers", "After miss-rate filter", "After MAF filter",
            "Final markers", "Input samples", "After call-rate filter",
            "After het filter", "Final samples"),
  Count = c(
    nrow(marker_qc),
    sum(marker_qc$miss_rate <= MISS_MK),
    sum(marker_qc$miss_rate <= MISS_MK & marker_qc$maf >= MAF_MIN),
    length(pass_markers),
    nrow(sample_qc),
    sum(sample_qc$call_rate_geno >= CR_SAMP),
    sum(sample_qc$call_rate_geno >= CR_SAMP & sample_qc$obs_het <= HET_SAMP),
    length(pass_samples)
  )
)
knitr::kable(filter_summary, align = "lr")
```

---

## 6  PCA

```{r pca_compute}
impute_mean <- function(mat) {
  row_means <- rowMeans(mat, na.rm = TRUE)
  idx       <- which(is.na(mat), arr.ind = TRUE)
  mat[idx]  <- row_means[idx[, 1]]
  mat
}

geno_imp <- impute_mean(geno_filt)
pca_res  <- prcomp(t(geno_imp), center = TRUE, scale. = FALSE)
pve      <- pca_res$sdev^2 / sum(pca_res$sdev^2) * 100

pca_dt <- as.data.table(pca_res$x[, 1:10])
pca_dt[, sample_id := rownames(pca_res$x)]
pca_dt <- merge(pca_dt, sample_qc[, .(sample_id, sample_type, call_rate_geno)],
                by = "sample_id")
```

### 6.1  Scree plot

```{r scree, fig.height=4, fig.width=6}
data.table(PC = 1:20, PVE = pve[1:20]) |>
  ggplot(aes(PC, PVE)) +
  geom_col(fill = "steelblue", width = 0.7) +
  geom_line(aes(group = 1), colour = "grey30") +
  geom_point(colour = "grey30") +
  labs(x = "Principal component", y = "Variance explained (%)",
       title = "Scree plot – top 20 PCs")
```

### 6.2  PC1 vs PC2 – parents vs NAM lines

```{r pca_pc12}
ggplot(pca_dt, aes(PC1, PC2, colour = sample_type, shape = sample_type)) +
  geom_point(data = pca_dt[sample_type == "NAM line"],
             alpha = 0.5, size = 1.4) +
  geom_point(data = pca_dt[sample_type == "Parent / check"],
             alpha = 1.0, size = 3) +
  scale_colour_manual(values = c("NAM line" = "steelblue",
                                  "Parent / check" = "firebrick")) +
  scale_shape_manual(values = c("NAM line" = 16, "Parent / check" = 17)) +
  labs(x = sprintf("PC1 (%.1f%%)", pve[1]),
       y = sprintf("PC2 (%.1f%%)", pve[2]),
       colour = NULL, shape = NULL,
       title = "PCA – PC1 vs PC2: parents highlighted")
```

### 6.3  PC1 vs PC3

```{r pca_pc13}
ggplot(pca_dt, aes(PC1, PC3, colour = sample_type, shape = sample_type)) +
  geom_point(data = pca_dt[sample_type == "NAM line"],  alpha = 0.5, size = 1.4) +
  geom_point(data = pca_dt[sample_type == "Parent / check"], alpha = 1.0, size = 3) +
  scale_colour_manual(values = c("NAM line" = "steelblue",
                                  "Parent / check" = "firebrick")) +
  scale_shape_manual(values = c("NAM line" = 16, "Parent / check" = 17)) +
  labs(x = sprintf("PC1 (%.1f%%)", pve[1]),
       y = sprintf("PC3 (%.1f%%)", pve[3]),
       colour = NULL, shape = NULL, title = "PCA – PC1 vs PC3")
```

### 6.4  PC1 vs PC2 – by call rate

```{r pca_callrate}
ggplot(pca_dt, aes(PC1, PC2, colour = call_rate_geno)) +
  geom_point(alpha = 0.65, size = 1.6) +
  scale_colour_viridis_c(option = "plasma", labels = scales::percent_format(),
                         name = "Call rate") +
  labs(x = sprintf("PC1 (%.1f%%)", pve[1]),
       y = sprintf("PC2 (%.1f%%)", pve[2]),
       title = "PCA – call rate overlay")
```

---

## 7  Linkage Disequilibrium

Pairwise r² within each chromosome. Subsampled to `MAX_LD_MARKERS_PER_CHR` per
chromosome if the marker count exceeds this limit to keep computation tractable.

```{r ld_compute, cache=TRUE}
set.seed(42)
marker_filt <- marker_qc[marker_id %in% pass_markers & !is.na(pos)]

ld_all <- rbindlist(lapply(sort(unique(marker_filt$chrom)), function(chr) {
  mk_chr <- marker_filt[chrom == chr]
  if (nrow(mk_chr) < 2) return(NULL)

  # Subsample if too many markers
  if (nrow(mk_chr) > MAX_LD_MARKERS_PER_CHR) {
    mk_chr <- mk_chr[sample(.N, MAX_LD_MARKERS_PER_CHR)]
    mk_chr <- mk_chr[order(pos)]
  }

  m   <- geno_filt[mk_chr$marker_id, , drop = FALSE]
  pos <- mk_chr$pos

  r2_mat <- cor(t(m), use = "pairwise.complete.obs")^2
  upper  <- upper.tri(r2_mat)
  idx    <- which(upper, arr.ind = TRUE)

  data.table(
    chrom   = chr,
    dist_bp = abs(pos[idx[, 2]] - pos[idx[, 1]]),
    r2      = r2_mat[upper]
  )
}))

ld_all[, dist_bin := floor(dist_bp / 2e6) * 2]
ld_bin <- ld_all[, .(
  med_r2  = median(r2, na.rm = TRUE),
  mean_r2 = mean(r2,   na.rm = TRUE),
  n_pairs = .N
), by = .(chrom, dist_bin)]
```

### 7.1  LD decay – all chromosomes

```{r ld_decay_all, fig.height=5}
ggplot(ld_bin[dist_bin <= 500], aes(dist_bin, med_r2, colour = chrom, group = chrom)) +
  geom_line(linewidth = 0.8, alpha = 0.85) +
  geom_point(size = 1.2, alpha = 0.7) +
  scale_colour_brewer(palette = "Set1", name = "Chromosome") +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Physical distance (Mb)", y = expression(paste("Median ", r^2)),
       title = "LD decay by chromosome (2 Mb bins)")
```

### 7.2  LD decay – faceted by chromosome

```{r ld_decay_facet, fig.height=6}
ggplot(ld_bin[dist_bin <= 500], aes(dist_bin, med_r2)) +
  geom_line(colour = "steelblue", linewidth = 0.8) +
  geom_point(colour = "steelblue", size = 1.0) +
  facet_wrap(~chrom, nrow = 2) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Physical distance (Mb)", y = expression(paste("Median ", r^2)),
       title = "LD decay per chromosome")
```

### 7.3  Distance at r² = 0.2

```{r ld_decay_point}
r2_thr <- 0.2

ld_decay_pt <- ld_bin[, {
  above <- .SD[med_r2 >= r2_thr]
  .(decay_dist_mb = if (nrow(above) == 0) NA_real_ else max(above$dist_bin) + 2)
}, by = chrom]

setorder(ld_decay_pt, chrom)
knitr::kable(ld_decay_pt,
             col.names = c("Chromosome", sprintf("Distance r² ≥ %.1f (Mb)", r2_thr)),
             align = "lr")
```

---

## 8  Save Outputs

```{r save_outputs}
marker_map_filt <- marker_map[marker_id %in% pass_markers]
sample_qc_filt  <- sample_qc[sample_id %in% pass_samples]

# ── .Rdata: numeric matrix + map + QC tables ──────────────────────────────────
rdata_path <- file.path(orei_root, "OREI_50K_geno.Rdata")
save(geno_filt, marker_map_filt, marker_qc, sample_qc, file = rdata_path)
cat("Saved:", rdata_path, "\n")

# ── Filtered HapMap (nucleotide calls preserved from original) ─────────────────
hmp_filt <- hmp[`rs#` %in% pass_markers, c(META_COLS, pass_samples), with = FALSE]
# Normalise chromosome column to match other platforms ("1H" format)
hmp_filt[, chrom := sub("^chr", "", chrom)]

hmp_path <- file.path(orei_root, "OREI_50K_hapmap_filtered.txt")
fwrite(hmp_filt, hmp_path, sep = "\t", quote = FALSE)
cat("Saved:", hmp_path, "\n")

cat(sprintf("\nFinal dataset: %d markers × %d samples\n",
            nrow(geno_filt), ncol(geno_filt)))
cat(sprintf("  NAM lines  : %d\n", sum(sample_qc_filt$sample_type == "NAM line")))
cat(sprintf("  Parents    : %d\n", sum(sample_qc_filt$sample_type == "Parent / check")))
```

---

## 9  Session Info

```{r session_info}
sessionInfo()
```
