---
title: "Solving Mixed Models"
author: "Kelly"
date: "2025-04-11"
output: html_document
---

# Setting up and solving mixed model equations.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS)

library(matlib)
```

### Technical Replication

We have discussed technical replication as an example of modeling covariance sturctures using mixed models. In this lab we will set up and solve mixed model equations that account for covariance between reepeated measures in the **R** covariance structure and using a mixed model with observation until as a random effect.


***Simulating the data**
```{r simulation}
set.seed(0)
#Setting the two treatment effects
trt=c(3,7)
# Setting up the residual covariance structure
Rb=matrix(c(1,.5,.5,.5,1,.5,.5,.5,1),3,3)
R=matrix(0,12,12)
R[1:3,1:3]=Rb
R[4:6,4:6]=Rb
R[7:9,7:9]=Rb
R[10:12,10:12]=Rb
heatmap(R)
#setting the mean to 0 and sampling residuals from a multivariate normal distribution
mu=rep(0,12)
res=MASS::mvrnorm(1,mu,R)

#We are simulating 12 phenotypes with the first 6 associated with treatment 1 and the other 6 associated with treatment 2
trt_effect=rep(0,12)
trt_effect[1:6]=trt[1]
trt_effect[7:12]=trt[2]

#Adding the treatment effects to the residuals
y=trt_effect+res

```

**The R covariance Structure**

```{r R structure}
R
```

From **R** you cand see the residual variance = 1 and there is a .5 correlation between groups of 3 replicated measurements taken from the same observational unit.

**True error variance of the mean**

```{r error variance}
l=c(1/6,1/6,1/6,1/6,1/6,1/6)

t(l)%*%R[1:6,1:6]%*%l

```


# **Lab assignment**
1) Use the provided **R** structure to solve for beta (the two treatment effects)
- Set of the design matrix **X** (1 pt)
```{r}
inp <- c(rep(1,6) , rep(0,6) )
rinp <- c(rep(0,6) , rep(1,6) )
X<- matrix(c(inp, rinp), ncol=2, nrow = 12)
X
y
```

- Estimate the treatment effects using the simulated data, **R**, and the design matrix **X** (2pts)

```{r}
Betas = solve(t(X) %*% solve(R) %*% X)  %*%  ((t(X) %*% solve(R) %*% y)) 
```

- Pull out the error variance for each beta estimate (take the inverse of the LHS and the error variances are on the diagonal) (2 pts)

```{r}
diag(solve(t(X) %*% solve(R) %*% X))
```

2) Fit a mixed model with a random effect for observational unit and solve for fixed effects for treatment 1 and 2.
- Based on the provided **R** structure create a new **Rd** covariance matrix and a **G** covariance matrix the will be equivalent. (2 pts)

```{r}
G= matrix(0, nrow=4, ncol=4)
diag(G)<- c(0.5, 0.5, 0.5, 0.5)

Rd <- matrix(0, nrow=12, ncol=12)
diag(Rd)<- 0.5

#first two in tr 1, secoond in tr 2. 
cc<-  c(rep(c(1,0,0,0),3), rep(c(0,1,0,0),3), rep(c(0,0,1,0),3), rep(c(0,0,0,1),3))
Z<- matrix(cc, nrow=12, ncol=4, byrow = TRUE)
R
X
```


- Set up and solve the mixed model equations using the **X**, **Z**, **G**, and **Rd** matrices. (no mixed model software, set it all up using the equations overed in class) (2 pts)

```{r}
LHS1 <- t(X)%*% solve(Rd) %*% X
LHS2<- t(X) %*% solve(Rd) %*% Z
LHS3<- t(Z) %*% solve(Rd) %*% X
LHS4<- t(Z) %*% solve(Rd) %*% Z + solve(G)
RHS1<- t(X) %*% solve(Rd) %*% y
RHS2<- t(Z) %*% solve(Rd) %*% y

RHS<- rbind(RHS1, RHS2)

LHS_up<- cbind(LHS1, LHS2) 
LHS_down<- cbind(LHS3, LHS4)
LHS<- rbind(LHS_up, LHS_down)


solutions<- solve(LHS) %*% RHS
treatment_effect <-solutions[1:2] 
g_effect<- solutions[3:6]
```

- Pull out the error variance for each beta estinate (take the inverse of the LHS and the error variances are on the diagonal) (2 pts)

```{r}
diag(solve(LHS))[1:2]
```


- Are the two models equivalent (look at both the estimates of the treatment effects and the error variance of the estimates)?




```{r}
treatment_effect
Beta



diag(solve(t(X) %*% solve(R) %*% X))
diag(solve(LHS))[1:2]
```

