---
title: "Master_models"
author: "sss322"
date: "2025-03-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```






### **M2: Single Phenomic Kernel (one best VI or integrated VI across time)**
```{r}
Mod_M2 <- asreml(
  fixed    = trait ~ 1,
  random   = ~ vm(GID, Pinv.sparse),    # Phenomic random effect
  residual = ~ dsum(~ units),
  data     = data
)
```


## 2. Add Multi-Kernel Integration (M4, M5)

### **M4/M5: G + P random effects in the same model**
```{r}
Mod_MultiKernel <- asreml(
  fixed    = trait ~ 1,
  random   = ~ vm(GID, Ginv.sparse)     # Genomic kernel
           + vm(GID, Pinv.sparse),      # Phenomic kernel
  residual = ~ dsum(~ units),
  data     = data
)
```

---

## 3. Move to Multivariate / Multi-Trait (M6 – M9)


```{r}
Mod_MultiTrait <- asreml(
  fixed  = cbind(trait1, trait2) ~ trait,  # or ~ trait - 1, depending on parameterization
  random = ~ corgh(trait):vm(GID, Ginv.sparse)
          + corgh(trait):vm(GID, Pinv.sparse),
  residual = ~ idv(trait):dsum(~ units),   # or corgh, us, etc., as appropriate
  data   = data
)
```


---

## 4. Incorporate Time-Series Phenomics (FPCA or PCA) + Multi-Kernel (M7, M9)


1. **Covariates** in the model’s fixed/random part, or  
2. **Used to construct a new phenomic kernel** (e.g., using the FPCA scores).


```{r}
# Suppose Pinv_fpca.sparse is a relationship matrix derived from the FPCA scores
Mod_FPCA <- asreml(
  fixed    = trait ~ 1,
  random   = ~ vm(GID, Ginv.sparse) 
           + vm(GID, Pinv_fpca.sparse),
  residual = ~ dsum(~ units),
  data     = data
)
```


## 5. Add Complex Interaction Structures (G×E, P×E) with Factor Analytic (FA)

### **Factor analytic (FA) with G kernel only** 

```{r}
Mod_FA_G <- asreml(
  fixed  = trait ~ 1,
  random = ~ vm(GID, Ginv.sparse):fa(LocYear, 2),  # FA(2) across environments
  residual = ~ dsum(~ units),
  data   = data
)
```
with 2 factors for the environment dimension.

### **Factor analytic (FA) with P kernel only**
```{r}
Mod_FA_P <- asreml(
  fixed  = trait ~ 1,
  random = ~ vm(GID, Pinv.sparse):fa(LocYear, 2),
  residual = ~ dsum(~ units),
  data   = data
)
```


### **Factor analytic (FA) with both G and P kernels**
```{r}
Mod_FA_G_P <- asreml(
  fixed  = trait ~ 1,
  random = ~ vm(GID, Ginv.sparse):fa(LocYear, 2)
         + vm(GID, Pinv.sparse):fa(LocYear, 2),
  residual = ~ dsum(~ units),
  data   = data
)
```

