---
title: "11.DK_GP"
author: "sss322"
date: "2025-10-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


CV1: New genotypes in a known environment (post-flight, full phenomics available).

## **0. Importing the dependencies**

```{r}
library(dplyr)
library(asreml)
library(ggplot2)
library(ASRgenomics)
library(readxl)
library(MASS)
library(tidyr)
library(dplyr)
library(caret)
library(asreml)
```





## **1. Importing the phenotypic data**

Importing the raw agronomic, mq and spectral 
```{r}

base_path <- if (.Platform$OS.type == "windows") {
  "C:/Users/Siim Sepp/NY_WMB"
} else {
  "~/Documents/GitHub/NY_WMB"
}

load(file.path(base_path, "data/data_ag.Rdata"))
load(file.path(base_path, "data/data_mq.Rdata"))
load(file.path(base_path, "data/WMB_master_geno.RData"))
data_ag<- data_ag[,-10] %>% na.omit()
data_mq<- data_mq[,-18] %>% na.omit()
```





## **2. Prepare data**


```{r}
#prep raw mat
load(file.path(base_path, "data/MT_raw_mat.Rdata"))
VIRC<- MT_sim_mat_raw[6:10]
VIRS<- MT_sim_mat_raw[1:5]
names(VIRS)<- c( "HELF24", "KET21", "MCG23","MCG25", "SNY22")
names(VIRC)<- c("MCG25", "HELF24", "MCG23", "SNY22", "KET21")
```



```{r}
#prep pred mat
load(file.path(base_path, "data/MT_pred_mat.Rdata"))
VIPC <- MT_sim_mat_pr[6:10]
VIPS <- MT_sim_mat_pr[1:5]
names(VIPS)<- c( "HELF24", "KET21", "MCG23","MCG25", "SNY22")
names(VIPC)<- c("MCG25", "HELF24", "MCG23", "SNY22", "KET21")
```



## **3. G+P kernel model function**

```{r}
# CV1 double-kernel cross-validation (post-flight phenomics available)
# - data: long table with columns Env, GID, <trait>, optionally Row/Column if you later add spatial residuals
# - traits: character vector of response names in 'data'
# - envs: optional subset of environments; defaults to all levels(data$Env)
# - A1: genomic relationship matrix (GRM), row/colnames = GID
# - VImat: named list of phenomic kernels by environment, e.g. VImat[["SNY22"]] = P_env matrix
# - reps, folds: CV settings
# Returns: list(results = CV rows, failures = errors)

run_doublekernel_cv1 <- function(data, A1, VImat, traits,
                                 envs = NULL, reps = 50, folds = 5,
                                 eig.tol = 1e-3, ridge = 1e-6, scale_kernels = TRUE) {

  stopifnot(all(c("Env", "GID") %in% names(data)))
  if (is.null(envs)) {
    if (is.factor(data$Env)) envs <- levels(data$Env) else envs <- sort(unique(data$Env))
  }

  # ---- helpers ---------------------------------------------------------------
  scale_to_unit_trace <- function(K) {
    tr <- sum(diag(K))
    if (!is.finite(tr) || tr <= 0) return(K)
    K * (nrow(K) / tr)
  }

  bend_and_inverse <- function(K, eig.tol = 1e-3, ridge = 1e-6) {
    # light ridge (keeps names)
    diag(K) <- diag(K) + ridge
    Gb <- G.tuneup(G = K, bend = TRUE, eig.tol = eig.tol)$Gb
    G.inverse(G = Gb, sparse = TRUE)$Ginv  # returns "dgCMatrix"
  }

  safe_cor <- function(x, y, method = "pearson") {
    ok <- stats::complete.cases(x, y)
    if (sum(ok) < 2) return(NA_real_)
    suppressWarnings(stats::cor(x[ok], y[ok], method = method))
  }

  res_all  <- data.frame()
  fail_log <- list()

  # ---- loop ------------------------------------------------------------------
  for (rep_run in seq_len(reps)) {
    set.seed(1000 + rep_run)

    for (tvar in traits) {
      if (!tvar %in% names(data)) {
        fail_log[[length(fail_log) + 1]] <- data.frame(
          RepRun = rep_run, Trait = tvar, Env = NA, Fold = NA,
          Stage = "precheck", Error = sprintf("Trait '%s' not in data", tvar)
        )
        next
      }

      for (env in envs) {
        message(sprintf("[rep %d] Trait=%s, Env=%s", rep_run, tvar, env))

        if (!env %in% names(VImat)) {
          fail_log[[length(fail_log) + 1]] <- data.frame(
            RepRun = rep_run, Trait = tvar, Env = env, Fold = NA,
            Stage = "precheck", Error = sprintf("No phenomic kernel for env '%s'", env)
          )
          next
        }

        # subset env data and keep only GIDs present in both kernels
        data_sub <- subset(data, Env == env)
        if (nrow(data_sub) < 3) {
          fail_log[[length(fail_log) + 1]] <- data.frame(
            RepRun = rep_run, Trait = tvar, Env = env, Fold = NA,
            Stage = "data_check", Error = "Too few rows in env subset"
          )
          next
        }

        P_env <- VImat[[env]]
        if (is.null(dim(P_env))) {
          fail_log[[length(fail_log) + 1]] <- data.frame(
            RepRun = rep_run, Trait = tvar, Env = env, Fold = NA,
            Stage = "precheck", Error = "P_env is not a matrix"
          )
          next
        }

        # align to common GIDs
        common <- Reduce(intersect, list(unique(as.character(data_sub$GID)),
                                         rownames(A1), rownames(P_env)))
        if (length(common) < max(5, folds + 1)) {
          fail_log[[length(fail_log) + 1]] <- data.frame(
            RepRun = rep_run, Trait = tvar, Env = env, Fold = NA,
            Stage = "id_intersect", Error = "Insufficient common GIDs across data, G, and P"
          )
          next
        }

        data_sub <- data_sub[data_sub$GID %in% common, , drop = FALSE]
        data_sub$GID  <- factor(as.character(data_sub$GID), levels = common)
        data_sub$GIDp <- data_sub$GID  # duplicate factor for P kernel

        G_sub <- A1[common, common, drop = FALSE]
        P_sub <- P_env[common, common, drop = FALSE]

        # optional scaling for stable VC estimation
        if (scale_kernels) {
          G_sub <- scale_to_unit_trace(G_sub)
          P_sub <- scale_to_unit_trace(P_sub)
        }

        # precompute inverses (same for all folds in CV1 with post-flight P available)
        Ginv.sparse <- tryCatch(bend_and_inverse(G_sub, eig.tol, ridge), error = identity)
        if (inherits(Ginv.sparse, "error")) {
          fail_log[[length(fail_log) + 1]] <- data.frame(
            RepRun = rep_run, Trait = tvar, Env = env, Fold = NA,
            Stage = "G_inverse", Error = Ginv.sparse$message
          )
          next
        }
        Pinv.sparse <- tryCatch(bend_and_inverse(P_sub, eig.tol, ridge), error = identity)
        if (inherits(Pinv.sparse, "error")) {
          fail_log[[length(fail_log) + 1]] <- data.frame(
            RepRun = rep_run, Trait = tvar, Env = env, Fold = NA,
            Stage = "P_inverse", Error = Pinv.sparse$message
          )
          next
        }

        # folds over genotypes (balanced-ish)
        n_g <- nrow(data_sub)
        fold_ids <- sample(rep(seq_len(folds), length.out = n_g))

        # build fixed formula
        formula_fixed <- as.formula(paste(tvar, "~ 1"))

        # per-fold CV
        for (f in seq_len(folds)) {
          train_data <- data_sub
          test_gids  <- as.character(train_data$GID[fold_ids == f])
          # mask phenotypes for test genotypes
          train_data[fold_ids == f, tvar] <- NA
          mod_cv <- tryCatch({
            asreml(
              fixed     = formula_fixed,
              random    = ~ vm(GID,  Ginv.sparse) + vm(GIDp, Pinv.sparse),
              residual  = ~ idv(units),
              data      = train_data,
              na.action = list(x = "include", y = "include"),
              ai.sing   = TRUE
            )
          }, error = identity)
          
          
          if (inherits(mod_cv, "error")) {
            fail_log[[length(fail_log) + 1]] <- data.frame(
              RepRun = rep_run, Trait = tvar, Env = env, Fold = f,
              Stage = "cv_model", Error = mod_cv$message
            )
            next
          }

          # EBVs for all GIDs
          p_gid <- tryCatch(predict(mod_cv, classify = "GID")$pvals, error = identity)
          if (inherits(p_gid, "error") || is.null(p_gid)) {
            fail_log[[length(fail_log) + 1]] <- data.frame(
              RepRun = rep_run, Trait = tvar, Env = env, Fold = f,
              Stage = "cv_predict", Error = if (inherits(p_gid, "error")) p_gid$message else "NULL preds"
            )
            next
          }
          names(p_gid)[names(p_gid) == "predicted.value"] <- "u_hat"

          # join with observed y for TEST genotypes only
          y_test <- data_sub[fold_ids == f, c("GID", tvar)]
          colnames(y_test)[2] <- "y_obs"
          out <- merge(y_test, p_gid[, c("GID", "u_hat")], by = "GID", all.x = TRUE)

          # metrics
          cor_p  <- safe_cor(out$u_hat, out$y_obs, "pearson")
          cor_s  <- safe_cor(out$u_hat, out$y_obs, "spearman")
          rmse   <- if (sum(stats::complete.cases(out$u_hat, out$y_obs)) >= 1) {
            sqrt(mean((out$y_obs - out$u_hat)^2, na.rm = TRUE))
          } else NA_real_

          res_all <- rbind(res_all, data.frame(
            RepRun = rep_run, Trait = tvar, Env = env, Fold = f,
            n_test = sum(fold_ids == f),
            Cor    = cor_p, Cor_s = cor_s, RMSE = rmse,
            P_available_for_test = TRUE
          ))
        } # folds
      } # env
    } # trait
  } # reps

  fail_df <- if (length(fail_log)) do.call(rbind, fail_log) else data.frame()
  list(results = res_all, failures = fail_df)
}
```


## **3. Genomic prediction of agronomic traits within, year**

```{r}

traits<- colnames(data_ag)[7:8]
GP_DK_results_ag_raw_sel <- run_doublekernel_cv1(data = data_ag, A1 = A1,VIPS, traits = traits, reps = 50)


save(GP_DK_results_ag_raw_sel, file= "GP_DK_results_ag_raw_sel.Rdata")
```


## **4. Genomic prediction of mq traits within, year**

```{r}
traits <- colnames(data_mq)[4:15]
envs<- unique(data_mq$Env)
GP_DK_results_mq_raw_sel <- run_doublekernel_cv1(data = data_mq, A1 = A1,VIPS, traits = traits,envs=envs, reps = 50)

hist(GP_DK_results_mq_raw_sel$results$Cor)
save(GP_DK_results_mq_raw_sel, file= "GP_DK_results_mq_raw_sel.Rdata")
```



## **4. Genomic prediction of agronomic BLUE within, year**

```{r}
load(file.path(base_path, "data/ag_BLUE_spatial.RData"))
preds<- preds %>% pivot_wider(names_from= "VI", values_from= "Predicted")
traits<- colnames(preds)[c(4,7)]
envs<- unique(preds$Env)

GP_DK_results_ag_pr <- run_doublekernel_cv1(data = preds, A1 = A1,VIPS, traits = traits,envs=envs, reps = 50)
save(GP_DK_results_ag_pr, file= "GP_DK_results_ag_pr.Rdata")
```


## **5. Genomic prediction of MQ BLUP traits within, year**
```{r}
load(file.path(base_path, "data/mq_BLUP_spatial.RData"))
traits <- unique(preds$VI)
preds<- preds[,-5] %>% pivot_wider(names_from = "VI", values_from = "Predicted")
envs<- unique(preds$Env)


GP_DK_results_mq_pr <- run_doublekernel_cv1(data = preds, A1 = A1,VImat, traits = traits,envs=envs, reps = 50)
save(GP_DK_results_mq_pr, file= "GP_DK_results_mq_pr.Rdata")
```


