---
title: "12. Prediction Plots"
author: "sss322"
date: "2025-10-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




```{r}
library(MASS)
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(stringr)
library(forcats)
```



# **1. Import phenomic prediction LM, RF, GLM and MM results **

Load files
```{r}
#LM, RF, EN, LASSO, 
# Define the base path - you can set this depending on your OS
base_path <- if (.Platform$OS.type == "windows") {
  "C:/Users/Siim Sepp/NY_WMBCL/output"
} else {
  "~/Documents/GitHub/NY_WMBCL/output"
}

# Load files with constructed paths
load(file.path(base_path, "PP_LM_results_ag_pred.Rdata"))
load(file.path(base_path, "PP_LM_results_ag_raw.Rdata"))
load(file.path(base_path, "PP_LM_results_mq_raw.Rdata"))

# PP MM
load(file.path(base_path, "PP_MM_results_ag_pred_comb.Rdata"))#
load(file.path(base_path, "PP_MM_results_ag_pred_sel.Rdata"))#
load(file.path(base_path, "PP_MM_results_ag_raw_comb.Rdata"))#
load(file.path(base_path, "PP_MM_results_ag_raw_sel.Rdata"))#
#load(file.path(base_path, "PP_MM_results_MQ_pred_comb.Rdata"))#
#load(file.path(base_path, "PP_MM_results_MQ_pred_sel.Rdata"))#
load(file.path(base_path, "PP_MM_results_MQ_raw_comb.Rdata"))#
load(file.path(base_path, "PP_MM_results_MQ_raw_sel.Rdata"))#

# GBLUP GP
load(file.path(base_path, "GP_MM_results_ag_pred.Rdata"))
load(file.path(base_path, "GP_MM_results_ag_raw.Rdata"))
#load(file.path(base_path, "GP_MM_results_mq_preds.Rdata"))
load(file.path(base_path, "GP_MM_results_mq_raw.Rdata"))

# MT GP
load(file.path(base_path, "MT_mq_res_raw_sp.Rdata"))
load(file.path(base_path, "MT_ag_res_raw_sp.Rdata"))

# double kernel GP
load(file.path(base_path, "GP_DK_results_ag_pr.Rdata"))
#load(file.path(base_path, "GP_DK_results_ag_pr_sel.Rdata"))
#load(file.path(base_path, "GP_DK_results_ag_raw.Rdata"))
load(file.path(base_path, "GP_DK_results_ag_raw_sel.Rdata"))
load(file.path(base_path, "GP_DK_results_mq_raw_sel.Rdata"))
#load(file.path(base_path, "GP_DK_results_mq_raw_sel.Rdata"))
#load(file.path(base_path, "GP_DK_results_mq_pr_c.Rdata"))
#load(file.path(base_path, "GP_DK_results_mq_pr_sel.Rdata"))
table(MT_ag_res_raw_sp$failures$Cor_Trait, MT_ag_res_raw_sp$failures$Trait)

```


# **2. Prepare and merge LM, RF, GLM and MM results **

**Merge MM**
```{r}
PP_MM_results_ag_pred_sel$met  <- "MM_ag_pred_sel"
PP_MM_results_ag_pred_comb$met <- "MM_ag_pred_comb"
PP_MM_results_ag_pred_sel$BLUE <- "Two"
PP_MM_results_ag_pred_comb$BLUE<- "Two"
PP_MM_results_ag_pred_sel$mod  <- "MM"
PP_MM_results_ag_pred_comb$mod <- "MM"
PP_MM_results_ag_pred_sel$sel  <- "sel"
PP_MM_results_ag_pred_comb$sel <- "comb"

PP_MM_results_ag_raw_comb$met  <- "MM_ag_raw_comb"
PP_MM_results_ag_raw_sel$met   <- "MM_ag_raw_sel"
PP_MM_results_ag_raw_comb$BLUE <- "One"
PP_MM_results_ag_raw_sel$BLUE  <- "One"
PP_MM_results_ag_raw_comb$mod  <- "MM"
PP_MM_results_ag_raw_sel$mod   <- "MM"
PP_MM_results_ag_raw_comb$sel  <- "comb"
PP_MM_results_ag_raw_sel$sel   <- "sel"

PP_MM_results_MQ_raw_comb$met  <- "MM_mq_raw_comb"
PP_MM_results_MQ_raw_sel$met   <- "MM_mq_raw_sel"
PP_MM_results_MQ_raw_comb$BLUE <- "One"
PP_MM_results_MQ_raw_sel$BLUE  <- "One"
PP_MM_results_MQ_raw_comb$mod  <- "MM"
PP_MM_results_MQ_raw_sel$mod   <- "MM"
PP_MM_results_MQ_raw_comb$sel  <- "comb"
PP_MM_results_MQ_raw_sel$sel   <- "sel"

#PP_MM_results_MQ_pred_comb$met <- "MM_mq_pred_comb"
#PP_MM_results_MQ_pred_sel$met  <- "MM_mq_pred_sel"
#PP_MM_results_MQ_pred_comb$BLUE<- "Two"
#PP_MM_results_MQ_pred_sel$BLUE <- "Two"
#PP_MM_results_MQ_pred_comb$mod <- "MM"
#PP_MM_results_MQ_pred_sel$mod  <- "MM"
#PP_MM_results_MQ_pred_comb$sel <- "comb"
#PP_MM_results_MQ_pred_sel$sel  <- "sel"

#PP_MM_results_ag_pred_sel<- PP_MM_results_ag_pred_sel[,-3]



PP_MM_master<- rbind(PP_MM_results_ag_pred_sel, PP_MM_results_ag_pred_comb,  PP_MM_results_ag_raw_comb,PP_MM_results_ag_raw_sel,  PP_MM_results_MQ_raw_comb, PP_MM_results_MQ_raw_sel)#, PP_MM_results_MQ_pred_comb) PP_MM_results_MQ_pred_sel,
```


**Merge LM RF, PCR** 
```{r}
names(PP_LM_results_ag_pred)<- names(PP_LM_results_ag_raw)
PP_LM_results_ag_pred$accuracy$met <- "LM_ag_pred"
PP_LM_results_ag_raw$accuracy$met  <- "LM_ag_raw"
PP_LM_results_mq_raw$accuracy$met  <- "LM_mq_raw"

PP_LM_results_ag_pred$accuracy$mod <- "LM"
PP_LM_results_ag_raw$accuracy$mod  <- "LM"
PP_LM_results_mq_raw$accuracy$mod  <- "LM"

PP_LM_results_ag_pred$accuracy$BLUE<- "Two"
PP_LM_results_ag_raw$accuracy$BLUE <- "One"
PP_LM_results_mq_raw$accuracy$BLUE <- "One"

PP_LM_results_ag_pred$accuracy$sel <- "comb"
PP_LM_results_ag_raw$accuracy$sel  <- "comb"
PP_LM_results_mq_raw$accuracy$sel  <- "comb"
PP_LM_master<- rbind(PP_LM_results_ag_pred$accuracy, PP_LM_results_ag_raw$accuracy, PP_LM_results_mq_raw$accuracy)

colnames(PP_MM_master)[3]<- "Accuracy"
colnames(PP_MM_master)[4]<- "Cycle"
PP_MM_master$Predictor_Set<- "NA"
PP_MM_master$Model <- "PBLUP"
PP_LM_MM_accuracy<- rbind(PP_LM_master, PP_MM_master)

PP_LM_MM_accuracy <- PP_LM_MM_accuracy %>% filter(!Trait == "BarleyColor.Agtron.")
PP_LM_MM_accuracy <- PP_LM_MM_accuracy %>% filter(!Trait == "WortClarity")
```




**Merge LM and MM.** 
```{r}
cv_summary <- PP_LM_MM_accuracy %>%
  group_by(Trait, met, Model, Predictor_Set) %>%
  summarise(
    mean_accuracy = round(mean(Accuracy, na.rm = TRUE),2),
    sd_accuracy   = round(sd(Accuracy, na.rm = TRUE),2),
    n  = n()
    )

cv_summary

trait_map <- c(
  KernelWeight.mg. = "KW",
  on6.64....       = "on6/64",
  MaltExtract...   = "ME",
  BarleyProtein... = "BP",
  WortProtein...   = "WP",
  S.T              = "S.T",
  DP..ASBC.       = "DP",
  Alpha.amylase.20.DU. = "AA",
  Beta.glucan.ppm. = "BG",
  HD_JD = "HD",
  yield_kgha = "GY",
  FAN.ppm.         = "FAN",
  AdjunctQualityScore = "Adj.Sc",
  All_MaltQualityScore = "All.Sc")

env_order<- c( "KET21", "SNY22" ,"MCG23","HELF24" , "MCG25" )


PP_LM_MM_accuracy <- PP_LM_MM_accuracy %>%
  mutate(Trait = recode(Trait, !!!trait_map))

pal <- c(
  KET21  = "cornsilk3",
  SNY22  = "palegreen2",
  MCG23  = "wheat2",
  HELF24 = "lightblue3",
  MCG25 = "grey")

pal1 <- c(
  EN  = "cornsilk3",
  Lin  = "palegreen2",
  PCR  = "wheat2",
  RR = "lightblue3",
  Lasso = "grey", 
  PBLUP = "pink2", 
  RF= "purple4")

trait_order<- c( "GY" , "HD",  "KW", "on6/64" , "FAN" , "DP", "AA", "S.T" , "BG" , "ME",  "BP", "WP", "Adj.Sc", "All.Sc" )

PP_LM_MM_accuracy  <- PP_LM_MM_accuracy  %>%  
  mutate( Environment = trimws(as.character(Environment)),
          Environment = fct_relevel(Environment, env_order)) %>% 
  mutate( Trait= trimws(as.character(Trait)),
          Trait = fct_relevel(Trait, trait_order))

unique(PP_LM_MM_accuracy$Model)
PP_LM_MM_accuracy <- PP_LM_MM_accuracy %>%
  mutate(Model = ifelse(grepl("lastic", Model), "EN", Model))

PP_LM_MM_accuracy <- PP_LM_MM_accuracy %>%
  mutate(Model = ifelse(grepl("asso", Model), "Lasso", Model))

PP_LM_MM_accuracy <- PP_LM_MM_accuracy %>%
  mutate(Model = ifelse(grepl("idge", Model), "RR", Model))

PP_LM_MM_accuracy <- PP_LM_MM_accuracy %>%
  mutate(Model = ifelse(grepl("Fores", Model), "RF", Model))

PP_LM_MM_accuracy <- PP_LM_MM_accuracy %>%
  mutate(Model = ifelse(grepl("Comp", Model), "PCR", Model))

PP_LM_MM_accuracy <- PP_LM_MM_accuracy %>%
  mutate(Model = ifelse(grepl("Linear", Model), "Lin", Model))


table(PP_LM_MM_accuracy$Environment, PP_LM_MM_accuracy$Trait)
```


**Merge GP**
```{r}
GP_MM_results_ag_pred$results$BLUE<- "Two"
GP_MM_results_ag_raw$results$BLUE<- "One"
GP_MM_results_mq_raw$results$BLUE<- "One"
#GP_MM_results_mq_preds$results$BLUE<- "Two"

GP_MM_results_ag_pred$results$met <- "GP"
GP_MM_results_ag_raw$results$met <- "GP"
GP_MM_results_mq_raw$results$met <- "GP"
#GP_MM_results_mq_preds$results$met <- "GP"


colnames(GP_MM_results_ag_pred$results)
#colnames(GP_MM_results_mq_preds$results)
colnames(GP_MM_results_ag_raw$results)
colnames(GP_MM_results_mq_raw$results)
colnames(GP_MM_results_ag_pred$results)
#colnames(GP_MM_results_mq_preds$results)[5]<- "Cor"

GP_MM_master <- rbind(GP_MM_results_ag_pred$results[,-5], GP_MM_results_ag_raw$results[,-5], GP_MM_results_mq_raw$results[,-5])#, #GP_MM_results_mq_preds$results$)



unique(GP_MM_master$Trait)

GP_MM_master <- GP_MM_master%>% filter(!Trait == "BarleyColor.Agtron.")
GP_MM_master <- GP_MM_master %>% filter(!Trait == "WortClarity")
GP_MM_master <- GP_MM_master %>% filter(!Trait == "winter_survival")
GP_MM_master <- GP_MM_master %>% filter(!Trait == "Ht")
str(GP_MM_master)
GP_MM_master<- GP_MM_master[,-1]
GP_MM_master<- GP_MM_master[-3]
GP_MM_master$Cor_Trait<- NA
GP_MM_master$type<- NA
```


**Merge MT GP**
```{r}
MT_res_raw<- rbind(MT_ag_res_raw_sp$results, MT_mq_res_raw_sp$results)
MT_res_raw<- MT_res_raw[-4:-5]
MT_res_raw$BLUE<- "One"
MT_res_raw$type<- NA
MT_res_raw$met<- "MT"
```


**merge DK GP**
```{r}
#GP_DK_results_ag_pr$results$type<- "comb"
GP_DK_results_ag_pr$results$type<- "sel"
#GP_DK_results_ag_raw_c$results$type<- "comb"
GP_DK_results_ag_raw_sel$results$type<- "sel"

#GP_DK_results_mq_pr_c$results$type<- "comb"
#GP_DK_results_mq_pr_sel$results$type<- "sel"
#GP_DK_results_mq_raw_c$results$type<- "comb"
GP_DK_results_mq_raw_sel$results$type<- "sel"

#failers
#GP_DK_results_ag_pr_c$failures$type<- "comb"
#GP_DK_results_ag_pr_sel$failures$type<- "sel"
#GP_DK_results_ag_raw_c$failures$type<- "comb"
#GP_DK_results_ag_raw_sel$failures$type<- "sel"

#GP_DK_results_mq_pr_c$failures$type<- "comb"
#GP_DK_results_mq_pr_sel$failures$type<- "sel"


GP_DK_res<- rbind(GP_DK_results_ag_pr$results,  GP_DK_results_ag_raw_sel$results,  GP_DK_results_mq_raw_sel$results)#, GP_DK_results_mq_pr_sel$results, GP_DK_results_mq_raw_c$results,GP_DK_results_ag_pr_c$results, GP_DK_results_ag_raw_c$results, GP_DK_results_mq_pr_c$results,)

#GP_DK_fail<- rbind(GP_DK_results_ag_pr_c$failures, GP_DK_results_ag_pr_sel$failures, GP_DK_results_ag_raw_c$failures, GP_DK_results_ag_raw_sel$failures, GP_DK_results_mq_pr_c$failures, GP_DK_results_mq_pr_sel$failures)

GP_DK_res<- GP_DK_res %>% filter(!Trait == "BarleyColor.Agtron.") %>% filter(!Trait == "WortClarity")
colnames(GP_DK_res)
GP_DK_res <- GP_DK_res[c(2,3,6,10)]
GP_DK_res$met<- "DK"
GP_DK_res$BLUE<- "One"
GP_DK_res$Cor_Trait<- NA
```



#Merge GP, MT, DK. 
```{r}
GP_DK_res<- GP_DK_res %>% filter(type== "sel")
GP_res<- rbind(GP_DK_res, GP_MM_master, MT_res_raw)


GP_res <- GP_res %>%
  mutate(Trait = recode(Trait, !!!trait_map))

GP_res <- GP_res  %>%  
  mutate( Environment = trimws(as.character(Env)),
          Environment = fct_relevel(Env, env_order)) %>%
  mutate( Trait= trimws(as.character(Trait)),
          Trait = fct_relevel(Trait, trait_order))


cv_summary <- GP_res %>%
  group_by(Trait, met) %>%
  summarise(
    mean_accuracy = round(mean(Cor, na.rm = TRUE),2),
    sd_accuracy   = round(sd(Cor, na.rm = TRUE),2),
    n  = n()
    )

cv_summary
```




# **3. Plot LM, RF, GLM and MM results **
```{r}
PP_LM_MM_accuracy1 <- PP_LM_MM_accuracy  %>% filter(BLUE == "One")

##############Trait##########################################################################
library(RColorBrewer)
base_pal <- brewer.pal(12, "Set3")
extra_pal <- brewer.pal(8, "Dark2")
pal_combined <- c(base_pal, extra_pal)



p<- ggplot(PP_LM_MM_accuracy1, aes(x = Trait, y = Accuracy, fill = Trait)) +
  geom_boxplot() +
   #facet_grid(~Trait, scales = "free") +
  labs(
    title = " ",
    y = "CV Prediction Accuracy",
    x = "Trait"
  ) +
  scale_fill_manual(values = pal_combined) +
  theme(
    strip.text = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 14),
    axis.text  = element_text(size = 14, angle = 90),
    legend.position = "bottom"
  )
print(p)

##############Environment#########################################################################
p<- ggplot(PP_LM_MM_accuracy1, aes(x = Environment, y = Accuracy, fill = Environment)) +
  geom_boxplot(notch=FALSE) +
   facet_grid(~Trait, scales = "free") +
  labs(
    title = " ",
    y = "CV Prediction Accuracy",
    x = "Environment"
  ) +
 scale_fill_manual(values = pal_combined) +
   coord_cartesian(ylim = c(-0.2, 1)) +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text  = element_text(size = 12, angle = 90),
    legend.position = "bottom", 
    legend.text = element_text(size = 12)
  )
print(p)

##############Model##########################################################################

# Accuracy by Model
PP_LM_MM_accuracy1 <- PP_LM_MM_accuracy %>%
  filter(BLUE == "One")

p<- ggplot(PP_LM_MM_accuracy1, aes(x = Model, y = Accuracy, fill = Model)) +
  geom_boxplot() +
  facet_grid(~Trait, scales = "free") +
  labs(
    title = " ",
    y = "CV Prediction Accuracy",
    x = "Model"
  ) +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text  = element_text(size = 12, angle = 90),
    legend.position = "bottom"
  )

print(p)

res <- PP_LM_MM_accuracy %>%
  filter(BLUE == "One") %>%  filter(Predictor_Set %in% c("NDVI", "PSRI", "NA")) %>% filter(!met %in% c("MM_ag_raw_comb", "MM_mq_raw_comb"))

##############trait env##########################################################################
cv_summary <- res %>%
  group_by(Trait, Environment) %>%
  summarise(
    mean_accuracy = mean(Accuracy, na.rm = TRUE),
    min = min(Accuracy, na.rm= TRUE), 
    max = max(Accuracy, na.rm = TRUE),
    sd_accuracy   = sd(Accuracy, na.rm = TRUE),
    n             = sum(!is.na(Accuracy)),
    se_accuracy   = sd_accuracy / sqrt(n),
    cv_accuracy   = ifelse(
      abs(mean_accuracy) > 1e-6,
      sd_accuracy / mean_accuracy*100,
      NA_real_
    ),
    .groups = "drop"
  ) %>% arrange(Trait, Environment)

cv_summary
mean(cv_summary$cv_accuracy)
##############trait##########################################################################
cv_summary <- res %>%
  group_by(Trait) %>%
  summarise(
    mean_accuracy = mean(Accuracy, na.rm = TRUE),
    min = min(Accuracy, na.rm= TRUE), 
    max = max(Accuracy, na.rm = TRUE),
    sd_accuracy   = sd(Accuracy, na.rm = TRUE),
    n             = sum(!is.na(Accuracy)),
    se_accuracy   = sd_accuracy / sqrt(n),
    cv_accuracy   = ifelse(
      (mean_accuracy) > 1e-1,
      sd_accuracy / mean_accuracy*100,
      NA_real_
    ),
    .groups = "drop"
  ) %>% arrange(mean_accuracy)

cv_summary

mean(cv_summary$cv_accuracy, na.rm=TRUE )
##############model##########################################################################

cv_summary<- res %>%
  group_by(Model) %>%
  summarise(
    mean_accuracy = mean(Accuracy, na.rm = TRUE),
    min = min(Accuracy, na.rm= TRUE), 
    max = max(Accuracy, na.rm = TRUE),
    sd_accuracy   = sd(Accuracy, na.rm = TRUE),
    n             = sum(!is.na(Accuracy)),
    se_accuracy   = sd_accuracy / sqrt(n),
    cv_accuracy   = ifelse(
      abs(mean_accuracy) > 1e-6,
      sd_accuracy / mean_accuracy*100,
      NA_real_
    ),
    .groups = "drop"
  ) %>%
  arrange(cv_accuracy)

mean(cv_summary$cv_accuracy)


PP_LM_MM_accuracy12<- PP_LM_MM_accuracy1 %>% filter(Trait %in% c("GY", "HD"))
PP_LM_MM_accuracy12 %>%
  group_by(Environment) %>%
  summarise(
    mean_accuracy = mean(Accuracy, na.rm = TRUE),
    min = min(Accuracy, na.rm= TRUE), 
    max = max(Accuracy, na.rm = TRUE),
    sd_accuracy   = sd(Accuracy, na.rm = TRUE),
    n             = sum(!is.na(Accuracy)),
    se_accuracy   = sd_accuracy / sqrt(n),
    cv_accuracy   = ifelse(
      abs(mean_accuracy) > 1e-6,
      sd_accuracy / mean_accuracy*100,
      NA_real_
    ),
    .groups = "drop"
  ) %>%
  arrange(Environment)


##############ANOVA##########################################################################
anova_mod <- aov(Accuracy ~ Trait + Environment + Model + 
                   Trait:Model + Environment:Model + Trait:Environment,
                 data = res)

anova_tab <- summary(anova_mod)[[1]]

anova_df <- as.data.frame(anova_tab)
anova_df$Effect <- rownames(anova_df)

# Remove residual row for % calculation
ss_total <- sum(anova_df$`Sum Sq`, na.rm = TRUE)

anova_df <- anova_df %>%
  mutate(
    perc_var = 100 * `Sum Sq` / ss_total
  )

```




# **4. Plot MM phenomic results **
```{r}
#One stage or two stage
PP_LM_MM_accuracy1 <- PP_LM_MM_accuracy  %>%   filter(mod == "MM")

#ENV
p<- ggplot(PP_LM_MM_accuracy1, aes(x = Environment, y = Accuracy, fill = Environment)) +
  geom_boxplot() +
   facet_grid(~Trait, scales = "free") +
  labs(
    title = " ",
    y = "CV Prediction Accuracy",
    x = "Model Method"
  ) +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text  = element_text(size = 10, angle = 90),
    legend.position = "bottom"
  )
print(p)
```




# **5. Plot GP results **
```{r}
#Accuracy by environment
GP_MM_master1<- GP_res  %>% filter(met == "GP") %>% filter(BLUE == "Two")
GP_MM_master2<- GP_res  %>% filter(met == "GP") %>% filter(BLUE == "One") %>% filter(!Trait %in% c("GY", "HD"))

GP_MM_master1<- rbind(GP_MM_master1, GP_MM_master2)

table(GP_MM_master1$Trait)
##############trait##########################################################################
ggplot(GP_MM_master1, aes(x =Trait, y = Cor, fill = Trait)) +
  geom_boxplot() +
  labs(title = " ",
       y = "Prediction Accuracy", x = "Model Method") +
  coord_cartesian(ylim = c(-0.2, 1)) +
  scale_fill_manual(values = pal_combined) +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text  = element_text(size = 12, angle = 90),
    legend.position = "bottom", 
    legend.text = element_text(size = 12)
  )



##############Env x trait##########################################################################
ggplot(GP_MM_master1, aes(x = Trait, y = Cor, fill = Trait)) +
  geom_boxplot() +
   facet_wrap(~  Env, scales = "free_x", ncol = 5) +
  labs(title = " ",
       y = "Prediction Accuracy", x = "") +
  scale_fill_manual(values = pal_combined) +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text  = element_text(size = 12, angle = 90),
    legend.position = "bottom", 
    legend.text = element_text(size = 12)
  )

#Summarize GP
mean(GP_MM_master1$Cor, na.rm= TRUE) 

##############Env##########################################################################
cv_summary <- GP_MM_master1 %>%
  group_by(Env) %>%
  summarise(
    mean_accuracy = mean(Cor, na.rm = TRUE),
    min = min(Cor, na.rm= TRUE), 
    max = max(Cor, na.rm = TRUE),
    sd_accuracy   = sd(Cor, na.rm = TRUE),
    n             = sum(!is.na(Cor)),
    se_accuracy   = sd_accuracy / sqrt(n),
    cv_accuracy   = ifelse(
      abs(mean_accuracy) > 1e-6,
      sd_accuracy / mean_accuracy*100,
      NA_real_
    ),
    .groups = "drop"
  ) %>% arrange(mean_accuracy)
cv_summary


##############trait##########################################################################
cv_summary1 <- GP_MM_master1 %>%
  group_by(Trait) %>%
  summarise(
    mean_accuracy = mean(Cor, na.rm = TRUE),
    min = min(Cor, na.rm= TRUE), 
    max = max(Cor, na.rm = TRUE),
    sd_accuracy   = sd(Cor, na.rm = TRUE),
    n             = sum(!is.na(Cor)),
    se_accuracy   = sd_accuracy / sqrt(n),
    cv_accuracy   = ifelse(
      (mean_accuracy) > 1e-2,
      sd_accuracy / mean_accuracy*100,
      NA_real_
    ),
    .groups = "drop"
  ) %>% arrange(mean_accuracy)


cv_summary1
dim(cv_summary1)
mean(cv_summary1$cv_accuracy[6:14,])

##############trait v env#########################################################################
cv_summary11 <- GP_MM_master1 %>%
  group_by(Trait, Env) %>%
  summarise(
    mean_accuracy = mean(Cor, na.rm = TRUE),
    min = min(Cor, na.rm= TRUE), 
    max = max(Cor, na.rm = TRUE),
    sd_accuracy   = sd(Cor, na.rm = TRUE),
    n             = sum(!is.na(Cor)),
    se_accuracy   = sd_accuracy / sqrt(n),
    cv_accuracy   = ifelse(
      abs(mean_accuracy) > 1e-6,
      sd_accuracy / mean_accuracy*100,
      NA_real_
    ),
    .groups = "drop"
  ) %>% arrange(desc(mean_accuracy))
cv_summary11

```




# **6. Plot MT GP results **

```{r}
#Accuracy by environment
MT_res_raw <- GP_res %>% filter(met =="MT") #>% filter(Cor_Trait =="FPC1")

table(MT_res_raw$Cor_Trait)

MT_res_raw <- MT_res_raw %>% filter(!Cor_Trait %in% c("FPC2", "FPC3"))

##############trait v env##########################################################################
ggplot(MT_res_raw, aes(x =Environment, y = Cor, fill = Environment)) +
  geom_boxplot() +
  facet_wrap(~  Trait, scales = "free_x", nrow= 2, ncol= 7) +
  labs(title = " ",
       y = "Prediction Accuracy", x = "Model Method") +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text  = element_text(size = 12, angle = 90),  legend.position = "bottom")


##############trait##########################################################################
ggplot(MT_res_raw, aes(x =Trait, y = Cor, fill = Trait)) +
  geom_boxplot() +
  #facet_wrap(~  Trait, scales = "free_x") +
  labs(title = " ",
       y = "Prediction Accuracy", x = "Model Method") +
  scale_fill_manual(values = pal_combined) +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text  = element_text(size = 12, angle = 90),
    legend.position = "bottom", 
    legend.text = element_text(size = 12)
  )


##############Env##########################################################################
#Summarize GP
cv_summary <- MT_res_raw %>%
  group_by(Env) %>%
  summarise(
    mean_accuracy = mean(Cor, na.rm = TRUE),
    min = min(Cor, na.rm= TRUE), 
    max = max(Cor, na.rm = TRUE),
    sd_accuracy   = sd(Cor, na.rm = TRUE),
    n             = sum(!is.na(Cor)),
    se_accuracy   = sd_accuracy / sqrt(n),
    cv_accuracy   = ifelse(
      abs(mean_accuracy) > 1e-6,
      sd_accuracy / mean_accuracy*100,
      NA_real_
    ),
    .groups = "drop"
  ) %>% arrange(mean_accuracy)
cv_summary
##############trait##########################################################################
cv_summary2 <- MT_res_raw %>%
  group_by(Trait) %>%
  summarise(
    mean_accuracy = mean(Cor, na.rm = TRUE),
    min = min(Cor, na.rm= TRUE), 
    max = max(Cor, na.rm = TRUE),
    sd_accuracy   = sd(Cor, na.rm = TRUE),
    n             = sum(!is.na(Cor)),
    se_accuracy   = sd_accuracy / sqrt(n),
    cv_accuracy   = ifelse(
      abs(mean_accuracy) > 1e-6,
      sd_accuracy / mean_accuracy*100,
      NA_real_
    ),
    .groups = "drop"
  ) %>% arrange(mean_accuracy)
cv_summary2


##############trait v env#########################################################################
cv_summary22 <- MT_res_raw %>%
  group_by(Trait, Env) %>%
  summarise(
    mean_accuracy = mean(Cor, na.rm = TRUE),
    min = min(Cor, na.rm= TRUE), 
    max = max(Cor, na.rm = TRUE),
    sd_accuracy   = sd(Cor, na.rm = TRUE),
    n             = sum(!is.na(Cor)),
    se_accuracy   = sd_accuracy / sqrt(n),
    cv_accuracy   = ifelse(
      abs(mean_accuracy) > 1e-6,
      sd_accuracy / mean_accuracy*100,
      NA_real_
    ),
    .groups = "drop"
  )  %>% arrange(mean_accuracy)
cv_summary22

cv_summary <- MT_res_raw %>%
  summarise(
       mean_accuracy = mean(Cor, na.rm = TRUE),
    sd_accuracy   = sd(Cor, na.rm = TRUE),
    se_accuracy   = sd_accuracy / sqrt(n()),
    cv_accuracy   = sd_accuracy / mean_accuracy,
    n = n()
    ) %>% arrange(mean_accuracy)
cv_summary

```



# **7. Plot DK GP results **

```{r}
#Accuracy by environment
GP_DK_res <- GP_res %>% filter(met =="DK") 

##############Env v trait##########################################################################
ggplot(GP_DK_res, aes(x =Environment, y = Cor, fill = Environment)) +
  geom_boxplot() +
 facet_wrap(~Trait, scales = "free_x", nrow = 2, ncol = 7)+
  labs(title = " ",
       y = " CV Prediction Accuracy", x = "") +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text  = element_text(size = 12, angle = 90),  legend.position = "bottom")

##############trait##########################################################################
ggplot(GP_DK_res, aes(x =Trait, y = Cor, fill = Trait)) +
  geom_boxplot() +
    scale_fill_manual(values = pal_combined) +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text  = element_text(size = 12, angle = 90),
    legend.position = "bottom", 
    legend.text = element_text(size = 12))+
   coord_cartesian(ylim = c(-0.2, 1))



##############Env##########################################################################
cv_summary <- GP_DK_res%>%
  group_by(Env) %>%
  summarise(
    mean_accuracy = mean(Cor, na.rm = TRUE),
    min = min(Cor, na.rm= TRUE), 
    max = max(Cor, na.rm = TRUE),
    sd_accuracy   = sd(Cor, na.rm = TRUE),
    n             = sum(!is.na(Cor)),
    se_accuracy   = sd_accuracy / sqrt(n),
    cv_accuracy   = ifelse(
      abs(mean_accuracy) > 1e-6,
      sd_accuracy / mean_accuracy*100,
      NA_real_
    ),
    .groups = "drop"
  )  %>% arrange(mean_accuracy)
cv_summary

##############trait##########################################################################
cv_summary3 <- GP_DK_res %>%
  group_by(Trait) %>%
  summarise(
    mean_accuracy = mean(Cor, na.rm = TRUE),
    min = min(Cor, na.rm= TRUE), 
    max = max(Cor, na.rm = TRUE),
    sd_accuracy   = sd(Cor, na.rm = TRUE),
    n             = sum(!is.na(Cor)),
    se_accuracy   = sd_accuracy / sqrt(n),
    cv_accuracy   = ifelse(
      abs(mean_accuracy) > 1e-2,
      sd_accuracy / mean_accuracy*100,
      NA_real_
    ),
    .groups = "drop"
  ) %>% arrange(mean_accuracy)
cv_summary3
max(cv_summary3$mean_accuracy)

##############trait env##########################################################################
cv_summary33 <- GP_DK_res %>%
  group_by(Trait, Env) %>%
    summarise(
    mean_accuracy = mean(Cor, na.rm = TRUE),
    min = min(Cor, na.rm= TRUE), 
    max = max(Cor, na.rm = TRUE),
    sd_accuracy   = sd(Cor, na.rm = TRUE),
    n             = sum(!is.na(Cor)),
    se_accuracy   = sd_accuracy / sqrt(n),
    cv_accuracy   = ifelse(
      abs(mean_accuracy) > 1e-6,
      sd_accuracy / mean_accuracy*100,
      NA_real_
    ),
    .groups = "drop"
  )  %>% arrange(mean_accuracy)
cv_summary33


cv_summary <- GP_DK_res %>%
  summarise(
       mean_accuracy = mean(Cor, na.rm = TRUE),
    sd_accuracy   = sd(Cor, na.rm = TRUE),
    se_accuracy   = sd_accuracy / sqrt(n()),
    cv_accuracy   = sd_accuracy / mean_accuracy,
    n = n()
    ) %>% arrange(mean_accuracy)
cv_summary

```



#Comparing MT_GP, DK. 

```{r}


cv_summary <- GP_res %>%
  group_by(Env, met, Trait) %>%
  summarise(
    mean_accuracy = mean(Cor, na.rm = TRUE),
    min = min(Cor, na.rm= TRUE), 
    max = max(Cor, na.rm = TRUE),
    sd_accuracy   = sd(Cor, na.rm = TRUE),
    n             = sum(!is.na(Cor)),
    se_accuracy   = sd_accuracy / sqrt(n),
    cv_accuracy   = ifelse(
      abs(mean_accuracy) > 1e-6,
      sd_accuracy / mean_accuracy*100,
      NA_real_
    ),
    .groups = "drop"
  )  %>% arrange(met, Trait)

cv_summary <- GP_res %>%
  group_by(Trait) %>%
    summarise(
    mean_accuracy = mean(Cor, na.rm = TRUE),
    min = min(Cor, na.rm= TRUE), 
    max = max(Cor, na.rm = TRUE),
    sd_accuracy   = sd(Cor, na.rm = TRUE),
    n             = sum(!is.na(Cor)),
    se_accuracy   = sd_accuracy / sqrt(n),
    cv_accuracy   = ifelse(
      abs(mean_accuracy) > 1e-6,
      sd_accuracy / mean_accuracy*100,
      NA_real_
    ),
    .groups = "drop"
  )  %>% arrange(cv_accuracy)
cv_summary 

cv_summary <- GP_res %>%
  group_by(Env, met) %>%
    summarise(
    mean_accuracy = mean(Cor, na.rm = TRUE),
    min = min(Cor, na.rm= TRUE), 
    max = max(Cor, na.rm = TRUE),
    sd_accuracy   = sd(Cor, na.rm = TRUE),
    n             = sum(!is.na(Cor)),
    se_accuracy   = sd_accuracy / sqrt(n),
    cv_accuracy   = ifelse(
      abs(mean_accuracy) > 1e-6,
      sd_accuracy / mean_accuracy*100,
      NA_real_
    ),
    .groups = "drop"
  )  %>% arrange(met, Env)
cv_summary

cv_summary <- GP_res %>%
  group_by(met) %>%
    summarise(
    mean_accuracy = mean(Cor, na.rm = TRUE),
    min = min(Cor, na.rm= TRUE), 
    max = max(Cor, na.rm = TRUE),
    sd_accuracy   = sd(Cor, na.rm = TRUE),
    n             = sum(!is.na(Cor)),
    se_accuracy   = sd_accuracy / sqrt(n),
    cv_accuracy   = ifelse(
      abs(mean_accuracy) > 1e-6,
      sd_accuracy / mean_accuracy*100,
      NA_real_
    ),
    .groups = "drop"
  )  %>% arrange(met, mean_accuracy)
cv_summary %>% filter(met == "GP")


cv_summary <- GP_res %>%
  group_by(Env) %>%
    summarise(
    mean_accuracy = mean(Cor, na.rm = TRUE),
    min = min(Cor, na.rm= TRUE), 
    max = max(Cor, na.rm = TRUE),
    sd_accuracy   = sd(Cor, na.rm = TRUE),
    n             = sum(!is.na(Cor)),
    se_accuracy   = sd_accuracy / sqrt(n),
    cv_accuracy   = ifelse(
      abs(mean_accuracy) > 1e-6,
      sd_accuracy / mean_accuracy*100,
      NA_real_
    ),
    .groups = "drop"
  ) 
cv_summary[,1:2]
str(res)



anova_mod <- aov(Cor ~ Trait + Env + met + 
                   Trait:met + Env:met + Trait:Env,
                 data = GP_res)

anova_tab <- summary(anova_mod)[[1]]

anova_df <- as.data.frame(anova_tab)
anova_df$Effect <- rownames(anova_df)

# Remove residual row for % calculation
ss_total <- sum(anova_df$`Sum Sq`, na.rm = TRUE)

anova_df <- anova_df %>%
  mutate(
    perc_var = 100 * `Sum Sq` / ss_total
  )

anova_plot_df <- anova_df %>%
  filter(Effect != "Residuals") %>%
  arrange(desc(perc_var))


ggplot(anova_plot_df, aes(x = reorder(Effect, perc_var), y = perc_var)) +
  geom_col(fill = "grey40") +
  coord_flip() +
  labs(
    x = "Effect",
    y = "% Variance Explained",
    title = "ANOVA Variance Decomposition of Prediction Accuracy"
  ) +
  theme_bw() +
  theme(
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 12)
  )


eta_sq <- eta_squared(anova_mod, partial = FALSE)
ggplot(eta_sq, aes(x = reorder(Parameter, Eta2), y = Eta2 * 100)) +
  geom_col(fill = "grey40") +
  labs(
    x = "Effect",
    y = "% Variance Explained (Eta²)"
  ) +
  theme_bw()
```


```{r}
met_order<- c("GP", "MT", "DK")

res<- GP_res %>% 
  mutate( Trait= trimws(as.character(Trait)),
          Trait = fct_relevel(Trait, trait_order)) %>% 
  mutate( met= trimws(as.character(met)),
          met = fct_relevel(met, met_order)) %>% 
  mutate( Environment = trimws(as.character(Environment)),
          Environment = fct_relevel(Environment, env_order)) 



cv_stability <- 
  res %>%
  group_by(Trait, met) %>%
    summarise(
    mean_accuracy = mean(Cor, na.rm = TRUE),
    min = min(Cor, na.rm= TRUE), 
    max = max(Cor, na.rm = TRUE),
    sd_accuracy   = sd(Cor, na.rm = TRUE),
    n             = sum(!is.na(Cor)),
    se_accuracy   = sd_accuracy / sqrt(n),
    cv_accuracy   = ifelse(
      (mean_accuracy) > 1e-1,
      sd_accuracy / mean_accuracy*100,
      NA_real_
    ),
    .groups = "drop"
  )  %>%  arrange( cv_accuracy)

mean(cv_stability$cv_accuracy, na.rm=TRUE)


cv_sum<-  res %>%
  group_by(Trait, met, Env) %>%
    summarise(
    mean_accuracy = mean(Cor, na.rm = TRUE),
    min = min(Cor, na.rm= TRUE), 
    max = max(Cor, na.rm = TRUE),
    sd_accuracy   = sd(Cor, na.rm = TRUE),
    n             = sum(!is.na(Cor)),
    se_accuracy   = sd_accuracy / sqrt(n),
    cv_accuracy   = ifelse(
      (mean_accuracy) > 1e-1,
      sd_accuracy / mean_accuracy*100,
      NA_real_
    ),
    .groups = "drop"
  )  

#####################################################################
cv_wide <- cv_sum %>%
  select(Trait, met, mean_accuracy, Env) %>%
  pivot_wider(names_from = met, values_from = mean_accuracy) 

cv_diff <- cv_wide %>%
  mutate(
    MT_vs_GP = MT - GP,
    DK_vs_GP = DK - GP,
    DK_vs_MT = DK - MT
  )
 

cv_diff <- cv_diff %>%
  mutate(
    MT_vs_GP_pct = (MT - GP) / abs(GP),
    DK_vs_GP_pct = (DK - GP) / abs(GP), 
    DK_vs_MT_pct = (DK - MT) / abs(MT)
  )

cv_summary <- cv_diff %>%
  group_by(Trait) %>%
  summarise(
    mean_MT_vs_GP = mean(MT_vs_GP, na.rm = TRUE),
    mean_DK_vs_GP = mean(DK_vs_GP, na.rm = TRUE),
    mean_DK_vs_MT = mean(DK_vs_MT, na.rm = TRUE), 
    mean_MT_vs_GPpct = mean(MT_vs_GP_pct, na.rm = TRUE),
    mean_DK_vs_GPpct = mean(DK_vs_GP_pct, na.rm = TRUE),
    mean_DK_vs_MTpct = mean(DK_vs_MT_pct, na.rm = TRUE),
    mean_GP = mean(GP, na.rm = TRUE),
    mean_MT = mean(MT, na.rm = TRUE),
    mean_DK = mean(DK, na.rm = TRUE), 
  ) %>% arrange(mean_DK_vs_GP)

cv_summary[,-5:-7]

mean(cv_summary$mean_DK_vs_GPpct)

mean(cv_summary$mean_MT_vs_GPpct)

```



# **8. Plot all GP together **

```{r}
str(GP_res)
cort<- c("NDVI" ,"PC1", "PC2",  "FPC2", "PC3",  "FPC3")
#cort<- c("FPC1")


#table(GP_res1$Cor_Trait)
GP_res1<- res %>%filter(!Cor_Trait %in% cort) #%>% filter(!type == "sel")



ggplot(GP_res1, aes(x = Environment, y = Cor, fill = met)) +
  geom_boxplot() +
   facet_grid(met~Trait, scales = "free") +
  labs(
    title = " ",
    y = "Prediction Accuracy",
    x = "Model Method"
  ) +
  scale_fill_manual(values = c("cornsilk3", "palegreen2" ,  "wheat2")) +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text  = element_text(size = 10, angle = 90),
    legend.position = "bottom"
  )

ggplot(GP_res1, aes(x = met, y = Cor, fill = met)) +
  geom_boxplot(
    width = 0.7,
    outlier.shape = NA,
    color = "black",        # strong outline for contrast
    alpha = 0.9
  ) +
  facet_grid(~Trait, scales = "free") +
  labs(
    title = " ",
    y = "CV Prediction Accuracy",
    x = "Model Method"
  ) + 
  coord_cartesian(ylim = c(-0.2, 1)) +
   scale_fill_manual(values = pal_combined) +
  theme(
    strip.text = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 16),
    axis.text  = element_text(size = 16, angle = 90),
    legend.position = "bottom", 
    legend.text = element_text(size = 14))


```

#


# **9. Plot PP with H <-GP **
```{r, eval=FALSE}
load("heritabilites.Rdata")
#load("analysis/heritabilites.Rdata")
str(H2ag)
H2s<- rbind(H2ag, H2mq)

H2s<-H2s %>%
  mutate(Trait = recode(Trait, !!!trait_map))

H2s <- H2s  %>%  
  mutate( Environment = trimws(as.character(Environment)),
          Environment = fct_relevel(Environment, env_order)) %>% 
  mutate( Trait= trimws(as.character(Trait)),
          Trait = fct_relevel(Trait, trait_order))



str(H2s)
str(PP_LM_MM_accuracy)

PP_LM_MM<- left_join(PP_LM_MM_accuracy, H2s, by= c("Trait", "Environment"))

PP_LM_MM$Cor<- PP_LM_MM$Accuracy * (PP_LM_MM$H2)



########################
PP_LM_MM1<- PP_LM_MM  %>% filter(BLUE == "One") #%>% filter(Trait %in% agtrait) 

ggplot(PP_LM_MM1, aes(x = Trait, y = Cor, fill = Trait)) +
  geom_boxplot() +
   facet_grid(~Environment, scales = "free") +
  labs(
    title = " ",
    y = "CV Prediction Accuracy",
    x = "Model Method"
  ) +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text  = element_text(size = 10, angle = 90),
    legend.position = "bottom"
  )


ggplot(PP_LM_MM1, aes(x = met, y = Cor, fill = met)) +
  geom_boxplot() +
   facet_grid(~Trait, scales = "free") +
  labs(
    title = " ",
    y = "CV Prediction Accuracy",
    x = "Model Method"
  ) +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text  = element_text(size = 10, angle = 90),
    legend.position = "bottom"
  )

PP_LM_MM1<- PP_LM_MM  %>% filter(BLUE == "One") 
table(PP_LM_MM$met)

ggplot(PP_LM_MM1, aes(x = Trait, y = Cor, fill = Trait)) +
  geom_boxplot() +
  labs(
    title = " ",
    y = "CV Prediction Accuracy",
    x = "Model Method"
  ) +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text  = element_text(size = 10, angle = 90),
    legend.position = "bottom"
  )




```



