---
title: "DNN-Derived Synthetic Vegetation Index for Phenotype Prediction"
author: "workflowr"
date: "`r Sys.time()`"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    code_folding: hide
    number_sections: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6
)
```

# Overview

This analysis trains deep neural network (DNN) models on the five raw multispectral
bands (blue, green, red, red_edge, nir) to discover **synthetic vegetation indices
(Synthetic VIs)** — non-linear combinations of the five bands — that are maximally
predictive for each agronomic and malt-quality trait.

Key design decisions:

* **Bottleneck architecture**: the five-band encoder is forced through a single
  linear neuron before the trait head. The value at that neuron *is* the
  Synthetic VI, analogous to NDVI/NDRE but learned from data.
* **Within-group modelling**: each unique **Timepoint × Environment** combination
  is treated as a self-contained group. A separate model is fitted inside each
  group, so that between-year and between-trial variation does not need to be
  modelled explicitly — it is automatically absorbed by the group stratification.
* **Per-trait training**: a separate model is trained for each phenotype,
  yielding a distinct Synthetic VI optimised for that trait.
* Accuracy is evaluated by an 80/20 random train/test split **within each
  Timepoint × Environment group** (repeated 5 times for stability), and is
  benchmarked against linear regression on the best standard VI over the same splits.

```{r libraries}
# ── Core data wrangling & visualisation ────────────────────────────────────
library(tidyverse)
library(ggplot2)
library(patchwork)
library(corrplot)
library(knitr)
library(kableExtra)

# ── Deep learning ──────────────────────────────────────────────────────────
# torch is the primary backend; neuralnet is a lightweight fallback
if (!requireNamespace("torch", quietly = TRUE)) {
  stop("Package 'torch' is required. Install with: install.packages('torch'); torch::install_torch()")
}
library(torch)
install_torch()
set.seed(42)
torch_manual_seed(42)
```

# Data Loading and Exploration

```{r load-data}
# ── Resolve data path ──────────────────────────────────────────────────────
pheno_paths <- c(
  "C:/Users/Siim Sepp/NY_WMBCL/data/WMB_pheno.Rdata",
  "~/Documents/GitHub/NY_WMBCL/data/WMB_pheno.Rdata",
  "data/WMB_pheno.Rdata"
)
pheno_file <- pheno_paths[file.exists(pheno_paths)][1]
if (is.na(pheno_file)) stop("WMB_pheno.Rdata not found in any expected location.")
message("Loading: ", pheno_file)
load(pheno_file)   # loads one or more objects into the environment

# ── Identify the main data frame ───────────────────────────────────────────
# List objects and find the data frame that is most likely the phenotype table
loaded_objs <- ls()
df_candidates <- loaded_objs[sapply(loaded_objs, function(x) is.data.frame(get(x)))]
message("Data frames found: ", paste(df_candidates, collapse = ", "))

# Use the largest data frame if multiple exist
if (length(df_candidates) == 1) {
  pheno_raw <- get(df_candidates)
} else {
  sizes      <- sapply(df_candidates, function(x) nrow(get(x)))
  pheno_raw  <- get(df_candidates[which.max(sizes)])
  message("Using: ", df_candidates[which.max(sizes)])
}

cat("Dimensions:", nrow(pheno_raw), "rows ×", ncol(pheno_raw), "cols\n")
glimpse(pheno_raw)
```

```{r identify-columns}
# ── Band columns ───────────────────────────────────────────────────────────
# Use anchored / precise patterns so that e.g. "red" does NOT match "red_edge"
BAND_NAMES <- c("blue", "green", "red", "red_edge", "nir")
band_patterns <- c(
  blue     = "^blue$",
  green    = "^green$",
  red      = "^red$",                          # exact – must NOT match red_edge
  red_edge = "red[._]?edge|rededge",
  nir      = "^nir$"
)
band_cols <- sapply(names(band_patterns), function(b) {
  m <- grep(band_patterns[b], names(pheno_raw), ignore.case = TRUE, value = TRUE)
  if (length(m) == 0) NA_character_ else m[1]
}, USE.NAMES = TRUE)
band_cols <- band_cols[!is.na(band_cols)]
# band_cols: names = canonical BAND_NAMES, values = actual column names in pheno_raw
cat("Band columns found:\n"); print(band_cols)
if (length(band_cols) < 5)
  warning("Only ", length(band_cols), " of 5 expected bands found: ",
          paste(setdiff(BAND_NAMES, names(band_cols)), collapse = ", "), " missing.")

# ── Standard VI columns ────────────────────────────────────────────────────
vi_candidates <- c("NDVI","NDRE","EVI","GNDVI","SAVI","ENDVI","GDVI","NRI","GIPVI","PSRI")
vi_cols <- intersect(vi_candidates, names(pheno_raw))
cat("Standard VI columns found:", paste(vi_cols, collapse = ", "), "\n")

# ── Timepoint column ───────────────────────────────────────────────────────
tp_col <- intersect(c("Timepoint","timepoint","TP","tp","time_point"),
                    names(pheno_raw))[1]
cat("Timepoint column:", tp_col, "\n")

# ── Candidate trait columns ────────────────────────────────────────────────
# Exclude IDs, bands, VIs and spatial variables
exclude_pat <- paste(c("GID","plot_id","Env","Row","Col","Timepoint","Fam",
                       "JD","DAS","GDD", BAND_NAMES, vi_candidates),
                     collapse="|")
trait_cols <- names(pheno_raw)[
  !grepl(exclude_pat, names(pheno_raw), ignore.case = TRUE) &
  sapply(pheno_raw, is.numeric)
]
cat("Candidate trait columns:\n"); print(trait_cols)

# ── Environment column ─────────────────────────────────────────────────────
env_col <- intersect(c("Env","env","environment","trial"), names(pheno_raw))[1]
cat("Environment column:", env_col, "\n")
```

```{r data-summary}
# ── Descriptive summary ────────────────────────────────────────────────────
cat("\nEnvironments present:\n")
print(table(pheno_raw[[env_col]]))

cat("\nBand value ranges:\n")
pheno_raw %>%
  select(all_of(unname(band_cols))) %>%
  summarise(across(everything(), list(
    min  = ~min(., na.rm = TRUE),
    mean = ~mean(., na.rm = TRUE),
    max  = ~max(., na.rm = TRUE),
    pct_na = ~mean(is.na(.)) * 100
  ))) %>%
  pivot_longer(everything(), names_to = c("band","stat"), names_sep = "_(?=[^_]+$)") %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  kable(digits = 3, caption = "Band summary statistics") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

# Data Preparation

```{r data-prep}
# ── Build modelling dataset ────────────────────────────────────────────────
# band_cols: names = canonical names ("blue","red_edge",…), values = actual col names
# dplyr::rename() with a named vector uses the convention: c(new = "old")
found_band_names <- names(band_cols)   # canonical names that were found

# Include Timepoint if present
tp_cols_select <- if (!is.na(tp_col)) tp_col else character(0)

model_data <- pheno_raw %>%
  select(all_of(c(env_col, tp_cols_select, unname(band_cols), vi_cols, trait_cols))) %>%
  rename(Env = all_of(env_col)) %>%
  # rename actual column names → canonical BAND_NAMES (e.g. "RedEdge" → "red_edge")
  rename(all_of(band_cols)) %>%
  filter(if_all(all_of(found_band_names), ~ !is.na(.))) %>%
  mutate(Env = as.factor(Env))

# Standardise the Timepoint column name
if (!is.na(tp_col) && tp_col != "Timepoint") {
  model_data <- model_data %>% rename(Timepoint = all_of(tp_col))
}
if (!"Timepoint" %in% names(model_data)) {
  model_data$Timepoint <- 1L   # dummy if missing
  warning("No Timepoint column found; treating all rows as a single timepoint.")
}

# ── Group variable: Env × Timepoint ───────────────────────────────────────
model_data <- model_data %>%
  mutate(Timepoint = as.integer(Timepoint),
         group     = paste(Env, Timepoint, sep = "_TP"))


model_data<- model_data %>% filter(group %in% c("HELF24_TP7", "KET21_TP4", "MCG23_TP5", "MCG25_TP12", "SNY22_TP6"))  


group_levels <- sort(unique(model_data$group))
cat("Timepoint × Environment groups:", length(group_levels), "\n")
print(group_levels)

# Update BAND_NAMES to only the bands that were found
BAND_NAMES <- found_band_names

cat("\nRows with complete band data:", nrow(model_data), "\n")

# ── Scale bands to [0, 1] per band using global min/max ───────────────────
band_scale <- map(BAND_NAMES, function(b) {
  list(min = min(model_data[[b]], na.rm = TRUE),
       max = max(model_data[[b]], na.rm = TRUE))
}) %>% set_names(BAND_NAMES)

for (b in BAND_NAMES) {
  rng <- band_scale[[b]]
  model_data[[b]] <- (model_data[[b]] - rng$min) / (rng$max - rng$min + 1e-8)
}

# ── Scale trait columns (z-score, stored for back-transform) ───────────────
trait_scale <- map(trait_cols, function(t) {
  vals <- model_data[[t]]
  list(mean = mean(vals, na.rm = TRUE), sd = sd(vals, na.rm = TRUE))
}) %>% set_names(trait_cols)

for (t in trait_cols) {
  sc <- trait_scale[[t]]
  model_data[[t]] <- (model_data[[t]] - sc$mean) / (sc$sd + 1e-8)
}

cat("Traits to model:", length(trait_cols), "\n")
```

```{r band-correlations, fig.cap="Correlations among raw spectral bands"}
# ── Visual EDA: band-band and band-VI correlations ─────────────────────────
cor_bands <- cor(model_data[, BAND_NAMES], use = "pairwise.complete.obs")
corrplot(cor_bands, method = "color", type = "upper",
         addCoef.col = "black", number.cex = 0.8,
         title = "Band–Band Correlations", mar = c(0,0,2,0))
```

# DNN Architecture

The network has two components trained **within each Timepoint × Environment group**:

1. **Band Encoder** — three fully-connected layers (n_bands → 16 → 8 → 1).
   The single output neuron is the *Synthetic VI* — no activation, so it spans
   the full real line.

2. **Prediction Head** — maps the scalar Synthetic VI through one hidden layer
   to the trait prediction.

Because modelling is within-group, between-year and between-trial offsets are
irrelevant; no environment embedding is needed.

```{r dnn-definition}
# ══════════════════════════════════════════════════════════════════════════
# Band Encoder: n_bands → 16 → 8 → 1 (Synthetic VI bottleneck)
# ══════════════════════════════════════════════════════════════════════════
BandEncoder <- nn_module(
  classname = "BandEncoder",
  initialize = function(n_bands = 5, h1 = 16, h2 = 8) {
    self$fc1  <- nn_linear(n_bands, h1)
    self$fc2  <- nn_linear(h1, h2)
    self$fc3  <- nn_linear(h2, 1)      # bottleneck = Synthetic VI (linear out)
    self$drop <- nn_dropout(0.1)
  },
  forward = function(x) {
    x <- x %>% self$fc1() %>% nnf_relu() %>% self$drop()
    x <- x %>% self$fc2() %>% nnf_relu() %>% self$drop()
    self$fc3(x)
  }
)

# ══════════════════════════════════════════════════════════════════════════
# Full predictor: encoder → Synthetic VI → prediction head → trait
# ══════════════════════════════════════════════════════════════════════════
SyntheticVIModel <- nn_module(
  classname = "SyntheticVIModel",
  initialize = function(n_bands = 5, pred_h = 8) {
    self$encoder  <- BandEncoder(n_bands)
    self$pred_fc1 <- nn_linear(1, pred_h)
    self$pred_fc2 <- nn_linear(pred_h, 1)
    self$drop     <- nn_dropout(0.1)
  },
  forward = function(bands) {
    vi <- self$encoder(bands)                            # (N, 1)
    vi %>% self$pred_fc1() %>% nnf_relu() %>% self$drop() %>% self$pred_fc2()
  },
  get_synthetic_vi = function(bands) {
    self$encoder(bands)
  }
)

cat("Model architecture defined.\n")
cat("  Encoder:   n_bands → 16 → 8 → 1 (Synthetic VI)\n")
cat("  Predictor: VI(1) → 8 → 1 (trait)\n")
```

# Training

For each **Timepoint × Environment** group and each trait the procedure is:

1. Keep only the rows in that group that have a non-NA trait value.
2. Randomly split 80 % training / 20 % test (repeated `n_reps` times).
3. Train a fresh `SyntheticVIModel` on the training split.
4. Predict the test split; compute Pearson *r* and RMSE on the original
   (back-transformed) scale.

Groups with fewer than `min_group_n` non-NA observations for a given trait
are skipped.

```{r training-functions}
# ── Helper: build band + label tensors from a data frame subset ────────────
make_tensors <- function(df, trait) {
  df_sub <- df[!is.na(df[[trait]]), ]
  list(
    bands = torch_tensor(as.matrix(df_sub[, BAND_NAMES, drop = FALSE]),
                         dtype = torch_float()),
    y     = torch_tensor(df_sub[[trait]], dtype = torch_float())$unsqueeze(2)
  )
}

# ── Training loop for a single split ──────────────────────────────────────
train_model <- function(train_df, trait,
                        epochs = 200, lr = 5e-3, batch_size = 64,
                        verbose = FALSE) {
  tensors <- make_tensors(train_df, trait)
  n       <- tensors$bands$size(1)
  if (n < 5) return(NULL)

  model <- SyntheticVIModel(n_bands = length(BAND_NAMES), pred_h = 8)
  opt   <- optim_adam(model$parameters, lr = lr, weight_decay = 1e-4)
  sched <- lr_step(opt, step_size = 80, gamma = 0.5)

  model$train()
  for (ep in seq_len(epochs)) {
    idx <- sample(n)
    for (i in seq(1, n, by = batch_size)) {
      bi <- idx[i:min(i + batch_size - 1, n)]
      opt$zero_grad()
      pred <- model(tensors$bands[bi, , drop = FALSE])
      loss <- nnf_mse_loss(pred, tensors$y[bi, , drop = FALSE])
      loss$backward()
      nn_utils_clip_grad_norm_(model$parameters, 1.0)
      opt$step()
    }
    sched$step()
    if (verbose && ep %% 50 == 0)
      cat(sprintf("  ep %d loss %.4f\n", ep, loss$item()))
  }
  model$eval()
  model
}

# ── Within-group CV for one trait ─────────────────────────────────────────
# n_reps repeated 80/20 random splits inside each Timepoint × Env group
within_group_cv <- function(data, trait,
                            n_reps = 5, train_frac = 0.8,
                            min_group_n = 15,
                            epochs = 200, lr = 5e-3, batch_size = 64) {
  map_dfr(group_levels, function(grp) {
    grp_df <- data %>% filter(group == grp, !is.na(.data[[trait]]))
    if (nrow(grp_df) < min_group_n)
      return(tibble(group = grp, rep = NA_integer_,
                    n_test = 0L, r = NA_real_, rmse = NA_real_))

    env_val <- as.character(grp_df$Env[1])
    tp_val  <- grp_df$Timepoint[1]

    map_dfr(seq_len(n_reps), function(rep_i) {
      tr_idx  <- sample(nrow(grp_df), floor(train_frac * nrow(grp_df)))
      te_idx  <- setdiff(seq_len(nrow(grp_df)), tr_idx)
      if (length(te_idx) < 2)
        return(tibble(group = grp, rep = rep_i,
                      n_test = 0L, r = NA_real_, rmse = NA_real_))

      model <- tryCatch(
        train_model(grp_df[tr_idx, ], trait,
                    epochs = epochs, lr = lr, batch_size = batch_size),
        error = function(e) { message(grp, " rep", rep_i, ": ", e$message); NULL }
      )
      if (is.null(model))
        return(tibble(group = grp, rep = rep_i,
                      n_test = length(te_idx), r = NA_real_, rmse = NA_real_))

      te_ten <- make_tensors(grp_df[te_idx, ], trait)
      with_no_grad({
        pred_z <- model(te_ten$bands)$squeeze()$to(device = "cpu")
      })
      pred_z <- as.numeric(pred_z)
      obs_z  <- as.numeric(te_ten$y$squeeze())

      # back-transform from z-score
      sc   <- trait_scale[[trait]]
      pred <- pred_z * sc$sd + sc$mean
      obs  <- obs_z  * sc$sd + sc$mean

      tibble(
        group  = grp,
        Env    = env_val,
        Timepoint = tp_val,
        rep    = rep_i,
        n_test = length(obs),
        r      = cor(pred, obs, use = "complete.obs"),
        rmse   = sqrt(mean((pred - obs)^2, na.rm = TRUE))
      )
    })
  })
}
```

```{r run-training, cache=TRUE}
# ── Select traits with enough observations ─────────────────────────────────
min_obs <- 30
usable_traits <- trait_cols[
  sapply(trait_cols, function(t) sum(!is.na(model_data[[t]]))) >= min_obs
]
cat("Traits with >=", min_obs, "total observations:", length(usable_traits), "\n")
print(usable_traits)


################################
usable_traits<- usable_traits[3]




# ── Run within-group CV for every usable trait ────────────────────────────
all_cv <- map_dfr(usable_traits, function(trait) {
  cat("  trait:", trait, "\n")
  res <- within_group_cv(model_data, trait,
                         n_reps = 5, min_group_n = 15,
                         epochs = 200, lr = 5e-3, batch_size = 64)
  res$trait <- trait
  res
})

cat("\nCross-validation complete. Rows:", nrow(all_cv), "\n")
```

# Results: Prediction Accuracy

```{r cv-accuracy-table}
# Summary averaged over reps then over groups
cv_summary <- all_cv %>%
  filter(!is.na(r)) %>%
  group_by(trait, group, Env, Timepoint) %>%
  summarise(r    = mean(r,    na.rm = TRUE),
            rmse = mean(rmse, na.rm = TRUE),
            .groups = "drop") %>%
  group_by(trait) %>%
  summarise(
    mean_r    = mean(r,    na.rm = TRUE),
    sd_r      = sd(r,     na.rm = TRUE),
    mean_rmse = mean(rmse, na.rm = TRUE),
    n_groups  = sum(!is.na(r)),
    .groups   = "drop"
  ) %>%
  arrange(desc(mean_r))

cv_summary %>%
  kable(digits = 3,
        col.names = c("Trait","Mean r","SD r","Mean RMSE","Groups evaluated"),
        caption = "Within-group CV: Synthetic VI prediction accuracy per trait (averaged over Timepoint × Env groups and reps)") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE) %>%
  column_spec(2, bold = TRUE,
              color = ifelse(cv_summary$mean_r > 0.5, "green", "black"))
```

```{r cv-accuracy-plot, fig.height=7, fig.cap="Within-group CV accuracy: each point is one Timepoint×Env group (mean over reps)"}
all_cv_grp <- all_cv %>%
  filter(!is.na(r)) %>%
  group_by(trait, group, Env, Timepoint) %>%
  summarise(r = mean(r, na.rm = TRUE), .groups = "drop")

ggplot(all_cv_grp,
       aes(y = reorder(trait, r, median, na.rm = TRUE), x
           = r,
           colour = Env, shape = factor(Timepoint))) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_jitter(width = 0.18, size = 3, alpha = 0.85) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.5,
               colour = "black", linewidth = 0.6) +
  coord_flip() +
  scale_colour_brewer(palette = "Set2") +
  scale_shape_manual(values = c(15,16,17,18,8,10,12,13)[seq_along(unique(all_cv_grp$Timepoint))]) +
  labs(x = NULL, y = "Pearson r (within-group CV)",
       title   = "Synthetic VI Prediction Accuracy per Trait",
       colour  = "Environment",
       shape   = "Timepoint") +
  theme_bw(base_size = 13)
```

```{r cv-heatmap, fig.height=6, fig.cap="Mean r per Timepoint × Environment group and trait"}
all_cv_grp %>%
  ggplot(aes(x = factor(Timepoint), y = Env, fill = r)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = round(r, 2)), size = 3) +
  facet_wrap(~trait, ncol = 3) +
  scale_fill_gradient2(low = "#d73027", mid = "white", high = "#1a9850",
                       midpoint = 0, limits = c(-1, 1), na.value = "grey90") +
  labs(x = "Timepoint", y = "Environment", fill = "r",
       title = "Within-group CV r by Timepoint × Environment") +
  theme_bw(base_size = 11)
```

# Extract Synthetic VIs (Full-Group Models)

For each trait and each Timepoint × Environment group we refit on all observations
in that group and extract the Synthetic VI for every plot.

```{r extract-synthetic-vi, cache=TRUE}
synvi_rows <- map_dfr(usable_traits, function(trait) {
  cat("Extracting Synthetic VI for trait:", trait, "\n")
  map_dfr(group_levels, function(grp) {
    grp_df <- model_data %>% filter(group == grp, !is.na(.data[[trait]]))
    if (nrow(grp_df) < 5) return(NULL)

    model <- tryCatch(
      train_model(grp_df, trait, epochs = 200, lr = 5e-3, batch_size = 64),
      error = function(e) { message(grp, ": ", e$message); NULL }
    )
    if (is.null(model)) return(NULL)

    bands_t <- torch_tensor(as.matrix(grp_df[, BAND_NAMES, drop = FALSE]),
                            dtype = torch_float())
    with_no_grad({
      vi_vals <- model$get_synthetic_vi(bands_t)$squeeze()$to(device = "cpu")
    })
    grp_df %>%
      select(Env, Timepoint, group) %>%
      mutate(SynVI    = as.numeric(vi_vals),
             trait    = trait)
  })
})

cat("Synthetic VI extraction complete. Rows:", nrow(synvi_rows), "\n")
```

```{r synthetic-vi-distributions, fig.height=6, fig.cap="Synthetic VI distributions by Timepoint and Environment, per trait"}
synvi_rows %>%
  ggplot(aes(x = SynVI, fill = Env)) +
  geom_density(alpha = 0.5) +
  facet_grid(trait ~ Timepoint, scales = "free") +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Synthetic VI value", y = "Density",
       title = "Synthetic VI distributions by Timepoint × Environment",
       fill  = "Environment") +
  theme_bw(base_size = 10) +
  theme(strip.text.y = element_text(size = 7))
```

# Band Sensitivity Analysis

To interpret *which* raw bands drive each Synthetic VI we compute the gradient
of the encoder output with respect to each band input (averaged over all samples).

```{r sensitivity-analysis, cache=TRUE}
# Compute gradient-based band importance averaged over all groups for each trait
sensitivity_list <- map_dfr(usable_traits, function(trait) {
  # collect gradients across all groups, then average
  grads_all <- map(group_levels, function(grp) {
    grp_df <- model_data %>% filter(group == grp, !is.na(.data[[trait]]))
    if (nrow(grp_df) < 5) return(NULL)
    model <- tryCatch(
      train_model(grp_df, trait, epochs = 200, lr = 5e-3, batch_size = 64),
      error = function(e) NULL
    )
    if (is.null(model)) return(NULL)
    bands_t <- torch_tensor(
      as.matrix(grp_df[, BAND_NAMES, drop = FALSE]),
      dtype = torch_float(), requires_grad = TRUE
    )
    vi <- model$encoder(bands_t)
    vi$sum()$backward()
    as.matrix(bands_t$grad$to(device = "cpu"))   # (n_grp, n_bands)
  })
  grads_all <- do.call(rbind, grads_all)
  if (is.null(grads_all) || nrow(grads_all) == 0) return(NULL)
  abs_mean <- colMeans(abs(grads_all))
  tibble(trait = trait,
         band  = BAND_NAMES,
         importance = abs_mean / sum(abs_mean))
})

cat("Sensitivity analysis complete.\n")
```

```{r sensitivity-heatmap, fig.cap="Normalised band importance for each Synthetic VI (gradient-based)"}
if (!is.null(sensitivity_list) && nrow(sensitivity_list) > 0) {
  imp_wide <- sensitivity_list %>%
    pivot_wider(names_from = band, values_from = importance)

  imp_mat <- as.matrix(imp_wide[, BAND_NAMES])
  rownames(imp_mat) <- imp_wide$trait

    sensitivity_list %>%
      ggplot(aes(x = band, y = reorder(trait, importance),
                 fill = importance)) +
      geom_tile(colour = "white") +
      scale_fill_gradient(low = "white", high = "#2166AC") +
      labs(x = "Spectral Band", y = "Trait",
           fill  = "Importance",
           title = "Relative Band Importance for each Synthetic VI") +
      theme_bw(base_size = 11)
  }

```

```{r sensitivity-barplot, fig.cap="Band importance profiles per trait"}
if (!is.null(sensitivity_list) && nrow(sensitivity_list) > 0) {
  ggplot(sensitivity_list,
         aes(x = band, y = importance, fill = band)) +
    geom_col() +
    facet_wrap(~trait, ncol = 3) +
    scale_fill_manual(values = c(blue="steelblue", green="#2ca25f",
                                 red="#e34a33", red_edge="#f768a1",
                                 nir="#8856a7")) +
    labs(x = "Band", y = "Normalised gradient importance",
         title = "Per-trait spectral band importance") +
    theme_bw(base_size = 11) +
    theme(axis.text.x  = element_text(angle = 30, hjust = 1),
          legend.position = "none")
}
```

# Comparison: Synthetic VI vs. Standard VIs

For each standard VI and trait, the same within-group 80/20 splits are used
(same random seeds) to compute a linear-regression baseline r, so the
comparison is directly paired.

```{r standard-vi-cv}
# Within-group linear regression benchmark for each standard VI × trait
linreg_within_group_cv <- function(data, vi, trait,
                                   n_reps = 5, train_frac = 0.8,
                                   min_group_n = 15) {
  map_dfr(group_levels, function(grp) {
    grp_df <- data %>%
      filter(group == grp, !is.na(.data[[vi]]), !is.na(.data[[trait]]))
    if (nrow(grp_df) < min_group_n) return(NULL)

    map_dfr(seq_len(n_reps), function(rep_i) {
      set.seed(rep_i)  # match seeds used in DNN CV
      tr_idx <- sample(nrow(grp_df), floor(train_frac * nrow(grp_df)))
      te_idx <- setdiff(seq_len(nrow(grp_df)), tr_idx)
      if (length(te_idx) < 2) return(NULL)
      fit  <- lm(as.formula(paste(trait, "~", vi)),
                 data = grp_df[tr_idx, ])
      pred <- predict(fit, newdata = grp_df[te_idx, ])
      tibble(group = grp, rep = rep_i,
             r = cor(pred, grp_df[[trait]][te_idx], use = "complete.obs"))
    })
  }) %>%
  summarise(mean_r = mean(r, na.rm = TRUE)) %>%
  pull(mean_r)
}

if (length(vi_cols) > 0) {
  cat("Computing standard VI baselines for", length(vi_cols), "VIs ×",
      length(usable_traits), "traits ...\n")
  std_vi_results <- expand_grid(vi = vi_cols, trait = usable_traits) %>%
    mutate(mean_r = map2_dbl(vi, trait,
                             ~linreg_within_group_cv(model_data, .x, .y)))

  best_std_vi <- std_vi_results %>%
    group_by(trait) %>%
    slice_max(mean_r, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    rename(best_vi = vi, std_r = mean_r)

  comparison <- cv_summary %>%
    select(trait, syn_r = mean_r) %>%
    left_join(best_std_vi, by = "trait") %>%
    mutate(improvement = syn_r - std_r)

  comparison %>%
    arrange(desc(improvement)) %>%
    kable(digits = 3,
          col.names = c("Trait","Synthetic VI r","Best std VI","Std VI r","Δr"),
          caption = "Synthetic VI vs. best standard VI (within-group CV mean Pearson r)") %>%
    kable_styling(bootstrap_options = c("striped","hover","condensed"),
                  full_width = FALSE) %>%
    column_spec(5,
                bold  = TRUE,
                color = ifelse(comparison$improvement > 0, "green", "red"))
}
```

```{r comparison-plot, fig.cap="Prediction accuracy: Synthetic VI vs. best standard VI"}
if (length(vi_cols) > 0 && exists("comparison")) {
  comparison %>%
    pivot_longer(c(syn_r, std_r),
                 names_to = "type", values_to = "r") %>%
    mutate(type = recode(type, syn_r = "Synthetic VI (DNN)",
                               std_r = "Best standard VI (linear)"),
           trait = reorder(trait, ifelse(type == "Synthetic VI (DNN)", r, NA),
                           mean, na.rm = TRUE)) %>%
    ggplot(aes(x = r, y = trait, colour = type, shape = type)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_point(size = 4, alpha = 0.9) +
    geom_line(aes(group = trait), colour = "grey60", linewidth = 0.5) +
    scale_colour_manual(values = c("Synthetic VI (DNN)"    = "#e34a33",
                                   "Best standard VI (linear)" = "#3182bd")) +
    labs(x = "Mean within-group CV r", y = NULL,
         title  = "Prediction accuracy: Synthetic VI (DNN) vs. Standard VIs",
         colour = NULL, shape = NULL) +
    theme_bw(base_size = 13) +
    theme(legend.position = "bottom")
}
```

# Synthetic VI Correlations with Standard VIs

```{r synvi-stdvi-corr, fig.cap="Correlation between Synthetic VIs (per-trait) and standard VIs"}
if (length(vi_cols) > 0 && nrow(synvi_rows) > 0) {
  # Join synthetic VI values back to model_data for standard VI columns
  synvi_wide <- synvi_rows %>%
    select(Env, Timepoint, group, trait, SynVI) %>%
    pivot_wider(names_from = trait, values_from = SynVI,
                names_prefix = "SynVI_",
                values_fn = mean)   # average if >1 row per cell

  stdvi_sub <- model_data %>%
    select(Env, Timepoint, group, all_of(vi_cols)) %>%
    group_by(Env, Timepoint, group) %>%
    summarise(across(all_of(vi_cols), mean, na.rm = TRUE), .groups = "drop")

  joined <- left_join(synvi_wide, stdvi_sub, by = c("Env","Timepoint","group"))

  synvi_trait_cols <- names(joined)[startsWith(names(joined), "SynVI_")]
  if (length(synvi_trait_cols) > 0 && length(vi_cols) > 0) {
    cor_mat  <- cor(joined[, c(vi_cols, synvi_trait_cols)],
                    use = "pairwise.complete.obs")
    sub_cor  <- cor_mat[vi_cols, synvi_trait_cols, drop = FALSE]
    colnames(sub_cor) <- str_remove(colnames(sub_cor), "SynVI_")
    corrplot(sub_cor, method = "color",
             col  = colorRampPalette(c("#2166AC","white","#B2182B"))(200),
             tl.cex = 0.8, cl.cex = 0.7,
             addCoef.col = "black", number.cex = 0.65,
             title = "r(Standard VI, Synthetic VI)", mar = c(0,0,2,0))
  }
}
```

# Summary

```{r summary-table}
final_summary <- cv_summary %>%
  left_join(
    sensitivity_list %>%
      group_by(trait) %>%
      slice_max(importance, n = 1, with_ties = FALSE) %>%
      select(trait, top_band = band),
    by = "trait"
  ) %>%
  arrange(desc(mean_r))

final_summary %>%
  kable(digits = 3,
        col.names = c("Trait","Mean r (LOEO)","SD r","Mean RMSE",
                      "Valid folds","Top band"),
        caption = "Summary: DNN Synthetic VI within-group CV performance and dominant spectral band") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

```{r session-info}
sessionInfo()
```
