---
title: "DNN Synthetic VI — Results Summary"
subtitle: "NY Winter Malting Barley · Trait: yield_kgha"
author: "workflowr"
date: "`r Sys.time()`"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    code_folding: hide
    number_sections: false
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 4.5)
library(tidyverse)
library(knitr)
library(kableExtra)
library(patchwork)
```

## Overview

A shallow bottleneck DNN (5 bands → 16 → 8 → **1 Synthetic VI** → 8 → 1)
was trained to predict **yield (kg/ha)** from five raw multispectral bands
(Blue, Green, Red, RedEdge, NIR) at one pre-selected optimal flight timepoint
per environment.
Five environments were used: **HELF24, KET21, MCG23, MCG25, SNY22**.

Two cross-validation scenarios were evaluated, both splitting by GID
(genotype ID) to prevent data leakage, with scalers fit on training data only:

| Scenario | What it measures |
|---|---|
| **LOEO** — leave one environment out | Cross-environment generalisation |
| **Within-env 5-fold CV** (by GID) | Within-trial predictive ability |

---

## Cross-Validation Results

```{r cv-comparison}
cv_tbl <- tibble(
  `CV scenario`  = c("LOEO", "Within-env 5-fold"),
  `Mean r`       = c(0.389, 0.728),
  `SD r`         = c(0.277, 0.044),
  `Mean RMSE`    = c(1656.3, 635.2)
)

cv_tbl %>%
  kable(digits = 3,
        caption = "DNN Synthetic VI — overall CV accuracy for yield_kgha") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE) %>%
  column_spec(2, bold = TRUE,
              color = ifelse(cv_tbl$`Mean r` > 0.5, "forestgreen", "darkorange"))
```

**Key takeaway:** within-environment predictive ability (r = 0.728) is
substantially higher than cross-environment generalisation (LOEO r = 0.389),
indicating that the learned band combination captures within-trial yield
variation well but is sensitive to between-environment shifts in spectral
response.

---

## Per-Environment Breakdown

```{r per-env, fig.height=5}
loeo <- tibble(
  Environment = c("HELF24","KET21","MCG23","MCG25","SNY22"),
  `LOEO r`    = c(0.151, 0.550, 0.092, 0.759, 0.392),
  `LOEO RMSE` = c(1963, 759, 1469, 3023, 1067),
  `N test`    = c(312L, 466L, 504L, 232L, 424L)
)

wenv <- tibble(
  Environment   = c("HELF24","KET21","MCG23","MCG25","SNY22"),
  `Within-env r`    = c(0.532, 0.639, 0.912, 0.776, 0.778),
  `Within-env SD r` = c(0.095, 0.044, 0.011, 0.042, 0.029),
  `Within-env RMSE` = c(700, 683, 571, 523, 700)
)

left_join(loeo, wenv, by = "Environment") %>%
  kable(digits = 3,
        caption = "Per-environment CV results — yield_kgha") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE) %>%
  column_spec(2,
              color = case_when(
                loeo$`LOEO r` >= 0.6  ~ "forestgreen",
                loeo$`LOEO r` >= 0.35 ~ "darkorange",
                TRUE                  ~ "firebrick"
              )) %>%
  column_spec(5,
              color = case_when(
                wenv$`Within-env r` >= 0.8 ~ "forestgreen",
                wenv$`Within-env r` >= 0.6 ~ "darkorange",
                TRUE                       ~ "firebrick"
              ))
```

```{r per-env-plot, fig.cap="Pearson r by environment for LOEO (cross-env) and within-env 5-fold CV"}
combined <- bind_rows(
  loeo %>% select(Environment, r = `LOEO r`) %>% mutate(CV = "LOEO"),
  wenv %>% select(Environment, r = `Within-env r`) %>% mutate(CV = "Within-env 5-fold")
)

ggplot(combined,
       aes(x = reorder(Environment, r, max), y = r,
           fill = CV)) +
  geom_col(position = position_dodge(0.7), width = 0.6) +
  geom_hline(yintercept = 0, linewidth = 0.4) +
  scale_fill_manual(values = c(LOEO = "#e34a33",
                               `Within-env 5-fold` = "#2c7bb6")) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  labs(x = "Environment", y = "Pearson r",
       title = "DNN Synthetic VI prediction accuracy by environment",
       fill = "CV scenario") +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom")
```

**Notable patterns:**

- **MCG23** shows the strongest within-environment signal (r = 0.912) but the
  weakest cross-environment generalisation (r = 0.092), suggesting a highly
  environment-specific spectral response in that trial year.
- **MCG25** delivers the best LOEO accuracy (r = 0.759) alongside a high
  within-env r (0.776), but has the largest RMSE (3 023 kg/ha) reflecting
  wide yield variation in that environment.
- **KET21** is the most balanced environment: LOEO r = 0.550, within-env r = 0.639.

---

## DNN Synthetic VI vs. Standard VIs (LOEO)

```{r vi-comparison}
vi_comp <- tibble(
  Trait        = "yield_kgha",
  `DNN Syn VI r (LOEO)` = 0.389,
  `Best std VI`         = "EVI",
  `Std VI r (LOEO)`     = 0.628,
  `Δr`                  = -0.239
)

vi_comp %>%
  kable(digits = 3,
        caption = "DNN Synthetic VI vs. best standard VI under LOEO") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE) %>%
  column_spec(5, bold = TRUE, color = "firebrick")
```

Under LOEO, the DNN Synthetic VI (r = 0.389) is outperformed by EVI
(r = 0.628, Δr = −0.239). This is consistent with the within-env vs.
cross-env gap observed above: the DNN learns environment-specific spectral
patterns that do not transfer well, whereas EVI is a fixed, calibration-free
index that generalises more robustly across environments.

---

## Band Importance

```{r band-importance, fig.height=3.5, fig.cap="Linearised band sensitivity (gradient of Synthetic VI at mean band values)"}
bands <- tibble(
  Band        = c("red","nir","blue","green","red_edge"),
  Sensitivity = c(-0.1943, 0.1435, -0.0760, 0.0305, -0.0097),
  Abs         = c( 0.1943, 0.1435,  0.0760, 0.0305,  0.0097)
)

p1 <- ggplot(bands, aes(x = reorder(Band, Abs), y = Sensitivity,
                         fill = Sensitivity > 0)) +
  geom_col(width = 0.6) +
  geom_hline(yintercept = 0, linewidth = 0.4) +
  coord_flip() +
  scale_fill_manual(values = c(`TRUE` = "#2c7bb6", `FALSE` = "#e34a33"),
                    labels = c(`TRUE` = "positive", `FALSE` = "negative")) +
  labs(x = NULL, y = "dSynVI / dBand",
       title = "Linearised sensitivity at mean band values",
       fill = "Direction") +
  theme_bw(base_size = 12) +
  theme(legend.position = "right")

p2 <- ggplot(bands, aes(x = reorder(Band, Abs), y = Abs, fill = Band)) +
  geom_col(width = 0.6) +
  coord_flip() +
  scale_fill_manual(values = c(blue = "steelblue", green = "#2ca25f",
                               red = "#e34a33", red_edge = "#f768a1",
                               nir = "#8856a7")) +
  labs(x = NULL, y = "|Sensitivity|",
       title = "Absolute importance") +
  theme_bw(base_size = 12) +
  theme(legend.position = "none")

p1 + p2
```

```{r band-table}
bands %>%
  arrange(desc(Abs)) %>%
  rename(`dVI/dBand` = Sensitivity, `|Sensitivity|` = Abs) %>%
  kable(digits = 4,
        caption = "Band sensitivity (linearised at mean; top band by gradient-based importance across all samples: red_edge)") %>%
  kable_styling(bootstrap_options = c("striped","condensed"),
                full_width = FALSE)
```

The learned Synthetic VI has a structure broadly similar to NDVI:
**NIR positive, Red negative** — the two bands with the largest absolute
sensitivity. Green contributes a small positive signal while RedEdge is
near-zero at the mean but ranks first by gradient-based importance averaged
over all samples (suggesting non-linear RedEdge sensitivity away from the mean).

The approximate linearised formula around mean band values is:

$$\text{SynVI}_{\text{yield}} \approx -0.194 \times \text{Red} + 0.144 \times \text{NIR} - 0.076 \times \text{Blue} + \text{const}$$

---

## Summary

```{r final-summary}
tibble(
  Metric                    = c("Trait modelled",
                                "Environments (N)",
                                "DNN architecture",
                                "LOEO mean r",
                                "LOEO mean RMSE (kg/ha)",
                                "Within-env 5-fold mean r",
                                "Within-env mean RMSE (kg/ha)",
                                "Best std VI (LOEO)",
                                "Best std VI r (LOEO)",
                                "DNN vs EVI Δr (LOEO)",
                                "Top band (gradient importance)",
                                "Top band (linearised sensitivity)"),
  Value                     = c("yield_kgha",
                                "5 (HELF24, KET21, MCG23, MCG25, SNY22)",
                                "5 → 16 → 8 → 1 (SynVI) → 8 → 1",
                                "0.389  (range: 0.092 – 0.759)",
                                "1 656  (range: 759 – 3 023)",
                                "0.728  (range: 0.532 – 0.912)",
                                "635  (range: 523 – 700)",
                                "EVI",
                                "0.628",
                                "−0.239  (DNN < EVI under LOEO)",
                                "RedEdge",
                                "Red (negative)")
) %>%
  kable(caption = "DNN Synthetic VI — key results for yield_kgha") %>%
  kable_styling(bootstrap_options = c("striped","hover"),
                full_width = FALSE)
```

```{r session-info, include=FALSE}
sessionInfo()
```
