---
title: "DNN-Derived Synthetic Vegetation Index for Phenotype Prediction"
author: "workflowr"
date: "`r Sys.time()`"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    code_folding: hide
    number_sections: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6
)
```

# Overview

This script uses a deep neural network (DNN) to derive **data-driven vegetation
indices** from raw multispectral drone imagery for the NY Winter Malting Barley
breeding programme.

## What it does

Standard vegetation indices (NDVI, NDRE, etc.) are fixed, hand-crafted
combinations of spectral bands designed for generic vegetation monitoring. Here,
we instead **let the data determine the optimal band combination** for each
agronomic or malt-quality trait. The result is a trait-specific **Synthetic VI**
— a scalar value computed from the five raw bands (blue, green, red, red_edge,
nir) that maximises predictive accuracy for that trait.

A separate model is trained for each phenotype trait, yielding a distinct
Synthetic VI per trait. After model training, the learned encoder weights are
extracted and presented as an explicit mathematical formula so the index can be
computed directly from band reflectances without re-running the DNN.

## Modelling approach

The DNN has a **bottleneck architecture** with two components:

1. **Band Encoder** (5 → 16 → 8 → 1): three fully-connected layers with ReLU
   activations compress the five bands into a single scalar — the Synthetic VI.
   The final layer has no activation, so the index spans the full real line.

2. **Prediction Head** (1 → 8 → 1): maps the Synthetic VI through one hidden
   layer to the trait value.

The bottleneck forces all spectral information through a single neuron before
prediction, mimicking the structure of hand-crafted indices (e.g. NDVI uses a
single derived value) while allowing non-linear, data-optimised band combinations.
Only five specific Timepoint × Environment groups are used, one per trial year,
representing the flight timing judged most informative per environment:
`HELF24_TP7`, `KET21_TP4`, `MCG23_TP5`, `MCG25_TP12`, `SNY22_TP6`.

## Cross-validation

Two complementary CV scenarios are implemented. In both, splits are by **GID**
(genotype ID) to prevent data leakage, and scalers are fit on training data only
then applied to test folds.

### Leave-One-Environment-Out (LOEO)

All observations from one environment are withheld; the model trains on the
remaining four. This measures **cross-environment generalisation**.

| Fold | Training environments | Validation environment |
|------|-----------------------|------------------------|
| 1 | KET21, MCG23, MCG25, SNY22 | HELF24 |
| 2 | HELF24, MCG23, MCG25, SNY22 | KET21 |
| 3 | HELF24, KET21, MCG25, SNY22 | MCG23 |
| 4 | HELF24, KET21, MCG23, SNY22 | MCG25 |
| 5 | HELF24, KET21, MCG23, MCG25 | SNY22 |

### Within-Environment 5-Fold CV (by GID)

Within each environment, GIDs are shuffled and assigned to 5 folds. The model
trains on 4/5 of GIDs and predicts the remaining 1/5. This measures
**within-environment predictive ability** for unseen genotypes.

Each scenario is repeated 3 times with different random seeds. Performance is
Pearson *r* and RMSE on the back-transformed trait scale, benchmarked against
linear regression on standard VIs under the same schemes.

```{r libraries}
# ── Core data wrangling & visualisation ────────────────────────────────────
library(tidyverse)
library(ggplot2)
library(patchwork)
library(corrplot)
library(knitr)
library(kableExtra)

# ── Deep learning ──────────────────────────────────────────────────────────
# torch is the primary backend; neuralnet is a lightweight fallback
if (!requireNamespace("torch", quietly = TRUE)) {
  stop("Package 'torch' is required. Install with: install.packages('torch'); torch::install_torch()")
}
library(torch)
if (!torch::torch_is_installed()) torch::install_torch()
set.seed(42)
torch_manual_seed(42)
```

# Data Loading and Exploration

```{r load-data}
# ── Resolve data path ──────────────────────────────────────────────────────
pheno_paths <- c(
  "C:/Users/Siim Sepp/NY_WMBCL/data/WMB_pheno.Rdata",
  "~/Documents/GitHub/NY_WMBCL/data/WMB_pheno.Rdata",
  "data/WMB_pheno.Rdata"
)
pheno_file <- pheno_paths[file.exists(pheno_paths)][1]
if (is.na(pheno_file)) stop("WMB_pheno.Rdata not found in any expected location.")
message("Loading: ", pheno_file)
load(pheno_file)   # loads one or more objects into the environment

# ── Identify the main data frame ───────────────────────────────────────────
# List objects and find the data frame that is most likely the phenotype table
loaded_objs <- ls()
df_candidates <- loaded_objs[sapply(loaded_objs, function(x) is.data.frame(get(x)))]
message("Data frames found: ", paste(df_candidates, collapse = ", "))

# Use the largest data frame if multiple exist
if (length(df_candidates) == 1) {
  pheno_raw <- get(df_candidates)
} else {
  sizes      <- sapply(df_candidates, function(x) nrow(get(x)))
  pheno_raw  <- get(df_candidates[which.max(sizes)])
  message("Using: ", df_candidates[which.max(sizes)])
}

cat("Dimensions:", nrow(pheno_raw), "rows ×", ncol(pheno_raw), "cols\n")
glimpse(pheno_raw)
```

```{r identify-columns}
# ── Band columns ───────────────────────────────────────────────────────────
# Use anchored / precise patterns so that e.g. "red" does NOT match "red_edge"
BAND_NAMES <- c("blue", "green", "red", "red_edge", "nir")
band_patterns <- c(
  blue     = "^blue$",
  green    = "^green$",
  red      = "^red$",                          # exact – must NOT match red_edge
  red_edge = "red[._]?edge|rededge",
  nir      = "^nir$"
)
band_cols <- sapply(names(band_patterns), function(b) {
  m <- grep(band_patterns[b], names(pheno_raw), ignore.case = TRUE, value = TRUE)
  if (length(m) == 0) NA_character_ else m[1]
}, USE.NAMES = TRUE)
band_cols <- band_cols[!is.na(band_cols)]
# band_cols: names = canonical BAND_NAMES, values = actual column names in pheno_raw
cat("Band columns found:\n"); print(band_cols)
if (length(band_cols) < 5)
  warning("Only ", length(band_cols), " of 5 expected bands found: ",
          paste(setdiff(BAND_NAMES, names(band_cols)), collapse = ", "), " missing.")

# ── Standard VI columns ────────────────────────────────────────────────────
vi_candidates <- c("NDVI","NDRE","EVI","GNDVI","SAVI","ENDVI","GDVI","NRI","GIPVI","PSRI")
vi_cols <- intersect(vi_candidates, names(pheno_raw))
cat("Standard VI columns found:", paste(vi_cols, collapse = ", "), "\n")

# ── Timepoint column ───────────────────────────────────────────────────────
tp_col <- intersect(c("Timepoint","timepoint","TP","tp","time_point"),
                    names(pheno_raw))[1]
cat("Timepoint column:", tp_col, "\n")

# ── Candidate trait columns ────────────────────────────────────────────────
# Exclude IDs, bands, VIs and spatial variables
exclude_pat <- paste(c("GID","plot_id","Env","Row","Col","Timepoint","Fam",
                       "JD","DAS","GDD", BAND_NAMES, vi_candidates),
                     collapse="|")
#trait_cols <- names(pheno_raw)[
#  !grepl(exclude_pat, names(pheno_raw), ignore.case = TRUE) &
#  sapply(pheno_raw, is.numeric)
#]

trait_cols<- colnames(pheno_raw)[19]
cat("Candidate trait columns:\n"); print(trait_cols)

# ── Environment column ─────────────────────────────────────────────────────
env_col <- intersect(c("Env","env","environment","trial"), names(pheno_raw))[1]
cat("Environment column:", env_col, "\n")

# ── GID column ─────────────────────────────────────────────────────────────
gid_col <- intersect(c("GID","gid","Genotype","genotype","EntryName","entry_name",
                       "Line","line","Name","name"),
                     names(pheno_raw))[1]
cat("GID column:", gid_col, "\n")
```

```{r data-summary}
# ── Descriptive summary ────────────────────────────────────────────────────
cat("\nEnvironments present:\n")
print(table(pheno_raw[[env_col]]))

cat("\nBand value ranges:\n")
pheno_raw %>%
  select(all_of(unname(band_cols))) %>%
  summarise(across(everything(), list(
    min  = ~min(., na.rm = TRUE),
    mean = ~mean(., na.rm = TRUE),
    max  = ~max(., na.rm = TRUE),
    pct_na = ~mean(is.na(.)) * 100
  ))) %>%
  pivot_longer(everything(), names_to = c("band","stat"), names_sep = "_(?=[^_]+$)") %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  kable(digits = 3, caption = "Band summary statistics") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

# Data Preparation

```{r data-prep}
# ── Build modelling dataset ────────────────────────────────────────────────
found_band_names <- names(band_cols)

tp_cols_select  <- if (!is.na(tp_col))  tp_col  else character(0)
gid_cols_select <- if (!is.na(gid_col)) gid_col else character(0)

model_data <- pheno_raw %>%
  select(all_of(c(gid_cols_select, env_col, tp_cols_select,
                  unname(band_cols), vi_cols, trait_cols))) %>%
  rename(Env = all_of(env_col)) %>%
  rename(all_of(band_cols)) %>%
  filter(if_all(all_of(found_band_names), ~ !is.na(.))) %>%
  mutate(Env = as.factor(Env))

# ── Standardise GID column name ────────────────────────────────────────────
if (!is.na(gid_col) && gid_col %in% names(model_data) && gid_col != "GID") {
  model_data <- model_data %>% rename(GID = all_of(gid_col))
} else if (!"GID" %in% names(model_data)) {
  model_data$GID <- as.character(seq_len(nrow(model_data)))
  warning("No GID column found; using row index as GID.")
}

# ── Standardise Timepoint column name ─────────────────────────────────────
if (!is.na(tp_col) && tp_col != "Timepoint") {
  model_data <- model_data %>% rename(Timepoint = all_of(tp_col))
}
if (!"Timepoint" %in% names(model_data)) {
  model_data$Timepoint <- 1L
  warning("No Timepoint column found; treating all rows as a single timepoint.")
}

# ── Group variable: Env × Timepoint ───────────────────────────────────────
model_data <- model_data %>%
  mutate(Timepoint = as.integer(Timepoint),
         group     = paste(Env, Timepoint, sep = "_TP"))

model_data <- model_data %>%
  filter(group %in% c("HELF24_TP7","KET21_TP4","MCG23_TP5","MCG25_TP12","SNY22_TP6"))

group_levels <- sort(unique(model_data$group))
cat("Timepoint × Environment groups:", length(group_levels), "\n")
print(group_levels)

BAND_NAMES <- found_band_names
cat("\nRows with complete band data:", nrow(model_data), "\n")

# ── Compute NDVI from raw band values (before global scaling) ─────────────
if (all(c("red", "nir") %in% names(model_data))) {
  model_data <- model_data %>%
    mutate(NDVI_raw = (nir - red) / (nir + red + 1e-8))
  cat("NDVI_raw computed: min =", round(min(model_data$NDVI_raw, na.rm=TRUE), 3),
      " mean =", round(mean(model_data$NDVI_raw, na.rm=TRUE), 3),
      " max =", round(max(model_data$NDVI_raw, na.rm=TRUE), 3), "\n")
} else {
  warning("Cannot compute NDVI_raw: 'red' or 'nir' not found in band columns.")
}

# ── Save unscaled copy for per-fold scaling inside CV functions ────────────
model_data_raw <- model_data

# ── Global scaling of model_data (used for full-data models & plots) ───────
band_scale <- map(BAND_NAMES, function(b) {
  list(min = min(model_data[[b]], na.rm = TRUE),
       max = max(model_data[[b]], na.rm = TRUE))
}) %>% set_names(BAND_NAMES)

for (b in BAND_NAMES) {
  rng <- band_scale[[b]]
  model_data[[b]] <- (model_data[[b]] - rng$min) / (rng$max - rng$min + 1e-8)
}

trait_scale <- map(trait_cols, function(t) {
  vals <- model_data[[t]]
  list(mean = mean(vals, na.rm = TRUE), sd = sd(vals, na.rm = TRUE))
}) %>% set_names(trait_cols)

for (t in trait_cols) {
  sc <- trait_scale[[t]]
  model_data[[t]] <- (model_data[[t]] - sc$mean) / (sc$sd + 1e-8)
}

cat("Traits to model:", length(trait_cols), "\n")
```

```{r band-correlations, fig.cap="Correlations among raw spectral bands"}
# ── Visual EDA: band-band and band-VI correlations ─────────────────────────
cor_bands <- cor(model_data[, BAND_NAMES], use = "pairwise.complete.obs")
corrplot(cor_bands, method = "color", type = "upper",
         addCoef.col = "black", number.cex = 0.8,
         title = "Band–Band Correlations", mar = c(0,0,2,0))
```




# DNN Architecture

The network has two components trained **within each Timepoint × Environment group**:

1. **Band Encoder** — three fully-connected layers (n_bands → 16 → 8 → 1).
   The single output neuron is the *Synthetic VI* — no activation, so it spans
   the full real line.

2. **Prediction Head** — maps the scalar Synthetic VI through one hidden layer
   to the trait prediction.

Because modelling is within-group, between-year and between-trial offsets are
irrelevant; no environment embedding is needed.

```{r dnn-definition}
# ══════════════════════════════════════════════════════════════════════════
# Band Encoder: n_bands → 16 → 8 → 1 (Synthetic VI bottleneck)
# ══════════════════════════════════════════════════════════════════════════
BandEncoder <- nn_module(
  classname = "BandEncoder",
  initialize = function(n_bands = 5, h1 = 16, h2 = 8) {
    self$fc1  <- nn_linear(n_bands, h1)
    self$fc2  <- nn_linear(h1, h2)
    self$fc3  <- nn_linear(h2, 1)      # bottleneck = Synthetic VI (linear out)
    self$drop <- nn_dropout(0.1)
  },
  forward = function(x) {
    x <- x %>% self$fc1() %>% nnf_relu() %>% self$drop()
    x <- x %>% self$fc2() %>% nnf_relu() %>% self$drop()
    self$fc3(x)
  }
)

# ══════════════════════════════════════════════════════════════════════════
# Full predictor: encoder → Synthetic VI → prediction head → trait
# ══════════════════════════════════════════════════════════════════════════
SyntheticVIModel <- nn_module(
  classname = "SyntheticVIModel",
  initialize = function(n_bands = 5, pred_h = 8) {
    self$encoder  <- BandEncoder(n_bands)
    self$pred_fc1 <- nn_linear(1, pred_h)
    self$pred_fc2 <- nn_linear(pred_h, 1)
    self$drop     <- nn_dropout(0.1)
  },
  forward = function(bands) {
    vi <- self$encoder(bands)                            # (N, 1)
    vi %>% self$pred_fc1() %>% nnf_relu() %>% self$drop() %>% self$pred_fc2()
  },
  get_synthetic_vi = function(bands) {
    self$encoder(bands)
  }
)

cat("Model architecture defined.\n")
cat("  Encoder:   n_bands → 16 → 8 → 1 (Synthetic VI)\n")
cat("  Predictor: VI(1) → 8 → 1 (trait)\n")
```

# Training

Two CV scenarios are run in parallel:

- **LOEO**: holds out one environment at a time, trains on the other four.
- **Within-env 5-fold CV**: within each environment, GIDs are assigned to 5 folds;
  the model trains on 4 folds and predicts the 5th.

In both, the band scaler (0–1 min–max) and trait scaler (z-score) are fit on
training data only and applied to the test fold — `model_data_raw` (unscaled) is
passed to the CV functions so scaling is done internally.

After CV, a final model trained on all data is used to extract the **Synthetic VI formula**.

```{r training-functions}
# ── Helper: build band + label tensors from a pre-scaled data frame ────────
make_tensors <- function(df, trait) {
  df_sub <- df[!is.na(df[[trait]]), ]
  list(
    bands = torch_tensor(as.matrix(df_sub[, BAND_NAMES, drop = FALSE]),
                         dtype = torch_float()),
    y     = torch_tensor(df_sub[[trait]], dtype = torch_float())$unsqueeze(2)
  )
}

# ── Scale df using stats from train_df; return scaled df + back-transform params
scale_from_train <- function(df, train_df, trait) {
  band_sc  <- map(BAND_NAMES, function(b)
    list(mn = min(train_df[[b]], na.rm = TRUE),
         mx = max(train_df[[b]], na.rm = TRUE))) %>% set_names(BAND_NAMES)
  trait_mn <- mean(train_df[[trait]], na.rm = TRUE)
  trait_sd <- sd(train_df[[trait]],   na.rm = TRUE)
  for (b in BAND_NAMES) {
    sc <- band_sc[[b]]
    df[[b]] <- (df[[b]] - sc$mn) / (sc$mx - sc$mn + 1e-8)
  }
  df[[trait]] <- (df[[trait]] - trait_mn) / (trait_sd + 1e-8)
  list(df = df, trait_mn = trait_mn, trait_sd = trait_sd)
}

# ── Core training loop ─────────────────────────────────────────────────────
train_model <- function(train_df, trait,
                        epochs = 300, lr = 3e-3, batch_size = 64,
                        seed = NULL, verbose = FALSE) {
  if (!is.null(seed)) torch_manual_seed(seed)
  tensors <- make_tensors(train_df, trait)
  n       <- tensors$bands$size(1)
  if (n < 10) return(NULL)

  model <- SyntheticVIModel(n_bands = length(BAND_NAMES), pred_h = 8)
  opt   <- optim_adam(model$parameters, lr = lr, weight_decay = 1e-4)
  sched <- lr_step(opt, step_size = 100, gamma = 0.5)

  model$train()
  for (ep in seq_len(epochs)) {
    idx <- sample(n)
    for (i in seq(1, n, by = batch_size)) {
      bi <- idx[i:min(i + batch_size - 1, n)]
      opt$zero_grad()
      pred <- model(tensors$bands[bi, , drop = FALSE])
      loss <- nnf_mse_loss(pred, tensors$y[bi, , drop = FALSE])
      loss$backward()
      nn_utils_clip_grad_norm_(model$parameters, 1.0)
      opt$step()
    }
    sched$step()
    if (verbose && ep %% 100 == 0)
      cat(sprintf("  ep %d | loss %.4f\n", ep, loss$item()))
  }
  model$eval()
  model
}

# ── LOEO: Leave-One-Environment-Out CV ────────────────────────────────────
# Takes raw (unscaled) data. Scaler fit on training environments only.
loeo_cv <- function(data_raw, trait, n_reps = 3,
                    epochs = 300, lr = 3e-3, batch_size = 64) {
  envs <- sort(unique(as.character(data_raw$Env)))
  map_dfr(envs, function(held_env) {
    train_raw <- data_raw %>% filter(Env != held_env, !is.na(.data[[trait]]))
    test_raw  <- data_raw %>% filter(Env == held_env,  !is.na(.data[[trait]]))
    if (nrow(train_raw) < 10 || nrow(test_raw) < 2) return(NULL)

    sc_out   <- scale_from_train(train_raw, train_raw, trait)
    train_sc <- sc_out$df
    trait_mn <- sc_out$trait_mn
    trait_sd <- sc_out$trait_sd
    test_sc  <- scale_from_train(test_raw, train_raw, trait)$df

    map_dfr(seq_len(n_reps), function(rep_i) {
      model <- tryCatch(
        train_model(train_sc, trait, epochs = epochs, lr = lr,
                    batch_size = batch_size, seed = rep_i * 100L),
        error = function(e) { message(held_env, " rep", rep_i, ": ", e$message); NULL }
      )
      if (is.null(model)) return(NULL)
      te_ten <- make_tensors(test_sc, trait)
      with_no_grad({
        pred_z <- model(te_ten$bands)$squeeze()$to(device = "cpu")
      })
      pred <- as.numeric(pred_z) * trait_sd + trait_mn
      obs  <- as.numeric(te_ten$y$squeeze()) * trait_sd + trait_mn
      tibble(held_env = held_env, rep = rep_i, n_test = length(obs),
             r    = cor(pred, obs, use = "complete.obs"),
             rmse = sqrt(mean((pred - obs)^2, na.rm = TRUE)))
    })
  })
}

# ── Within-environment 5-fold CV (split by GID) ───────────────────────────
# Takes raw (unscaled) data. GIDs assigned to folds; scaler fit on train GIDs.
within_env_cv <- function(data_raw, trait, n_folds = 5, n_reps = 3,
                          epochs = 300, lr = 3e-3, batch_size = 64) {
  envs <- sort(unique(as.character(data_raw$Env)))
  map_dfr(envs, function(env_i) {
    env_df <- data_raw %>% filter(Env == env_i, !is.na(.data[[trait]]))
    gids   <- unique(env_df$GID)
    if (length(gids) < n_folds) {
      message("Env ", env_i, ": ", length(gids), " GIDs < n_folds — skipping")
      return(NULL)
    }
    map_dfr(seq_len(n_reps), function(rep_i) {
      set.seed(rep_i * 77L)
      shuffled <- sample(gids)
      fold_id  <- rep(seq_len(n_folds), length.out = length(shuffled))
      gid_fold <- tibble(GID = shuffled, fold = fold_id)

      map_dfr(seq_len(n_folds), function(f) {
        test_gids   <- gid_fold$GID[gid_fold$fold == f]
        train_raw_f <- env_df %>% filter(!GID %in% test_gids)
        test_raw_f  <- env_df %>% filter( GID %in% test_gids)
        if (nrow(train_raw_f) < 10 || nrow(test_raw_f) < 2) return(NULL)

        sc_out   <- scale_from_train(train_raw_f, train_raw_f, trait)
        train_sc <- sc_out$df
        trait_mn <- sc_out$trait_mn
        trait_sd <- sc_out$trait_sd
        test_sc  <- scale_from_train(test_raw_f, train_raw_f, trait)$df

        model <- tryCatch(
          train_model(train_sc, trait, epochs = epochs, lr = lr,
                      batch_size = batch_size, seed = rep_i * 100L + f),
          error = function(e) NULL
        )
        if (is.null(model)) return(NULL)
        te_ten <- make_tensors(test_sc, trait)
        with_no_grad({
          pred_z <- model(te_ten$bands)$squeeze()$to(device = "cpu")
        })
        pred <- as.numeric(pred_z) * trait_sd + trait_mn
        obs  <- as.numeric(te_ten$y$squeeze()) * trait_sd + trait_mn
        tibble(Env = env_i, fold = f, rep = rep_i, n_test = length(obs),
               r    = cor(pred, obs, use = "complete.obs"),
               rmse = sqrt(mean((pred - obs)^2, na.rm = TRUE)))
      })
    })
  })
}

# ── Linear-model LOEO CV (for NDVI baseline) ─────────────────────────────
linreg_loeo_cv <- function(data_raw, predictor, trait) {
  envs <- sort(unique(as.character(data_raw$Env)))
  map_dfr(envs, function(held_env) {
    tr <- data_raw %>% filter(Env != held_env,
                              !is.na(.data[[predictor]]), !is.na(.data[[trait]]))
    te <- data_raw %>% filter(Env == held_env,
                              !is.na(.data[[predictor]]), !is.na(.data[[trait]]))
    if (nrow(tr) < 10 || nrow(te) < 2) return(NULL)
    fit  <- lm(as.formula(paste(trait, "~", predictor)), data = tr)
    pred <- predict(fit, newdata = te)
    obs  <- te[[trait]]
    tibble(held_env = held_env, n_test = nrow(te),
           r    = cor(pred, obs, use = "complete.obs"),
           rmse = sqrt(mean((pred - obs)^2, na.rm = TRUE)))
  })
}

# ── Linear-model within-env 5-fold CV by GID (for NDVI baseline) ──────────
linreg_wenv_cv <- function(data_raw, predictor, trait,
                           n_folds = 5, n_reps = 3) {
  envs <- sort(unique(as.character(data_raw$Env)))
  map_dfr(envs, function(env_i) {
    env_df <- data_raw %>% filter(Env == env_i,
                                  !is.na(.data[[predictor]]), !is.na(.data[[trait]]))
    gids <- unique(env_df$GID)
    if (length(gids) < n_folds) return(NULL)
    map_dfr(seq_len(n_reps), function(rep_i) {
      set.seed(rep_i * 77L)
      shuffled <- sample(gids)
      fold_id  <- rep(seq_len(n_folds), length.out = length(shuffled))
      gid_fold <- tibble(GID = shuffled, fold = fold_id)
      map_dfr(seq_len(n_folds), function(f) {
        test_gids <- gid_fold$GID[gid_fold$fold == f]
        tr <- env_df %>% filter(!GID %in% test_gids)
        te <- env_df %>% filter( GID %in% test_gids)
        if (nrow(tr) < 5 || nrow(te) < 2) return(NULL)
        fit  <- lm(as.formula(paste(trait, "~", predictor)), data = tr)
        pred <- predict(fit, newdata = te)
        obs  <- te[[trait]]
        tibble(Env = env_i, fold = f, rep = rep_i, n_test = nrow(te),
               r    = cor(pred, obs, use = "complete.obs"),
               rmse = sqrt(mean((pred - obs)^2, na.rm = TRUE)))
      })
    })
  })
}
```

```{r run-training, cache=TRUE}
# ── Select traits with enough observations ────────────────────────────────
min_obs <- 30
usable_traits <- trait_cols[
  sapply(trait_cols, function(t) sum(!is.na(model_data_raw[[t]]))) >= min_obs
]
cat("Traits with >=", min_obs, "observations:", length(usable_traits), "\n")
print(usable_traits)

usable_traits <- usable_traits[1]   # user selection – expand as needed

# ── LOEO CV (raw data → per-fold scaling inside loeo_cv) ─────────────────
cat("\nRunning LOEO CV...\n")
cv_loeo <- map_dfr(usable_traits, function(trait) {
  cat("  trait:", trait, "\n")
  res       <- loeo_cv(model_data_raw, trait, n_reps = 3,
                       epochs = 300, lr = 3e-3, batch_size = 64)
  res$trait <- trait
  res
})
cat("LOEO CV done. Rows:", nrow(cv_loeo), "\n")

# ── Within-environment 5-fold CV (raw data → per-fold scaling inside) ────
cat("\nRunning within-environment 5-fold CV...\n")
cv_wenv <- map_dfr(usable_traits, function(trait) {
  cat("  trait:", trait, "\n")
  res       <- within_env_cv(model_data_raw, trait, n_folds = 5, n_reps = 3,
                              epochs = 300, lr = 3e-3, batch_size = 64)
  res$trait <- trait
  res
})
cat("Within-env CV done. Rows:", nrow(cv_wenv), "\n")
```

# Results: Cross-Validation Accuracy

## Leave-One-Environment-Out (LOEO)

```{r loeo-summary-table}
loeo_fold <- cv_loeo %>%
  filter(!is.na(r)) %>%
  group_by(trait, held_env) %>%
  summarise(r      = mean(r,      na.rm = TRUE),
            rmse   = mean(rmse,   na.rm = TRUE),
            n_test = first(n_test),
            .groups = "drop")

loeo_summary <- loeo_fold %>%
  group_by(trait) %>%
  summarise(mean_r    = mean(r,    na.rm = TRUE),
            sd_r      = sd(r,     na.rm = TRUE),
            mean_rmse = mean(rmse, na.rm = TRUE),
            n_folds   = n(),
            .groups   = "drop") %>%
  arrange(desc(mean_r))

loeo_summary %>%
  kable(digits = 3,
        col.names = c("Trait","Mean r","SD r","Mean RMSE","Folds"),
        caption   = "LOEO CV: DNN Synthetic VI accuracy (mean over 5 held-out environments, 3 training reps each)") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE) %>%
  column_spec(2, bold = TRUE,
              color = ifelse(loeo_summary$mean_r > 0.5, "green", "black"))

cat("\nPer-environment LOEO results:\n")
loeo_fold %>%
  arrange(trait, held_env) %>%
  kable(digits = 3,
        col.names = c("Trait","Held-out env","Mean r","Mean RMSE","N test"),
        caption   = "LOEO CV per held-out environment") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

```{r loeo-accuracy-plot, fig.height=5, fig.cap="LOEO CV: dots = training repetitions, crossbar = mean r"}
cv_loeo %>%
  filter(!is.na(r)) %>%
  ggplot(aes(x = r, y = reorder(held_env, r, mean, na.rm = TRUE),
             colour = held_env)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_jitter(height = 0.12, size = 2.5, alpha = 0.7) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.45,
               colour = "black", linewidth = 0.7) +
  facet_wrap(~trait, ncol = 2) +
  scale_colour_brewer(palette = "Set2") +
  labs(x = "Pearson r (LOEO)", y = "Held-out environment",
       title  = "DNN Synthetic VI – LOEO Accuracy",
       colour = "Environment") +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom")
```

## Within-Environment 5-Fold CV (by GID)

```{r wenv-summary-table}
wenv_fold_summ <- cv_wenv %>%
  filter(!is.na(r)) %>%
  group_by(trait, Env, fold) %>%
  summarise(r    = mean(r,    na.rm = TRUE),
            rmse = mean(rmse, na.rm = TRUE),
            .groups = "drop")

wenv_env_summary <- wenv_fold_summ %>%
  group_by(trait, Env) %>%
  summarise(mean_r    = mean(r,    na.rm = TRUE),
            sd_r      = sd(r,     na.rm = TRUE),
            mean_rmse = mean(rmse, na.rm = TRUE),
            n_folds   = n(),
            .groups   = "drop")

wenv_summary <- wenv_env_summary %>%
  group_by(trait) %>%
  summarise(mean_r    = mean(mean_r,    na.rm = TRUE),
            sd_r      = mean(sd_r,      na.rm = TRUE),
            mean_rmse = mean(mean_rmse, na.rm = TRUE),
            .groups   = "drop")

wenv_summary %>%
  kable(digits = 3,
        col.names = c("Trait","Mean r","Mean within-env SD r","Mean RMSE"),
        caption   = "Within-environment 5-fold CV (by GID): mean Pearson r across all environments and folds") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE) %>%
  column_spec(2, bold = TRUE,
              color = ifelse(wenv_summary$mean_r > 0.5, "green", "black"))

cat("\nPer-environment within-env CV results:\n")
wenv_env_summary %>%
  arrange(trait, Env) %>%
  kable(digits = 3,
        col.names = c("Trait","Environment","Mean r","SD r","Mean RMSE","N folds"),
        caption   = "Within-environment 5-fold CV: per-environment summary") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

```{r wenv-accuracy-plot, fig.height=5, fig.cap="Within-env CV: dots = individual folds × reps, crossbar = mean r per environment"}
cv_wenv %>%
  filter(!is.na(r)) %>%
  ggplot(aes(x = r, y = reorder(Env, r, mean, na.rm = TRUE),
             colour = Env)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_jitter(height = 0.12, size = 2, alpha = 0.6) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.45,
               colour = "black", linewidth = 0.7) +
  facet_wrap(~trait, ncol = 2) +
  scale_colour_brewer(palette = "Set2") +
  labs(x = "Pearson r (within-env 5-fold CV by GID)", y = "Environment",
       title  = "DNN Synthetic VI – Within-Environment CV Accuracy",
       colour = "Environment") +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom")
```

## CV Scenario Comparison

```{r cv-scenario-comparison}
cv_scenario_tbl <- bind_rows(
  loeo_summary  %>% select(trait, mean_r, sd_r, mean_rmse) %>% mutate(CV = "LOEO"),
  wenv_summary  %>% select(trait, mean_r, sd_r, mean_rmse) %>% mutate(CV = "Within-env 5-fold")
)

cv_scenario_tbl %>%
  arrange(trait, CV) %>%
  kable(digits = 3,
        col.names = c("Trait","Mean r","SD r","Mean RMSE","CV scenario"),
        caption   = "DNN Synthetic VI: LOEO vs. within-environment CV") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

```{r cv-scenario-plot, fig.height=4, fig.cap="LOEO vs. within-environment CV: mean Pearson r (error bar = SD across folds/envs)"}
cv_scenario_tbl %>%
  ggplot(aes(x = mean_r,
             y = reorder(trait, mean_r, max, na.rm = TRUE),
             colour = CV, shape = CV)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_errorbarh(aes(xmin = mean_r - sd_r, xmax = mean_r + sd_r),
                 height = 0.2, alpha = 0.6) +
  geom_point(size = 4) +
  scale_colour_manual(values = c(LOEO = "#e34a33",
                                 `Within-env 5-fold` = "#2c7bb6")) +
  labs(x = "Mean Pearson r", y = NULL,
       title  = "LOEO vs. within-environment CV accuracy",
       colour = "CV scenario", shape = "CV scenario") +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom")
```

# Full-Data Models and Synthetic VI Extraction

One final model per trait is trained on **all five groups** (no held-out fold).
These models are used to (a) extract the Synthetic VI for every observation and
(b) derive the VI formula from the learned encoder weights.

```{r extract-synthetic-vi, cache=TRUE}
full_models <- map(usable_traits, function(trait) {
  cat("Training full model – trait:", trait, "\n")
  tryCatch(
    train_model(model_data, trait, epochs = 300, lr = 3e-3, batch_size = 64,
                seed = 42L),
    error = function(e) { message(e$message); NULL }
  )
}) %>% set_names(usable_traits)

synvi_rows <- map_dfr(usable_traits, function(trait) {
  model <- full_models[[trait]]
  if (is.null(model)) return(NULL)
  df <- model_data %>% filter(!is.na(.data[[trait]]))
  bands_t <- torch_tensor(as.matrix(df[, BAND_NAMES, drop = FALSE]),
                          dtype = torch_float())
  with_no_grad({
    vi_vals <- model$get_synthetic_vi(bands_t)$squeeze()$to(device = "cpu")
  })
  df %>%
    select(Env, Timepoint, group) %>%
    mutate(SynVI = as.numeric(vi_vals), trait = trait)
})

cat("Synthetic VI extraction complete. Rows:", nrow(synvi_rows), "\n")
```

```{r synthetic-vi-distributions, fig.height=6, fig.cap="Synthetic VI distributions by Timepoint and Environment, per trait"}
synvi_rows %>%
  ggplot(aes(x = SynVI, fill = Env)) +
  geom_density(alpha = 0.5) +
  facet_grid(trait ~ Timepoint, scales = "free") +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Synthetic VI value", y = "Density",
       title = "Synthetic VI distributions by Timepoint × Environment",
       fill  = "Environment") +
  theme_bw(base_size = 10) +
  theme(strip.text.y = element_text(size = 7))
```

# Band Sensitivity Analysis

To interpret *which* raw bands drive each Synthetic VI we compute the gradient
of the encoder output with respect to each band input (averaged over all samples).

```{r sensitivity-analysis, cache=TRUE}
# Use the full-data models for gradient-based band importance
sensitivity_list <- map_dfr(usable_traits, function(trait) {
  model <- full_models[[trait]]
  if (is.null(model)) return(NULL)
  df <- model_data %>% filter(!is.na(.data[[trait]]))
  bands_t <- torch_tensor(
    as.matrix(df[, BAND_NAMES, drop = FALSE]),
    dtype = torch_float(), requires_grad = TRUE
  )
  vi <- model$encoder(bands_t)
  vi$sum()$backward()
  grads    <- as.matrix(bands_t$grad$to(device = "cpu"))
  abs_mean <- colMeans(abs(grads))
  tibble(trait = trait,
         band  = BAND_NAMES,
         importance = abs_mean / sum(abs_mean))
})

cat("Sensitivity analysis complete.\n")
```

```{r sensitivity-heatmap, fig.cap="Normalised band importance for each Synthetic VI (gradient-based)"}
if (!is.null(sensitivity_list) && nrow(sensitivity_list) > 0) {
  imp_wide <- sensitivity_list %>%
    pivot_wider(names_from = band, values_from = importance)

  imp_mat <- as.matrix(imp_wide[, BAND_NAMES])
  rownames(imp_mat) <- imp_wide$trait

    sensitivity_list %>%
      ggplot(aes(x = band, y = reorder(trait, importance),
                 fill = importance)) +
      geom_tile(colour = "white") +
      scale_fill_gradient(low = "white", high = "#2166AC") +
      labs(x = "Spectral Band", y = "Trait",
           fill  = "Importance",
           title = "Relative Band Importance for each Synthetic VI") +
      theme_bw(base_size = 11)
  }

```

```{r sensitivity-barplot, fig.cap="Band importance profiles per trait"}
if (!is.null(sensitivity_list) && nrow(sensitivity_list) > 0) {
  ggplot(sensitivity_list,
         aes(x = band, y = importance, fill = band)) +
    geom_col() +
    facet_wrap(~trait, ncol = 3) +
    scale_fill_manual(values = c(blue="steelblue", green="#2ca25f",
                                 red="#e34a33", red_edge="#f768a1",
                                 nir="#8856a7")) +
    labs(x = "Band", y = "Normalised gradient importance",
         title = "Per-trait spectral band importance") +
    theme_bw(base_size = 11) +
    theme(axis.text.x  = element_text(angle = 30, hjust = 1),
          legend.position = "none")
}
```

# Synthetic VI Formula

The Synthetic VI is defined by the three learned linear layers of the encoder:

$$
\mathbf{h}_1 = \text{ReLU}(\mathbf{W}_1 \mathbf{x} + \mathbf{b}_1), \quad
\mathbf{h}_2 = \text{ReLU}(\mathbf{W}_2 \mathbf{h}_1 + \mathbf{b}_2), \quad
\text{SynVI} = \mathbf{w}_3^\top \mathbf{h}_2 + b_3
$$

where $\mathbf{x} = [\text{blue, green, red, red\_edge, nir}]^\top$ are the
min-max scaled band values. The tables below show the extracted weights for each
trait's full-data model.

```{r vi-formula, results="asis"}
for (trait in usable_traits) {
  model <- full_models[[trait]]
  if (is.null(model)) next

  cat("\n\n## Trait:", trait, "\n\n")

  # ── Extract layer weights ──────────────────────────────────────────────
  W1 <- as.matrix(model$encoder$fc1$weight$to(device="cpu")$detach())
  b1 <- as.numeric(model$encoder$fc1$bias$to(device="cpu")$detach())
  W2 <- as.matrix(model$encoder$fc2$weight$to(device="cpu")$detach())
  b2 <- as.numeric(model$encoder$fc2$bias$to(device="cpu")$detach())
  W3 <- as.numeric(model$encoder$fc3$weight$to(device="cpu")$detach())
  b3 <- as.numeric(model$encoder$fc3$bias$to(device="cpu")$detach())

  # W1: shape (16, n_bands)
  colnames(W1) <- BAND_NAMES
  rownames(W1) <- paste0("h1[", seq_len(nrow(W1)), "]")

  cat("**Layer 1 weights** W1 (16 × n_bands) and bias b1:\n\n")
  print(kable(round(cbind(W1, bias = b1), 4), format = "html") %>%
        kable_styling(bootstrap_options = c("striped","condensed"),
                      full_width = FALSE, font_size = 10))

  # W2: shape (8, 16)
  rownames(W2) <- paste0("h2[", seq_len(nrow(W2)), "]")
  colnames(W2) <- paste0("h1[", seq_len(ncol(W2)), "]")

  cat("\n\n**Layer 2 weights** W2 (8 × 16) and bias b2:\n\n")
  print(kable(round(cbind(W2, bias = b2), 4), format = "html") %>%
        kable_styling(bootstrap_options = c("striped","condensed"),
                      full_width = FALSE, font_size = 10))

  # W3: shape (1, 8) → output weights
  w3_df <- tibble(node = paste0("h2[", seq_along(W3), "]"),
                  weight = round(W3, 4))
  cat("\n\n**Layer 3 weights** w3 (1 × 8) and bias b3 =", round(b3, 4), ":\n\n")
  print(kable(w3_df, format = "html",
              col.names = c("Hidden node h2","Output weight")) %>%
        kable_styling(bootstrap_options = c("striped","condensed"),
                      full_width = FALSE, font_size = 10))

  # ── Effective linear sensitivity (first-order Taylor approx) ───────────
  # Compute mean input, propagate through network, get gradient
  df_trait  <- model_data %>% filter(!is.na(.data[[trait]]))
  mean_bands <- colMeans(df_trait[, BAND_NAMES])
  bands_t    <- torch_tensor(matrix(mean_bands, nrow = 1),
                             dtype = torch_float(), requires_grad = TRUE)
  vi_val <- model$encoder(bands_t)
  vi_val$backward()
  sens <- as.numeric(bands_t$grad$to(device="cpu"))
  sens_df <- tibble(Band = BAND_NAMES,
                    `dVI/dBand (at mean)` = round(sens, 4),
                    `|Sensitivity|`       = round(abs(sens), 4)) %>%
    arrange(desc(`|Sensitivity|`))

  cat("\n\n**Linearised sensitivity** (gradient at mean band values):\n\n")
  print(kable(sens_df, format = "html") %>%
        kable_styling(bootstrap_options = c("striped","condensed"),
                      full_width = FALSE))

  # ── Approximate explicit formula (top 3 bands, linear approx) ──────────
  top3 <- sens_df$Band[1:min(3, nrow(sens_df))]
  coef_str <- paste(
    sprintf("%+.4f × %s", sens_df$`dVI/dBand (at mean)`[match(top3, sens_df$Band)], top3),
    collapse = " "
  )
  cat(sprintf(
    "\n\n**Approximate linear formula** (linearised around band means):\n\n$$\\text{SynVI}_{%s} \\approx %.4f %s + \\text{const}$$\n\n",
    trait, 0, coef_str
  ))
}
```

# Comparison: Synthetic VI vs. Standard VIs

For each standard VI and trait the same LOGO scheme is used (train on 4 groups,
test on 5th) with a simple linear regression baseline.

```{r standard-vi-cv}
# LOGO linear regression benchmark for each standard VI × trait
linreg_logo_cv <- function(data, vi, trait) {
  map_dfr(group_levels, function(held_out) {
    tr <- data %>% filter(group != held_out,
                          !is.na(.data[[vi]]), !is.na(.data[[trait]]))
    te <- data %>% filter(group == held_out,
                          !is.na(.data[[vi]]), !is.na(.data[[trait]]))
    if (nrow(tr) < 10 || nrow(te) < 2) return(NULL)
    fit  <- lm(as.formula(paste(trait, "~", vi)), data = tr)
    pred <- predict(fit, newdata = te)
    tibble(held_out = held_out,
           r = cor(pred, te[[trait]], use = "complete.obs"))
  }) %>%
  summarise(mean_r = mean(r, na.rm = TRUE)) %>%
  pull(mean_r)
}

if (length(vi_cols) > 0) {
  cat("Computing standard VI baselines for", length(vi_cols), "VIs ×",
      length(usable_traits), "traits ...\n")
  std_vi_results <- expand_grid(vi = vi_cols, trait = usable_traits) %>%
    mutate(mean_r = map2_dbl(vi, trait,
                             ~linreg_logo_cv(model_data, .x, .y)))

  best_std_vi <- std_vi_results %>%
    group_by(trait) %>%
    slice_max(mean_r, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    rename(best_vi = vi, std_r = mean_r)

  comparison <- loeo_summary %>%
    select(trait, syn_r = mean_r) %>%
    left_join(best_std_vi, by = "trait") %>%
    mutate(improvement = syn_r - std_r)

  comparison %>%
    arrange(desc(improvement)) %>%
    kable(digits = 3,
          col.names = c("Trait","DNN Synthetic VI r","Best std VI","Std VI r","Δr"),
          caption = "DNN Synthetic VI vs. best standard VI (LOEO mean Pearson r)") %>%
    kable_styling(bootstrap_options = c("striped","hover","condensed"),
                  full_width = FALSE) %>%
    column_spec(5,
                bold  = TRUE,
                color = ifelse(comparison$improvement > 0, "green", "red"))
}
```

```{r comparison-plot, fig.cap="Prediction accuracy: Synthetic VI vs. best standard VI"}
if (length(vi_cols) > 0 && exists("comparison")) {
  comparison %>%
    pivot_longer(c(syn_r, std_r),
                 names_to = "type", values_to = "r") %>%
    mutate(type = recode(type, syn_r = "Synthetic VI (DNN)",
                               std_r = "Best standard VI (linear)"),
           trait = reorder(trait, ifelse(type == "Synthetic VI (DNN)", r, NA),
                           mean, na.rm = TRUE)) %>%
    ggplot(aes(x = r, y = trait, colour = type, shape = type)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_point(size = 4, alpha = 0.9) +
    geom_line(aes(group = trait), colour = "grey60", linewidth = 0.5) +
    scale_colour_manual(values = c("Synthetic VI (DNN)"    = "#e34a33",
                                   "Best standard VI (linear)" = "#3182bd")) +
    labs(x = "Mean LOEO r", y = NULL,
         title  = "Prediction accuracy: DNN Synthetic VI vs. Standard VIs (LOEO)",
         colour = NULL, shape = NULL) +
    theme_bw(base_size = 13) +
    theme(legend.position = "bottom")
}
```

# Synthetic VI Correlations with Standard VIs

```{r synvi-stdvi-corr, fig.cap="Correlation between Synthetic VIs (per-trait) and standard VIs"}
if (length(vi_cols) > 0 && nrow(synvi_rows) > 0) {
  # Join synthetic VI values back to model_data for standard VI columns
  synvi_wide <- synvi_rows %>%
    select(Env, Timepoint, group, trait, SynVI) %>%
    pivot_wider(names_from = trait, values_from = SynVI,
                names_prefix = "SynVI_",
                values_fn = mean)   # average if >1 row per cell

  stdvi_sub <- model_data %>%
    select(Env, Timepoint, group, all_of(vi_cols)) %>%
    group_by(Env, Timepoint, group) %>%
    summarise(across(all_of(vi_cols), mean, na.rm = TRUE), .groups = "drop")

  joined <- left_join(synvi_wide, stdvi_sub, by = c("Env","Timepoint","group"))

  synvi_trait_cols <- names(joined)[startsWith(names(joined), "SynVI_")]
  if (length(synvi_trait_cols) > 0 && length(vi_cols) > 0) {
    cor_mat  <- cor(joined[, c(vi_cols, synvi_trait_cols)],
                    use = "pairwise.complete.obs")
    sub_cor  <- cor_mat[vi_cols, synvi_trait_cols, drop = FALSE]
    colnames(sub_cor) <- str_remove(colnames(sub_cor), "SynVI_")
    corrplot(sub_cor, method = "color",
             col  = colorRampPalette(c("#2166AC","white","#B2182B"))(200),
             tl.cex = 0.8, cl.cex = 0.7,
             addCoef.col = "black", number.cex = 0.65,
             title = "r(Standard VI, Synthetic VI)", mar = c(0,0,2,0))
  }
}
```

# Summary

```{r summary-table}
final_summary <- loeo_summary %>%
  left_join(
    wenv_summary %>% select(trait, wenv_r = mean_r),
    by = "trait"
  ) %>%
  left_join(
    sensitivity_list %>%
      group_by(trait) %>%
      slice_max(importance, n = 1, with_ties = FALSE) %>%
      select(trait, top_band = band),
    by = "trait"
  ) %>%
  arrange(desc(mean_r))

final_summary %>%
  kable(digits = 3,
        col.names = c("Trait","LOEO r","SD r","LOEO RMSE","N folds",
                      "Within-env r","Top band"),
        caption = "Summary: DNN Synthetic VI – LOEO and within-environment CV accuracy with dominant spectral band") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

```{r session-info}
sessionInfo()
```
