---
title: "Population Structure and Linkage Disequilibrium Analysis"
author: "workflowr"
date: "`r Sys.time()`"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    code_folding: hide
    number_sections: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6
)
```

# Overview

This script characterises the **population structure** and **linkage disequilibrium
(LD)** of the NY Winter Malting Barley (NY_WMB) marker data. The genotype input is:

| File | Format | Description |
|------|--------|-------------|
| `LTT_geno_50K.csv` | Semicolon-delimited, markers × individuals | AA / AB / BB calls, converted to 0 / 1 / 2 additive coding |

SNP chromosome and physical position information is sourced from the Morex V3
assembly annotation.

**Population structure** is explored via:

* Principal Component Analysis (PCA) on the centred marker matrix
* Discriminant Analysis of Principal Components (DAPC)
* Genomic kinship heatmap

**Linkage disequilibrium** is characterised via:

* Genome-wide and per-chromosome LD decay (r² vs physical distance)
* Per-chromosome LD heatmaps on a marker subsample
* Estimation of the distance at which r² drops to half its maximum (LD half-life)

```{r libraries}
library(tidyverse)
library(ggplot2)
library(patchwork)
library(RSpectra)       # fast truncated SVD/eigen for large matrices
library(factoextra)     # PCA visualisation helpers
library(corrplot)
library(knitr)
library(kableExtra)
library(readxl)

# Population genetics
if (requireNamespace("adegenet", quietly = TRUE)) library(adegenet)

# LD heatmap (optional — install with: install.packages("LDheatmap"))
ld_heatmap_avail <- requireNamespace("LDheatmap", quietly = TRUE)

set.seed(42)
```

# Data Loading

```{r load-geno}
# ── Resolve project root ───────────────────────────────────────────────────
proj_roots <- c(
  "C:/Users/Siim Sepp/NY_WMBCL",
  "C:/Users/Siim Sepp/NY_WMBCL",
  "~/Documents/GitHub/NY_WMBCL"
)
proj_root <- proj_roots[dir.exists(proj_roots)][1]
message("Project root: ", proj_root)

# ── Read LTT_geno_50K.csv ─────────────────────────────────────────────────
# Format: semicolon-delimited, markers as rows, individuals as columns.
# First column = marker_name; genotypes coded as AA / AB / BB.
geno_file <- file.path(proj_root, "LTT_geno_50K.csv")
if (!file.exists(geno_file)) stop("LTT_geno_50K.csv not found at: ", geno_file)

raw <- read.csv(geno_file, sep = ";", header = TRUE,
                check.names = FALSE, stringsAsFactors = FALSE)

marker_names <- raw[[1]]          # first column = marker IDs
ind_names    <- colnames(raw)[-1] # remaining columns = individual IDs
geno_char    <- as.matrix(raw[, -1])   # markers × individuals (character)
rownames(geno_char) <- marker_names

# ── Convert AA / AB / BB → 0 / 1 / 2 (additive coding) ──────────────────
# AA = homozygous reference → 0
# AB / BA = heterozygous   → 1
# BB = homozygous alternate → 2
# anything else (e.g. "--", "NC", "") → NA
aabb_to_012 <- function(x) {
  x <- trimws(toupper(x))
  out <- rep(NA_real_, length(x))
  out[x == "AA"]              <- 0
  out[x %in% c("AB", "BA")] <- 1
  out[x == "BB"]              <- 2
  out
}

geno_012 <- apply(geno_char, 2, aabb_to_012)   # still markers × individuals
rownames(geno_012) <- marker_names
colnames(geno_012) <- ind_names

# Transpose to individuals × markers (standard orientation)
geno_a <- t(geno_012)    # additive 0/1/2 matrix

n_ind <- nrow(geno_a)
n_snp <- ncol(geno_a)
cat(sprintf("Individuals : %d\n", n_ind))
cat(sprintf("Markers     : %d\n", n_snp))
cat(sprintf("Missing (%%) : %.2f%%\n", mean(is.na(geno_a)) * 100))
cat(sprintf("Genotype counts — AA: %d  AB: %d  BB: %d  NA: %d\n",
            sum(geno_012 == 0, na.rm = TRUE),
            sum(geno_012 == 1, na.rm = TRUE),
            sum(geno_012 == 2, na.rm = TRUE),
            sum(is.na(geno_012))))
```

```{r load-map}
# ── SNP chromosome / position map ─────────────────────────────────────────
map_paths <- c(
  file.path(proj_root, "data/geno/SNPPositions_50k_on_MorexV3.xlsx"),
  file.path(proj_root, "data/SNPPositions_50k_on_MorexV3.xlsx")
)
map_file <- map_paths[file.exists(map_paths)][1]

if (!is.na(map_file)) {
  snp_map <- read_xlsx(map_file)
  # Standardise column names regardless of exact header text
  names(snp_map) <- tolower(names(snp_map))
  snp_map$chr <- gsub("H", "", snp_map$`refseq name`)
  snp_map$chr <- gsub("chr", "", snp_map$chr)
  # Detect rs/marker, chrom, pos columns
  marker_col <- colnames(snp_map)[1]
  chr_col    <- grep("chr|chrom",          names(snp_map), value = TRUE)[1]
  pos_col    <- colnames(snp_map)[3]
  snp_map <- snp_map %>%
    rename(Marker = all_of(marker_col),
           Chr    = all_of(chr_col),
           Pos    = all_of(pos_col)) %>%
    mutate(Chr = as.character(Chr),
           Pos = as.numeric(Pos)) %>%
    filter(Marker %in% colnames(geno_a),
           !Chr %in% c("0","Un","NA",NA))   # drop unplaced markers
  cat("Markers with map position:", nrow(snp_map),
      "on", n_distinct(snp_map$Chr), "chromosomes\n")
} else {
  warning("SNP position file not found — LD decay will use marker index order only.")
  snp_map <- tibble(
    Marker = colnames(geno_a),
    Chr    = "unknown",
    Pos    = seq_len(ncol(geno_a))
  )
}
```

# Data Quality Summary

```{r qc-summary}
# ── Per-marker MAF and missing rate ───────────────────────────────────────
maf_vec <- apply(geno_a, 2, function(x) {
  p <- mean(x, na.rm = TRUE) / 2      # freq of alt allele (0/1/2 coding)
  min(p, 1 - p)
})
miss_vec <- colMeans(is.na(geno_a))

qc_df <- tibble(MAF = maf_vec, Missing = miss_vec)

p_maf  <- ggplot(qc_df, aes(MAF)) +
  geom_histogram(bins = 50, fill = "#3182bd", colour = "white") +
  labs(title = "Minor Allele Frequency distribution",
       x = "MAF", y = "Count") +
  theme_bw(base_size = 13)

p_miss <- ggplot(qc_df, aes(Missing)) +
  geom_histogram(bins = 50, fill = "#e34a33", colour = "white") +
  labs(title = "Per-marker missing rate",
       x = "Missing proportion", y = "Count") +
  theme_bw(base_size = 13)

p_maf + p_miss

tibble(
  Statistic = c("Individuals","Markers","Mean MAF","Markers MAF < 0.05",
                "Mean missing rate","Markers > 10% missing"),
  Value = c(n_ind, n_snp,
            round(mean(maf_vec, na.rm = TRUE), 3),
            sum(maf_vec < 0.05, na.rm = TRUE),
            round(mean(miss_vec), 4),
            sum(miss_vec > 0.10))
) %>%
  kable(caption = "Genotype data summary") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

```{r marker-density, fig.cap="SNP density across the barley genome (Morex V3)"}
if (any(snp_map$Chr != "unknown")) {
  snp_map %>%
    mutate(Pos_Mb = Pos / 1e6) %>%
    ggplot(aes(x = Pos_Mb)) +
    geom_histogram(bins = 100, fill = "#2ca25f", colour = "blue") +
    facet_wrap(~Chr, ncol = 4, scales = "free_x") +
    labs(x = "Physical position (Mb)", y = "SNP count",
         title = "SNP density per chromosome") +
    theme_bw(base_size = 11)
}
```

# Population Structure

## PCA

```{r pca, cache=TRUE}
# ── Impute missing values with column mean before PCA ─────────────────────
geno_imp <- geno_a
col_means <- colMeans(geno_a, na.rm = TRUE)
for (j in seq_len(ncol(geno_imp))) {
  nas <- is.na(geno_imp[, j])
  if (any(nas)) geno_imp[nas, j] <- col_means[j]
}

# ── Centre markers (subtract column mean) ─────────────────────────────────
geno_ctr <- scale(geno_imp, center = TRUE, scale = FALSE)

# ── Truncated SVD via RSpectra (fast for large matrices) ──────────────────
n_pc <- min(20, n_ind - 1)
svd_res <- RSpectra::svds(geno_ctr, k = n_pc)

# PC scores (n_ind × n_pc)
pc_scores <- svd_res$u %*% diag(svd_res$d)
rownames(pc_scores) <- rownames(geno_a)
colnames(pc_scores) <- paste0("PC", seq_len(n_pc))

# Variance explained
var_exp <- svd_res$d^2 / sum(svd_res$d^2) * 100

cat("Variance explained by PC1–5:\n")
print(round(var_exp[1:5], 2))
```

```{r pca-scree, fig.height=4, fig.cap="Scree plot — variance explained by each PC"}
tibble(PC = seq_len(n_pc), VarExp = var_exp) %>%
  ggplot(aes(x = PC, y = VarExp)) +
  geom_col(fill = "#3182bd") +
  geom_line(aes(group = 1), colour = "black") +
  geom_point(size = 2) +
  labs(x = "Principal Component", y = "Variance explained (%)",
       title = "Scree plot") +
  theme_bw(base_size = 13)
```

```{r pca-scatter, fig.height=8, fig.cap="PCA scatter plots — PC1–4"}
pc_df <- as_tibble(pc_scores[, 1:min(6, n_pc)]) %>%
  mutate(ID = rownames(pc_scores))

# Attempt to join population labels if available
pop_file <- file.path(proj_root, "data/geno/wmb_pedigree.Rdata")
if (file.exists(pop_file)) {
  load(pop_file)
  ped_obj <- ls()[sapply(ls(), function(x) is.data.frame(get(x)) &&
                           any(grepl("GID|id|pop|fam", names(get(x)), ignore.case=TRUE)))]
  if (length(ped_obj) > 0) {
    ped <- get(ped_obj[1])
    id_col  <- grep("GID|id",  names(ped), ignore.case=TRUE, value=TRUE)[1]
    pop_col <- grep("pop|fam|type|group", names(ped), ignore.case=TRUE, value=TRUE)[1]
    if (!is.na(id_col) && !is.na(pop_col)) {
      pc_df <- pc_df %>%
        left_join(ped %>% select(ID = all_of(id_col),
                                 Population = all_of(pop_col)), by = "ID")
    }
  }
}
if (!"Population" %in% names(pc_df)) pc_df$Population <- "All"

colour_scale <- if (n_distinct(pc_df$Population) <= 9) {
  scale_colour_brewer(palette = "Set1")
} else {
  scale_colour_viridis_d()
}

p12 <- ggplot(pc_df, aes(PC1, PC2, colour = Population)) +
  geom_point(alpha = 0.7, size = 2) + colour_scale +
  labs(x = sprintf("PC1 (%.1f%%)", var_exp[1]),
       y = sprintf("PC2 (%.1f%%)", var_exp[2])) +
  theme_bw(base_size = 12) + theme(legend.position = "right")

p13 <- ggplot(pc_df, aes(PC1, PC3, colour = Population)) +
  geom_point(alpha = 0.7, size = 2) + colour_scale +
  labs(x = sprintf("PC1 (%.1f%%)", var_exp[1]),
       y = sprintf("PC3 (%.1f%%)", var_exp[3])) +
  theme_bw(base_size = 12) + theme(legend.position = "none")

p23 <- ggplot(pc_df, aes(PC2, PC3, colour = Population)) +
  geom_point(alpha = 0.7, size = 2) + colour_scale +
  labs(x = sprintf("PC2 (%.1f%%)", var_exp[2]),
       y = sprintf("PC3 (%.1f%%)", var_exp[3])) +
  theme_bw(base_size = 12) + theme(legend.position = "none")

p14 <- ggplot(pc_df, aes(PC1, PC4, colour = Population)) +
  geom_point(alpha = 0.7, size = 2) + colour_scale +
  labs(x = sprintf("PC1 (%.1f%%)", var_exp[1]),
       y = sprintf("PC4 (%.1f%%)", var_exp[4])) +
  theme_bw(base_size = 12) + theme(legend.position = "none")

(p12 + p13) / (p23 + p14) +
  plot_layout(guides = "collect") &
  theme(legend.position = "right")
```

```{r pca-cumvar, fig.height=4, fig.cap="Cumulative variance explained"}
tibble(PC = seq_len(n_pc), CumVar = cumsum(var_exp)) %>%
  ggplot(aes(PC, CumVar)) +
  geom_line() + geom_point(size = 2) +
  geom_hline(yintercept = c(50, 80), linetype = "dashed", colour = c("orange","red")) +
  annotate("text", x = n_pc * 0.8, y = 52, label = "50%", colour = "orange") +
  annotate("text", x = n_pc * 0.8, y = 82, label = "80%", colour = "red") +
  labs(x = "Number of PCs", y = "Cumulative variance explained (%)",
       title = "Cumulative variance explained by PCs") +
  theme_bw(base_size = 13)
```

## Discriminant Analysis of Principal Components (DAPC)

DAPC (Jombart et al. 2010) finds the linear combination of PCs that best
discriminates pre-defined or inferred groups. We use k-means clustering on the
PC scores to infer K groups, then run DAPC.

```{r dapc, cache=TRUE, fig.cap="DAPC — discriminant components coloured by inferred cluster"}
if (requireNamespace("adegenet", quietly = TRUE)) {

  # ── Choose number of clusters via BIC on k-means ──────────────────────
  max_k  <- min(10, floor(n_ind / 10))
  bic_k  <- map_dbl(2:max_k, function(k) {
    km <- kmeans(pc_scores[, 1:min(10, n_pc)], centers = k,
                 nstart = 10, iter.max = 50)
    km$tot.withinss   # use within-SS as proxy (BIC not directly from kmeans)
  })

  # Elbow method: rate of change of within-SS
  k_opt <- which.min(diff(bic_k)) + 2   # +2 because we start from k=2
  cat("Optimal K (elbow method):", k_opt, "\n")

  p_bic <- tibble(K = 2:max_k, WithinSS = bic_k) %>%
    ggplot(aes(K, WithinSS)) +
    geom_line() + geom_point(size = 2) +
    geom_vline(xintercept = k_opt, linetype = "dashed", colour = "red") +
    labs(x = "Number of clusters (K)", y = "Total within-cluster SS",
         title = sprintf("K-means elbow plot (selected K = %d)", k_opt)) +
    theme_bw(base_size = 13)
  print(p_bic)

  # ── K-means with optimal K ─────────────────────────────────────────────
  km_fit  <- kmeans(pc_scores[, 1:min(10, n_pc)], centers = k_opt,
                    nstart = 25, iter.max = 100)
  pop_vec <- as.factor(km_fit$cluster)

  # ── DAPC ──────────────────────────────────────────────────────────────
  # Convert to genind-free numeric DAPC using adegenet's low-level interface
  n_da  <- k_opt - 1
  n_pca <- min(50, n_ind - 1)
  dapc_res <- adegenet::dapc(geno_imp, grp = pop_vec,
                              n.pca = n_pca, n.da = n_da)

  # ── Membership probability bar chart ──────────────────────────────────
  memb <- as_tibble(dapc_res$posterior) %>%
    mutate(ID = rownames(geno_a), Cluster = as.character(km_fit$cluster)) %>%
    arrange(Cluster) %>%
    mutate(Order = row_number()) %>%
    pivot_longer(starts_with("V") | matches("^[0-9]+$"),
                 names_to = "Ancestry", values_to = "Prob")

  p_memb <- ggplot(memb, aes(x = Order, y = Prob, fill = Ancestry)) +
    geom_col(width = 1) +
    scale_fill_brewer(palette = "Set2") +
    labs(x = "Individual (sorted by cluster)", y = "Membership probability",
         title = "DAPC – cluster membership probabilities") +
    theme_bw(base_size = 12) +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
  print(p_memb)

  # ── DAPC scatter (DA1 vs DA2 if ≥2 DAs) ──────────────────────────────
  da_df <- as_tibble(dapc_res$ind.coord) %>%
    mutate(Cluster = as.factor(km_fit$cluster))
  if (ncol(dapc_res$ind.coord) >= 2) {
    p_da <- ggplot(da_df, aes(.data[[names(da_df)[1]]],
                               .data[[names(da_df)[2]]],
                               colour = Cluster)) +
      geom_point(alpha = 0.7, size = 2) +
      scale_colour_brewer(palette = "Set1") +
      labs(x = "DA1", y = "DA2",
           title = "DAPC scatter plot") +
      theme_bw(base_size = 13)
    print(p_da)
  }

} else {
  cat("Package 'adegenet' not available. Install with: install.packages('adegenet')\n")
}
```

## Genomic Kinship Heatmap

```{r kinship-heatmap, fig.height=7, fig.cap="Genomic kinship matrix (VanRaden method) — individuals sorted by PC1"}
# Compute kinship matrix from centred markers
# K = Z Z' / (2 Σ p_j(1-p_j))
p_j    <- colMeans(geno_imp, na.rm = TRUE) / 2   # allele freq
denom  <- 2 * sum(p_j * (1 - p_j))
Z      <- sweep(geno_imp, 2, 2 * p_j, "-")       # centre by 2p
K_mat  <- tcrossprod(Z) / denom                   # n_ind × n_ind

# Sort individuals by PC1 for better visual structure
pc1_order <- order(pc_scores[, 1])
K_sorted  <- K_mat[pc1_order, pc1_order]

# Subsample if > 200 individuals for display
max_disp <- 200
if (nrow(K_sorted) > max_disp) {
  idx      <- round(seq(1, nrow(K_sorted), length.out = max_disp))
  K_sorted <- K_sorted[idx, idx]
  cat("Note: kinship heatmap shows", max_disp, "individuals (subsampled).\n")
}

K_cor <- cov2cor(K_sorted)

corrplot(K_cor,
         tl.pos = "n",
         cl.cex = 0.8,
         title  = "Genomic kinship (sorted by PC1)",
         mar    = c(0, 0, 2, 0))

hist(K_sorted[,1])
# Diagonal (self-kinship) summary
cat(sprintf("\nKinship diagonal (self-kinship): mean = %.3f, range [%.3f, %.3f]\n",
            mean(diag(K_mat)), min(diag(K_mat)), max(diag(K_mat))))
cat(sprintf("Off-diagonal kinship           : mean = %.3f, range [%.3f, %.3f]\n",
            mean(K_mat[lower.tri(K_mat)]),
            min(K_mat[lower.tri(K_mat)]),
            max(K_mat[lower.tri(K_mat)])))
```

# Linkage Disequilibrium

## LD Computation

```{r ld-setup}
# ── Restrict to mapped markers only ───────────────────────────────────────
mapped_markers <- intersect(colnames(geno_a), snp_map$Marker)
cat("Mapped markers available for LD:", length(mapped_markers), "\n")

geno_mapped <- geno_imp[, mapped_markers, drop = FALSE]
map_sub     <- snp_map %>%
  filter(Marker %in% mapped_markers) %>%
  arrange(Chr, Pos)

chromosomes <- sort(unique(map_sub$Chr))
cat("Chromosomes:", paste(chromosomes, collapse = ", "), "\n")

# ── Fast r² function for two marker vectors ────────────────────────────────
r2_pair <- function(x, y) {
  ok   <- complete.cases(x, y)
  if (sum(ok) < 10) return(NA_real_)
  x <- x[ok]; y <- y[ok]
  # r² via correlation of allele counts
  cor(x, y)^2
}

# ── Sample pairs within a distance window and compute r² ──────────────────
compute_ld_decay <- function(chrom, n_pairs = 5000, max_dist_mb = 300) {
  chr_map  <- map_sub %>% filter(Chr == chrom) %>% arrange(Pos)
  if (nrow(chr_map) < 10) return(NULL)
  markers  <- chr_map$Marker
  G_chr    <- geno_mapped[, markers, drop = FALSE]
  n_m      <- length(markers)

  # Random pairs sampling
  n_sample <- min(n_pairs, n_m * (n_m - 1) / 2)
  pairs    <- matrix(sample.int(n_m, 2 * n_sample, replace = TRUE),
                     ncol = 2)
  pairs    <- pairs[pairs[,1] != pairs[,2], , drop = FALSE]
  pairs    <- pairs[1:min(n_sample, nrow(pairs)), , drop = FALSE]

  map_dfr(seq_len(nrow(pairs)), function(k) {
    i <- pairs[k, 1]; j <- pairs[k, 2]
    dist_mb <- abs(chr_map$Pos[i] - chr_map$Pos[j]) / 1e6
    if (dist_mb > max_dist_mb) return(NULL)
    r2 <- r2_pair(G_chr[, i], G_chr[, j])
    tibble(Chr = chrom, dist_mb = dist_mb, r2 = r2)
  })
}
```

```{r ld-decay-compute, cache=TRUE}
cat("Computing LD decay per chromosome ...\n")
ld_decay <- map_dfr(chromosomes, function(chr) {
  cat(" Chr", chr, "\n")
  compute_ld_decay(chr, n_pairs = 5000)
})
cat("Total SNP pairs computed:", nrow(ld_decay), "\n")
```

## Genome-wide LD Decay

```{r ld-decay-plot, fig.cap="Genome-wide LD decay: r² vs physical distance. LOESS smoother shown in red."}
ld_decay %>%
  filter(!is.na(r2)) %>%
  ggplot(aes(x = dist_mb, y = r2)) +
  geom_point(alpha = 0.05, size = 0.6, colour = "grey40") +
  geom_smooth(method = "loess", span = 0.1,
              colour = "#e34a33", linewidth = 1.2, se = TRUE,
              fill   = "#fdbb84") +
  scale_x_continuous(limits = c(0, NA)) +
  labs(x = "Physical distance (Mb)", y = expression(r^2),
       title = "Genome-wide LD decay") +
  theme_bw(base_size = 13)
```

```{r ld-halflife}
# ── Estimate LD half-life (distance where smoothed r² = 0.5 × max r²) ────
ld_smooth <- ld_decay %>%
  filter(!is.na(r2), dist_mb <= 100) %>%
  arrange(dist_mb)

loess_fit  <- loess(r2 ~ dist_mb, data = ld_smooth, span = 0.1)
dist_grid  <- seq(0, max(ld_smooth$dist_mb, na.rm = TRUE), by = 0.1)
r2_hat     <- predict(loess_fit, newdata = data.frame(dist_mb = dist_grid))

r2_max     <- max(r2_hat, na.rm = TRUE)
half_r2    <- r2_max / 2
half_dist  <- dist_grid[which.min(abs(r2_hat - half_r2))]

cat(sprintf("Maximum smoothed r²         : %.3f\n", r2_max))
cat(sprintf("LD half-life distance        : %.1f Mb  (r² = %.3f)\n",
            half_dist, half_r2))

# Per-chromosome half-life
ld_hl_chr <- map_dfr(chromosomes, function(chr) {
  d <- ld_decay %>% filter(Chr == chr, !is.na(r2), dist_mb <= 100)
  if (nrow(d) < 50) return(tibble(Chr = chr, halflife_Mb = NA_real_))
  fit   <- tryCatch(loess(r2 ~ dist_mb, data = d, span = 0.15),
                    error = function(e) NULL)
  if (is.null(fit)) return(tibble(Chr = chr, halflife_Mb = NA_real_))
  r2h   <- predict(fit, newdata = data.frame(dist_mb = dist_grid))
  mx    <- max(r2h, na.rm = TRUE)
  hl    <- dist_grid[which.min(abs(r2h - mx / 2))]
  tibble(Chr = chr, max_r2 = round(mx, 3), halflife_Mb = round(hl, 2))
})

ld_hl_chr %>%
  kable(col.names = c("Chromosome","Max r²","LD half-life (Mb)"),
        caption = "Per-chromosome LD decay statistics") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

## Per-chromosome LD Decay

```{r ld-decay-per-chr, fig.height=10, fig.cap="Per-chromosome LD decay curves"}
ld_decay %>%
  filter(!is.na(r2)) %>%
  ggplot(aes(x = dist_mb, y = r2)) +
  geom_point(alpha = 0.07, size = 0.5, colour = "grey50") +
  geom_smooth(method = "loess", span = 0.15,
              colour = "#3182bd", linewidth = 1, se = FALSE) +
  facet_grid(~Chr) +
  labs(x = "Physical distance (Mb)", y = expression(r^2),
       title = "Per-chromosome LD decay") +
  theme_bw(base_size = 11)
```

```{r ld-halflife-barplot, fig.height=4, fig.cap="LD half-life distance per chromosome"}
ld_hl_chr %>%
  filter(!is.na(halflife_Mb)) %>%
  ggplot(aes(x = Chr, y = halflife_Mb, fill = Chr)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(halflife_Mb, " Mb")),
            vjust = -0.3, size = 3.5) +
  scale_fill_brewer(palette = "Set3") +
  labs(x = "Chromosome", y = "LD half-life (Mb)",
       title = "Distance at which r² declines to half its maximum") +
  theme_bw(base_size = 13)
```

## LD Heatmaps (per chromosome, subsampled)

LD heatmaps are computed on a thinned set of markers per chromosome
(max `n_heatmap` markers equally spaced along the chromosome) to keep
computation tractable.

```{r ld-heatmaps, fig.height=7, cache=TRUE}
n_heatmap <- 200   # markers per chromosome for heatmap

for (chr in chromosomes) {
  chr_map <- map_sub %>% filter(Chr == chr) %>% arrange(Pos)
  if (nrow(chr_map) < 5) next

  # Thin to n_heatmap equally spaced markers
  idx_thin <- round(seq(1, nrow(chr_map), length.out = min(n_heatmap, nrow(chr_map))))
  chr_map  <- chr_map[idx_thin, ]
  G_chr    <- geno_mapped[, chr_map$Marker, drop = FALSE]
  n_m      <- ncol(G_chr)

  # r² matrix
  r2_mat <- matrix(NA_real_, n_m, n_m)
  for (i in seq_len(n_m)) {
    for (j in seq_len(i)) {
      r2_mat[i, j] <- r2_mat[j, i] <- r2_pair(G_chr[, i], G_chr[, j])
    }
  }
  diag(r2_mat) <- 1
  rownames(r2_mat) <- colnames(r2_mat) <- chr_map$Marker

  if (ld_heatmap_avail) {
    library(LDheatmap)
    LDheatmap(r2_mat,
              genetic.distances = chr_map$Pos / 1e6,
              distances         = "physical",
              LDmeasure         = "r",          # LDheatmap uses r not r²
              main              = paste0("LD heatmap – Chr ", chr,
                                         "  (n = ", n_m, " markers)"),
              color             = heat.colors(20),
              flip              = TRUE)
  } else {
    # Fallback: corrplot heatmap
    corrplot(r2_mat,
             method   = "color",
             col      = colorRampPalette(c("white","#FC8D59","#D73027"))(50),
             tl.pos   = "n",
             cl.cex   = 0.7,
             title    = paste0("r² heatmap – Chr ", chr,
                                " (n = ", n_m, " markers)"),
             mar      = c(0, 0, 2, 0))
  }
}
```

# Focal-SNP LD Plot

Set the three parameters below to plot r² between **one focal SNP** and every
other SNP within a chosen physical window on its chromosome. This is useful for
inspecting LD around a candidate marker (e.g. a GWAS hit or a QTL peak).

```{r focal-snp-params}
# ════════════════════════════════════════════════════════════
# USER PARAMETERS — edit these three lines
focal_snp    <- colnames(geno_a)[1]   # marker name, e.g. "BK_03"
window_mb    <- 10                    # ± distance in Mb around the focal SNP
r2_threshold <- 0.2                   # highlight markers above this r² (dashed line)
# ════════════════════════════════════════════════════════════

# Print available markers on each chromosome so the user can pick one
snp_map %>%
  group_by(Chr) %>%
  summarise(n_markers = n(),
            example_markers = paste(head(Marker, 3), collapse = ", "),
            .groups = "drop") %>%
  kable(caption = "Available chromosomes — pick a focal_snp from any of these") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

```{r focal-snp-ld, fig.height=5, fig.cap="r² between the focal SNP and neighbours within the chosen window"}
# ── Validate focal SNP ─────────────────────────────────────────────────────
if (!focal_snp %in% snp_map$Marker) {
  stop("focal_snp '", focal_snp, "' not found in the SNP map. ",
       "Check the marker name or choose another.")
}

focal_row  <- snp_map %>% filter(Marker == focal_snp)
focal_chr  <- focal_row$Chr
focal_pos  <- focal_row$Pos
window_bp  <- window_mb * 1e6

cat(sprintf("Focal SNP  : %s\n", focal_snp))
cat(sprintf("Chromosome : %s\n", focal_chr))
cat(sprintf("Position   : %.3f Mb\n", focal_pos / 1e6))
cat(sprintf("Window     : ± %.1f Mb  [%.3f – %.3f Mb]\n",
            window_mb,
            (focal_pos - window_bp) / 1e6,
            (focal_pos + window_bp) / 1e6))

# ── Select markers in window ───────────────────────────────────────────────
win_map <- snp_map %>%
  filter(Chr  == focal_chr,
         Pos  >= focal_pos - window_bp,
         Pos  <= focal_pos + window_bp,
         Marker %in% colnames(geno_a)) %>%
  arrange(Pos)

cat(sprintf("Markers in window: %d\n", nrow(win_map)))
if (nrow(win_map) < 2) stop("Fewer than 2 mapped markers in window — widen window_mb.")

# ── Compute r² between focal SNP and every window marker ──────────────────
focal_geno <- geno_imp[, focal_snp]

ld_focal <- win_map %>%
  mutate(
    r2       = map_dbl(Marker, function(m) r2_pair(focal_geno, geno_imp[, m])),
    dist_mb  = (Pos - focal_pos) / 1e6,    # signed: negative = upstream
    Pos_Mb   = Pos / 1e6,
    is_focal = Marker == focal_snp
  )

# ── Manhattan-style LD plot ───────────────────────────────────────────────
p_ld <- ggplot(ld_focal, aes(x = Pos_Mb, y = r2, colour = r2)) +
  # window boundary lines
  geom_vline(xintercept = c((focal_pos - window_bp) / 1e6,
                             (focal_pos + window_bp) / 1e6),
             linetype = "dotted", colour = "grey70") +
  # r² threshold line
  geom_hline(yintercept = r2_threshold, linetype = "dashed",
             colour = "grey40", linewidth = 0.6) +
  annotate("text", x = min(ld_focal$Pos_Mb),
           y = r2_threshold + 0.02, hjust = 0, size = 3, colour = "grey40",
           label = paste0("r² = ", r2_threshold)) +
  # all markers
  geom_point(size = 2.5, alpha = 0.85) +
  # focal SNP highlighted
  geom_point(data = filter(ld_focal, is_focal),
             size = 4, shape = 23, fill = "black", colour = "white") +
  # focal SNP label
  geom_text(data = filter(ld_focal, is_focal),
            aes(label = Marker), vjust = -1, size = 3, colour = "black") +
  scale_colour_gradientn(
    colours = c("steelblue","#fdae61","#d73027"),
    limits  = c(0, 1), name = expression(r^2)
  ) +
  scale_y_continuous(limits = c(0, 1.02), breaks = seq(0, 1, 0.2)) +
  labs(
    x     = sprintf("Position on Chr %s (Mb)", focal_chr),
    y     = expression(r^2),
    title = sprintf("LD around %s (Chr %s : %.3f Mb) | window ± %.1f Mb",
                    focal_snp, focal_chr, focal_pos / 1e6, window_mb)
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "right")

print(p_ld)

# ── Supplementary: r² vs signed distance ─────────────────────────────────
p_decay <- ggplot(ld_focal, aes(x = dist_mb, y = r2, colour = r2)) +
  geom_vline(xintercept = 0, linetype = "solid", colour = "grey30",
             linewidth = 0.4) +
  geom_hline(yintercept = r2_threshold, linetype = "dashed",
             colour = "grey40", linewidth = 0.6) +
  geom_point(size = 2.5, alpha = 0.85) +
  geom_point(data = filter(ld_focal, is_focal),
             size = 4, shape = 23, fill = "black", colour = "white") +
  geom_smooth(method = "loess", span = 0.3, se = FALSE,
              colour = "black", linewidth = 0.7, linetype = "dashed") +
  scale_colour_gradientn(
    colours = c("steelblue","#fdae61","#d73027"),
    limits  = c(0, 1), name = expression(r^2)
  ) +
  scale_y_continuous(limits = c(0, 1.02), breaks = seq(0, 1, 0.2)) +
  labs(
    x     = "Distance from focal SNP (Mb)",
    y     = expression(r^2),
    title = "r² vs signed distance from focal SNP"
  ) +
  theme_bw(base_size = 13)

print(p_decay)

# ── Table: top markers in high LD with focal SNP ─────────────────────────
ld_focal %>%
  filter(!is_focal, !is.na(r2)) %>%
  arrange(desc(r2)) %>%
  head(20) %>%
  mutate(Pos_Mb   = round(Pos_Mb,   4),
         dist_mb  = round(dist_mb,  3),
         r2       = round(r2,       3)) %>%
  select(Marker, Chr, Pos_Mb, dist_mb, r2) %>%
  kable(col.names = c("Marker","Chr","Position (Mb)","Distance (Mb)","r²"),
        caption   = sprintf("Top 20 markers in highest LD with %s", focal_snp)) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

# Summary

```{r summary-table}
tibble(
  Analysis    = c("Markers total","Markers mapped","Individuals",
                  "Mean MAF","Missing rate",
                  "PCs explaining 50% var","PCs explaining 80% var",
                  "Optimal K (DAPC)","Genome-wide LD half-life (Mb)"),
  Value = c(
    n_snp,
    length(mapped_markers),
    n_ind,
    round(mean(maf_vec, na.rm = TRUE), 3),
    round(mean(miss_vec) * 100, 2),
    sum(cumsum(var_exp) <= 50) + 1,
    sum(cumsum(var_exp) <= 80) + 1,
    if (exists("k_opt")) k_opt else NA,
    round(half_dist, 2)
  )
) %>%
  kable(caption = "Analysis summary") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

```{r session-info}
sessionInfo()
```
