---
title: "2. Trait heritability analysis"
author: "sss322"
date: "2025-04-11"
output: html_document
---


```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning=FALSE, results = "hide")
```


## **Introduction**
This document provides cotains the heritability calculation for multi-spectral drone imagery vegetative indices, maltinq quality and agronomic traits. The goal is to assess the heritability of all traits and choose the ones to proceed with.


## **0. Import packages and data**
```{r, message=FALSE}
pkgs <- c("MASS","dplyr","tidyr","asreml","ASRtriala","ASRgenomics",
          "ggplot2","openxlsx","workflowr","readxl", "forcats", "scales")
suppressPackageStartupMessages(invisible(lapply(pkgs, require, character.only = TRUE)))
```

```{r}
paths <- c("C:/Users/Siim Sepp/NY_WMB/data",
           "~/Documents/GitHub/NY_WMB/data",
           "data")

files <- c("data_ag.Rdata","data_mq.Rdata","data_sp.Rdata","VImat.Rdata")

load_first <- function(fname, dirs = paths) {
  for (d in dirs) {
    p <- file.path(path.expand(d), fname)
    if (file.exists(p)) { load(p, envir = .GlobalEnv); return(invisible(p)) }
  }
  stop("File not found in any known location: ", fname)
}
invisible(lapply(files, load_first))
```





## **2. Calculating heritabilites.** 


### **Calculating heritability for agronomic traits**
```{r, message=FALSE}
#Select traits and environments
data_ag<- data_ag %>% arrange(Env)
traits <- colnames(data_ag)[7]
envs   <- levels(data_ag$Env)[1:2]
data<- data_ag 
data$GID<- as.character(data$GID)

#######################################################
#Braod sense heritability for GY 21 and 24
######################################################
model_list <- list()
ratio_list <- list()

# Loop over each trait
for (tr in traits) {
  
  # Nested list/vector for storing results across environments
  model_env  <- list()
  ratio_env  <- numeric(length(envs)) 
  # Loop over each environment
  for (i in seq_along(envs)) {
    e <- envs[i]
    # Create a temporary subset for environment e
    data_sub <- data %>% 
      dplyr::filter(Env == e)
    data_sub$GID<- as.factor(data_sub$GID)
    # Center the trait by subtracting its mean (within env)
    data_sub[[tr]] <- data_sub[[tr]] - mean(data_sub[[tr]], na.rm = TRUE)
    # Fit the model
    mod_M1 <- tryCatch({
      asreml(
        fixed    = as.formula(paste(tr, "~  1")),
        random   = ~ GID  +  ~ar1(Row):ar1(Column),,
        residual = ~ units,
        data     = data_sub   )
    }, error = function(err) {
      
      message(paste("Error fitting trait:", tr, "in environment:", e))
      message("Error message:", err$message)
      return(NA)
    })
    # Store the model in a nested list
    model_env[[e]] <- mod_M1
    plot(mod_M1)
    # If successful, compute the variance ratio
    if (inherits(mod_M1, "asreml")) {
      vc <- summary(mod_M1)$varcomp
      rep_counts <- table(data_sub$GID)
      r <- mean(rep_counts)
      # Assuming first row is genetic and second row is residual
      ratio_env[i] <- vc[1, 1] / (vc[1, 1]  + vc[2, 1]+ vc[5, 1]/r)

    } else {
      ratio_env[i] <- NA
    }
  }
  model_list[[tr]] <- model_env
  ratio_list[[tr]] <- ratio_env
}
# Create a data frame from the ratio_list
ratio_df <- data.frame(
  Trait       = rep(traits, each = length(envs)),
  Environment = rep(envs, times = length(traits)),
  Ratio       = unlist(ratio_list),
  stringsAsFactors = FALSE
)


#######################################################
#Broad sense heritability for GY 22, 23 and 25
######################################################
envs   <- levels(data_ag$Env)[3:5]
model_list <- list()
ratio_list <- list()

# Loop over each trait
for (tr in traits) {
  
  # Nested list/vector for storing results across environments
  model_env  <- list()
  ratio_env  <- numeric(length(envs)) 
  
  # Loop over each environment
  for (i in seq_along(envs)) {
    e <- envs[i]
    
    # Create a temporary subset for environment e
    data_sub <- data %>% 
      dplyr::filter(Env == e)
    data_sub$GID<- as.factor(data_sub$GID)
    # Center the trait by subtracting its mean (within env)
    data_sub[[tr]] <- data_sub[[tr]] - mean(data_sub[[tr]], na.rm = TRUE)
    
    # Fit the model
    mod_M1 <- tryCatch({
      
      asreml(
        fixed    = as.formula(paste(tr, "~  Row + Column")),
        random   = ~ GID,
        residual = ~ units,
        data     = data_sub   )
      
          }, error = function(err) {
            message(paste("Error fitting trait:", tr, "in environment:", e))
      message("Error message:", err$message)
      return(NA)
    })
        # Store the model in a nested list
    model_env[[e]] <- mod_M1
    plot(mod_M1)
    # If successful, compute the variance ratio
    if (inherits(mod_M1, "asreml")) {
      vc <- summary(mod_M1)$varcomp
      rep_counts <- table(data_sub$GID)
      r <- mean(rep_counts)
      print(r)
      # Assuming first row is genetic and second row is residual
      ratio_env[i] <- vc[1, 1] / (vc[1, 1]  + vc[2, 1]/r)
    } else {
      ratio_env[i] <- NA
    }  }
  # Store all environment-specific models for this trait
  model_list[[tr]] <- model_env
  # Store the vector of ratio values across environments
  ratio_list[[tr]] <- ratio_env
}


# Create a data frame from the ratio_list
ratio_df1 <- data.frame(
  Trait       = rep(traits, each = length(envs)),
  Environment = rep(envs, times = length(traits)),
  Ratio       = unlist(ratio_list),
  stringsAsFactors = FALSE
)

H2ag<-rbind(ratio_df, ratio_df1)
```


### **Calculating heritability for heading traits**
```{r, message=FALSE}
#Select traits and environments
traits <- colnames(data_ag)[c(8,9)]
envs   <- levels(data_ag$Env)

#######################################################
#Broad sense heritability for HD 21-25
######################################################
model_list <- list()
ratio_list <- list()

# Loop over each trait
for (tr in traits) {
  
  # Nested list/vector for storing results across environments
  model_env  <- list()
  ratio_env  <- numeric(length(envs)) 
  # Loop over each environment
  for (i in seq_along(envs)) {
    e <- envs[i]
    # Create a temporary subset for environment e
    data_sub <- data %>% 
      dplyr::filter(Env == e)
    data_sub$GID<- as.factor(data_sub$GID)
    # Fit the model
    mod_M1 <- tryCatch({
      asreml(
        fixed    = as.formula(paste(tr, "~ 1")),
        random   = ~ GID + ~ar1(Row):ar1(Column),
        residual = ~ units,
        data     = data_sub   )
    }, error = function(err) {
      
      message(paste("Error fitting trait:", tr, "in environment:", e))
      message("Error message:", err$message)
      return(NA)
    })
    # Store the model in a nested list
    model_env[[e]] <- mod_M1
    plot(mod_M1)
    # If successful, compute the variance ratio
    if (inherits(mod_M1, "asreml")) {
      vc <- summary(mod_M1)$varcomp
      rep_counts <- table(data_sub$GID)
      r <- mean(rep_counts)
      # Assuming first row is genetic and second row is residual
      ratio_env[i] <- vc[1, 1] / (vc[1, 1] + vc[5, 1]/r)

    } else {
      ratio_env[i] <- NA
    }
  }
  model_list[[tr]] <- model_env
  ratio_list[[tr]] <- ratio_env
}
# Create a data frame from the ratio_list
ratio_df1 <- data.frame(
  Trait       = rep(traits, each = length(envs)),
  Environment = rep(envs, times = length(traits)),
  Ratio       = unlist(ratio_list),
  stringsAsFactors = FALSE
)

H2ag <-rbind(H2ag, ratio_df1)
colnames(H2ag)[3]<- "H2"
H2ag$H2<- round(H2ag$H2, 2)
H2ag$h2<- "NA"
H2ag$H2[7]<- 0
```



### **Calculating narrow sense heritability for ag traits**
```{r, message=FALSE}
#Select traits and environments
traits <- colnames(data_ag)[7:9]
envs   <- levels(data_ag$Env)

#######################################################
#Narrow sense heritability  21-25
######################################################
model_list <- list()
ratio_list <- list()

# Loop over each trait
for (tr in traits) {
  
  # Nested list/vector for storing results across environments
  model_env  <- list()
  ratio_env  <- numeric(length(envs)) 
  # Loop over each environment
  for (i in seq_along(envs)) {
    e <- envs[i]
    # Create a temporary subset for environment e
    data_sub <- data %>% 
      dplyr::filter(Env == e)
    load("data/geno/WMB_master_geno.RData")
    indp = data_sub$GID %in% rownames(A1)
    data_sub = data_sub[indp,]
    indg = rownames(A1) %in% data_sub$GID
    A1 = A1[indg, indg]
    Ghat.bend <- G.tuneup(G = A1, bend = TRUE, eig.tol = 1e-03)$Gb
    Ginv.sparse <- G.inverse(G = Ghat.bend, sparse = TRUE)$Ginv
    ASRgenomics::match.kinship2pheno(K = A1, pheno.data = data_sub, indiv = "GID")
    data_sub$GID <- as.factor(data_sub$GID)
    
    # Fit the model
    mod_M1 <- tryCatch({
      asreml(
        fixed    = as.formula(paste(tr, "~ 1")),
        random   = ~ vm(GID, Ginv.sparse) + ~ar1(Row):ar1(Column),
        residual = ~ units,
        data     = data_sub   )
    }, error = function(err) {
      
      message(paste("Error fitting trait:", tr, "in environment:", e))
      message("Error message:", err$message)
      return(NA)
    })
    # Store the model in a nested list
    model_env[[e]] <- mod_M1
    plot(mod_M1)
    # If successful, compute the variance ratio
    if (inherits(mod_M1, "asreml")) {
      vc <- summary(mod_M1)$varcomp
      rep_counts <- table(data_sub$GID)
      r <- mean(rep_counts)
      # Assuming first row is genetic and second row is residual
      ratio_env[i] <- vc[1, 1] / (vc[1, 1] + vc[5, 1]/r)

    } else {
      ratio_env[i] <- NA
    }
  }
  model_list[[tr]] <- model_env
  ratio_list[[tr]] <- ratio_env
}
# Create a data frame from the ratio_list
ratio_df1 <- data.frame(
  Trait       = rep(traits, each = length(envs)),
  Environment = rep(envs, times = length(traits)),
  Ratio       = unlist(ratio_list),
  stringsAsFactors = FALSE
)

H2ag$h2<- ratio_df1$Ratio
H2ag1<- H2ag
```


### **Heritability for MQ traits**
```{r}
data_mq$Env<- as.character(data_mq$Env)
data<- data_mq
data$Env<- as.factor(data$Env)
traits <- colnames(data_mq)[c(4:15)]
envs   <- levels(data$Env)#

#######################################################
#Broad sense heritability for MQ 23-24
######################################################
ratio_rows <- list()

for (tr in traits) {
  model_env <- list()
  
  for (e in envs) {
    data_sub <- data %>%
      filter(Env == e) %>%
      filter(.data[[tr]] > 0)
    data_sub<- data_sub %>% filter(!Fam == "NA") %>% filter(!GID == "NA")
    data_sub$Fam<- as.factor(data_sub$Fam)
    # Preprocessing as before...
    load("data/geno/WMB_master_geno.RData")
    indp = data_sub$GID %in% rownames(A1)
    data_sub = data_sub[indp,]
    indg = rownames(A1) %in% data_sub$GID
    A1 = A1[indg, indg]
    Ghat.bend <- G.tuneup(G = A1, bend = TRUE, eig.tol = 1e-03)$Gb
    Ginv.sparse <- G.inverse(G = Ghat.bend, sparse = TRUE)$Ginv
    ASRgenomics::match.kinship2pheno(K = A1, pheno.data = data_sub, indiv = "GID")
    data_sub$GID <- as.factor(data_sub$GID)
    
    # Center trait
    data_sub[[tr]] <- data_sub[[tr]] - mean(data_sub[[tr]], na.rm = TRUE)
    
    mod_M1 <- tryCatch({
      asreml(
        fixed    = as.formula(paste(tr, "~ 1")),
        random   = ~ vm(GID, Ginv.sparse) + ~Fam,
        residual = ~ units,
        data     = data_sub,
        ai.sing  = TRUE
      )
    }, error = function(err) {
      message(paste("Error fitting trait:", tr, "in environment:", e))
      return(NA)
    })
    
    model_env[[e]] <- mod_M1
    
    if (inherits(mod_M1, "asreml")) {
      vc <- summary(mod_M1)$varcomp
      r <- mean(table(data_sub$GID))
      H2 <- (vc[1, 1] + + vc[2, 1]) / (vc[1, 1] + vc[2, 1] + vc[3, 1] / r)
      h2 <- vc[1, 1]  / (vc[1, 1] + vc[2, 1] + vc[3, 1] / r)
      
      ratio_rows[[length(ratio_rows) + 1]] <- data.frame(
        Trait = tr,
        Environment = e,
        H2 = H2,
        h2 = h2
      )
    }
  }
  
  model_list[[tr]] <- model_env
}

# Combine all rows into a single data frame
H2mq <- bind_rows(ratio_rows)
H2mq$H2<- round(H2mq$H2, 2)
H2mq$h2<- round(H2mq$h2, 2)
```




### **Heritability for spectral traits**
```{r}
data<- merge(data_sp, data_ag[,c(1,5,6)], by= "plot_id")
colnames(data)[4]<- "timepoint"
traits <- colnames(data)[5:13][c(1,9)]
data$Env<- as.factor(data$Env)
envs   <- levels(data$Env)[-2]
data$GID<- as.character(data$GID)

# Prepare lists to store models and ratio values
model_list <- list()
ratio_list <- list()

#######################################################
#Broad sense heritability for VI 22-25
######################################################
## ------------------------------------------------------------------
## outer loop : traits
## ------------------------------------------------------------------
for (tr in traits) {

  model_env  <- list()
  ratio_env  <- list()

  ## ---------------------------------------------------------------
  ## middle loop : environments
  ## ---------------------------------------------------------------
  for (e in envs) {

    data_env <- data %>% 
      filter(Env == e, .data[[tr]] > 0)           # keep only positive values

    ## nothing to do if env has no data for this trait
    if (nrow(data_env) == 0) next

    ## time‑points present in this env × trait subset
    tps <- sort(unique(data_env$timepoint))
    

    model_tp <- list()
    ratio_tp <- numeric(length(tps))

    ## ------------------------------------------------------------
    ## inner loop : time‑points
    ## ------------------------------------------------------------
    for (j in seq_along(tps)) {

      tp <- tps[j]

      data_sub <- data_env %>% 
        filter(timepoint == tp) %>% 
        mutate(
          GID      = factor(GID),
          !!tr     := .data[[tr]] - mean(.data[[tr]], na.rm = TRUE)  # centre
        )


      mod_M1 <- tryCatch(
        asreml(
          fixed    = reformulate("1", response = tr),
          random   = ~GID +  ~ar1(Row):ar1(Column),
          residual = ~ units,
          data     = data_sub,
          ai.sing  = TRUE
        ),
        error = function(err) {
          message("Error fitting trait ", tr,
                  " | env ", e, " | tp ", tp, "\n  ",
                  err$message)
          NA
        }
      )

      model_tp[[as.character(tp)]] <- mod_M1

      ## --------- variance ratio if model converged ---------------
      if (inherits(mod_M1, "asreml")) {
        vc          <- summary(mod_M1)$varcomp
        rep_counts  <- table(data_sub$GID)
        r           <- mean(rep_counts)
        ratio_tp[j] <- vc[1, 1] / (vc[1, 1] + vc[5, 1] / r)
      } else {
        ratio_tp[j] <- NA
      }
    } # end time‑point loop

    names(model_tp) <- names(ratio_tp) <- as.character(tps)
    model_env [[e]] <- model_tp
    ratio_env [[e]] <- ratio_tp
  } # end environment loop

  model_list[[tr]] <- model_env
  ratio_list[[tr]] <- ratio_env
} # end trait loop

## ------------------------------------------------------------------
## flatten the ratio_list to a data‑frame
## ------------------------------------------------------------------
ratio_df <- purrr::imap_dfr(ratio_list, ~{
  tr_name <- .y
  purrr::imap_dfr(.x, ~{
    env_name <- .y
    tibble(
      Trait       = tr_name,
      Environment = env_name,
      Timepoint   = as.numeric(names(.x)),
      Ratio       = .x
    )
  })
})

H2sp <- ratio_df



#######################################################
#Broad sense heritability for VI 21
######################################################
envs   <- levels(data$Env)[2]
data$GID<- as.character(data$GID)

# Prepare lists to store models and ratio values
model_list <- list()
ratio_list <- list()

## ------------------------------------------------------------------
## outer loop : traits
## ------------------------------------------------------------------
for (tr in traits) {

  model_env  <- list()
  ratio_env  <- list()

  ## ---------------------------------------------------------------
  ## middle loop : environments
  ## ---------------------------------------------------------------
  for (e in envs) {

    data_env <- data %>% 
      filter(Env == e, .data[[tr]] > 0)           # keep only positive values

    ## nothing to do if env has no data for this trait
    if (nrow(data_env) == 0) next

    ## time‑points present in this env × trait subset
    tps <- sort(unique(data_env$timepoint))
    

    model_tp <- list()
    ratio_tp <- numeric(length(tps))

    ## ------------------------------------------------------------
    ## inner loop : time‑points
    ## ------------------------------------------------------------
    for (j in seq_along(tps)) {

      tp <- tps[j]

      data_sub <- data_env %>% 
        filter(timepoint == tp) %>% 
        mutate(
          GID      = factor(GID),
          !!tr     := .data[[tr]] - mean(.data[[tr]], na.rm = TRUE)  # centre
        )


      mod_M1 <- tryCatch(
        asreml(
          fixed    = reformulate("1", response = tr),
          random   = ~GID ,#+  ~ar1(Row):ar1(Column),
          residual = ~ units,
          data     = data_sub,
          ai.sing  = TRUE
        ),
        error = function(err) {
          message("Error fitting trait ", tr,
                  " | env ", e, " | tp ", tp, "\n  ",
                  err$message)
          NA
        }
      )

      model_tp[[as.character(tp)]] <- mod_M1

      ## --------- variance ratio if model converged ---------------
      if (inherits(mod_M1, "asreml")) {
        vc          <- summary(mod_M1)$varcomp
        rep_counts  <- table(data_sub$GID)
        r           <- mean(rep_counts)
        ratio_tp[j] <- vc[1, 1] / (vc[1, 1] + vc[2, 1] / r)
        #ratio_tp[j] <- vc[1, 1] / (vc[1, 1] + vc[5, 1] / r)
      } else {
        ratio_tp[j] <- NA
      }
    } # end time‑point loop

    names(model_tp) <- names(ratio_tp) <- as.character(tps)
    model_env [[e]] <- model_tp
    ratio_env [[e]] <- ratio_tp
  } # end environment loop

  model_list[[tr]] <- model_env
  ratio_list[[tr]] <- ratio_env
} # end trait loop

## ------------------------------------------------------------------
## flatten the ratio_list to a data‑frame
## ------------------------------------------------------------------
ratio_df <- purrr::imap_dfr(ratio_list, ~{
  tr_name <- .y
  purrr::imap_dfr(.x, ~{
    env_name <- .y
    tibble(
      Trait       = tr_name,
      Environment = env_name,
      Timepoint   = as.numeric(names(.x)),
      Ratio       = .x
    )
  })
})

H2sp <- rbind(H2sp, ratio_df)
###########################################################################################################
colnames(H2sp)
colnames(H2sp)[4]<- c("H2")
H2sp$h<- "NULL"
H2sp$H2 <- round(H2sp$H2, 2)
```



### **Heritability for the PC-s**
```{r, eval=TRUE}
load("data/F_PCA_scoresG.Rdata")
F_PC_scores<- merge(PCA_scores_df, FPCA_scores_df, by= c("Env","VI", "plot_id"))
colnames(F_PC_scores)
F_PC_scores<- F_PC_scores[,c(1:6,13:15)]
# Assuming your dataframe is called df_pcs
F_PC_wide <- F_PC_scores %>%
  pivot_longer(
    cols = contains("PC"),
    names_to = "Component",
    values_to = "Loadings"
  ) %>%
  unite("VI_Component", VI, Component, sep = "_") %>%
  pivot_wider(
    names_from = VI_Component,
    values_from = Loadings
  )


df_ag <- merge(data_ag, F_PC_wide, by= c("plot_id", "Env"))
data_ag<- df_ag %>% arrange(Env)

#Select traits and environments
traits <- colnames(F_PC_wide)[c(33:38, 45:50)]
envs   <- unique(df_ag$Env)
data<- data_ag 
#######################################################
#Broad sense heritability for FPCA and PCA 22 and 25
######################################################
model_list <- list()
ratio_list <- list()

# Loop over each trait
for (tr in traits) {
  
  # Nested list/vector for storing results across environments
  model_env  <- list()
  ratio_env  <- numeric(length(envs)) 
  
  # Loop over each environment
  for (i in seq_along(envs)) {
    e <- envs[i]
    
    # Create a temporary subset for environment e
    data_sub <- data %>% 
      dplyr::filter(Env == e)
    data_sub$GID<- as.factor(data_sub$GID)

    
    # Center the trait by subtracting its mean (within env)
    data_sub[[tr]] <- data_sub[[tr]] - mean(data_sub[[tr]], na.rm = TRUE)
    
    # Fit the model
    mod_M1 <- tryCatch({
      
      asreml(
        fixed    = as.formula(paste(tr, "~ 1")),
        random   = ~ GID + ~ar1(Row):ar1(Column),
        residual = ~ units,
        data     = data_sub,
        ai.sing  = TRUE   # Helps diagnose singularities
      )
      
      
    }, error = function(err) {
      
      message(paste("Error fitting trait:", tr, "in environment:", e))
      message("Error message:", err$message)
      return(NA)
    })
    
    # Store the model in a nested list
    model_env[[e]] <- mod_M1
    
    # If successful, compute the variance ratio
    if (inherits(mod_M1, "asreml")) {
      vc <- summary(mod_M1)$varcomp
      rep_counts <- table(data_sub$GID)
      r <- mean(rep_counts)
      print(r)
      # Assuming first row is genetic and second row is residual
      ratio_env[i] <- vc[1, 1] / (vc[1, 1] + vc[5, 1]/r)
    } else {
      ratio_env[i] <- NA
    }
  }
  
  # Store all environment-specific models for this trait
  model_list[[tr]] <- model_env
  # Store the vector of ratio values across environments
  ratio_list[[tr]] <- ratio_env
}

# Create a data frame from the ratio_list
ratio_df <- data.frame(
  Trait       = rep(traits, each = length(envs)),
  Environment = rep(envs, times = length(traits)),
  Ratio       = unlist(ratio_list),
  stringsAsFactors = FALSE
)

H2PC1 <- ratio_df

#######################################################
#Broad sense heritability for FPCA and PCA 21, 23, 24
#######################################################

envs   <- unique(df_ag$Env)[c(1,3,4)]
data<- data_ag 

model_list <- list()
ratio_list <- list()

# Loop over each trait
for (tr in traits) {
  
  # Nested list/vector for storing results across environments
  model_env  <- list()
  ratio_env  <- numeric(length(envs)) 
  
  # Loop over each environment
  for (i in seq_along(envs)) {
    e <- envs[i]
    
    # Create a temporary subset for environment e
    data_sub <- data %>% 
      dplyr::filter(Env == e)
    data_sub$GID<- as.factor(data_sub$GID)

    
    # Center the trait by subtracting its mean (within env)
    data_sub[[tr]] <- data_sub[[tr]] - mean(data_sub[[tr]], na.rm = TRUE)
    
    # Fit the model
    mod_M1 <- tryCatch({
      
      asreml(
        fixed    = as.formula(paste(tr, "~ 1")),
        random   = ~ GID ,#+ ~ar1(Row):ar1(Column),
        residual = ~ units,
        data     = data_sub,
        ai.sing  = TRUE   # Helps diagnose singularities
      )
      
      
    }, error = function(err) {
      
      message(paste("Error fitting trait:", tr, "in environment:", e))
      message("Error message:", err$message)
      return(NA)
    })
    
    # Store the model in a nested list
    model_env[[e]] <- mod_M1
    
    # If successful, compute the variance ratio
    if (inherits(mod_M1, "asreml")) {
      vc <- summary(mod_M1)$varcomp
      rep_counts <- table(data_sub$GID)
      r <- mean(rep_counts)
      print(r)
      # Assuming first row is genetic and second row is residual
      ratio_env[i] <- vc[1, 1] / (vc[1, 1] + vc[2, 1]/r)
    } else {
      ratio_env[i] <- NA
    }
  }
  
  # Store all environment-specific models for this trait
  model_list[[tr]] <- model_env
  # Store the vector of ratio values across environments
  ratio_list[[tr]] <- ratio_env
}

# Create a data frame from the ratio_list
ratio_df <- data.frame(
  Trait       = rep(traits, each = length(envs)),
  Environment = rep(envs, times = length(traits)),
  Ratio       = unlist(ratio_list),
  stringsAsFactors = FALSE
)
H2PC <- ratio_df
H2PC<- rbind(H2PC, H2PC1)
###########################################################################################################
colnames(H2PC)[3]<- c("H2")
H2PC$h<- "NULL"
H2PC$H2<- round(H2PC$H2, 2)
table(H2PC$Environment)
############################################################################
h2s<- list(H2ag, H2sp, H2mq, H2PC)
writexl::write_xlsx(h2s, path= "heritabilites.xlsx")
save(H2sp, H2ag, H2mq, H2PC, file = "heritabilites.Rdata")
```




## **3. Plots for heritability**


```{r}
load("C:/Users/Siim Sepp/NY_WMB/heritabilites.Rdata")
```


### **VI heritability**
```{r, results='markup'}
#Heading date
H2sp<- H2sp %>% filter(Trait %in% c( "NDVI", "PSRI"))
H2sp<- H2sp %>% filter(Trait == "NDVI")
H2sp$Environment<- as.factor(H2sp$Environment)

# Desired order
env_order <- c("KET21", "SNY22", "MCG23", "HELF24", "MCG25")

# Reorder factor levels in the data frame
H2sp <- H2sp %>%
  mutate(Environment = factor(Environment, levels = env_order))

load("data/wmb_multi_alldata.Rdata")
#load("~/Documents/GitHub/NY_WMB/data/wmb_multi_alldata.Rdata")
timepoint_JD <- multi[,c(1, 3, 9, 10, 30, 31)] %>%
  distinct()  # Keep only unique combinations

colnames(timepoint_JD)[1] <- "Environment"

H2sp <- H2sp %>%
  left_join(timepoint_JD, by = c("Environment","Timepoint"))

summary(H2sp)
H2sp$JD<- as.numeric(H2sp$DAS)
###############################################################################
H2sp_summary <- H2sp %>%
  group_by(Environment, Trait) %>%
  summarize(
    mean_H = mean(H2, na.rm = TRUE),
    min_H = min(H2, na.rm = TRUE), 
    ma_H = max(H2, na.rm = TRUE),
    n      = n()) %>% 
  arrange(Trait)


H2sp_summary1 <- H2sp1 %>%
  group_by(Trait) %>%
  summarize(
    mean_H = mean(H2, na.rm = TRUE),
    min_H = min(H2, na.rm = TRUE), 
    ma_H = max(H2, na.rm = TRUE),
    n      = n()) %>% 
  arrange(Trait)

H2sp1<- H2sp %>% filter(H2 > 0)

#######################################################################
env_order<- c( "KET21", "SNY22" ,"MCG23","HELF24" , "MCG25" )

pal <- c(
  KET21  = "cornsilk3",
  SNY22  = "palegreen2",
  MCG23  = "wheat2",
  HELF24 = "lightblue3", 
  MCG25 = "yellow3"
)
H2sp
H2sp<- H2sp %>% arrange(Trait,Environment,  Timepoint)
#H2sp$Timepoint[13:21] <- H2sp$Timepoint[13:21] - 11
#H2sp$Timepoint[1:12] <- H2sp$Timepoint[1:12] - 1

H2sp$Timepoint[13:24] <- H2sp$Timepoint[13:24] - 1
H2sp$Timepoint[25:33] <- H2sp$Timepoint[25:33] - 11
H2sp$Timepoint[45:54] <- H2sp$Timepoint[45:54] - 1
H2sp$Timepoint[55:58] <- H2sp$Timepoint[55:58] - 11

H2sp <- H2sp %>%  mutate( Environment = trimws(as.character(Environment)),
    Environment = fct_relevel(Environment, env_order)  
  )

ggplot(H2sp, aes(x = Timepoint, y = H2, fill = Environment)) +
  geom_col() +
  scale_fill_manual(values = pal, drop =  FALSE) +
  facet_grid(Trait~Environment, scales= "free_x") +
  labs(x = "Flight", y = "H") +
  scale_y_continuous(breaks = c(0, 0.25,0.5, 0.75, 1)) +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text  = element_text(size = 10, angle = 90),  legend.position = "bottom"
  )

ggplot(H2sp, aes(x = Environment, y = H2)) +
  geom_boxplot() +
  scale_fill_manual(values = pal, drop =  FALSE) +
  facet_grid(~Trait, scales= "free_x") +
  labs(x = "Flight", y = "H") +
  scale_y_continuous(breaks = c(0, 0.25,0.5, 0.75, 1)) +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text  = element_text(size = 10, angle = 90),  legend.position = "bottom"
  )
```



### **MQ and ag heritability**
```{r, results='markup'}
str(H2ag)
str(H2mq)
H2ag<- H2ag[,-4]
H2mq<- H2mq[,-4]
H2s<- rbind(H2ag,H2mq)

unique(H2ag$Trait) 

trait_map <- c(
  "winter_survival"  = "WS",
  "KernelWeight.mg." = "KW",
  "on6.64...."       = "on6/64", 
  "MaltExtract..."   = "ME",
  "BarleyProtein..." = "BP",
  "WortProtein..."   = "WP",
  "S.T"              = "S.T", 
  "DP..ASBC."       = "DP", 
  "Alpha.amylase.20.DU." = "AA", 
  "Beta.glucan.ppm." = "BG",
  "HD_JD" = "HD", 
  "yield_kgha" = "GY", 
  "FAN.ppm."         = "FAN", 
  "AdjunctQualityScore" = "Adj.Sc",
  "All_MaltQualityScore" = "All.Sc")


H2s <- H2s %>%
  mutate(Trait = recode(Trait, !!!trait_map))

pal <- c(
  KET21  = "cornsilk3",
  SNY22  = "palegreen2",
  MCG23  = "violet",
  HELF24 = "lightblue3", 
  MCG25 = "grey"
)

pal <- c(
  KET21  = "cornsilk3",
  SNY22  = "palegreen2",
  MCG23  = "wheat2",
  HELF24 = "lightblue3", 
  MCG25 = "yellow3"
)

trait_order<- c( "GY" , "HD",  "WS", "KW", "on6/64" , "FAN" , "DP", "AA", "S.T" , "BG" , "ME",  "BP", "WP", "Adj.Sc", "All.Sc" )
env_order<- c( "KET21", "SNY22" ,"MCG23","HELF24" , "MCG25" )
H2s <- H2s %>%  mutate( Environment = trimws(as.character(Environment)),
    Environment = fct_relevel(Environment, env_order)  
  ) %>% mutate( Trait= trimws(as.character(Trait)),
    Trait = fct_relevel(Trait, trait_order)  
  )

H2s$H2 <- as.numeric(H2s$H2)

ggplot(H2s, aes(x=Environment,  y = H2, fill = Environment)) +
  geom_col()+
  scale_fill_manual(values= pal, drop= FALSE)+
  facet_grid(~Trait, scales= "free_x") +
  labs(x = "Environment", y = "H^2") +
  scale_y_continuous(breaks = c(0, 0.25,0.5, 0.75, 1)) +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text  = element_text(size = 8, angle = 90),
        legend.position = "bottom"
  )


H2s<- H2s %>% filter(!Trait == "All.Sc")

ggplot(H2s, aes(x=Environment,  y = h2, fill = Environment)) +
  geom_col()+
  scale_fill_manual(values= pal, drop= FALSE)+
  facet_grid(~Trait, scales= "free_x") +
  labs(x = "Environment", y = "H^2") +
  scale_y_continuous(breaks = c(0, 0.25,0.5, 0.75, 1)) +
  theme(
    strip.text = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 16),
    axis.text  = element_text(size = 12, angle = 90,), 
    legend.position = "bottom"
  )


H2s_summary <- H2s %>%
  group_by(Trait) %>%
  summarize(
    mean_H = mean(H2, na.rm = TRUE),
    min_H = min(H2, na.rm = TRUE), 
    max_H = max(H2, na.rm = TRUE),
    mean_h = mean(h2, na.rm = TRUE),
    min_h = min(h2, na.rm = TRUE), 
    max_h = max(h2, na.rm = TRUE),
    n      = n()) %>% 
  arrange(mean_H)


H2sp_summary1 <- H2sp1 %>%
  group_by(Trait) %>%
  summarize(
    mean_H = mean(H2, na.rm = TRUE),
    min_H = min(H2, na.rm = TRUE), 
    ma_H = max(H2, na.rm = TRUE),
    n      = n()) %>% 
  arrange(Trait)

```



###  **Plots for PC and FPC heritability**
```{r, results='markup'}
#Broad sene heritability of VI_PC-s over Envs. 
H2PC$Environment <- factor(H2PC$Environment, levels = env_order)
H2PC<- H2PC %>% arrange(Environment)
ips <- unique(H2PC$Trait)[31:36]
H2PC<- H2PC %>% filter(Trait %in% ips)

ggplot(H2PC, aes(x = Trait, y = H2, fill = Environment)) +
  geom_col() +
  scale_fill_manual(values = pal, drop =  FALSE) +
  facet_grid(~Environment, scales = "free") +
  labs(x = "Trait", y = "H") +
  scale_y_continuous(breaks = c(0, 0.25,0.5, 0.75, 1)) +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text  = element_text(size = 10, angle = 90),  legend.position = "bottom"
  )
```

