---
title: "3. Spectral data Spatial BLUE"
output: html_document
author: "sss322"
date: "2024-12_06"
---


```{r setup, include=FALSE, warning=FALSE, eval=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# Introduction
This document provides an overview of the spatial BLUE wth the multi-spectral drone imagery data, agronomic and malting quality data. The goal is to conduct the first BLUE stage with the vegetation indices and explore data distribution.



## **1. Importing Dependencies **

Loading dependencies
```{r, message=FALSE}
library(readr)
library(readxl)
library(dplyr)
library(MASS)
library(ggplot2)
library(tidyr)
library(ASRgenomics)
library(factoextra)
library(fda)
library(ASRtriala)
library(writexl)
library(openxlsx)
library(jsonlite)
library(data.table)
library(asreml)
library(dplyr)
library(purrr)
Sys.setlocale("LC_TIME", "C")
```



## **2. Setting the working directory and importing the data**

```{r, message=FALSE}
#load("C:/Users/Siim Sepp/NY_WMB/data/WMB_pheno_all.Rdata")
load("C:/Users/Siim Sepp/NY_WMB/data/data_sp.Rdata")
load("C:/Users/Siim Sepp/NY_WMB/data/data_mq.Rdata")
load("C:/Users/Siim Sepp/NY_WMB/data/data_ag.Rdata")

#load("~/Documents/GitHub/NY_WMB/data/WMB_pheno_all.Rdata")
table(data_sp$Env, data_sp$Timepoint)
table(data_mq$Env)
data_mq1<- data_mq %>% drop_na()
```

```{r, message=FALSE}
data <- data_sp %>% arrange(Timepoint) #%>%
  #filter(
   # !(Env == "SNY22" & Timepoint == "7"),
  #  !(Env == "MCG23" & Timepoint == "1"))
data$GID<- as.factor(data$GID)
colnames(data_sp)
```



## **3. Fit simple BLUE model for spec**

```{r}
envs    <-  unique(data_sp$Env)
traits <- colnames(data_sp)[c(5,13)]
raw_dat <- data_sp %>% left_join(data_ag[,c(1,5,6)], by= "plot_id")
raw_dat<- raw_dat[,c(2,3, 4, 5, 13, 15, 16)]
## -------------------------------------------------------------
##  containers
## -------------------------------------------------------------
fit_list  <- list()   
pred_list <- list()   
data$GID<- as.character(data$GID)
timepoints <- unique(raw_dat$Timepoint)
## -----------------------------------------------------------------
##  loops
## -----------------------------------------------------------------
for (tr in traits) {

  fits_vi <- list()

  for (e in envs) {
    
    dat_sub <- raw_dat %>%
      filter(Env == e, !is.na(.data[[tr]])) %>%
      mutate(
        GID = factor(GID),
        Row = factor(Row),
        Col = factor(Column)
      )
    
    timepoints <- unique(dat_sub$Timepoint)

    for (tp in timepoints) {
      dat_sub1 <- dat_sub %>% filter(Timepoint == tp) %>% na.omit()

      ## model -----------------------------------------------------------------
      mod <- tryCatch(
        asreml(
          fixed    = reformulate("GID", response = tr),
          random   = ~ar1(Row):ar1(Column),
          residual = ~ units,
          data     = dat_sub1
        ),
        error = function(err) {
          message("asreml failed for ", tr, " | ", e, " | tp ", tp,
                  " : ", err$message)
          NULL
        }
      )

      ## store fit -------------------------------------------------------------
      fits_vi[[paste(e, tp, sep = "_")]] <- mod

      ## predictions -----------------------------------------------------------
      if (!is.null(mod)) {
        pred_result <- tryCatch({
          pred <- predict(mod, classify = "GID")$pvals %>%
            as_tibble() %>%
            transmute(
              VI        = tr,
              Env       = e,
              Timepoint = tp,
              GID,
              Predicted = predicted.value,
              SE        = std.error
            )

          pred_list[[length(pred_list) + 1]] <- pred
        }, error = function(err) {
          message("Prediction failed for ", tr, " | ", e, " | tp ", tp,
                  " : ", err$message)
          NULL
        })
      }
    } # time‑point loop
  }   # environment loop

  fit_list[[tr]] <- fits_vi
}     # trait loop

## -----------------------------------------------------------------
##  tidy table of all predictions
## -----------------------------------------------------------------
preds <- bind_rows(pred_list)
preds_m<- preds
table(preds$Env, preds$VI)
##  optional saves
save(fit_list, file = "spec_BLUE_spatial_fits.RData")
save(preds,  file = "spec_BLUE_spatial.RData")
```



## **4. Adding flight time data to the predicitons and filter for QC**

```{r}
load("C:/Users/Siim Sepp/NY_WMB/data/wmb_multi_alldata.Rdata")
colnames(multi)
# Create a mapping of timepoint to JD
timepoint_JD <- multi[,c(1,3, 10,31)] %>%
  distinct()  
colnames(timepoint_JD)[2] <- "Timepoint"
timepoint_JD
```



## **5. Fit simple BLUE model for ag**

```{r, eval=TRUE}
envs    <- unique(data_ag$Env)
traits <- colnames(data_ag)[c(7:9)]
raw_dat <- data <-  data_ag                                   

## -------------------------------------------------------------
fit_list  <- list()
pred_list <- list() 
data$GID<- as.character(data$GID)
## -----------------------------------------------------------------
for (tr in traits) {

  fits_vi <- list()

  for (e in envs) {
    
          dat_sub <- raw_dat %>%
        filter(Env == e,
               !is.na(.data[[tr]])) %>%
        mutate(
          GID = factor(GID),
          Row = factor(Row),
          Col = factor(Column)
        )

      ## model -----------------------------------------------------------------
      mod <- tryCatch(
        asreml(
          fixed    = reformulate("GID", response = tr),   #
          random   = ~Row + ~Column,
          residual = ~ units,              
          data     = dat_sub
        ),
        error = function(err) {
          message("asreml failed for ", tr, " | ", e,
                  " : ", err$message)
          NULL
        }
      )

      ## store fit -------------------------------------------------------------
      fits_vi[[paste(e,  sep = "_")]] <- mod

      ## predictions -----------------------------------------------------------
      if (!is.null(mod)) {
        pred <- predict(mod,
                        classify = "GID")$pvals %>%
          as_tibble() %>%
          transmute(
            VI          = tr,
            Env = e,
            GID,
            Predicted   = predicted.value,
            SE          = std.error
          )

        pred_list[[length(pred_list) + 1]] <- pred
      }
  }   # environment loop

  fit_list[[tr]] <- fits_vi
}     # trait loop

## -----------------------------------------------------------------
##  tidy table of all predictions
## -----------------------------------------------------------------
preds <- bind_rows(pred_list)
table(preds$VI, preds$Env)
##  optional saves
save(fit_list, file = "data/ag_BLUE_spatial_fits.RData")
save(preds,  file = "data/ag_BLUE_spatial.RData")
```



## **6. Fit simple BLUP model for MQ**

```{r}
envs    <- c("MCG23","HELF24")#, "MCG25")
traits <- colnames(data_mq)[4:15]

#data_mq<- left_join(data_mq, data_ag[,c(2,5,6)], by= "plot_id")
raw_dat <- data <-  data_mq %>% na.omit() %>% droplevels()
str(raw_dat)
## -------------------------------------------------------------
fit_list  <- list()   
pred_list <- list()
data$GID<- as.character(data$GID)

## -----------------------------------------------------------------

for (tr in traits) {

  fits_vi <- list()

  for (e in envs) {
    
          dat_sub <- raw_dat %>%
        filter(Env == e) %>%
        mutate(
          GID = factor(GID),
          Row = factor(Row),
          Col = factor(Column)
        )

      ## model -----------------------------------------------------------------
      mod <- tryCatch(
        asreml(
          fixed    = reformulate("1", response = tr),   #  ~ 1
          random   = ~GID + ~ar1(Row):ar1(Column),                             # genotype effect
          residual = ~ units,               # spatial correction
          data     = dat_sub
        ),
        error = function(err) {
          message("asreml failed for ", tr, " | ", e,
                  " : ", err$message)
          NULL
        }
      )

      ## store fit -------------------------------------------------------------
      fits_vi[[paste(e,  sep = "_")]] <- mod

      ## predictions -----------------------------------------------------------
      if (!is.null(mod)) {
  pred <- tryCatch({
    predict(mod, classify = "GID")$pvals %>%
      as_tibble() %>%
      transmute(
        VI        = tr,
        Env       = e,
        GID,
        Predicted = predicted.value,
        SE        = std.error
      )
  },
  error = function(err) {
    message("asreml predict failed for ", tr, " | ", e,
            " : ", err$message)
    NULL
  })

  if (!is.null(pred)) {
    pred_list[[length(pred_list) + 1]] <- pred
  }
}
  }   # environment loop

  fit_list[[tr]] <- fits_vi
}     # trait loop

## -----------------------------------------------------------------
##  tidy table of all predictions
## -----------------------------------------------------------------
preds <- bind_rows(pred_list)
table(preds$Env)
##  optional saves
save(fit_list, file = "data/mq_BLUP_spatial_fits.RData")
save(preds,  file = "data/mq_BLUP_spatial.RData")
```



## **7. Create a VI matrix of VI BLUPS. **

```{r}
load("spec_BLUE_spatial.RData")

VImats <- list()
multi<- preds
VIs <- unique(multi$VI)
# Prepare an empty list to hold pivoted data per environment
VImats <- list()

str(multi)


# Step 1: Make the dataframe wider so each VI is a column
multi_wide <- multi[,1:5] %>%
  pivot_wider(
    names_from = VI,
    values_from = Predicted
  )

VImats <- list()

for (e in unique(multi_wide$Env)) {
  VImats[[e]] <- list()
  for (i in VIs) {
    subdata <- multi_wide[multi_wide$Env == e, c("Timepoint", i, "GID")]
    wide_df <- tidyr::pivot_wider(
      data        = subdata,
      names_from  = Timepoint,
      values_from = i
    )
    VImats[[e]][[i]] <- wide_df
  }
}

wide_list <- multi_wide %>%
  group_split(Env) %>%
  set_names(unique(multi_wide$Env)) %>%
  map(~ pivot_wider(.x,
                    id_cols = GID,
                    names_from = Timepoint,
                    values_from = c(NDVI, PSRI),
                    names_glue = "{.value}_TP{Timepoint}"
  ))


VImats$MCG25[["combined"]]<- wide_list$MCG25
VImats$HELF24[["combined"]]<- wide_list$HELF24
VImats$MCG23[["combined"]]<- wide_list$MCG23
VImats$SNY22[["combined"]]<- wide_list$SNY22
VImats$KET21[["combined"]]<- wide_list$KET21
save(VImats, file= "data/VImatBLUE.Rdata")
```

