---
title: "4. Exploratory Phenotype analysis"
author: "sss322"
date: "2025-04-09"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

## **Import packages and data**
```{r, message=FALSE}
pkgs <- c("dplyr","tidyr","readr","readxl","lubridate","stringr", "ggplot2", "forcats", "ASRtriala", "openxlsx")
suppressPackageStartupMessages(invisible(lapply(pkgs, require, character.only = TRUE)))

base_path <- if (.Platform$OS.type == "windows") {
  "C:/Users/Siim Sepp/NY_WMB"
} else {
  "~/Documents/GitHub/NY_WMB"
}

load(file.path(base_path, "data/data_ag.Rdata"))
load(file.path(base_path, "data/data_mq.Rdata"))
load(file.path(base_path, "data/data_sp.Rdata"))
Sys.setenv(LANGUAGE = "en")
```




# **1. Calc HD, GDD and sowing date**

```{r}
#long VI table if you need downstream
traits <- names(data_sp)[5:13]
data_sp_long <- tidyr::pivot_longer(data_sp, all_of(traits), names_to="trait", values_to="value")

# sowing dates → map to Env and Year (Oct sowing → next spring season)
sow_map <- tibble::tibble(
  Environment = c("KET21","SNY22","MCG23","HELF24","MCG25"),
  sd          = as.Date(c("2020-10-08","2021-10-11","2022-10-11","2023-10-03","2024-09-27"))
) %>%
  mutate(Year = year(sd) + 1)

# flights (ensure Date parsed, strip trailing " UTC" if present)
flight_d <- readxl::read_excel("data/Flights/flightdate.xlsx") %>%
  mutate(Date = as.Date(str_remove(as.character(Date), " UTC$")))

# attach sowing dates and compute days-from-sowing (JDD) and from Jan 1 (JDD1)
JD_map <- flight_d %>%
  left_join(sow_map, by = "Year") %>%
  mutate(
    JDD  = as.integer(Date - sd),
    JDD1 = as.integer(Date - as.Date(paste0(Year, "-01-01")))
  )

# weather → daily mean temp and cumulative GDD per year
base_temp <- 40
weather <- readr::read_csv("data/weather_dataq_ithaca.csv", show_col_types = FALSE) %>%
  mutate(
    DATE    = as.Date(DATE),
    year    = year(DATE),
    t_avg   = (TMAX + TMIN) / 2,
    gdd_day = pmax(t_avg - base_temp, 0)
  ) %>%
  arrange(year, DATE) %>%
  group_by(year) %>%
  mutate(GDD_cum = cumsum(gdd_day)) %>%
  ungroup()

# GDD on flight dates: join cumulative GDD by exact date & year
JD_map <- JD_map %>%
  left_join(weather %>% transmute(Year = year, Date = DATE, GDD = GDD_cum),
            by = c("Year","Date"))

# average HD_JD per year (use Env→Year map so it matches sow_map)
year_map <- sow_map %>% dplyr::select(Env = Environment, Year)
data_ag  <- data_ag %>% mutate(Env = as.character(Env)) %>% left_join(year_map, by = "Env")

avg_HD_by_year <- data_ag %>%
  group_by(Year) %>%
  summarise(mean_HD_JD = mean(HD_JD, na.rm = TRUE), .groups = "drop") %>%
  mutate(JD_date = as.Date(paste0(Year, "-01-01")) + mean_HD_JD)

# GDD at the mean heading date (vectorized join to weather)
avg_HD_by_year <- avg_HD_by_year %>%
  left_join(weather %>% transmute(Year = year, JD_date = DATE, GDD_JD = GDD_cum),
            by = c("Year","JD_date"))

# merge heading-date info to flight map and compute distance from HD
JD_map <- JD_map %>%
  left_join(avg_HD_by_year, by = "Year") %>%
  mutate(d_f_HD = as.integer(Date - JD_date))

save(JD_map, file= "data/JDmap.Rdata")
```

```{r}
weather <- weather %>% #filter(!year == "2020") 
  mutate( year = year(DATE),doy = yday(DATE), month = month(DATE) ) 

weather_filtered <- weather %>% filter(month(DATE) <= 6) %>% filter(month(DATE) > 3) %>% 
  mutate( year = year(DATE), month_day = as.Date(format(DATE, "2000-%m-%d")))


JD_map_filtered <- JD_map %>% filter(month(Date) <= 6) %>% filter(month(Date) > 3) %>% 
  mutate( year = year(Date), month_day = as.Date(format(Date, "2000-%m-%d")), 
          JD_date_norm = as.Date(format(JD_date, "2000-%m-%d") ))
```


```{r}
# GDD and flight timepoint GDD. 

JD_map_filtered[,1:2]

ggplot(weather_filtered, aes(x = month_day, y = GDD_cum)) +
  geom_line(color = "darkgreen", size= 1) +
  geom_vline(data = JD_map_filtered, aes(xintercept = month_day), 
             color = "red", linetype = "dashed",size= 1) +
   geom_vline(data = JD_map_filtered, aes(xintercept = JD_date_norm), 
             color = "black",size= 1) +
  facet_grid(rows= vars(year))+
  scale_x_date(
    date_labels = "%b%d", date_breaks = "1 week",
    limits = as.Date(c("2000-04-15", "2000-06-30"))
  ) +
  labs(
   # title = "Cumulative GDD by Year with Flight Dates",
    x = "Date",
    y = "Cumulative GDD"
  ) +
  #theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 24),
    strip.text = element_text(size = 14),
    legend.text = element_text(size = 14),
    axis.text = element_text(size=12), 
    axis.title = element_text(size=14))
```




# **2. Plotting the distribution of traits**


### *Plotting the agronomic and the MQ traits*


```{r}
colnames(data_ag)
colnames(data_ag)[9]<- "WS"
colnames(data_mq)
dag<- data_ag[,c(1:3,7,8,9)]

dmq<- data_mq[,c(2,4:15)]
damq<- left_join(dag, dmq, by= "plot_id")

traits<- colnames(damq)[c(-1,-2, -3)]
damq$Alpha.amylase.20.DU. <- as.double(damq$Alpha.amylase.20.DU.) 
damql<- damq  %>%
  pivot_longer(
    cols = all_of(traits),
    names_to = "trait",
    values_to = "value"
  )

env_order<- c( "KET21", "SNY22" ,"MCG23","HELF24" , "MCG25" )
trait_map <- c(
  "KernelWeight.mg." = "KW",
  "on6.64...."       = "on6/64", 
  "MaltExtract..."   = "ME",
  "BarleyProtein..." = "BP",
  "WortProtein..."   = "WP",
  "S.T"              = "S.T", 
  "DP..ASBC."       = "DP", 
  "Alpha.amylase.20.DU." = "AA", 
  "Beta.glucan.ppm." = "BG",
  "HD_JD" = "HD", 
  "yield_kgha" = "GY", 
  "FAN.ppm."         = "FAN", 
  "AdjunctQualityScore" = "Adj.Sc",
  "All_MaltQualityScore" = "All.Sc")



pal <- c(
  KET21  = "cornsilk3",
  SNY22  = "palegreen2",
  MCG23  = "wheat2",
  HELF24 = "lightblue3", 
  MCG25 = "yellow3"
)

trait_order<- c( "GY" , "HD", "WS",  "KW", "on6/64" , "FAN" , "DP", "AA", "S.T" , "BG" , "ME",  "BP", "WP", "Adj.Sc", "All.Sc" )

colnames(damql)[c(3,4)] <- c("Environment", "Trait")
damql <- damql %>%
  mutate(Trait = recode(Trait, !!!trait_map))

damql <- damql %>%  mutate( Environment = trimws(as.character(Environment)),
    Environment = fct_relevel(Environment, env_order)  
  ) %>% mutate( Trait= trimws(as.character(Trait)),
    Trait = fct_relevel(Trait, trait_order)  
  )


p <- ggplot(damql, aes(x = value,  fill = Environment)) +
    geom_density(alpha= 0.5,adjust = 1.5) +
  scale_fill_manual(values= pal, drop= FALSE)+
    facet_wrap(~Trait,  scales = "free", ncol=5, nrow=3) +
    theme(legend.position = "bottom",
    strip.text = element_text(size = 12, face = "bold"))+
  ylab("Density") +
  xlab("")

p


p <- ggplot(damql, aes(y = value,  fill = Environment)) +
    geom_boxplot(alpha= 0.5) +
  scale_fill_manual(values= pal, drop= FALSE)+
    facet_wrap(~Trait,  scales = "free", ncol=5, nrow=3) +
    theme(
    strip.text = element_text(size = 12, face = "bold"))+
  ylab("Density") +
  xlab("")

p

summary(damq$HD_JD)

```


### *Plotting the spectral traits*
```{r}

colnames(JD_map)[6]<- "Env"
colnames(JD_map)[c(2,3, 5,6,10)]
###########################################################################################
data_sp_long <- left_join(data_sp_long, JD_map[c(2,3, 5,6,10)], by = c("Env" ,"Timepoint" ))
###########################################################################################
table(data_sp_long$Env, data_sp_long$Timepoint, data_sp_long$trait)

data_sub <- data_sp_long %>% filter(DAS >150) %>% filter(!trait == "GDVI") %>% filter(trait %in% c("NDVI", "PSRI"))

data_sub1 <- data_sub %>%  mutate( Environment = trimws(as.character(Env)),
    Environment = fct_relevel(Environment, env_order)  
  ) %>% mutate( Trait= trimws(as.character(trait)),
    Trait = fct_relevel(Trait, trait_order)  
  )


#Smooth plot GDD VI
p <- ggplot(data_sub1, aes(x = GDD, y = value, color = Environment)) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 4), se = FALSE, linewidth = 2, alpha = 0.5) +
  scale_color_manual(values = pal, drop = FALSE)+
  facet_wrap(~trait,  scales = "free_y") +
  labs(
    #title = paste("Smooth VI-s over locations"),
    x = "GDD",                         
    y = "Value",
    fill = "Trait")+
    #color = "Trait") +
  theme(
    strip.text = element_text(size = 16, face = "bold"),
    plot.title = element_text(hjust = 0.5, size = 22, face = "bold"),
    axis.text = element_text(size = 16),
    axis.title = element_text(size = 16),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 16), 
    legend.position = "bottom"
  )

print(p)

#Smooth plot date VI
data_sub2 <- data_sub1 %>%
  dplyr::mutate(Date = as.Date(sprintf("2000-%02d-%02d",
                                       lubridate::month(Date),
                                       lubridate::day(Date))))

p <- ggplot(data_sub2, aes(x = Date, y = value, color = Environment)) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 4), se = FALSE, linewidth = 2, alpha = 0.5) +
  scale_color_manual(values = pal, drop = FALSE)+
  facet_wrap(~trait,  scales = "free_y") +
  labs(
    #title = paste("Smooth VI-s over locations"),
    x = "Date",                         
    y = "Value",
    fill = "Trait")+
  theme(
    strip.text = element_text(size = 16, face = "bold"),
    plot.title = element_text(hjust = 0.5, size = 22, face = "bold"),
    axis.text = element_text(size = 16),
    axis.title = element_text(size = 16),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 16), 
    legend.position = "bottom"
  )
Sys.setenv(LANGUAGE = "en")
p
```




### *Plotting the histograms PC and FPC scores*
```{r}
load("data/F_PCA_scoresG.Rdata")
F_PC_scores<- FPCA_scores_df
colnames(F_PC_scores)

# Assuming your dataframe is called df_pcs
F_PC_long <- F_PC_scores %>%
  pivot_longer(
    cols = contains("PC"),
    names_to = "Component",
    values_to = "Loadings")

colnames(F_PC_long)
colnames(data_ag)
df_ag <- left_join(F_PC_long, data_ag[,-3], by= c("plot_id"))
df_mq <- left_join(F_PC_long, data_mq[,-3], by= c("plot_id"))
```


```{r}
df_ag<- df_ag %>% arrange(Env)

df_ag1 <- df_ag %>% filter(Component %in% c( "FPC1"  ,"FPC2", "FPC3")) %>% mutate( Env = fct_relevel(Env, env_order))

p <- ggplot(df_ag1, aes(x = Loadings, fill = Env)) +
    geom_density(alpha = 0.3, adjust = 1.5) +
    facet_wrap(~Component, scales= "free") +
  scale_fill_manual(values = pal, drop = FALSE) +

    #theme_minimal() +
  theme(
    strip.text = element_text(size = 16, face = "bold"),
    plot.title = element_text(hjust = 0.5, size = 22, face = "bold"),
     axis.text = element_text(size = 16),     
    axis.title = element_text(size = 16),
    legend.title = element_text(size = 16), 
    legend.text = element_text(size = 16), 
    legend.position = "bottom")+
  ggtitle("") +
  ylab("Density") +
  xlab("")

print(p)
```




# **3. Calculating Summary Statistics, CV and overlap**
```{r}
#yield
data_ag$GID<- as.factor(data_ag$GID)
sumst_ag <- audit.met(data_ag, gen= "GID",resp= "yield_kgha", trial= "Env")
print(sumst_ag)

#Malting quality
traits<- colnames(data_mq)[4:15]
data_mq$GID<- as.factor(data_mq$GID)
data_mq$Alpha.amylase.20.DU <- as.numeric(data_mq$Alpha.amylase.20.DU.)
str(data_mq)
MQstats<- list()
for (tr in traits){
  MQstats[tr]<- audit.met(data_mq, gen= "GID",resp= tr, trial= "Env")
}


#Spectral data
stats_sp<- data.frame()

traits<- colnames(data_sp)[5:13]
spec_stats <- list()

data_sp$Env<- as.factor(data_sp$Env)

for (tr in traits) {
  for (e in unique(data_sp$Env)) {
    

    data_sub <- data_sp %>% filter(Env == e)
    data_sub$GID<- as.factor(data_sub$GID)
    data_sub$Timepoint <- as.factor(data_sub$Timepoint)
    

    result <- audit.met(data_sub, gen = "GID", resp = tr, trial = "Timepoint")
    result$met.stats$trait<- tr
    result$met.stats$Env<- e
    stats_sp<- rbind(stats_sp, result$met.stats)
  }
}
#str(stats_sp)

stats_sp <- stats_sp %>% filter(!Env == "MCG23" & !Timepoint == "2" & !trait == "PSRI")
boxplot(stats_sp$CVp~stats_sp$Timepoint)

stats_sp$trait <- as.factor(stats_sp$trait)


p <- ggplot(stats_sp, aes(y = CVp, group = Timepoint)) +
      geom_boxplot(alpha = 0.2) +
      facet_wrap(~ Env) +
      labs(
        title = paste("Timepoint:", "| VI:", ~Env) ) +
      #theme_minimal() +
      theme(
        plot.title = element_text(hjust = 0.5, size = 14),
        strip.text = element_text(size = 12)
      )
    
print(p)
p <- ggplot(stats_sp, aes(y = CVp, group = Timepoint)) +
      geom_boxplot(alpha = 0.2) +
      facet_wrap(~ trait) +
      labs(
        #title = paste("Timepoint:", tp, "| VI:", vi, e),
       # x = ~timepoint,
        #y = vi
      ) +
      #theme_minimal() +
      theme(
        plot.title = element_text(hjust = 0.5, size = 14),
        strip.text = element_text(size = 12)
      )
    
print(p)
```




# **4. Correlation analysis**


### *Printing the plots per PC:Agronomic combination, all  within Envs*

```{r, message=FALSE, warning=FALSE}
######################################################################################################
#ag traits
######################################################################################################
data<- df_ag
traits<- unique(data$Component)
ag_traits<- colnames(data)[10:11]

df_long <- data %>%
  pivot_longer(
    cols = ag_traits,
    names_to = "Trait",
    values_to = "Value")

corr_summary <- df_long %>%
        group_by(Env, Component, VI, Trait) %>%
        summarise(
          correlation = cor(Loadings, Value, use = "complete.obs"),
          .groups = "drop")
cs_ag<- corr_summary

# scatterplot for PC loadaing : value. wrap over componenets
for (t in traits) {
    df_trait_plot <- df_long %>% filter(Component == t) %>% na.omit()
    
    p <- ggplot(df_trait_plot, aes(x = Value, y = Loadings, color= Env, fill= Env)) +
      geom_point(alpha = 0.12) +
      geom_smooth(method = "lm", se = FALSE, color = "blue", linewidth = 0.8) +
      facet_wrap(~Trait, scales = "free_x") +
      labs(
        title = paste("Trait:", t),
        x = paste(~Trait, "Value"),
        y = "Loading"
      ) +
      #theme_minimal() +
      theme(
        plot.title = element_text(hjust = 0.5, size = 14),
        strip.text = element_text(size = 10)
      ) +
      ggpubr::stat_cor(method = "pearson", label.x.npc = "left", label.y.npc = "top")
    
    print(p)
}

#############################################################################################################
#Malting quality traits
#############################################################################################################
data<- df_mq# %>% filter(!WortColor == "n.d.") %>% na.omit()

traits<- unique(data$Component)
mq_traits<- colnames(data)[7:18]
# Correlation results storage
str(data)
df_long <- data[,c(1,2,3,4,5,7:18)] %>%
  pivot_longer(
    cols = mq_traits,
    names_to = "Trait",
    values_to = "Value")


corr_summary <- df_long %>%
  group_by(Env, Component, VI, Trait) %>%
  summarise(
    n_pair = sum(complete.cases(Loadings, Value)),
    correlation = if (n_pair >= 2)
      suppressWarnings(cor(Loadings, Value, use = "complete.obs"))
    else NA_real_,
    .groups = "drop"
  ) %>%
  filter(n_pair >= 2)


cs_mq<- corr_summary
# scatterplot for PC loadaing : value. wrap over componenets
for (t in traits) {
    df_trait_plot <- df_long %>% filter(Component == t) %>% na.omit()
    
    p <- ggplot(df_trait_plot, aes(x = Value, y = Loadings, color= Env, fill= Env)) +
      geom_point(alpha = 0.12) +
      geom_smooth(method = "lm", se = FALSE, color = "blue", linewidth = 0.8) +
      facet_wrap(~Trait, scales = "free_x") +
      labs(
        title = paste("Trait:", t),
        x = paste(~Trait, "Value"),
        y = "Loading"
      ) +
      theme_minimal() +
      theme(
        plot.title = element_text(hjust = 0.5, size = 14),
        strip.text = element_text(size = 10)
      ) +
      ggpubr::stat_cor(method = "pearson", label.x.npc = "left", label.y.npc = "top")
    
    print(p)
}

# scatterplot for PC loadaing : value. wrap over componenets and VIs
for(i in unique(df_long$VI)){
for (t in traits) {
    df_trait_plot <- df_long %>% filter(Component == t) %>% filter(VI == i) %>% na.omit()
    
    p <- ggplot(df_trait_plot, aes(x = Value, y = Loadings, color= Env, fill= Env)) +
      geom_point(alpha = 0.2) +
      geom_smooth(method = "lm", se = FALSE, color = "blue", linewidth = 0.8) +
      facet_wrap(~Trait, scales = "free_x") +
      labs(
        title = paste("Trait:", t, i),
        x = paste(~Trait, "Value"),
        y = "Loading"
      ) +
      theme_minimal() +
      theme(
        plot.title = element_text(hjust = 0.5, size = 14),
        strip.text = element_text(size = 10)
      ) +
      ggpubr::stat_cor(method = "pearson", label.x.npc = "left", label.y.npc = "top")
    
    print(p)
}}
table(data_mq$Env, data_mq$Row)
```




### *Heatmap for PC loading and trait*
```{r}
#Agronomic traits and loading
data$Env<- as.character(data$Env)
ENVs <- unique(data$Env)

heatmap_plot <- ggplot(cs_ag, aes(x = factor(Component), y = Trait, fill = correlation)) +
      geom_tile(color = "white") +
  facet_wrap(~Env)+
        scale_fill_gradient2(low = "blue", high = "red", midpoint = 0, limits = c(-1, 1)) +
        labs(
        title = paste("Correlation Heatmap"),
        x = "Timepoint",
        y = "Trait",
        fill = "r"
      ) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 14),
        axis.text = element_text(size = 10)
      )
print(heatmap_plot)
    
#################################################################################
#Malting quality traits and loading
heatmap_plot <- ggplot(cs_mq, aes(x = factor(Component), y = Trait, fill = correlation)) +
      geom_tile(color = "white") +
  facet_wrap(~Env)+
        scale_fill_gradient2(low = "blue", high = "red", midpoint = 0, limits = c(-1, 1)) +
        labs(
        title = paste("Correlation Heatmap"),
        x = "Timepoint",
        y = "Trait",
        fill = "r"
      ) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 14),
        axis.text = element_text(size = 10)
      )
print(heatmap_plot)

#################################################################################
heatmap_plot <- ggplot(cs_mq, aes(x = factor(Env), y = Trait, fill = correlation)) +
      geom_tile(color = "white") +
  facet_wrap(~Component)+
        scale_fill_gradient2(low = "blue", high = "red", midpoint = 0, limits = c(-1, 1)) +
        labs(
        title = paste("Correlation Heatmap"),
        x = "Timepoint",
        y = "Trait",
        fill = "r"
      ) +
      #theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 14),
        axis.text = element_text(size = 10)
      )
print(heatmap_plot)
```



```{r}
#differnt plots. 
library(broom)
######################################################################################
data<- df_ag
traits<- unique(data$Component)
ag_traits<- colnames(data)[10:11]

df_long <- data %>%
  pivot_longer(
    cols = ag_traits,
    names_to = "Trait",
    values_to = "Value")


# 1) Correlations (r) and p-values per Env × Trait × Component
corr_df <- df_long %>%
  group_by(Env, Trait, Component) %>%
  nest() %>%
  mutate(
    test = purrr::map(data, ~{
      dd <- .x %>% drop_na(Value, Loadings)
      if (nrow(dd) < 3) return(tibble(estimate = NA_real_, p.value = NA_real_, n = nrow(dd)))
      ct <- cor.test(dd$Value, dd$Loadings, method = "pearson")
      tibble(estimate = unname(ct$estimate), p.value = ct$p.value, n = nrow(dd))
    })
  ) %>%
  #select(-data) %>%
  unnest(test) %>%
  mutate(
    r = estimate,
    p_adj = p.adjust(p.value, method = "BH"),
    sig = p_adj < 0.05
  )

# 2) Heatmap (rows = Trait, cols = Component), faceted by Env
ggplot(corr_df, aes(x = Component, y = Trait, fill = r, alpha = sig)) +
  geom_tile(color = "grey90") +
  # Optional: label r on tiles
  geom_text(aes(label = ifelse(is.finite(r), sprintf("%.2f", r), "")), size = 6) +
  scale_fill_gradient2(limits = c(-1, 1), midpoint = 0,
                       low = "#2c7bb6", mid = "white", high = "#d7191c",
                       name = "Pearson r") +
  scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.35), guide = "none") +
  facet_wrap(~ Env, scales = "free_x") +
  labs(x = "Predictor (Component)", y = "Trait")+
       #title = "Correlation of Trait Values vs. PCA Loadings by Environment",
       #subtitle = "Tiles shaded by Pearson r; transparency indicates FDR significance (BH)") +
  theme_minimal(base_size = 16) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 16)
  )

######################################################################################
data<- df_mq# %>% filter(!WortColor == "n.d.") %>% na.omit()

traits<- unique(data$Component)
mq_traits<- colnames(data)[7:18]
# Correlation results storage
str(data)
df_long <- data[,c(1,2,4,5,7:18)] %>%
  pivot_longer(
    cols = mq_traits,
    names_to = "Trait",
    values_to = "Value")



# 1) Correlations (r) and p-values per Env × Trait × Component
corr_df <- df_long %>%
  group_by(Trait, Component) %>%
  nest() %>%
  mutate(
    test = purrr::map(data, ~{
      dd <- .x %>% drop_na(Value, Loadings)
      if (nrow(dd) < 3) return(tibble(estimate = NA_real_, p.value = NA_real_, n = nrow(dd)))
      ct <- cor.test(dd$Value, dd$Loadings, method = "pearson")
      tibble(estimate = unname(ct$estimate), p.value = ct$p.value, n = nrow(dd))
    })
  ) %>%
  #select(-data) %>%
  unnest(test) %>%
  mutate(
    r = estimate,
    p_adj = p.adjust(p.value, method = "BH"),
    sig = p_adj < 0.05
  )
mqe<- c("MCG23", "HELF24", "MCG25")
#corr_df<- corr_df %>% filter(Env %in% mqe)
# 2) Heatmap (rows = Trait, cols = Component), faceted by Env
ggplot(corr_df, aes(x = Component, y = Trait, fill = r, alpha = sig)) +
  geom_tile(color = "grey90") +
  # Optional: label r on tiles
  geom_text(aes(label = ifelse(is.finite(r), sprintf("%.2f", r), "")), size = 6) +
  scale_fill_gradient2(limits = c(-1, 1), midpoint = 0,
                       low = "#2c7bb6", mid = "white", high = "#d7191c",
                       name = "Pearson r") +
  scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.35), guide = "none") +
  #facet_wrap(~ Env, scales = "free_x") +
  labs(x = "Predictor (Component)", y = "Trait")+
       #title = "Correlation of Trait Values vs. PCA Loadings by Environment",
       #subtitle = "Tiles shaded by Pearson r; transparency indicates FDR significance (BH)") +
  theme_minimal(base_size = 16) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10)
  )
```



```{r}

load("data/ag_BLUE_spatial.RData")
ag1<- preds
load("data/mq_BLUP_spatial.RData")
mq1<- preds
load("data/spec_BLUE_spatial.RData")
sp1<- preds

mfp<- left_join(sp1, ag1, by= c("Env", "GID", "VI"))
#Agronomic traits and loading
data$Env<- as.character(data$Env)
ENVs <- unique(data$Env)

heatmap_plot <- ggplot(cs_ag, aes(x = factor(Component), y = Trait, fill = correlation)) +
      geom_tile(color = "white") +
  facet_wrap(~Env)+
        scale_fill_gradient2(low = "blue", high = "red", midpoint = 0, limits = c(-1, 1)) +
        labs(
        title = paste("Correlation Heatmap"),
        x = "Timepoint",
        y = "Trait",
        fill = "r"
      ) +
     # theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 14),
        axis.text = element_text(size = 10)
      )
print(heatmap_plot)
```




# **5. PCA biplot — agronomic and malting quality traits**

```{r pca_biplot_ag_mq, fig.width = 10, fig.height = 8, warning = FALSE, message = FALSE}
# ---- Prepare trait matrix ----
# damq joins agronomic (GY, HD, WS) and malt quality traits.
# Restrict to complete cases so every trait contributes equally to the PCA.
trait_cols <- colnames(damq)[-(1:3)]          # all trait columns (excl. plot_id, Env, GID)
pca_mat    <- damq[, c("plot_id", "Env", trait_cols)]
pca_mat$Alpha.amylase.20.DU. <- as.double(pca_mat$Alpha.amylase.20.DU.)
pca_mat    <- pca_mat[complete.cases(pca_mat[, trait_cols]), ]

# ---- Run PCA (centered and scaled) ----
pca_res <- prcomp(pca_mat[, trait_cols], center = TRUE, scale. = TRUE)
pve     <- round(summary(pca_res)$importance[2, ] * 100, 1)   # % variance per PC

# ---- Scores (one point per observation) ----
scores     <- as.data.frame(pca_res$x[, 1:2])
scores$Env <- factor(pca_mat$Env, levels = env_order)

# ---- Loadings (one arrow per trait) ----
loadings           <- as.data.frame(pca_res$rotation[, 1:2])
loadings$trait_raw <- rownames(loadings)

# Map to short names; fall back to the raw column name for any unmapped trait
trait_labels   <- c(trait_map, "WS" = "WS")
mapped_labels  <- trait_labels[loadings$trait_raw]
loadings$label <- ifelse(is.na(mapped_labels), loadings$trait_raw, mapped_labels)

# Scale arrows so they span ~70 % of the score cloud radius
scale_fac     <- max(abs(range(scores[, 1:2]))) /
                 max(sqrt(loadings$PC1^2 + loadings$PC2^2)) * 0.70
loadings$xend <- loadings$PC1 * scale_fac
loadings$yend <- loadings$PC2 * scale_fac

# ---- Biplot ----
p_pca <- ggplot() +
  # reference axes
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey65", linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey65", linewidth = 0.4) +
  # observation scores coloured by environment
  geom_point(data  = scores,
             aes(x = PC1, y = PC2, colour = Env),
             alpha = 5.45, size = 4.8) +
  # trait loading arrows
  geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = xend, yend = yend),
               arrow     = arrow(length = unit(0.22, "cm"), type = "closed"),
               colour    = "black", linewidth = 0.75, linejoin = "mitre") +
  # trait labels nudged 13 % beyond the arrowhead
  geom_text(data = loadings,
            aes(x = xend * 1.13, y = yend * 1.13, label = label),
            size     = 3.6,
            fontface = "bold",
            colour   = "black") +
  scale_colour_manual(values = pal, drop = FALSE) +
  labs(
    x       = paste0("PC1  (", pve[1], "% variance explained)"),
    y       = paste0("PC2  (", pve[2], "% variance explained)"),
    colour  = "Environment",
    caption = paste0("n = ", nrow(pca_mat),
                     " observations with complete agronomic + malting quality data.")
  ) +
  theme(
    axis.title       = element_text(size = 13, face = "bold"),
    axis.text        = element_text(size = 11),
    legend.title     = element_text(size = 12, face = "bold"),
    legend.text      = element_text(size = 11),
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    plot.caption     = element_text(size = 9, colour = "grey50")
  )

print(p_pca)
```


### *PCA biplot within each environment*

```{r pca_biplot_per_env, fig.width = 9, fig.height = 7, warning = FALSE, message = FALSE}
# pca_mat, trait_cols and trait_labels are carried over from the chunk above.
# Run a separate PCA for each environment and plot its biplot.

for (e in env_order) {

  df_e <- pca_mat[pca_mat$Env == e, ]

  if (nrow(df_e) < 10) {
    message("Skipping ", e, ": fewer than 10 complete observations.")
    next
  }

  # ---- PCA within environment ----
  pca_e <- prcomp(df_e[, trait_cols], center = TRUE, scale. = TRUE)
  pve_e <- round(summary(pca_e)$importance[2, ] * 100, 1)

  # ---- Scores ----
  sc_e     <- as.data.frame(pca_e$x[, 1:2])

  # ---- Loadings ----
  ld_e           <- as.data.frame(pca_e$rotation[, 1:2])
  ld_e$trait_raw <- rownames(ld_e)
  mapped_e       <- trait_labels[ld_e$trait_raw]
  ld_e$label     <- ifelse(is.na(mapped_e), ld_e$trait_raw, mapped_e)

  # Scale arrows to ~70 % of score cloud radius
  sf_e      <- max(abs(range(sc_e[, 1:2]))) /
               max(sqrt(ld_e$PC1^2 + ld_e$PC2^2)) * 0.70
  ld_e$xend <- ld_e$PC1 * sf_e
  ld_e$yend <- ld_e$PC2 * sf_e

  # ---- Biplot ----
  p_e <- ggplot() +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey65", linewidth = 0.4) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey65", linewidth = 0.4) +
    geom_point(data = sc_e,
               aes(x = PC1, y = PC2),
               colour = pal[e], alpha = 0.55, size = 2) +
    geom_segment(data = ld_e,
                 aes(x = 0, y = 0, xend = xend, yend = yend),
                 arrow     = arrow(length = unit(0.22, "cm"), type = "closed"),
                 colour    = "black", linewidth = 0.75, linejoin = "mitre") +
    geom_text(data = ld_e,
              aes(x = xend * 1.13, y = yend * 1.13, label = label),
              size = 3.6, fontface = "bold", colour = "black") +
    labs(
      title   = e,
      x       = paste0("PC1  (", pve_e[1], "% variance explained)"),
      y       = paste0("PC2  (", pve_e[2], "% variance explained)"),
      caption = paste0("n = ", nrow(df_e), " observations with complete ag + MQ data.")
    ) +
    theme(
      plot.title       = element_text(size = 14, face = "bold", hjust = 0.5),
      axis.title       = element_text(size = 12, face = "bold"),
      axis.text        = element_text(size = 10),
      panel.grid.minor = element_blank(),
      plot.caption     = element_text(size = 9, colour = "grey50")
    )

  print(p_e)
}
```










































# **5. Printing the plots per VI:TP all MQ scatter plots** 

```{r, message=FALSE, warning=FALSE}
load("data/data_ag.Rdata")
load("data/data_mq.Rdata")
load("data/data_sp.Rdata")
data_ag<- data_ag[c(1:2,3, 7,8, 10)]
data_mq<- data_mq[c(2, 4:15)]
data_sp<- data_sp[c(1,4:13)]
data<- left_join(data_sp, data_mq, by = "plot_id")
data<- left_join(data, data_ag, by = "plot_id")


traits<- colnames(data_mq)[-1]
VIs<-  colnames(data_sp)[-1:-2]

df_trait_plot <- data %>% pivot_longer(
          cols = all_of(traits),
          names_to = "MQtrait",
          values_to = "MQtrait_value"
        )

df_trait_plot<- df_trait_plot%>% pivot_longer(
          cols = all_of(VIs),
          names_to = "VI",
          values_to = "VI_value"
        )

for (i in unique(df_trait_plot$MQtrait)) {
  for (vi in unique(df_trait_plot$VI)) {

df_trait_plot1 <- df_trait_plot %>% filter(MQtrait == i, VI == vi)

p <- ggplot(df_trait_plot1, aes(x = MQtrait_value, y = VI_value, color=Env, fill= Env)) +
      geom_point(alpha = 0.2) +
      geom_smooth(method = "lm", se = FALSE, color = "blue", linewidth = 0.8) +
      facet_wrap(~Timepoint, scales = "free_x") +
      labs(
        title = paste(i, vi),
        x = paste(i, "Value"),
        y = "Vegetative Index Value"
      ) +
      theme_minimal() +
      theme(
        plot.title = element_text(hjust = 0.5, size = 14),
        strip.text = element_text(size = 10)
      ) +
      ggpubr::stat_cor(method = "pearson", label.x.npc = "left", label.y.npc = "top")
    
print(p)

}}
```






#######################################################################################################3

### *plot traits against VIs at timepoints.*

```{r}
traits<- colnames(data)[c(-1:-11,-24, -25, -28)]
# Assuming your original data has one column per VI (wide format)
df_long <- data %>%
  pivot_longer(
    cols = all_of(traits),    
    names_to = "trait",
    values_to = "trait_value"
  )


df_long_vi <- df_long %>%
  pivot_longer(
    cols = all_of(VIs),         
    names_to = "VI",
    values_to = "VI_value"
  )

unique_traits <- unique(df_long_vi$trait)
unique_VIs <- unique(df_long_vi$VI)

ENVs<- unique(df_long_vi$Env)
ENVs<- ENVs[-3]
df_long_vi$trait<- as.factor(df_long_vi$trait)

#for (tr in unique_traits) {
for (e in ENVs) {
  for (vi in unique_VIs) {
  
  df_sub <- df_long_vi %>%
    filter(VI == vi, Env == e)
  
  p <- ggplot(df_sub, aes(x = VI_value, y = trait_value)) +
    
    # Add separate linear model line for each timepoint
    geom_point(alpha = 0.1, aes(group = Timepoint, color = Timepoint))+
    geom_smooth(
      aes(group = Timepoint, color = Timepoint, fill = Timepoint),
      method = "lm", se = FALSE, linewidth = 0.6
    ) +
    facet_wrap(~ trait, scales = "free_y") +
    
    labs(
      title = paste("MQ traits of timepoints of VI:", vi, "- across environment", e),
      x = paste(vi, "value"),
      y = vi,
      color = "Timepoint"
    ) +
    
    theme_minimal() +
    theme(
      strip.text = element_text(size = 6),
      axis.text = element_text(size = 6),
      plot.title = element_text(hjust = 0.5, size = 10),
      legend.position = "bottom"
    )
  
  print(p)
  }}
#}


```


### *Heatmap for VI and trait correlation*
```{r}
timepoints <- unique(data$Timepoint)
data$Env<- as.character(data$Env)
ENVs <- unique(data$Env)
table(data$Env, data$Timepoint)
# 1) make sure columns are numeric
data <- data %>%
  mutate(across(all_of(c(traits, VIs)), ~ suppressWarnings(as.numeric(.))))

# 2) safe correlation helper
safe_cor <- function(x, y, method = "pearson") {
  ok <- is.finite(x) & is.finite(y) & !is.na(x) & !is.na(y)
  if (sum(ok) < 2L) return(NA_real_)
  if (sd(x[ok]) == 0 || sd(y[ok]) == 0) return(NA_real_)
  suppressWarnings(cor(x[ok], y[ok], method = method))
}

# 3) compute correlations for ALL VIs at once, per Env×Timepoint×Trait
cor_results_df <- data %>%
  filter(Env %in% ENVs) %>%
  pivot_longer(all_of(traits), names_to = "Trait", values_to = "TraitValue") %>%
  group_by(Env, Timepoint, Trait) %>%
  summarise(
    across(all_of(VIs), ~ safe_cor(TraitValue, .x), .names = "cor_{.col}"),
    .groups = "drop"
  ) %>%
  pivot_longer(starts_with("cor_"), names_to = "VI", values_to = "correlation") %>%
  mutate(VI = sub("^cor_", "", VI))


table(cor_results_df$Env)
ENVs<- unique(cor_results_df$Env)  

cor_results_df1<- cor_results_df %>% filter(Env == ENVs[1])


heatmap_plot <- ggplot(cor_results_df1, aes(x = factor(Timepoint), y = Trait, fill = correlation)) +
      geom_tile(color = "white") +
  facet_wrap(~VI)+
        scale_fill_gradient2(low = "blue", high = "red", midpoint = 0, limits = c(-1, 1)) +
        labs(
        title = paste("Correlation Heatmap for",  ENVs[1]),
        x = "Timepoint",
        y = "Trait",
        fill = "r"
      ) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 14),
        axis.text = element_text(size = 10)
      )
print(heatmap_plot)


###########################################################    
cor_results_df1<- cor_results_df %>% filter(Env == ENVs[2])


heatmap_plot <- ggplot(cor_results_df1, aes(x = factor(Timepoint), y = Trait, fill = correlation)) +
      geom_tile(color = "white") +
  facet_wrap(~VI)+
        scale_fill_gradient2(low = "blue", high = "red", midpoint = 0, limits = c(-1, 1)) +
        labs(
        title = paste("Correlation Heatmap for", "in", ENVs[2]),
        x = "Timepoint",
        y = "Trait",
        fill = "r"
      ) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 14),
        axis.text = element_text(size = 10)
      )
print(heatmap_plot)

cor_results_df1<- cor_results_df %>% filter(Env == ENVs[3])


heatmap_plot <- ggplot(cor_results_df1, aes(x = factor(Timepoint), y = Trait, fill = correlation)) +
      geom_tile(color = "white") +
  facet_wrap(~VI)+
        scale_fill_gradient2(low = "blue", high = "red", midpoint = 0, limits = c(-1, 1)) +
        labs(
        title = paste("Correlation Heatmap for",  ENVs[3]),
        x = "Timepoint",
        y = "Trait",
        fill = "r"
      ) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 14),
        axis.text = element_text(size = 10)
      )
print(heatmap_plot)
####################################################
cor_results_df1<- cor_results_df %>% filter(Env == ENVs[4])


heatmap_plot <- ggplot(cor_results_df1, aes(x = factor(Timepoint), y = Trait, fill = correlation)) +
      geom_tile(color = "white") +
  facet_wrap(~VI)+
        scale_fill_gradient2(low = "blue", high = "red", midpoint = 0, limits = c(-1, 1)) +
        labs(
        title = paste("Correlation Heatmap for",  ENVs[4]),
        x = "Timepoint",
        y = "Trait",
        fill = "r"
      ) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 14),
        axis.text = element_text(size = 10)
      )
print(heatmap_plot)
####################################################
cor_results_df1<- cor_results_df %>% filter(Env == ENVs[5])


heatmap_plot <- ggplot(cor_results_df1, aes(x = factor(Timepoint), y = Trait, fill = correlation)) +
      geom_tile(color = "white") +
  facet_wrap(~VI)+
        scale_fill_gradient2(low = "blue", high = "red", midpoint = 0, limits = c(-1, 1)) +
        labs(
        title = paste("Correlation Heatmap for", ENVs[5]),
        x = "Timepoint",
        y = "Trait",
        fill = "r"
      ) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 14),
        axis.text = element_text(size = 10)
      )
print(heatmap_plot)
```




# **6. Summarizing correlations over VI: componenet: trait**

```{r, echo= FALSE}
ggplot(cs_mq, aes(x = Trait, y = correlation)) +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~Component , scales = "free_y") +
  labs(
    x = "FPC",
    y = "Correlation with Trait",
    title = "Distribution of Correlations: FPCs by VI"
  ) +
  #theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )



ggplot(cs_mq, aes(x = Trait, y = correlation)) +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~VI , scales = "free_y") +
  labs(
    x = "FPC",
    y = "Correlation with Trait",
    title = "Distribution of Correlations: FPCs by VI"
  ) +
  #theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )

ggplot(cs_ag, aes(x = Component, y = correlation, fill = VI)) +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~ Trait, scales = "free_y") +
  labs(
    x = "FPC",
    y = "Correlation with Trait",
    title = "Distribution of Correlations: FPCs by VI"
  ) +
  #theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )
```


### *env, #componenet, vi, #trait*
```{r}
################################################################################
#testing the component
ggplot(cs_ag, aes(x = Component, y = correlation)) +
  geom_boxplot(outlier.size = 0.5) +
  labs(
    x = "FPC",
    y = "Correlation with Trait",
    title = "Distribution of Correlations: FPCs by VI"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom")

ggplot(cs_ag, aes(x = Component, y = correlation, fill = Env)) +
  geom_boxplot(outlier.size = 0.5) +
  labs(
    x = "FPC",
    y = "Correlation with Trait",
    title = "Distribution of Correlations: FPCs by VI"
  ) +
  #theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom")
# fill no diff: VI
# fill diff: trait and env in FPC1 an PC1, 



################################################################################
#testing the trait
ggplot(cs_ag, aes(x = Trait, y = correlation)) +
  geom_boxplot(outlier.size = 0.5) +
  labs(
    x = "FPC",
    y = "Correlation with Trait",
    title = "Distribution of Correlations: FPCs by VI"
  ) +
  #theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom")

ggplot(cs_ag, aes(x = Trait, y = correlation, fill = VI)) +
  geom_boxplot(outlier.size = 0.5) +
  labs(
    x = "FPC",
    y = "Correlation with Trait",
    title = "Distribution of Correlations: FPCs by VI"
  ) +
  #theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom")

# fill no diff: env, vi
# fill diff:compoenent in some

################################################################################
#testing the VI
ggplot(cs_ag, aes(x = VI, y = correlation)) +
  geom_boxplot(outlier.size = 0.5) +
  labs(
    x = "FPC",
    y = "Correlation with Trait",
    title = "Distribution of Correlations: FPCs by VI"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom")

ggplot(cs_ag, aes(x = VI, y = correlation, fill = Trait)) +
  geom_boxplot(outlier.size = 0.5) +
  labs(
    x = "FPC",
    y = "Correlation with Trait",
    title = "Distribution of Correlations: FPCs by VI"
  ) +
  #theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom")

# fill no diff: env(23 and 24 have more var), trait
# fill diff: FPC1 and PC1 have some varianc.e

################################################################################
#testing the env
ggplot(cs_ag, aes(x = Env, y = correlation)) +
  geom_boxplot(outlier.size = 0.5) +
  labs(
    x = "FPC",
    y = "Correlation with Trait",
    title = "Distribution of Correlations: FPCs by VI"
  ) +
  #theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom")

ggplot(cs_ag, aes(x = Env, y = correlation, fill = VI)) +
  geom_boxplot(outlier.size = 0.5) +
  labs(
    x = "FPC",
    y = "Correlation with Trait",
    title = "Distribution of Correlations: FPCs by VI"
  ) +
  #theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom")

# fill no diff:VI
# fill diff: FPC1 and PC1 have some variance,  trait(23 and 24 have more var and high cor)

#23 and 22 have stronger corrs
#env : trait
#FPC1 PC1 better than other. 
```



### *the corrs of PCs with VIs, traits summary*
```{r}

#print(pp, n= 24)
# Env- how a VI or componenet correaltes over years
cs_ag %>%
  group_by(Env, , Trait) %>%
  summarise(mean = mean(correlation), var= var(correlation), min= min(correlation), sd= sd(correlation), n = n())%>% arrange(Trait)

# VI- how a componenet or Env correaltes over VIs
ip<- cs_ag %>%
  group_by(VI, , Trait) %>%
  summarise(mean = mean(correlation), var= var(correlation),min= min(correlation), sd= sd(correlation), n = n())%>% arrange(Trait)

print(ip, n= 40)

# Componenet- how a VI or Env correaltes over componenet
pp<- cs_ag %>%
  group_by(Component, Trait) %>%
  summarise(mean = mean(correlation), var= var(correlation),  min= min(correlation), sd= sd(correlation), n = n())%>% arrange(Trait) %>% arrange(mean)

cs_ag %>%
  group_by(Component, Trait) %>%
  summarise(mean = mean(correlation), var= var(correlation),  min= min(correlation), sd= sd(correlation), n = n())%>% arrange(Trait) %>% arrange(mean)

###################################################################################################################
# Env- how a VI or componenet correaltes over years
ii<- cs_mq %>%
  group_by(Env,  Trait) %>%
  summarise(mean = mean(correlation), var= var(correlation), min= min(correlation), sd= sd(correlation), n = n())%>% arrange(Trait)

print(ii, n= 34)

# VI- how a componenet or Env correaltes over VIs
ip <- cs_mq %>%
  group_by(VI, , Trait) %>%
  summarise(mean = mean(correlation), var= var(correlation),min= min(correlation), sd= sd(correlation), n = n()) %>% arrange(Trait)

# Componenet- how a VI or Env correaltes over componenet
ip<- cs_mq %>%
  group_by(Component, Trait) %>%
  summarise(mean = mean(correlation), var= var(correlation),  min= min(correlation), sd= sd(correlation), n = n()) %>% arrange(Trait) %>% arrange(mean)

print(ip, n= 170)

tp<- rbind(ip, pp)

openxlsx::write.xlsx(tp, file= "analysis/CoR_MQ_ag.xlsx")
```




# **7. Variable Importance score **


## *. Variable Importance score for yield*

```{r, eval=FALSE}

#21 and 4
#22 and 6
#23 and 5
#24 and 7
#25 and 10

load(file.path(base_path, "data/data_sp.Rdata"))
traits <- names(data_sp)[c(5:13)]
data_sp_long <- tidyr::pivot_longer(data_sp, all_of(traits), names_to="trait", values_to="value")

#data_sp_long <- data_sp_long[,-5:-11]

#extracting raw spec for MT GP
##############
d21<- data_sp_long %>% filter(Env == "KET21") %>% filter(Timepoint  == "4")

d22<- data_sp_long %>% filter(Env == "SNY22") %>% filter(Timepoint  == "6")

d23<- data_sp_long %>% filter(Env == "MCG23") %>% filter(Timepoint == "5" )

d24<- data_sp_long %>% filter(Env == "HELF24") %>% filter(Timepoint =="7")

d25<- data_sp_long %>% filter(Env ==  "MCG25" ) %>% filter(Timepoint =="12")


MT_data<- rbind(d21, d22, d23, d24, d25)

table(MT_data$Env, MT_data$Timepoint)



################################################################################################ 

# extract MT from BLUP
load("data/spec_BLUE_spatial.RData")

d21<- preds %>% filter(Env == "KET21") %>% filter(Timepoint == "4" )

d22<- preds %>% filter(Env == "SNY22") %>% filter(Timepoint == "6" )

d23<- preds %>% filter(Env == "MCG23") %>% filter(Timepoint == "5" )

d24<- preds %>% filter(Env == "HELF24") %>% filter(Timepoint =="7")

d25<- preds %>% filter(Env ==  "MCG25" ) %>% filter(Timepoint =="12")

colnames(d21)[5]<- "value"
colnames(d22)[5]<- "value"
colnames(d23)[5]<- "value"
colnames(d24)[5]<- "value"
colnames(d25)[5]<- "value"

MT_data_preds<- rbind(d21,d22, d23, d24, d25)
####################################

str(MT_data)
str(data_ag)
MT_ag_raw <- left_join(MT_data, data_ag[,c(1,3, 5,6)], by= c( "Env", "plot_id"))
table(MT_ag_raw$Env, MT_ag_raw$Timepoint)

save(MT_ag_raw, file= "data/MT_ag_raw.Rdata")

###################################
# joining the selected timepoint for GP with BLues for yield and VIs.
load("data/ag_BLUE_spatial.RData")

tr<- c("HD_JD", "yield_kgha")
preds<- preds[,-5] %>% filter(VI %in% tr )
ag_pred <- preds %>%  pivot_wider(names_from = VI, values_from = Predicted)
MT_ag_preds <- left_join(MT_data_preds, ag_pred, by= c("GID", "Env"))
save(MT_ag_preds, file= "data/MT_ag_preds.Rdata")

###################################
# joining the selected timepoint for GP for raw  MQ and VIs.
MT_mq_raw <- left_join(MT_data, data_mq, by= c("plot_id")) %>% na.omit
save(MT_mq_raw, file= "data/MT_mq_raw.Rdata")
```

