---
title: "P_kernel_prep"
author: "sss322"
date: "2025-02-07"
output: html_document
---

```{r}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
knitr::opts_chunk$set(dev = "png", dpi = 150, fig.keep = "all", fig.show = "asis")
```
# **Load Dependencies**

```{r, message=FALSE}
library(MASS)
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
```



Calculating a Relationship matrix

- RAW:
 - ALL different VIs from the specific timepoints 
 - ALL different VIs from all timepoints
 - input: MT_ag_raw and MT_ag_preds
 
 
- BLUE:
 - ALL different VIs from the specific timepoints 
 - ALL different VIs from all timepoints
 - input: VImat and VImatBLUE combined


```{r}

base_path <- if (.Platform$OS.type == "windows") {
  "C:/Users/Siim Sepp/NY_WMB"
} else {
  "~/Documents/GitHub/NY_WMB"
}

load(file.path(base_path, "data/MT_ag_preds.Rdata"))
load(file.path(base_path, "data/MT_ag_raw.Rdata"))
```

# **1. Prepare traits to predict**

**Separate agronomic data to predict**
```{r}
#Integrated specralBLU w ag BLUE and raw spectral BLUE with ag raw. long verisions
MT_preds <- MT_ag_preds[,-6] %>% drop_na() #%>% filter(VI %in% vis) 

MT_raw<-  MT_ag_raw[,c(1,3,4,6,7)] %>% drop_na() #%>% filter(trait %in% vis) 

############################################################
MT_raw <- MT_raw %>%
  dplyr::mutate(Env_TP = paste(Env,Timepoint, sep = "_"))
MT_raw<- MT_raw[,-2:-3]
colnames(MT_raw)

# Split into a named list of dataframes by Env_Timepoint
MT_raw_w <- MT_raw %>%  split(.$Env_TP) %>%
  map(~ .x %>%  
        dplyr::select(-Env_TP) %>%
        pivot_wider(names_from = trait, values_from = value) %>%
        column_to_rownames("plot_id"))


#Separate into flight TP:ENV dataframes with ag traits. Lining up the columns for LM 
#############################################################
MT_preds <- MT_preds %>%
  dplyr::mutate(Env_TP = paste(Env, Timepoint, sep = "_"))
MT_preds <- MT_preds[,c(1,4,5,8)]
str(MT_preds)


# Split into a named list of dataframes by Env_Timepoint
MT_pr_w <- MT_preds %>% #[,c(-2,-3, -6, -7)] %>%
  split(.$Env_TP) %>%
  map(~ .x %>%
        dplyr::select(-Env_TP) %>%
        pivot_wider(names_from = VI, values_from = value) %>%
        column_to_rownames("GID"))
#############################################################
str(MT_preds)
str(MT_pr_w)
str(MT_raw_w)
str(MT_raw)
#MT are long dataframes of the predicted or raw VI with timepoint, trait and Env
#MT w are wide dataframes widened with the VI.made into a list per timepoint:Env 
```


**Adding the wide to the end of the list**
```{r}
load("data/VImat.Rdata")
MT_raw_w[["MCG25C"]]<- VImats$MCG25$combined %>%  column_to_rownames("plot_id") 
MT_raw_w[["HELF24C"]]<- VImats$HELF24$combined %>%  column_to_rownames("plot_id") 
MT_raw_w[["MCG23C"]]<- VImats$MCG23$combined %>%  column_to_rownames("plot_id") 
MT_raw_w[["SNY22C"]]<- VImats$SNY22$combined %>%  column_to_rownames("plot_id") 
MT_raw_w[["KET21C"]]<- VImats$KET21$combined %>%  column_to_rownames("plot_id") 
save(MT_raw_w, file= "data/MT_raw_wide.Rdata")

#########################################

load("data/VImatBLUE.Rdata")

MT_pr_w[["MCG25C"]]<- VImats$MCG25$combined %>%  column_to_rownames("GID")
MT_pr_w[["HELF24C"]]<- VImats$HELF24$combined %>%  column_to_rownames("GID")
MT_pr_w[["MCG23C"]]<- VImats$MCG23$combined   %>%  column_to_rownames("GID")
MT_pr_w[["SNY22C"]]<- VImats$SNY22$combined   %>%  column_to_rownames("GID")
MT_pr_w[["KET21C"]]<- VImats$KET21$combined   %>%  column_to_rownames("GID")

save(MT_pr_w, file= "data/MT_pred_wide.Rdata")


#QC check
MT_pr_w[[1]][1:3,1:3]
MT_raw_w[[1]][1:3,1:3]
```




# **2. Creating a relationship matrix for each list item**

```{r}
# Initialize output list
MT_sim_mat_pr <- list()

# Loop through each environment × timepoint entry in the list
for (name in names(MT_pr_w)) {
  mat <- as.matrix(MT_pr_w[[name]])
  
  # Euclidean distance matrix
  eu_dist <- dist(mat, method = "euclidean")
  
  # Convert to full matrix and compute similarity
  sim_mat <- 1 / (1 + as.matrix(eu_dist))
  
  # Preserve GID as row and column names
  rownames(sim_mat) <- rownames(mat)
  colnames(sim_mat) <- rownames(mat)
  
  # Save to output list
  MT_sim_mat_pr[[name]] <- sim_mat
}
MT_sim_mat_pr[[4]][1:3,1:3]



#######################################################
# Initialize output list
MT_sim_mat_raw <- list()

# Loop through each environment × timepoint entry in the list
for (name in names(MT_raw_w)) {
  mat <- as.matrix(MT_raw_w[[name]])
  
  # Euclidean distance matrix
  eu_dist <- dist(mat, method = "euclidean")
  
  # Convert to full matrix and compute similarity
  sim_mat <- 1 / (1 + as.matrix(eu_dist))
  
  # Preserve GID as row and column names
  rownames(sim_mat) <- rownames(mat)
  colnames(sim_mat) <- rownames(mat)
  
  # Save to output list
  MT_sim_mat_raw[[name]] <- sim_mat
}


library(Matrix)

M <- MT_sim_mat_pr$KET21C

M[is.na(M)] <- 0

MT_sim_mat_pr$KET21C<- as.matrix(nearPD(M, corr = FALSE, keepDiag = TRUE)$mat)
MT_sim_mat_pr$KET21_4<- as.matrix(nearPD(MT_sim_mat_pr$KET21_4, corr = FALSE, keepDiag = TRUE)$mat)

MT_sim_mat_raw$KET21C<- as.matrix(nearPD(MT_sim_mat_raw$KET21C, corr = FALSE, keepDiag = TRUE)$mat)
MT_sim_mat_raw$KET21_4<- as.matrix(nearPD(MT_sim_mat_raw$KET21_4, corr = FALSE, keepDiag = TRUE)$mat)

save(MT_sim_mat_raw, file= "data/MT_raw_mat.Rdata")
save(MT_sim_mat_pr, file= "data/MT_pred_mat.Rdata")
```



# **3. Heatmap Plots**

```{r, fig.keep='all', fig.show='asis', dev='png', out.width='80%', fig.align='center', warning=FALSE, message=FALSE}

heatmap(as.matrix(MT_sim_mat_pr[[1]]), main = paste0("Phenomic Relationship Matrix BLUP  ", names(MT_sim_mat_pr)[1]))
heatmap(as.matrix(MT_sim_mat_pr[[2]]), main = paste0("Phenomic Relationship Matrix BLUP  ", names(MT_sim_mat_pr)[2]))
heatmap(as.matrix(MT_sim_mat_pr[[3]]), main = paste0("Phenomic Relationship Matrix BLUP  ", names(MT_sim_mat_pr)[3]))
heatmap(as.matrix(MT_sim_mat_pr[[4]]), main = paste0("Phenomic Relationship Matrix BLUP  ", names(MT_sim_mat_pr)[4]))
heatmap(as.matrix(MT_sim_mat_pr[[5]]), main = paste0("Phenomic Relationship Matrix BLUP  ", names(MT_sim_mat_pr)[5]))
heatmap(as.matrix(MT_sim_mat_pr[[6]]), main = paste0("Phenomic Relationship Matrix BLUP  ", names(MT_sim_mat_pr)[6]))
heatmap(as.matrix(MT_sim_mat_pr[[7]]), main = paste0("Phenomic Relationship Matrix BLUP  ", names(MT_sim_mat_pr)[7]))
heatmap(as.matrix(MT_sim_mat_pr[[8]]), main = paste0("Phenomic Relationship Matrix BLUP  ", names(MT_sim_mat_pr)[8]))
heatmap(as.matrix(MT_sim_mat_pr[[9]]), main = paste0("Phenomic Relationship Matrix BLUP  ", names(MT_sim_mat_pr)[9]))
heatmap(as.matrix(MT_sim_mat_pr[[10]]), main = paste0("Phenomic Relationship Matrix BLUP  ", names(MT_sim_mat_pr)[10]))

length(MT_sim_mat_pr)
length(MT_sim_mat_raw)
MT_sim_mat_raw[[1]][1:3,1:3]
MT_sim_mat_raw[[2]][1:3,1:3]
MT_sim_mat_raw[[3]][1:3,1:3]


#########################################################################################################

heatmap(as.matrix(MT_sim_mat_raw[[1]]), main = paste0("Phenomic Relationship Matrix Raw  ", names(MT_sim_mat_raw)[1]))
heatmap(as.matrix(MT_sim_mat_raw[[2]]), main = paste0("Phenomic Relationship Matrix Raw  ", names(MT_sim_mat_raw)[2]))
heatmap(as.matrix(MT_sim_mat_raw[[3]]), main = paste0("Phenomic Relationship Matrix Raw  ", names(MT_sim_mat_raw)[3]))
heatmap(as.matrix(MT_sim_mat_raw[[4]]), main = paste0("Phenomic Relationship Matrix Raw  ", names(MT_sim_mat_raw)[4]))
heatmap(as.matrix(MT_sim_mat_raw[[5]]), main = paste0("Phenomic Relationship Matrix Raw  ", names(MT_sim_mat_raw)[5]))
heatmap(as.matrix(MT_sim_mat_raw[[6]]), main = paste0("Phenomic Relationship Matrix Raw  ", names(MT_sim_mat_raw)[6]))
heatmap(as.matrix(MT_sim_mat_raw[[7]]), main = paste0("Phenomic Relationship Matrix Raw  ", names(MT_sim_mat_raw)[7]))
heatmap(as.matrix(MT_sim_mat_raw[[8]]), main = paste0("Phenomic Relationship Matrix Raw  ", names(MT_sim_mat_raw)[8]))
heatmap(as.matrix(MT_sim_mat_raw[[9]]), main = paste0("Phenomic Relationship Matrix Raw  ", names(MT_sim_mat_raw)[9]))
heatmap(as.matrix(MT_sim_mat_raw[[10]]), main = paste0("Phenomic Relationship Matrix Raw  ", names(MT_sim_mat_raw)[10]))
```



