---
title: "7. Phenomic prediction models"
author: "sss322"
date: "2025-02-07"
output: html_document
---



```{r}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```






# **1. Import dependencies and data **


**Import dependencies**

```{r}
library(MASS)
library(dplyr)
library(tidyr)
library(caret)
library(glmnet)
#library(asreml)
library(ggplot2)
library(readr)
library(stringr)
```


**Import data**

```{r setup, include=FALSE}
base_path <- if (.Platform$OS.type == "windows") {
  "C:/Users/Siim Sepp/NY_WMBCL"
} else {
  "~/Documents/GitHub/NY_WMBCL"
}

load(file.path(base_path, "data/data_ag.Rdata"))
load(file.path(base_path, "data/data_mq.Rdata"))
load(file.path(base_path, "data/VImat.Rdata"))
```






# **2. Predict traits with rf, lm, glm and pcr **



**Write the function**

```{r}
run_pheno_models_long <- function(data_long, VImats, model_list = c("rf", "lm", "glmnet", "pcr"), cycles = 50) {


  VIs <- setdiff(names(VImats[[1]]), "plot_id")

  #storage
  accuracy_list <- list()
  var_importance_list <- list()
  prediction_list <- list()

  acc_idx <- varimp_idx <- pred_idx <- 1

  for (env_idx in names(VImats)) {
    for (trait_name in unique(data_long$VI)) {

      response_data <- data_long %>%
        filter(Env == env_idx, VI == trait_name)

      for (p in VIs) {

        predictor_data <- VImats[[env_idx]][[p]]

        merged_data <- inner_join(response_data, predictor_data, by = "plot_id") %>%
          rename(S.T = Predicted) %>%
          na.omit()

        cols <- c("S.T", colnames(predictor_data)[-1])
        cols <- cols[cols %in% colnames(merged_data)]
        merged_data <- merged_data[, cols]

        for (r in 1:cycles) {
          set.seed(r)

          ind <- createDataPartition(merged_data$S.T, p = 0.8, list = FALSE)
          train <- merged_data[ind, ]
          test  <- merged_data[-ind, ]

          nzv <- nearZeroVar(train, saveMetrics = TRUE)
          train <- train[, !nzv$nzv]

          if (ncol(train) <= 1) next

          custom <- trainControl(method = "repeatedcv", number = 5, repeats = 1)

          for (model_type in model_list) {
            if (model_type == "glmnet") {
              for (alpha_val in c(0, 0.5, 1)) {
                model <- tryCatch(train(S.T ~ ., data = train, method = "glmnet",
                                        tuneGrid = expand.grid(alpha = alpha_val, lambda = seq(0.0001, 1, length = 10)),
                                        trControl = custom), error = function(e) NULL)
                if (!is.null(model)) {
                  model_pred <- predict(model, test)
                  model_acc <- cor(model_pred, test$S.T, use = "complete.obs")

                  model_name <- c("0" = "Ridge", "0.5" = "ElasticNet", "1" = "Lasso")[as.character(alpha_val)]

                  accuracy_list[[acc_idx]] <- data.frame(Environment = env_idx, Cycle = r,
                                                         Trait = trait_name, Predictor_Set = p,
                                                         Model = model_name, Accuracy = model_acc)
                  acc_idx <- acc_idx + 1

                  prediction_list[[pred_idx]] <- data.frame(Environment = env_idx, Cycle = r,
                                                            Trait = trait_name, Predictor_Set = p,
                                                            Model = model_name, True = test$S.T,
                                                            Predicted = model_pred)
                  pred_idx <- pred_idx + 1

                  varimp <- tryCatch(varImp(model)$importance, error = function(e) NULL)
                  if (!is.null(varimp)) {
                    var_importance_list[[varimp_idx]] <- data.frame(Environment = env_idx, Cycle = r,
                                                                    Trait = trait_name, Predictor_Set = p,
                                                                    Model = model_name,
                                                                    Predictor = rownames(varimp),
                                                                    Importance = varimp$Overall)
                    varimp_idx <- varimp_idx + 1
                  }
                }
              }
            } else {
              if (model_type == "pcr" && nrow(train) <= (ncol(train) - 1)) next
              model <- tryCatch(train(S.T ~ ., data = train, method = model_type, trControl = custom), error = function(e) NULL)
              if (!is.null(model)) {
                model_pred <- predict(model, test)
                model_acc <- cor(model_pred, test$S.T, use = "complete.obs")

                model_name <- c("rf" = "Random Forest", "lm" = "Linear Regression", "pcr" = "PCR")[model_type]

                accuracy_list[[acc_idx]] <- data.frame(Environment = env_idx, Cycle = r,
                                                       Trait = trait_name, Predictor_Set = p,
                                                       Model = model_name, Accuracy = model_acc)
                acc_idx <- acc_idx + 1

                prediction_list[[pred_idx]] <- data.frame(Environment = env_idx, Cycle = r,
                                                          Trait = trait_name, Predictor_Set = p,
                                                          Model = model_name, True = test$S.T,
                                                          Predicted = model_pred)
                pred_idx <- pred_idx + 1

                if (model_type %in% c("rf", "lm")) {
                  varimp <- tryCatch(varImp(model)$importance, error = function(e) NULL)
                  if (!is.null(varimp)) {
                    var_importance_list[[varimp_idx]] <- data.frame(Environment = env_idx, Cycle = r,
                                                                    Trait = trait_name, Predictor_Set = p,
                                                                    Model = model_name,
                                                                    Predictor = rownames(varimp),
                                                                    Importance = varimp$Overall)
                    varimp_idx <- varimp_idx + 1
                  }
                } else if (model_type == "pcr") {
                  var_importance_list[[varimp_idx]] <- data.frame(Environment = env_idx, Cycle = r,
                                                                  Trait = trait_name, Predictor_Set = p,
                                                                  Model = model_name,
                                                                  Predictor = NA, Importance = NA)
                  varimp_idx <- varimp_idx + 1
                }
              }
            }
          }
        }
      }
    }
  }

  return(list(
    accuracy = do.call(rbind, accuracy_list),
    importance = do.call(rbind, var_importance_list),
    predictions = do.call(rbind, prediction_list)
  ))
}

```




**Prepare phenotype data**
```{r}
#prepare agronomic traits
traits_ag<- colnames(data_ag)[7:8]
data_ag<- data_ag %>%
  pivot_longer(cols = all_of(traits_ag), names_to = "VI", values_to = "Predicted")

#prepare maltinq quality traits
traits_mq<- colnames(data_mq)[4:15]
data_mq<- data_mq%>%
  pivot_longer(cols = all_of(traits_mq), names_to = "VI", values_to = "Predicted")

#prepare agronomic BLUE traits
load("data/ag_BLUE_spatial.RData")

data_ag_pr <- preds[,-5] %>% filter(VI %in% traits_ag )%>% pivot_wider(names_from = "VI", values_from = "Predicted")

#prepare mq BLUE traits
#load("data/mq_BLUp_spatial.RData")
#load("data/mq_BLUP_spatial.RData")

#data_mq_pr <- preds[,-5] %>% filter(VI %in% traits_mq )%>% pivot_wider(names_from = "VI", values_from = "Predicted")
```




**Run the predicton for agronomic traits RAW**

```{r}
PP_LM_results_ag_raw  <- run_pheno_models_long(data_ag, VImats, model_list = c("rf", "lm", "glmnet", "pcr"), cycles = 5)

#hist(PP_LM_results_ag_raw$accuracy$Accuracy)
save(PP_LM_results_ag_raw, file= "PP_LM_results_ag_raw.Rdata")
#table(PP_LM_results_ag_raw$accuracy$Environment, PP_LM_results_ag_raw$accuracy$Trait)
#boxplot(PP_LM_results_ag_raw$accuracy$Accuracy~ PP_LM_results_ag_raw$accuracy$Trait)
```




**Run the predicton for MQ traits RAW**

```{r}
PP_LM_results_mq_raw  <- run_pheno_models_long(data_mq, VImats[c(1,3)], model_list = c("rf", "lm", "glmnet", "pcr"), cycles = 50)

save(PP_LM_results_mq_raw, file= "PP_LM_results_mq_raw.Rdata")
```




**Run the predicton for agronomic traits PRED**

```{r}
# Define parameters
load("~/data/VImatBLUE.Rdata")
#load("C:/Users/Siim Sepp/NY_WMB/data/VImatBLUE.Rdata")
VIs<- names(VImats$KET21)[-3]

cycles <- 50

# Models to run
model_list <- c("rf", "lm", "glmnet", "pcr")

# Initialize storage lists
accuracy_list <- list()
var_importance_list <- list()
prediction_list <- list()

# Initialize counters
acc_idx <- 1
varimp_idx <- 1
pred_idx <- 1


# Start loops
for (env_idx in names(VImats)) {
  for (t in traits_ag) {
    
    response_data <- data_ag_pr %>%
      filter(Env == env_idx)
    
    for (p in VIs) {
      
      predictor_data <- VImats[[env_idx]][[p]] #%>% na.omit()
      
      merged_data <- inner_join(response_data, predictor_data, by = "GID") %>%
        rename(S.T = all_of(t)) #%>% na.omit()
      
      #merged_data <- merged_data %>% select(where(~ mean(!is.na(.)) > 0.5)) %>% na.omit()
      
      cols <- c("S.T", colnames(predictor_data)[-1])
      cols <- cols[cols %in% colnames(merged_data)]
      merged_data <- merged_data[, cols]
      
      for (r in 1:cycles) {
        set.seed(r)
        
        ind <- createDataPartition(merged_data$S.T, p = 0.8, list = FALSE)
        train <- merged_data[ind, ]  
        test  <- merged_data[-ind, ] 
        
        # Pre-clean predictors
        nzv <- nearZeroVar(train, saveMetrics = TRUE)
        train <- train[, !nzv$nzv]
        
        if (ncol(train) <= 1) next  # 
        
        custom <- trainControl(method = "repeatedcv", number = 5, repeats = 1)
        
        for (model_type in model_list) {
          
          if (model_type == "glmnet") {
            for (alpha_val in c(0, 0.5, 1)) {
              model <- tryCatch(
                train(
                  S.T ~ ., data = train, method = "glmnet",
                  tuneGrid = expand.grid(alpha = alpha_val,
                                         lambda = seq(0.0001, 1, length = 10)),
                  trControl = custom
                ),
                error = function(e) NULL
              )
              
              if (!is.null(model)) {
                model_pred <- tryCatch(predict(model, test), error = function(e) rep(NA, nrow(test)))
                
                if (!all(is.na(model_pred))) {
                  model_acc <- cor(model_pred, test$S.T, use = "complete.obs")
                } else {
                  model_acc <- NA
                }
                
                model_name <- switch(
                  as.character(alpha_val),
                  "0" = "Ridge Regression",
                  "0.5" = "Elastic Net",
                  "1" = "Lasso Regression"
                )
                
                # Save accuracy
                accuracy_list[[acc_idx]] <- data.frame(
                  Environment = env_idx, Cycle = r, Trait = t,
                  Predictor_Set = p, Model = model_name, Accuracy = model_acc
                )
                acc_idx <- acc_idx + 1
                
                # Save predictions (optional)
                prediction_list[[pred_idx]] <- data.frame(
                  Environment = env_idx, Cycle = r, Trait = t,
                  Predictor_Set = p, Model = model_name,
                  True = test$S.T, Predicted = model_pred
                )
                pred_idx <- pred_idx + 1
                
                # Save variable importance
                varimp <- tryCatch(varImp(model)$importance, error = function(e) NULL)
                if (!is.null(varimp)) {
                  varimp_df <- data.frame(
                    Environment = env_idx,
                    Cycle = r,
                    Trait = t,
                    Predictor_Set = p,
                    Model = model_name,
                    Predictor = rownames(varimp),
                    Importance = varimp$Overall
                  )
                  var_importance_list[[varimp_idx]] <- varimp_df
                  varimp_idx <- varimp_idx + 1
                }
              }
            }
          } else {
            # Skip if nrow too small compared to ncol
            if (model_type == "pcr" && nrow(train) <= (ncol(train) - 1)) next
            
            model <- tryCatch(
              train(S.T ~ ., data = train, method = model_type, trControl = custom),
              error = function(e) NULL
            )
            
            if (!is.null(model)) {
              model_pred <- tryCatch(predict(model, test), error = function(e) rep(NA, nrow(test)))
              
              if (!all(is.na(model_pred))) {
                model_acc <- cor(model_pred, test$S.T, use = "complete.obs")
              } else {
                model_acc <- NA
              }
              
              model_name <- switch(
                model_type,
                "rf" = "Random Forest",
                "lm" = "Linear Regression",
                "pcr" = "Principal Component Regression",
                model_type
              )
              
              # Save accuracy
              accuracy_list[[acc_idx]] <- data.frame(
                Environment = env_idx, Cycle = r, Trait = t,
                Predictor_Set = p, Model = model_name, Accuracy = model_acc
              )
              acc_idx <- acc_idx + 1
              
              # Save predictions (optional)
              prediction_list[[pred_idx]] <- data.frame(
                Environment = env_idx, Cycle = r, Trait = t,
                Predictor_Set = p, Model = model_name,
                True = test$S.T, Predicted = model_pred
              )
              pred_idx <- pred_idx + 1
              
              # Save variable importance
              if (model_type %in% c("rf", "lm")) {
                varimp <- tryCatch(varImp(model)$importance, error = function(e) NULL)
                if (!is.null(varimp)) {
                  varimp_df <- data.frame(
                    Environment = env_idx,
                    Cycle = r,
                    Trait = t,
                    Predictor_Set = p,
                    Model = model_name,
                    Predictor = rownames(varimp),
                    Importance = varimp$Overall
                  )
                  var_importance_list[[varimp_idx]] <- varimp_df
                  varimp_idx <- varimp_idx + 1
                }
              } else if (model_type == "pcr") {
                # PCR: no var importance, add placeholder
                varimp_df <- data.frame(
                  Environment = env_idx,
                  Cycle = r,
                  Trait = t,
                  Predictor_Set = p,
                  Model = model_name,
                  Predictor = NA,
                  Importance = NA
                )
                var_importance_list[[varimp_idx]] <- varimp_df
                varimp_idx <- varimp_idx + 1
              }
            }
          }
        }
      }
    }
  }
}

# Combine the results
accuracy_results <- do.call(rbind, accuracy_list)
var_importance_results <- do.call(rbind, var_importance_list)
prediction_results <- do.call(rbind, prediction_list)

PP_LM_results_ag_pred <- list(accuracy_results, var_importance_results, prediction_results) 
save(PP_LM_results_ag_pred, file= "PP_LM_results_ag_pred.Rdata")
```




**Run the predicton for mq traits PRED**

```{r}
# Define parameters
#load("~/data/VImatBLUE.Rdata")
load("C:/Users/Siim Sepp/NY_WMB/data/VImatBLUE.Rdata")

cycles <- 5

# Models to run
model_list <- c("rf", "lm", "glmnet", "pcr")

# Initialize storage lists
accuracy_list <- list()
var_importance_list <- list()
prediction_list <- list()

# Initialize counters
acc_idx <- 1
varimp_idx <- 1
pred_idx <- 1

traits_mq<- colnames(data_mq_pr)[3:12]

#data_mq_pr <- data_mq_pr %>% na.omit()
names(VImats)
# Start loops
for (env_idx in names(VImats)[4]) {
  for (t in traits_mq) {
    
    response_data <- data_mq_pr %>%
      filter(Env == env_idx)
    
    for (p in VIs) {
      
      predictor_data <- VImats[[env_idx]][[p]]
      
      merged_data <- inner_join(response_data, predictor_data, by = "GID") %>%
        rename(ST = all_of(t)) #%>% na.omit()
      
      cols <- c("ST", colnames(predictor_data)[-1])
      cols <- cols[cols %in% colnames(merged_data)]
      merged_data <- merged_data[, cols]
      
      for (r in 1:cycles) {
        set.seed(r)
        
        ind <- createDataPartition(merged_data$ST, p = 0.8, list = FALSE)
        train <- merged_data[ind, ]
        test  <- merged_data[-ind, ]
        
        # Pre-clean predictors
        nzv <- nearZeroVar(train, saveMetrics = TRUE)
        train <- train[, !nzv$nzv]
        
        if (ncol(train) <= 1) next  # 
        
        custom <- trainControl(method = "repeatedcv", number = 5, repeats = 1)
        
        for (model_type in model_list) {
          
          if (model_type == "glmnet") {
            for (alpha_val in c(0, 0.5, 1)) {
              model <- tryCatch(
                train(
                  S.T ~ ., data = train, method = "glmnet",
                  tuneGrid = expand.grid(alpha = alpha_val,
                                         lambda = seq(0.0001, 1, length = 10)),
                  trControl = custom
                ),
                error = function(e) NULL
              )
              
              if (!is.null(model)) {
                model_pred <- tryCatch(predict(model, test), error = function(e) rep(NA, nrow(test)))
                
                if (!all(is.na(model_pred))) {
                  model_acc <- cor(model_pred, test$ST, use = "complete.obs")
                } else {
                  model_acc <- NA
                }
                
                model_name <- switch(
                  as.character(alpha_val),
                  "0" = "Ridge Regression",
                  "0.5" = "Elastic Net",
                  "1" = "Lasso Regression"
                )
                
                # Save accuracy
                accuracy_list[[acc_idx]] <- data.frame(
                  Environment = env_idx, Cycle = r, Trait = t,
                  Predictor_Set = p, Model = model_name, Accuracy = model_acc
                )
                acc_idx <- acc_idx + 1
                
                # Save predictions (optional)
                prediction_list[[pred_idx]] <- data.frame(
                  Environment = env_idx, Cycle = r, Trait = t,
                  Predictor_Set = p, Model = model_name,
                  True = test$ST, Predicted = model_pred
                )
                pred_idx <- pred_idx + 1
                
                # Save variable importance
                varimp <- tryCatch(varImp(model)$importance, error = function(e) NULL)
                if (!is.null(varimp)) {
                  varimp_df <- data.frame(
                    Environment = env_idx,
                    Cycle = r,
                    Trait = t,
                    Predictor_Set = p,
                    Model = model_name,
                    Predictor = rownames(varimp),
                    Importance = varimp$Overall
                  )
                  var_importance_list[[varimp_idx]] <- varimp_df
                  varimp_idx <- varimp_idx + 1
                }
              }
            }
          } else {
            # Skip PCR if nrow too small compared to ncol
            if (model_type == "pcr" && nrow(train) <= (ncol(train) - 1)) next
            
            model <- tryCatch(
              train(ST ~ ., data = train, method = model_type, trControl = custom),
              error = function(e) NULL
            )
            
            if (!is.null(model)) {
              model_pred <- tryCatch(predict(model, test), error = function(e) rep(NA, nrow(test)))
              
              if (!all(is.na(model_pred))) {
                model_acc <- cor(model_pred, test$ST, use = "complete.obs")
              } else {
                model_acc <- NA
              }
              
              model_name <- switch(
                model_type,
                "rf" = "Random Forest",
                "lm" = "Linear Regression",
                "pcr" = "Principal Component Regression",
                model_type
              )
              
              # Save accuracy
              accuracy_list[[acc_idx]] <- data.frame(
                Environment = env_idx, Cycle = r, Trait = t,
                Predictor_Set = p, Model = model_name, Accuracy = model_acc
              )
              acc_idx <- acc_idx + 1
              
              # Save predictions (optional)
              prediction_list[[pred_idx]] <- data.frame(
                Environment = env_idx, Cycle = r, Trait = t,
                Predictor_Set = p, Model = model_name,
                True = test$ST, Predicted = model_pred
              )
              pred_idx <- pred_idx + 1
              
              # Save variable importance
              if (model_type %in% c("rf", "lm")) {
                varimp <- tryCatch(varImp(model)$importance, error = function(e) NULL)
                if (!is.null(varimp)) {
                  varimp_df <- data.frame(
                    Environment = env_idx,
                    Cycle = r,
                    Trait = t,
                    Predictor_Set = p,
                    Model = model_name,
                    Predictor = rownames(varimp),
                    Importance = varimp$Overall
                  )
                  var_importance_list[[varimp_idx]] <- varimp_df
                  varimp_idx <- varimp_idx + 1
                }
              } else if (model_type == "pcr") {
                # PCR: no var importance, add placeholder
                varimp_df <- data.frame(
                  Environment = env_idx,
                  Cycle = r,
                  Trait = t,
                  Predictor_Set = p,
                  Model = model_name,
                  Predictor = NA,
                  Importance = NA
                )
                var_importance_list[[varimp_idx]] <- varimp_df
                varimp_idx <- varimp_idx + 1
              }
            }
          }
        }
      }
    }
  }
}

# Combine the results
accuracy_results <- do.call(rbind, accuracy_list)
var_importance_results <- do.call(rbind, var_importance_list)
prediction_results <- do.call(rbind, prediction_list)

PP_LM_results_mq_pred <- list(accuracy_results, var_importance_results, prediction_results) 
save(PP_LM_results_mq_pred, file= "PP_LM_results_mq_pred.Rdata")
```





# **3. Plot agronomic model variable importance and accuracy for RF, LM; GLM, PCR**


**Ordering and cleaning the var importance scores**
```{r}
#load("C:/Users/Siim Sepp/NY_WMB/analysis/PP_var_importance_scores_ag.RData")
load("C:/Users/Siim Sepp/NY_WMB/analysis/PP_LM_results_ag_pred.RData")
load("C:/Users/Siim Sepp/NY_WMB/analysis/PP_LM_results_mq_pred.RData")
load("C:/Users/Siim Sepp/NY_WMB/analysis/PP_LM_results_mq_raw.Rdata")
load("C:/Users/Siim Sepp/NY_WMB/analysis/PP_LM_results_ag_raw.Rdata")

PP_LM_results_ag_raw$importance$met<- "raw" 
PP_LM_results_mq_raw$importance$met<- "raw" 
PP_LM_results_ag_pred$importance$met<- "BLUP"
PP_LM_results_mq_pred$importance$met<- "BLUP"

names(PP_LM_results_mq_pred) <-  names(PP_LM_results_ag_raw)
names(PP_LM_results_ag_pred) <-  names(PP_LM_results_ag_raw)





var_importance_results<- rbind(PP_LM_results_ag_raw$importance, PP_LM_results_mq_raw$importance)
acc_results<- rbind(PP_LM_results_ag_raw$accuracy, PP_LM_results_mq_raw$accuracy)
pred_results<- rbind(PP_LM_results_ag_raw$predictions, PP_LM_results_mq_raw$predictions)

var_importance_results<- rbind(PP_LM_results_ag_pred$importance, PP_LM_results_mq_pred$importance)
acc_results<- rbind(PP_LM_results_ag_pred$accuracy, PP_LM_results_mq_pred$accuracy)
pred_results<- rbind(PP_LM_results_ag_pred$predictions, PP_LM_results_mq_pred$predictions)

###########################################################################
#ordering  and cleaning the Predictor nr for plots. 
var_importance_results <- var_importance_results %>%
  mutate(
    Predictor = as.character(Predictor),
    Predictor = str_replace_all(Predictor, "[`\\\\]", "") |> str_trim(),
    Predictor = factor(Predictor)
  )


unique(var_importance_results$Predictor)
var_importance_results <- var_importance_results %>% filter(Predictor_Set %in% c("NDVI", "PSRI")) %>% drop_na()

                                                            
ip<- levels(var_importance_results$Predictor)#[1:17]

#var_importance_results<- var_importance_results %>% filter(Predictor %in% ip ) 


var_importance_results <- var_importance_results %>%
  mutate(
    Predictor_num = parse_number(as.character(Predictor))
  )

var_importance_results <- var_importance_results %>%
  mutate(
    Predictor = factor(Predictor,
                       levels = Predictor[order(Predictor_num)],
                       labels = Predictor[order(Predictor_num)]))


var_importance_results$Predictor<-  as.numeric(var_importance_results$Predictor)
envs<- c( "KET21" , "SNY22",  "MCG23", "HELF24", "MCG25") 
var_importance_results <- var_importance_results %>%
  mutate( Environment = factor(Environment, levels = envs)) %>% arrange(Environment)
```




**Plotting Smooth varimp scores over envs, models, VIs and traits. **
```{r}

load("C:/Users/Siim Sepp/NY_WMB/JD_map.Rdata")
colnames(var_importance_results)[6]<- "Timepoint"

var_importance_results<- var_importance_results %>% left_join(JD_map, by= c("Environment", "Timepoint"))

head(var_importance_results)
colnames(JD_map)


var_importance_results1 <- var_importance_results %>%# filter(Trait == "yield_kgha") %>%
  filter(Predictor_Set %in% c("NDVI", "PSRI")) %>% filter(Model == "Linear Regression")# %>%
  #filter(Predictor_Set == "NDVI")
#########################################################################################

var_importance_plot <- var_importance_results1 %>%
  mutate(
    Date = as.Date(Date),
    Year = year(Date),
    DOY = yday(Date)
  )

# Verify different years per environment
cat("Years by Environment:\n")
print(var_importance_plot %>% 
  group_by(Env) %>% 
  summarise(Years = paste(unique(Year), collapse = ", ")))

# Create plot with DOY for overlap
month_labels <- data.frame(
  DOY = c(15, 45, 75, 105, 135, 165, 195, 225, 255, 285, 315, 345),
  Month = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
            "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
)

p <- ggplot(var_importance_plot,
       aes(x = DOY, y = Importance, color = Environment, fill = Environment)) +
  geom_smooth(se = FALSE, linewidth = 2, method = "loess") +
  scale_x_continuous(
    breaks = month_labels$DOY,
    labels = month_labels$Month,
    limits = c(min(var_importance_plot$DOY) - 5, 
               max(var_importance_plot$DOY) + 5)
  ) +
  scale_color_brewer(palette = "Set3") +
  labs(
    x = "Month",
    y = "Variable Importance Score",
    color = "Environment", 
    fill = "Environment"
  ) +
  theme(
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 16),
    legend.position = "bottom", legend.text = element_text(size=12)
  )

print(p)

```


```{r}
# ==============================================================================
# ANOVA FOR VARIABLE IMPORTANCE SCORES
# ==============================================================================

library(dplyr)
library(car)  # For Anova() with Type II/III SS
library(emmeans)  # For post-hoc tests
library(ggplot2)

# ==============================================================================
# 1. DATA PREPARATION
# ==============================================================================
str(var_importance_results)
# Convert to factors
var_importance_anova <- var_importance_results1 %>%
  mutate(
    Env = factor(Env),
    Date = factor(Date),
    Predictor_Set = factor(Predictor_Set),
    Model = factor(Model), 
    Trait = factor(Trait),
    Year = factor(Year)
  )

library(car)


# Prepare factors
var_importance_anova <- var_importance_results1 %>%
  mutate(across(c(Env, Predictor_Set, Model, Trait, Date), factor))

str(var_importance_anova)
# Run ANOVA
model <- lm(Importance ~ Env + Predictor_Set + Model + Trait, 
            data = var_importance_anova)

# Results
Anova(model, type = 2)

# Means
var_importance_anova %>%
  group_by(Model) %>%
  summarise(Mean = mean(Importance), SD = sd(Importance))

```



# **4. Phenomic prediction of MQ and agronomic  traits with phenomic relationship kernel CV1**



**Write mixed models function with phenomic kernel**

```{r}
run_blup_cv <- function(preds, MT_sim_mat, n_repeats = 50) {
  traits <- unique(preds$VI)
  envs <- unique(preds$Env)
  
  results <- data.frame()
  
  for (trait in traits) {
    cat("Processing trait:", trait, "\n")
    
    preds_trait <- preds %>% filter(VI == trait)
    
    for (env in envs) {
      cat("  Processing environment:", env, "\n")
      
      df <- preds_trait %>% filter(Env == env)
      
      if (!env %in% names(MT_sim_mat)) {
        cat("    Environment", env, "not found in relationship matrix list. Skipping.\n")
        next
      }
      
      # Rename 
      if ("plot_id" %in% names(df)) {
        df <- df %>% rename(plotID = plot_id)
      } else if ("GID" %in% names(df)) {
        df <- df %>% rename(plotID = GID)
      }
      
      K <- MT_sim_mat[[env]]
      gid <- as.character(df$plotID)
      
      common_gids <- intersect(gid, rownames(K))
      if (length(common_gids) < 10) {
        cat("    Too few common GIDs. Skipping.\n")
        next
      }
      
      df <- df %>% filter(plotID %in% common_gids)
      df$plotID <- factor(df$plotID, levels = common_gids)
      K <- K[common_gids, common_gids]


      
      all_cv_cor <- c()
      
      for (rep in 1:n_repeats) {
        set.seed(100 + rep)
        folds <- createFolds(df$Predicted, k = 5)
        cv_cor <- c()
        
        for (i in seq_along(folds)) {
          test_idx <- folds[[i]]
          train <- df[-test_idx, ]
          test  <- df[test_idx, ]
          
          train$plotID <- factor(train$plotID, levels = rownames(K))
          
          mod_cv <- tryCatch({
            if (all(c("Row", "Column") %in% names(train))) {
              asreml(
                fixed = Predicted ~ 1,
                random = ~ vm(plotID, K) + ar1(Row):ar1(Column),
                data = train,
                trace = FALSE
              )
            } else {
              asreml(
                fixed = Predicted ~ 1,
                random = ~ vm(plotID, K),
                data = train,
                trace = FALSE
              )
            }
          }, error = function(e) return(NULL))
          
          if (is.null(mod_cv)) next
          
          pred <- tryCatch({
            predict(mod_cv, classify = "plotID")
          }, error = function(e) return(NULL))
          
          if (is.null(pred)) next
          
          gebvs <- pred$pvals
          colnames(gebvs)[1] <- "plotID"
          
          gebvs <- gebvs[gebvs$plotID %in% test$plotID, ]
          test_sub <- test[test$plotID %in% gebvs$plotID, ]
          
          merged <- merge(test_sub, gebvs, by = "plotID")
          if (nrow(merged) > 0) {
  # Check if we have at least two complete (non-NA) pairs
  complete_cases <- complete.cases(merged$Predicted, merged$predicted.value)
  if (sum(complete_cases) > 1) {
    cor_val <- tryCatch({
      cor(merged$Predicted, merged$predicted.value, use = "complete.obs")
    }, warning = function(w) {
      cat("      Correlation warning:", conditionMessage(w), "\n")
      return(NA)
    }, error = function(e) {
      cat("      Correlation error:", conditionMessage(e), "\n")
      return(NA)
    })
    if (!is.na(cor_val)) {
      cv_cor <- c(cv_cor, cor_val)
    }
  } else {
    cat("      Not enough complete cases to compute correlation.\n")
  }
}

        }
        
        if (length(cv_cor) > 0) {
          all_cv_cor <- c(all_cv_cor, mean(cv_cor, na.rm = TRUE))
        }
      }
      
      if (length(all_cv_cor) > 0) {
        results <- rbind(results, data.frame(
          Trait = trait,
          Environment = env,
          CV_PredictionAccuracy = all_cv_cor,
          N_Repeats = length(all_cv_cor)
        ))
      } else {
        cat("    No valid CV predictions. Skipping result logging.\n")
      }
    }
  }
  
  return(results)
}
```






**Prepare Similarity matrices and pheno data for the MM function**
```{r}
load("data/ag_BLUE_spatial.RData")
preds_ag<- preds

#prep pred mat for combined and sel VIs
load("data/MT_pred_mat.Rdata")

names(MT_sim_mat_pr)
MT_sim_mat_pr_comb <- MT_sim_mat_pr[6:10]
names(MT_sim_mat_pr_comb)<- c("MCG25","HELF24", "MCG23","SNY22", "KET21")

MT_sim_mat_pr_sel <- MT_sim_mat_pr[c(4,1,3,5,2)]
names(MT_sim_mat_pr_sel)<- c("MCG25","HELF24", "MCG23","SNY22", "KET21")

#prep raw mat for combined and sel VIs
load("data/MT_raw_mat.Rdata")

names(MT_sim_mat_raw)

MT_sim_mat_raw_comb <- MT_sim_mat_raw[6:10]
names(MT_sim_mat_raw_comb)<- c("MCG25","HELF24", "MCG23","SNY22", "KET21")

MT_sim_mat_raw_sel <- MT_sim_mat_raw[c(4,1,3,5,2)]
names(MT_sim_mat_raw_sel)<-c("MCG25","HELF24", "MCG23","SNY22", "KET21")

```




**PP within env selected VIs for agronomic traits with PREDS comb**

```{r}
tr<- unique(preds$VI)[-2:-3]
preds_ag<- preds_ag %>% filter(VI %in% tr)

PP_MM_results_ag_pred_comb <- run_blup_cv(preds_ag, MT_sim_mat_pr_comb)
save(PP_MM_results_ag_pred_comb, file = "PP_MM_results_ag_pred_comb.Rdata")
```




#**PP within env selected VIs for agronomic traits with PREDS selected**

```{r}
PP_MM_results_ag_pred_sel <- run_blup_cv(preds_ag, MT_sim_mat_pr_sel)

save(PP_MM_results_ag_pred_sel, file = "PP_MM_results_ag_pred_sel.Rdata")
```




#**PP within env VIs for agronomic traits with RAW comb**

```{r}
load("C:/Users/Siim Sepp/NY_WMB/data/data_ag.Rdata")

tr<- colnames(data_ag)[c(7,8)]

data_ag<- data_ag[,c(-2,-4,-9,-10)] %>% pivot_longer(values_to = "Predicted", names_to = "VI", cols = tr)

PP_MM_results_ag_raw_comb <- run_blup_cv(data_ag, MT_sim_mat_raw_comb)

save(PP_MM_results_ag_raw_comb, file = "PP_MM_results_ag_raw_comb.Rdata")
```




**PP within env VIs for agronomic traits with RAW selected**

```{r}
PP_MM_results_ag_raw_sel <- run_blup_cv(data_ag, MT_sim_mat_raw_sel)
save(PP_MM_results_ag_raw_sel, file = "PP_MM_results_ag_raw_sel.Rdata")
```




**PP within env  VIs for MQ traits with RAW comb**

```{r}
load("C:/Users/Siim Sepp/NY_WMB/data/data_mq.Rdata")

tr<- colnames(data_mq)[4:15]
data_mq<- data_mq %>% pivot_longer(values_to = "Predicted", names_to = "VI", cols = tr)

PP_MM_results_MQ_raw_comb <- run_blup_cv(data_mq, MT_sim_mat_raw_comb)
save(PP_MM_results_MQ_raw_comb, file = "PP_MM_results_MQ_raw_comb.Rdata")
```




**PP within env  VIs for MQ traits with RAW sel**

```{r}
PP_MM_results_MQ_raw_sel <- run_blup_cv(data_mq, MT_sim_mat_raw_sel)
save(PP_MM_results_MQ_raw_sel, file = "PP_MM_results_MQ_raw_sel.Rdata")


#PP_MM_results_MQ_pred_sel <- run_blup_cv(preds, MT_sim_mat_pr_sel)
#save(PP_MM_results_MQ_pred_sel, file = "PP_MM_results_MQ_pred_sel.Rdata")

#PP_MM_results_MQ_pred_comb <- run_blup_cv(preds, MT_sim_mat_pr_comb)
#save(PP_MM_results_MQ_pred_comb, file = "PP_MM_results_MQ_pred_comb.Rdata")
```




# **5. Phenomic prediction of MQ and agronomic traits with phenomic relationship kernel CV2**



**Write mixed models function with phenomic kernel**

```{r}
run_blup_cv2_phenoK_globalK <- function(preds, MT_sim_mat,
                                       n_repeats = 50, folds = 5,
                                       mask_frac = 1,          # 1 = heavy, <1 = light sparse testing
                                       trace = FALSE,
                                       globalK = NULL,          # optionally pass a precomputed global kernel
                                       globalK_envs = NULL,     # envs to use when building global K (default = all in MT_sim_mat)
                                       min_gids = 10) {

  # helpers
  safe_cor <- function(x, y) {
    ok <- stats::complete.cases(x, y)
    if (sum(ok) < 2) return(NA_real_)
    suppressWarnings(stats::cor(x[ok], y[ok]))
  }

  # standardize ID column name
  df0 <- preds
  if ("plot_id" %in% names(df0)) {
    df0 <- dplyr::rename(df0, plotID = plot_id)
  } else if ("GID" %in% names(df0)) {
    df0 <- dplyr::rename(df0, plotID = GID)
  }
  if (!("plotID" %in% names(df0))) stop("preds must contain plot_id, GID, or plotID.")
  if (!all(c("VI", "Env", "Predicted") %in% names(df0))) stop("preds must contain VI, Env, Predicted.")

  # -------------------------------------------------------------------
  # Build ONE GLOBAL kernel K (once), to be reused for all traits
  # -------------------------------------------------------------------
  if (is.null(globalK)) {

    # which env kernels to use to build global K?
    if (is.null(globalK_envs)) globalK_envs <- names(MT_sim_mat)
    globalK_envs <- intersect(globalK_envs, names(MT_sim_mat))
    if (length(globalK_envs) < 1) stop("No environments available to build global K from MT_sim_mat.")

    K_list_all <- MT_sim_mat[globalK_envs]

    # global ID set = intersection across chosen env kernels (most conservative & safest)
    global_gids <- Reduce(intersect, lapply(K_list_all, rownames))
    if (length(global_gids) < min_gids) {
      stop("Too few common GIDs across env kernels to form a global kernel. Consider fewer envs or ensure consistent IDs.")
    }
    global_gids <- sort(global_gids)

    # average env-specific kernels -> global K
    Ksum <- matrix(0, nrow = length(global_gids), ncol = length(global_gids),
                   dimnames = list(global_gids, global_gids))

    for (e in globalK_envs) {
      Ke <- K_list_all[[e]][global_gids, global_gids, drop = FALSE]
      Ksum <- Ksum + Ke
    }
    globalK <- Ksum / length(globalK_envs)

  } else {
    # user supplied a global kernel
    if (is.null(rownames(globalK)) || is.null(colnames(globalK))) stop("globalK must have rownames and colnames.")
    if (!identical(sort(rownames(globalK)), sort(colnames(globalK)))) stop("globalK row/col names must match.")
    global_gids <- sort(rownames(globalK))
    globalK <- globalK[global_gids, global_gids, drop = FALSE]
  }

  traits <- unique(df0$VI)
  results <- data.frame()
  failures <- list()

  for (trait in traits) {
    cat("\nProcessing trait:", trait, "\n")

    df_trait <- df0 %>% dplyr::filter(VI == trait)

    # restrict to envs present in preds AND also present in MT_sim_mat (optional but recommended)
    envs <- intersect(unique(df_trait$Env), names(MT_sim_mat))
    if (length(envs) < 2) {
      cat("  <2 environments with kernels available in MT_sim_mat for this trait. Skipping.\n")
      next
    }

    # filter phenotype data to GLOBAL gids (so the same K applies)
    df_trait <- df_trait %>%
      dplyr::filter(Env %in% envs) %>%
      dplyr::mutate(plotID = as.character(plotID)) %>%
      dplyr::filter(plotID %in% global_gids)

    gids_in_df <- sort(unique(df_trait$plotID))
    if (length(gids_in_df) < min_gids) {
      cat("  Too few GIDs after filtering to global kernel. Skipping.\n")
      next
    }

    # subset K to the gids actually present for this trait (still a slice of the global kernel)
    K <- globalK[gids_in_df, gids_in_df, drop = FALSE]

    # factor levels
    df_trait$plotID <- factor(as.character(df_trait$plotID), levels = gids_in_df)
    df_trait$Env    <- factor(df_trait$Env, levels = envs)

    has_spatial <- all(c("Row", "Column") %in% names(df_trait))

    # ---- CV2: folds assigned at genotype level, masking within env only ----
    gids <- levels(df_trait$plotID)

    for (rep in seq_len(n_repeats)) {
      set.seed(100 + rep)

      fold_id <- sample(rep(seq_len(folds), length.out = length(gids)))
      names(fold_id) <- gids

      for (f in seq_len(folds)) {
        test_gids_all <- names(fold_id)[fold_id == f]

        train_data <- df_trait

        for (env in envs) {
          env_gids <- unique(as.character(df_trait$plotID[df_trait$Env == env]))
          test_gids <- intersect(test_gids_all, env_gids)
          if (length(test_gids) == 0) next

          # light sparse testing: only mask a fraction of these
          if (!is.null(mask_frac) && mask_frac < 1) {
            n_take <- max(1, floor(length(test_gids) * mask_frac))
            test_gids <- sample(test_gids, n_take)
          }

          idx_mask <- train_data$Env == env & as.character(train_data$plotID) %in% test_gids
          train_data$Predicted[idx_mask] <- NA
        }

        rand_formula <- if (has_spatial) {
          ~ vm(plotID, K) + ar1(Row):ar1(Column)
        } else {
          ~ vm(plotID, K)
        }

        mod_cv <- tryCatch({
          asreml(
            fixed     = Predicted ~ Env,
            random    = rand_formula,
            residual  = ~ idv(units),
            data      = train_data,
            na.action = list(x = "include", y = "include"),
            trace     = trace,
            ai.sing   = TRUE
          )
        }, error = function(e) e)

        if (inherits(mod_cv, "error")) {
          failures[[length(failures) + 1]] <- data.frame(
            Trait = trait, RepRun = rep, Fold = f, Stage = "cv_model",
            Error = mod_cv$message
          )
          next
        }

        pred <- tryCatch(predict(mod_cv, classify = "plotID")$pvals, error = function(e) e)
        if (inherits(pred, "error") || is.null(pred)) {
          failures[[length(failures) + 1]] <- data.frame(
            Trait = trait, RepRun = rep, Fold = f, Stage = "cv_predict",
            Error = if (inherits(pred, "error")) pred$message else "NULL preds"
          )
          next
        }
        colnames(pred)[colnames(pred) == "predicted.value"] <- "u_hat"

        # Evaluate ONLY on masked cells
        masked_idx <- is.na(train_data$Predicted) & !is.na(df_trait$Predicted)
        test_rows <- df_trait[masked_idx, c("plotID", "Env", "Predicted")]
        if (nrow(test_rows) < 2) next

        out <- merge(test_rows, pred[, c("plotID", "u_hat")], by = "plotID", all.x = TRUE)

        results <- rbind(results, data.frame(
          Trait  = trait,
          RepRun = rep,
          Fold   = f,
          n_test = nrow(out),
          Cor    = safe_cor(out$u_hat, out$Predicted),
          Sparse = ifelse(mask_frac >= 1, "heavy", paste0("light_", mask_frac))
        ))
      }
    }
  }

  failures_df <- if (length(failures)) do.call(rbind, failures) else data.frame()
  list(results = results,
       failures = failures_df,
       globalK = globalK,
       global_gids = global_gids)
}

```



#**Agronomic pred comb f1**
```{r}
cv2_global_f1 <- run_blup_cv2_phenoK_globalK(
  preds_ag, MT_sim_mat_pr_comb,
  n_repeats = 50, folds = 5,
  mask_frac = 1 )

boxplot(cv2_global_f1$results$Cor)
PP_MM_results_ag_pred_comb_CV2_f1 <- cv2_global_f1
save(PP_MM_results_ag_pred_comb_CV2_f1, file = "PP_MM_results_ag_pred_comb_CV2_f1.Rdata")
```


#**Agronomic pred comb f0.5**
```{r}
cv2_global_f05 <- run_blup_cv2_phenoK_globalK(
  preds_ag, MT_sim_mat_pr_comb,
  n_repeats = 50, folds = 5,
  mask_frac = 0.5 )

PP_MM_results_ag_pred_comb_CV2_f05 <- cv2_global_f05
save(PP_MM_results_ag_pred_comb_CV2_f05, file = "PP_MM_results_ag_pred_comb_CV2_f05.Rdata")
```


#**Agronomic pred sel f1**
```{r}
cv2_global_f1 <- run_blup_cv2_phenoK_globalK(
  preds_ag, MT_sim_mat_pr_sel,
  n_repeats = 50, folds = 5,
  mask_frac = 1 )

PP_MM_results_ag_pred_sel_CV2_f1 <- cv2_global_f1
save(PP_MM_results_ag_pred_sel_CV2_f1, file = "PP_MM_results_ag_pred_sel_CV2_f1.Rdata")
```


#**Agronomic pred sel f0.5**
```{r}
cv2_global_f05 <- run_blup_cv2_phenoK_globalK(
  preds_ag, MT_sim_mat_pr_sel,
  n_repeats = 50, folds = 5,
  mask_frac = 0.5 )

PP_MM_results_ag_pred_sel_CV2_f05 <- cv2_global_f05
save(PP_MM_results_ag_pred_sel_CV2_f05, file = "PP_MM_results_ag_pred_sel_CV2_f05.Rdata")
```




# **6. Ensemble models: stacking and averaging across base learners**


**Load raw prediction results**

```{r}
analysis_path <- if (.Platform$OS.type == "windows") {
  "C:/Users/Siim Sepp/NY_WMBCL/analysis/"
} else {
  "~/Documents/GitHub/NY_WMBCL/analysis/"
}

load(file.path(analysis_path, "PP_LM_results_ag_raw.Rdata"))
load(file.path(analysis_path, "PP_LM_results_mq_raw.Rdata"))
```




**Ensemble builder function**

Builds three ensembles from the per-cycle held-out predictions stored in any `run_pheno_models_long` result:
- **Mean**: unweighted average of all base model predictions
- **Weighted**: accuracy-weighted average (weights = mean r across cycles, zero-floored)
- **Stacked**: ridge regression meta-learner trained on the first half of cycles, evaluated on the second half

```{r}
build_ensemble_accuracy <- function(pred_df, acc_df) {

  # Align test-set rows across models via obs_id within each (Env, Cycle, Trait, PredSet, Model)
  pred_wide <- pred_df %>%
    group_by(Environment, Cycle, Trait, Predictor_Set, Model) %>%
    mutate(obs_id = row_number()) %>%
    ungroup() %>%
    pivot_wider(
      id_cols    = c(Environment, Cycle, Trait, Predictor_Set, obs_id, True),
      names_from  = Model,
      values_from = Predicted
    )

  model_cols <- setdiff(names(pred_wide),
                        c("Environment", "Cycle", "Trait", "Predictor_Set", "obs_id", "True"))
  model_cols <- model_cols[sapply(pred_wide[model_cols], is.numeric)]

  if (length(model_cols) < 2) {
    warning("Need >= 2 models for ensemble; returning NULL.")
    return(NULL)
  }

  # Weights proportional to mean accuracy (zero-floored, normalised)
  w_tbl <- acc_df %>%
    group_by(Model) %>%
    summarise(w = pmax(mean(Accuracy, na.rm = TRUE), 0), .groups = "drop") %>%
    filter(Model %in% model_cols)

  weight_vec <- setNames(w_tbl$w, w_tbl$Model)[model_cols]
  weight_vec[is.na(weight_vec)] <- 0
  total_w <- sum(weight_vec)
  if (total_w == 0) weight_vec[] <- 1 / length(model_cols) else weight_vec <- weight_vec / total_w

  # NA-safe weighted row mean
  weighted_row_mean <- function(mat, wv) {
    apply(mat, 1, function(x) {
      ok <- !is.na(x)
      if (!any(ok)) return(NA_real_)
      sum(x[ok] * wv[ok]) / sum(wv[ok])
    })
  }

  pred_mat <- as.matrix(pred_wide[, model_cols])
  pred_wide$Ensemble_Mean     <- rowMeans(pred_mat, na.rm = TRUE)
  pred_wide$Ensemble_Weighted <- weighted_row_mean(pred_mat, weight_vec)

  # Stacked ensemble: ridge meta-learner trained on first half of cycles
  all_cycles         <- sort(unique(pred_wide$Cycle))
  split_idx          <- ceiling(length(all_cycles) / 2)
  stacker_train_cyc  <- all_cycles[seq_len(split_idx)]
  stacker_test_cyc   <- all_cycles[(split_idx + 1):length(all_cycles)]

  pred_wide$Ensemble_Stacked <- NA_real_

  groups <- pred_wide %>%
    distinct(Trait, Predictor_Set, Environment) %>%
    unite("grp_key", Trait, Predictor_Set, Environment, sep = "|", remove = FALSE)

  for (i in seq_len(nrow(groups))) {
    g <- groups[i, ]
    idx_all   <- pred_wide$Trait == g$Trait &
                 pred_wide$Predictor_Set == g$Predictor_Set &
                 pred_wide$Environment == g$Environment
    idx_train <- idx_all & pred_wide$Cycle %in% stacker_train_cyc
    idx_test  <- idx_all & pred_wide$Cycle %in% stacker_test_cyc

    df_tr <- pred_wide[idx_train, c(model_cols, "True")] %>% tidyr::drop_na()
    df_te <- pred_wide[idx_test,  model_cols, drop = FALSE]

    if (nrow(df_tr) < 10 || nrow(df_te) < 2) next

    X_tr       <- as.matrix(df_tr[, model_cols])
    y_tr       <- df_tr$True
    complete_te <- complete.cases(df_te)
    X_te       <- as.matrix(df_te[complete_te, , drop = FALSE])

    meta <- tryCatch(
      glmnet::cv.glmnet(X_tr, y_tr, alpha = 0, nfolds = 5),
      error = function(e) NULL
    )

    if (!is.null(meta)) {
      stacked_preds             <- predict(meta, newx = X_te, s = "lambda.min")[, 1]
      test_row_idx              <- which(idx_test)[complete_te]
      pred_wide$Ensemble_Stacked[test_row_idx] <- stacked_preds
    }
  }

  # Compute r per (Env, Cycle, Trait, PredSet) for each ensemble type
  ensemble_cols <- c("Ensemble_Mean", "Ensemble_Weighted", "Ensemble_Stacked")

  acc_ens <- pred_wide %>%
    group_by(Environment, Cycle, Trait, Predictor_Set) %>%
    summarise(
      across(
        all_of(ensemble_cols),
        ~ tryCatch(cor(.x, True, use = "complete.obs"), error = function(e) NA_real_)
      ),
      .groups = "drop"
    ) %>%
    pivot_longer(all_of(ensemble_cols), names_to = "Model", values_to = "Accuracy")

  acc_ens
}
```




**Run ensembles on agronomic and malt quality raw results**

```{r}
ensemble_acc_ag <- build_ensemble_accuracy(
  PP_LM_results_ag_raw$predictions,
  PP_LM_results_ag_raw$accuracy
)

ensemble_acc_mq <- build_ensemble_accuracy(
  PP_LM_results_mq_raw$predictions,
  PP_LM_results_mq_raw$accuracy
)

# Combine base-model and ensemble accuracies into a single tidy frame
all_acc <- bind_rows(
  mutate(PP_LM_results_ag_raw$accuracy, dataset = "Agronomic"),
  mutate(ensemble_acc_ag,               dataset = "Agronomic"),
  mutate(PP_LM_results_mq_raw$accuracy, dataset = "Malt Quality"),
  mutate(ensemble_acc_mq,               dataset = "Malt Quality")
) %>%
  mutate(is_ensemble = grepl("^Ensemble", Model))

save(ensemble_acc_ag, ensemble_acc_mq, all_acc,
     file = file.path(analysis_path, "ensemble_results.Rdata"))
```




**Plot 1 – Accuracy boxplot: base models vs ensemble methods, faceted by trait**

```{r}
model_order <- all_acc %>%
  group_by(Model) %>%
  summarise(med = median(Accuracy, na.rm = TRUE), .groups = "drop") %>%
  arrange(med) %>%
  pull(Model)

p1 <- all_acc %>%
  filter(!is.na(Accuracy)) %>%
  mutate(Model = factor(Model, levels = model_order)) %>%
  ggplot(aes(x = Model, y = Accuracy, fill = is_ensemble)) +
  geom_boxplot(outlier.size = 0.4, linewidth = 0.35) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_fill_manual(
    values = c("FALSE" = "#5B9BD5", "TRUE" = "#ED7D31"),
    labels = c("Base model", "Ensemble")
  ) +
  facet_wrap(~ Trait, scales = "free_x", ncol = 3) +
  coord_flip() +
  labs(
    x = NULL, y = "Prediction accuracy (r)",
    fill  = "Model type",
    title = "Prediction accuracy: base models vs ensemble methods"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.text      = element_text(face = "bold"),
    legend.position = "bottom",
    axis.text.y     = element_text(size = 8)
  )

print(p1)
```




**Plot 2 – Mean accuracy heatmap: trait × model**

```{r}
base_model_levels <- c("Linear Regression", "Ridge", "ElasticNet",
                        "Lasso", "PCR", "Random Forest")
ens_model_levels  <- c("Ensemble_Mean", "Ensemble_Weighted", "Ensemble_Stacked")

mean_acc <- all_acc %>%
  group_by(dataset, Trait, Model) %>%
  summarise(mean_r = mean(Accuracy, na.rm = TRUE), .groups = "drop") %>%
  mutate(Model = factor(Model, levels = c(base_model_levels, ens_model_levels)))

p2 <- ggplot(mean_acc, aes(x = Model, y = Trait, fill = mean_r)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.2f", mean_r)), size = 2.8, colour = "black") +
  scale_fill_gradient2(
    low      = "#d73027",
    mid      = "#ffffbf",
    high     = "#1a9641",
    midpoint = 0,
    limits   = c(-0.3, 1),
    name     = "Mean r"
  ) +
  facet_wrap(~ dataset, scales = "free_y") +
  labs(
    x     = NULL, y = NULL,
    title = "Mean prediction accuracy: traits × models"
  ) +
  theme_bw(base_size = 11) +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1),
    strip.text   = element_text(face = "bold"),
    panel.grid   = element_blank()
  )

print(p2)
```




**Plot 3 – Ensemble gain over best individual base model, per trait**

```{r}
ensemble_gain <- all_acc %>%
  group_by(dataset, Trait, Model) %>%
  summarise(mean_r = mean(Accuracy, na.rm = TRUE), .groups = "drop") %>%
  mutate(is_ensemble = grepl("^Ensemble", Model)) %>%
  group_by(dataset, Trait) %>%
  summarise(
    best_base      = max(mean_r[!is_ensemble], na.rm = TRUE),
    best_ens       = max(mean_r[is_ensemble],  na.rm = TRUE),
    best_ens_model = Model[is_ensemble][which.max(mean_r[is_ensemble])],
    gain           = best_ens - best_base,
    .groups = "drop"
  ) %>%
  arrange(gain)

p3 <- ggplot(ensemble_gain,
             aes(x = reorder(Trait, gain), y = gain, fill = gain > 0)) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label  = sprintf("%+.3f", gain),
        hjust  = ifelse(gain >= 0, -0.1, 1.1)),
    size = 3
  ) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(
    values = c("TRUE" = "#2ca02c", "FALSE" = "#d62728"),
    labels = c("Loss", "Gain"),
    name   = NULL
  ) +
  facet_wrap(~ dataset, scales = "free_y") +
  coord_flip() +
  labs(
    x     = NULL,
    y     = "Accuracy gain (best ensemble − best base model)",
    title = "Ensemble gain over best individual model"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.text      = element_text(face = "bold"),
    legend.position = "bottom"
  )

print(p3)
```




**Plot 4 – Per-predictor-set comparison: mean accuracy by model across traits**

```{r}
top_vi <- c("NDVI", "PSRI")

p4 <- all_acc %>%
  filter(!is.na(Accuracy), Predictor_Set %in% top_vi) %>%
  group_by(dataset, Trait, Model, Predictor_Set, is_ensemble) %>%
  summarise(mean_r = mean(Accuracy, na.rm = TRUE), .groups = "drop") %>%
  mutate(Model = factor(Model, levels = rev(c(base_model_levels, ens_model_levels)))) %>%
  ggplot(aes(x = Model, y = mean_r,
             colour = is_ensemble, shape = Predictor_Set)) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  scale_colour_manual(
    values = c("FALSE" = "#4E79A7", "TRUE" = "#E15759"),
    labels = c("Base model", "Ensemble")
  ) +
  facet_wrap(~ Trait + dataset, scales = "free_x", ncol = 4) +
  coord_flip() +
  labs(
    x      = NULL, y = "Mean accuracy (r)",
    colour = "Model type", shape = "Predictor set",
    title  = "Mean accuracy by model and VI predictor set"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.text      = element_text(face = "bold", size = 8),
    legend.position = "bottom",
    axis.text.y     = element_text(size = 8)
  )

print(p4)
```

