---
title: "8. GP_ag_mq"
author: "sss322"
date: "2025-03-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```





## **0. Importing the dependencies**

```{r}
library(dplyr)
library(asreml)
library(ggplot2)
library(ASRgenomics)
library(readxl)
library(MASS)
library(tidyr)
library(dplyr)
library(caret)
library(asreml)
```





## **1. Importing the phenotypic data**

Importing the raw agronomic, mq and spectral 
```{r}
base_path <- if (.Platform$OS.type == "windows") {
  "C:/Users/Siim Sepp/NY_WMB"
} else {
  "~/Documents/GitHub/NY_WMB"
}

load(file.path(base_path, "data/data_ag.Rdata"))
load(file.path(base_path, "data/data_mq.Rdata"))
load(file.path(base_path, "data//WMB_master_geno.Rdata"))

data_ag<- data_ag %>% na.omit()
data_mq<- data_mq %>% na.omit()
```





## **2. Importing the genotypic data**

```{r}
ASRgenomics::kinship.diagnostics(A1)
ASRgenomics::kinship.heatmap(A1)
ASRgenomics::kinship.pca(A1)
str(data_ag)
```



```{r}
run_gblup_cv <- function(data, A1, traits, envs = NULL, reps = 50, folds = 5,
                         eig.tol = 1e-3) {
  if (is.null(envs)) envs <- if (is.factor(data$Env)) levels(data$Env) else sort(unique(data$Env))

  res_all  <- data.frame()
  fail_log <- list()

  safe_cor <- function(x, y) {
    ok <- stats::complete.cases(x, y)
    if (sum(ok) < 2) return(NA_real_)
    suppressWarnings(stats::cor(x[ok], y[ok]))
  }

  for (rep_run in seq_len(reps)) {
    set.seed(1000 + rep_run)
    message("\nStarting repetition #", rep_run)

    for (tvar in traits) {
      for (env in envs) {
        message("\n   Processing Trait = ", tvar, ", Environment = ", env)

        data_sub <- subset(data, Env == env)
        if (nrow(data_sub) < folds) {
          fail_log[[length(fail_log) + 1]] <- data.frame(
            RepRun = rep_run, Trait = tvar, Env = env, Fold = NA,
            Stage = "data_check", Error = "Too few rows for CV"
          )
          next
        }

        # align to G
        keep_gids <- intersect(unique(as.character(data_sub$GID)), rownames(A1))
        data_sub  <- data_sub[data_sub$GID %in% keep_gids, , drop = FALSE]
        if (length(keep_gids) < folds + 1) {
          fail_log[[length(fail_log) + 1]] <- data.frame(
            RepRun = rep_run, Trait = tvar, Env = env, Fold = NA,
            Stage = "id_intersect", Error = "Insufficient common GIDs with A1"
          )
          next
        }
        A1_sub <- A1[keep_gids, keep_gids, drop = FALSE]

        # bend + inverse
        Ghat.bend   <- G.tuneup(G = A1_sub, bend = TRUE, eig.tol = eig.tol)$Gb
        Ginv.sparse <- G.inverse(G = Ghat.bend, sparse = TRUE)$Ginv

        # factors
        data_sub$GID <- factor(as.character(data_sub$GID), levels = keep_gids)

        # detect spatial columns
        has_spatial <- all(c("Row", "Column") %in% colnames(data_sub))

        # CV split at the genotype level
        gids <- unique(as.character(data_sub$GID))
        fold_gids <- sample(rep(seq_len(folds), length.out = length(gids)))
        names(fold_gids) <- gids

        # fixed model
        formula_fixed <- as.formula(paste(tvar, "~ 1"))

        for (f in seq_len(folds)) {
          test_gids <- names(fold_gids)[fold_gids == f]
          train_data <- data_sub

          # mask ALL records of the test GIDs
          train_data[[tvar]][train_data$GID %in% test_gids] <- NA

          # model: AR1xAR1 in RANDOM part if present; otherwise just vm()
          rand_formula <- if (has_spatial) {
            ~ vm(GID, Ginv.sparse) + ar1(Row):ar1(Column)
          } else {
            ~ vm(GID, Ginv.sparse)
          }

          mod_cv <- tryCatch({
            asreml(
              fixed     = formula_fixed,
              random    = rand_formula,
              residual  = ~ idv(units),
              data      = train_data,
              na.action = list(x = "include", y = "include"),
              ai.sing   = TRUE
            )
          }, error = function(e) e)

          if (inherits(mod_cv, "error")) {
            fail_log[[length(fail_log) + 1]] <- data.frame(
              RepRun = rep_run, Trait = tvar, Env = env, Fold = f,
              Stage = "cv_model", Error = mod_cv$message
            )
            next
          }

          # EBVs for all GIDs
          p_gid <- tryCatch(predict(mod_cv, classify = "GID")$pvals, error = function(e) e)
          if (inherits(p_gid, "error") || is.null(p_gid)) {
            fail_log[[length(fail_log) + 1]] <- data.frame(
              RepRun = rep_run, Trait = tvar, Env = env, Fold = f,
              Stage = "cv_predict", Error = if (inherits(p_gid, "error")) p_gid$message else "NULL preds"
            )
            next
          }
          names(p_gid)[names(p_gid) == "predicted.value"] <- "u_hat"

          # observed phenotypes for TEST GIDs only (from the original, unmasked data_sub)
          y_test <- data_sub[data_sub$GID %in% test_gids, c("GID", tvar)]
          colnames(y_test)[2] <- "y_obs"

          out <- merge(y_test, p_gid[, c("GID", "u_hat")], by = "GID", all.x = TRUE)

          res_all <- rbind(res_all, data.frame(
            RepRun = rep_run,
            Trait  = tvar,
            Env    = env,
            Fold   = f,
            n_test = length(test_gids),
            Cor    = safe_cor(out$u_hat, out$y_obs)   # <-- ONLY cor(u_hat, y_obs)
          ))
        }
      }
    }
  }

  fail_df <- if (length(fail_log)) do.call(rbind, fail_log) else data.frame()
  list(results = res_all, failures = fail_df)
}
```





## **3. Genomic prediction of agronomic traits within, year**

```{r}
traits<- colnames(data_ag)[7:8]
GP_MM_results_ag_raw <- run_gblup_cv(data = data_ag, A1 = A1, traits = traits, reps = 50)
save(GP_MM_results_ag_raw, file= "GP_MM_results_ag_raw.Rdata")
```



## **4. Genomic prediction of agronomic BLUE within, year**

```{r}
#load("C:/Users/Siim Sepp/NY_WMB/data/ag_BLUE_spatial.RData")
preds<- preds %>% pivot_wider(names_from= "VI", values_from= "Predicted")

traits<- colnames(preds)[4:5]
envs<- unique(preds$Env)


GP_MM_results_ag_pred <- run_gblup_cv(data = preds, A1 = A1, traits = traits, envs = envs , reps = 50)
save(GP_MM_results_ag_pred, file= "GP_MM_results_ag_pred.Rdata")
```


## **5. Genomic prediction of MQ BLUP traits within, year**
#```{r}
load("C:/Users/Siim Sepp/NY_WMB/data/mq_BLUP_spatial.RData")

mqtraits <- unique(preds$VI)
preds<- preds[,-5] %>% pivot_wider(names_from = "VI", values_from = "Predicted")
envs<- unique(preds$Env)

GP_MM_results_mq_preds <- run_gblup_cv(data = preds, A1 = A1,envs = envs, traits = mqtraits, reps = 50)
save(GP_MM_results_mq_preds, file= "GP_MM_results_mq_preds.Rdata")
#```


## **6. Genomic prediction of MQ raw traits within, year**

```{r}
mqtraits <- colnames(data_mq)[4:15]
envs<- unique(data_mq$Env)

GP_MM_results_mq_raw <- run_gblup_cv(data = data_mq, A1 = A1,envs = envs, traits = mqtraits, reps = 50)
save(GP_MM_results_mq_raw, file= "GP_MM_results_mq_raw.Rdata")
table(GP_MM_results_mq_raw$results$Env)
```





## **7. Genomic prediction of MQ and agronomic traitswith CV2**




```{r}
run_gblup_cv_cv2 <- function(data, A1, traits, envs = NULL,
                             reps = 50, folds = 5,
                             eig.tol = 1e-3,
                             mask_frac = NULL) {
  if (is.null(envs)) envs <- if (is.factor(data$Env)) levels(data$Env) else sort(unique(data$Env))

  res_all  <- data.frame()
  fail_log <- list()

  safe_cor <- function(x, y) {
    ok <- stats::complete.cases(x, y)
    if (sum(ok) < 2) return(NA_real_)
    suppressWarnings(stats::cor(x[ok], y[ok]))
  }

  for (rep_run in seq_len(reps)) {
    set.seed(2000 + rep_run)
    message("\nStarting repetition #", rep_run)

    for (tvar in traits) {
      message("\n Trait = ", tvar)

      # creating a copy
      df_all <- data

      # align to G (globally)
      keep_gids <- intersect(unique(as.character(df_all$GID)), rownames(A1))
      df_all    <- df_all[df_all$GID %in% keep_gids, , drop = FALSE]
      if (length(keep_gids) < folds + 2) {
        fail_log[[length(fail_log) + 1]] <- data.frame(
          RepRun = rep_run, Trait = tvar, Env = NA, Fold = NA,
          Stage = "id_intersect", Error = "Insufficient common GIDs with A1"
        )
        next
      }

      A1_sub <- A1[keep_gids, keep_gids, drop = FALSE]

      # bend + inverse
      Ghat.bend   <- G.tuneup(G = A1_sub, bend = TRUE, eig.tol = eig.tol)$Gb
      Ginv.sparse <- G.inverse(G = Ghat.bend, sparse = TRUE)$Ginv

      # factors
      df_all$GID <- factor(as.character(df_all$GID), levels = keep_gids)

      # detect spatial columns (must exist in ALL envs you want AR1xAR1 for)
      has_spatial <- all(c("Row", "Column") %in% colnames(df_all))

      # ---------- CV2 masking scheme ----------
      # Create folds at the GENOTYPE level, but apply masking WITHIN EACH ENV:
      gids <- unique(as.character(df_all$GID))
      fold_gids <- sample(rep(seq_len(folds), length.out = length(gids)))
      names(fold_gids) <- gids

      # Optional: only mask a fraction of genotypes per env per fold (sparser masking)
      # If mask_frac is NULL, masks all genotypes assigned to that fold.
      # If mask_frac is in (0,1],subsamples the fraction of the fold's genotypes in each env.
      if (!is.null(mask_frac)) {
        if (!is.numeric(mask_frac) || mask_frac <= 0 || mask_frac > 1) {
          stop("mask_frac must be NULL or a number in (0,1].")
        }
      }

      # fixed model
      formula_fixed <- as.formula(paste(tvar, "~ 1"))

      for (env in envs) {
        message("\n   Environment = ", env)

        df_env <- df_all[df_all$Env == env, , drop = FALSE]
        if (nrow(df_env) < folds) {
          fail_log[[length(fail_log) + 1]] <- data.frame(
            RepRun = rep_run, Trait = tvar, Env = env, Fold = NA,
            Stage = "data_check", Error = "Too few rows in this env for CV"
          )
          next
        }

        for (f in seq_len(folds)) {
          # genotypes intended for masking in this env/fold
          fold_gid_set <- names(fold_gids)[fold_gids == f]

          # only keep those that actually occur in this environment
          env_gids <- unique(as.character(df_env$GID))
          test_gids <- intersect(fold_gid_set, env_gids)

          if (length(test_gids) < 2) {
            fail_log[[length(fail_log) + 1]] <- data.frame(
              RepRun = rep_run, Trait = tvar, Env = env, Fold = f,
              Stage = "fold_split", Error = "Too few test GIDs in this env/fold"
            )
            next
          }

          # optionally thin masking within this env/fold
          if (!is.null(mask_frac) && mask_frac < 1) {
            n_take <- max(1, floor(length(test_gids) * mask_frac))
            test_gids <- sample(test_gids, n_take)
          }

          train_data <- df_all

          # MASK ONLY THIS ENV'S RECORDS for the test_gids (CV2 sparse testing)
          idx_mask <- train_data$Env == env & as.character(train_data$GID) %in% test_gids
          train_data[[tvar]][idx_mask] <- NA

          # random part: keep your spatial structure if available
          rand_formula <- if (has_spatial) {
            ~ vm(GID, Ginv.sparse) + ar1(Row):ar1(Column)
          } else {
            ~ vm(GID, Ginv.sparse)
          }

          mod_cv <- tryCatch({
            asreml(
              fixed     = formula_fixed,
              random    = rand_formula,
              residual  = ~ idv(units),
              data      = train_data,
              na.action = list(x = "include", y = "include"),
              ai.sing   = TRUE
            )
          }, error = function(e) e)

          if (inherits(mod_cv, "error")) {
            fail_log[[length(fail_log) + 1]] <- data.frame(
              RepRun = rep_run, Trait = tvar, Env = env, Fold = f,
              Stage = "cv_model", Error = mod_cv$message
            )
            next
          }

          # EBVs for all GIDs
          p_gid <- tryCatch(predict(mod_cv, classify = "GID")$pvals, error = function(e) e)
          if (inherits(p_gid, "error") || is.null(p_gid)) {
            fail_log[[length(fail_log) + 1]] <- data.frame(
              RepRun = rep_run, Trait = tvar, Env = env, Fold = f,
              Stage = "cv_predict", Error = if (inherits(p_gid, "error")) p_gid$message else "NULL preds"
            )
            next
          }
          names(p_gid)[names(p_gid) == "predicted.value"] <- "u_hat"

          # observed phenotypes ONLY for the masked cells (env + test_gids) from the original (unmasked) df_all
          y_test <- df_all[df_all$Env == env & as.character(df_all$GID) %in% test_gids, c("GID", tvar)]
          colnames(y_test)[2] <- "y_obs"

          out <- merge(y_test, p_gid[, c("GID", "u_hat")], by = "GID", all.x = TRUE)

          res_all <- rbind(res_all, data.frame(
            RepRun = rep_run,
            Trait  = tvar,
            Env    = env,
            Fold   = f,
            n_test = length(unique(as.character(y_test$GID))),
            Cor    = safe_cor(out$u_hat, out$y_obs)  # cor(u_hat, y_obs) for masked cells
          ))
        }
      }
    }
  }

  fail_df <- if (length(fail_log)) do.call(rbind, fail_log) else data.frame()
  list(results = res_all, failures = fail_df)
}


```



```{r}
cv2_out_f1 <- run_gblup_cv_cv2(
  data  = data_ag,
  A1    = A1,
  traits = c("yield_kgha"),
  reps  = 50,
  folds = 5
)

head(cv2_out$results)
boxplot(cv2_out$results$Cor)
save(cv2_out_f1, file= "GP_ag_cv2_res_f1.Rdata")
```


```{r}
cv2_ag_out_f05 <- run_gblup_cv_cv2(
  data  = data_ag,
  A1    = A1,
  traits = c("yield_kgha"),
  reps  = 50,
  folds = 5,
   mask_frac = 0.5
)

save(cv2_ag_out_f05, file= "GP_ag_cv2_res_f05.Rdata")
```





## **8. Genomic prediction of MQ and agronomic traits within year, based on the UAV based predictons of these traits**

- Predict original BV
- Predict phenotype and then predict the BV and compare 

```{r}

run_uav_genomic_cv <- function(
    data, VImats, A1, traits, env_levels, VIs = NULL,
    n_reps = 50, k_folds = 5
) {
  
  data$Env <- as.factor(data$Env)
  if (is.null(VIs)) VIs <- names(VImats[[env_levels[1]]])
  
  res_final <- data.frame()

  for (rep_run in seq_len(n_reps)) {
    message("\n==== Starting repetition #", rep_run, " ====\n")

    ## A) UAV-BASED PHENOMIC PREDICTION
    uav_pred_list <- list()

    for (env in env_levels) {
      data_sub <- data %>% filter(Env == env)

      for (tvar in traits) {
        for (i in VIs) {
          predictor_data <- VImats[[env]][[i]]
          merged_df <- merge(data_sub, predictor_data, by = "plot_id") %>% na.omit()

          set.seed(rep_run)
          inTrain <- createDataPartition(merged_df[[tvar]], p = 0.8, list = FALSE)
          trainDF <- merged_df[inTrain, ]
          trainDF <- trainDF[, c(tvar, colnames(predictor_data))]

          ctrl <- trainControl(method = "cv", number = 5)
          trainDF$plot_id <- as.character(trainDF$plot_id)
          merged_df$plot_id <- as.character(merged_df$plot_id)
          

          # Prepare predictor set 
          train_features <- trainDF[,-2] 
          test_features  <- merged_df[,-2]
          # Train the model
          rf_model <- train(
              formula(paste(tvar, "~ .")),
                data = train_features,
                method = "rf",
                trControl = ctrl
              )
        
          # Predict
          UAV_pred <- predict(rf_model, newdata = test_features)


          tmp_out <- data.frame(
            Env = env, Trait = tvar, VI = i,
            GID = merged_df$GID,
            plot_id = merged_df$plot_id,
            UAV_pred = UAV_pred
            )
          uav_pred_list[[length(uav_pred_list) + 1]] <- tmp_out
        }
      }
    }
    uav_pred_df <- do.call(rbind, uav_pred_list)

    ## A0) TRUE BV ESTIMATION
    true_bv_list <- list()

    for (env in env_levels) {
      real_sub <- data %>% filter(Env == env)

      for (tvar in traits) {
        real_trait_sub <- real_sub[, c("GID", tvar)] %>% na.omit()
        real_trait_sub <- real_trait_sub[real_trait_sub$GID %in% rownames(A1), ]
        A1_sub <- A1[rownames(A1) %in% real_trait_sub$GID, rownames(A1) %in% real_trait_sub$GID]

        Ghat.bend <- G.tuneup(G = A1_sub, bend = TRUE, eig.tol = 1e-03)$Gb
        Ginv.sparse <- G.inverse(G = Ghat.bend, sparse = TRUE)$Ginv

        real_trait_sub$GID <- as.factor(real_trait_sub$GID)
        form_real <- as.formula(paste(tvar, "~ 1"))

        mod_full_real <- tryCatch({
          asreml(
            fixed = form_real,
            random = ~ vm(GID, Ginv.sparse),
            residual = ~ units,
            data = real_trait_sub,
            ai.sing = TRUE
          )
        }, error = function(e) NULL)

        if (!is.null(mod_full_real)) {
          bv_df <- predict(mod_full_real, classify = "GID")$pvals
          colnames(bv_df)[colnames(bv_df) == "predicted.value"] <- "BV_true"
          bv_df$Env <- env; bv_df$Trait <- tvar
          true_bv_list[[length(true_bv_list) + 1]] <- bv_df[, c("GID", "BV_true", "Env", "Trait")]
        }
      }
    }
    true_bv_df <- do.call(rbind, true_bv_list)

    ## B) GENOMIC PREDICTION FROM UAV-DERIVED PHENOTYPES
    for (env in env_levels) {
      for (tvar in traits) {
        for (i in VIs) {
          message("   [Rep=", rep_run, "] UAV-based -> Genomic: Env=", env, " Trait=", tvar, " VI=", i)

          sub_pred <- uav_pred_df %>% filter(Env == env, Trait == tvar, VI == i)
          sub_true <- true_bv_df %>% filter(Env == env, Trait == tvar)

          merged_pheno <- merge(sub_pred, sub_true, by = "GID")
          merged_pheno <- merged_pheno[merged_pheno$GID %in% rownames(A1), ]
          A1_sub <- A1[rownames(A1) %in% merged_pheno$GID, rownames(A1) %in% merged_pheno$GID]

          Ghat.bend <- G.tuneup(G = A1_sub, bend = TRUE, eig.tol = 1e-03)$Gb
          Ginv.sparse <- G.inverse(G = Ghat.bend, sparse = TRUE)$Ginv

          merged_pheno$Y <- merged_pheno$UAV_pred
          merged_pheno$GID <- as.factor(merged_pheno$GID)

          set.seed(rep_run)
          fold_ids <- sample(rep(1:k_folds, length.out = nrow(merged_pheno)))
          fold_list <- vector("list", length = k_folds)

          for (f in seq_len(k_folds)) {
            train_df <- merged_pheno
            train_df$Y[fold_ids == f] <- NA

            mod_cv <- tryCatch({
              asreml(
                fixed = Y ~ 1,
                random = ~ vm(GID, Ginv.sparse),
                residual = ~ units,
                data = train_df,
                na.action = list(x = "include", y = "include"),
                ai.sing = TRUE
              )
            }, error = function(e) NULL)

            pred_cv <- tryCatch({
              if (!is.null(mod_cv)) predict(mod_cv, classify = "GID")$pvals else NULL
            }, error = function(e) NULL)

            if (!is.null(pred_cv)) {
              colnames(pred_cv)[colnames(pred_cv) == "predicted.value"] <- "BV_cv"
              merged_test <- merge(pred_cv[, c("GID", "BV_cv")], merged_pheno[, c("GID", "BV_true")], by = "GID")
              test_gids <- train_df$GID[fold_ids == f]
              merged_test <- merged_test %>% filter(GID %in% test_gids)
              cor_val <- cor(merged_test$BV_cv, merged_test$BV_true, use = "complete.obs")
            } else {
              cor_val <- NA
            }

            fold_list[[f]] <- data.frame(
              Rep = rep_run, Env = env, Trait = tvar, VI = i, Fold = f, CorVal = cor_val
            )
          }
          res_final <- rbind(res_final, do.call(rbind, fold_list))
        }
      }
    }
  }

  return(res_final)
}

```


test MQ

```{r}
load("C:/Users/Siim Sepp/NY_WMB/data/VImat.Rdata")
mqtraits <- colnames(data_mq)[4:15]
envs<- unique(data_mq$Env)
VIs<- names(VImats$KET21)

GP_PP_MM_mq <- run_uav_genomic_cv(data = data_mq, VImats = VImats, A1 = A1, traits = mqtraits, env_levels = envs, VIs = NULL,
    n_reps = 5, k_folds = 5
)

save(GP_PP_MM_mq, file = "GP_PP_MM_results_mq.Rdata")
```


test ag

```{r}
agtraits <- colnames(data_ag)[7:10]
envs<- unique(data_ag$Env)
VIs<- names(VImats$KET21)

GP_PP_MM_ag <- run_uav_genomic_cv(data = data_ag, VImats = VImats, A1 = A1, traits = agtraits, env_levels = envs, VIs = NULL,
    n_reps = 50, k_folds = 5
)

save(GP_PP_MM_ag, file = "GP_PP_MM_results_ag.Rdata")

#boxplot(GP_PP_MM_ag$CorVal ~GP_PP_MM_ag$Trait)
#boxplot(GP_PP_MM_mq$CorVal ~GP_PP_MM_mq$Trait)
```

