---
title: "2. Spectral data Spatial BLUP"
output: html_document
author: "sss322"
date: "2024-12_06"
---


```{r setup, include=FALSE, warning=FALSE, eval=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# Introduction
This document provides an overview of the spatial BLUP wth the multi-spectral drone imagery data. The goal is to conduct the first BLUP stage with the vegetation indices and explore data distribution.



## **1. Importing Dependencies **
Loading dependencies
```{r, message=FALSE}
library(readr)
library(readxl)
library(dplyr)
library(MASS)
library(ggplot2)
library(tidyr)
library(ASRgenomics)
library(factoextra)
library(fda)
library(ASRtriala)
library(writexl)
library(openxlsx)
library(jsonlite)
library(data.table)
Sys.setlocale("LC_TIME", "C")
```


## **2. Setting the working directory and importing the data**
```{r, message=FALSE}
load("data/data_sp.Rdata")
load("~/Documents/GitHub/NY_WMB/data/data_sp.Rdata")
table(data_sp$Env, data_sp$Timepoint)
```

```{r, message=FALSE}
data<- data_sp %>% arrange(Timepoint)
data$GID<- as.factor(data$GID)
colnames(data_sp)
```




## **3. Generating genotype BLUP-s to eliminate spatial effects **
Model Selection Fit models based on statistical goodness-of-fit measures.
```{r}
#Setup the models to fit for each TP and VI
model.setup <- ASRtriala::model.setup
specs.single <- model.setup %>%
  filter(add.rep == FALSE) %>%
  filter(add.ibk == FALSE)
```

```{r, eval=FALSE}
# Create an empty list to store all models stats for all indices
all_indices_models_stats <- list()
all_models_stats <- list()

# List of timepoints to loop through
timepoints <- (unique(data$timepoint))
timepoints <- "2"


# List of indices to loop through
indices<- c("NDVI", "NDRE", "EVI", "GDVI", "GNDVI", "NRI", "SAVI", "ENDVI", "GIPVI")
indices<- c("NDVI")

# Get unique levels of Env
env_levels <- unique(data$Env)
env_levels <- "SNY22"

# Create an empty list to store models stats for each Env
all_env_models_stats <- list()

# Outer loop through each Env level
for (env in env_levels) {
  # Filter data for the current Env level
  env_data <- data[data$Env == env, ]
  
  # Create an empty list to store models stats for each index within this Env
  all_indices_models_stats <- list()
  
  # Get unique indices for the current Env
  #indices <- unique(env_data$index_column) # Replace 'index_column' with the column name for indices

  # Loop through each index
  for (index in indices) {
    # Create an empty list to store models stats for each timepoint for the current index
    all_models_stats <- list()
    
    # Get unique timepoints for the current index
    timepoints <- unique(env_data$timepoint)

    # Loop through each timepoint
    for (tp in timepoints) {
      st.data <- env_data[env_data$timepoint == tp, ]
      
      # Use the current index in the response variable
      models.stats <- select.single(
        data = st.data,
        gen = "GID",
        row = "Row",
        col = "Column",
        resp = index,
        type.gen = "random",
        data.model = specs.single,
        threshold = 3.5,
        pworkspace = "16gb",
        criteria = "logDopt"

      )
      
      # Store the goodness-of-fit statistics for the current timepoint and index
      all_models_stats[[tp]] <- models.stats$gof.stats
    }
    
    # Store the model stats for this index in the overall list
    all_indices_models_stats[[index]] <- all_models_stats
  }
  
  # Store the model stats for this Env in the overall list
  all_env_models_stats[[env]] <- all_indices_models_stats
  
  # Save the models stats for the current Env level to a separate file
  # save(all_indices_models_stats, file = paste0("model_stats_", env, ".Rdata"))
}

# Save the overall stats for all Envs to a single file
# save(all_env_models_stats, file = "all_env_models_stats.Rdata")

#all_env_models_stats$HELF24$NDVI

#save the results
save(all_indices_models_stats, file= "spec_mod_stats_n.Rdata")
 

```





## **4. Select the best-of-fit models**
```{r, eval=FALSE}
#load("blup1_fit_stats/mod_stats.Rdata")
load("data/mod_statsNY.Rdata")
all_models_stats<- mod_statsNY[[1]]
timepoints <- 1:6

all_models_stats$NDVI[[7]][1:14,]
model_selections_list <- list(
  ENDVI  = c( 13, 3, 1, 3, 6, 1, 1, 1),#
  GIPVI  = c( 1, 1, 1, 4, 1, 1, 1, 1),#
  SAVI   = c( 3, 3, 5, 2, 6, 7, 1, 1),#
  GDVI   = c( 3, 4, 4, 1, 5, 1, 1, 1),#
  GNDVI  = c( 3, 1, 1, 3, 2, 1, 1, 1),#
  EVI    = c( 1, 1, 6, 1, 1, 1, 1, 1),#
  NDVI   = c( 1, 5, 12, 1, 1, 1, 1, 1),#
  NDRE   = c( 11, 2, 2, 5, 5, 3, 1, 1),#
  PSRI   = c( 1, 1, 1, 1, 1, 1, 1, 1),#
  NRI    = c( 1, 1, 5, 3, 1, 1, 1, 1)#
)


# Initialize a data frame to store the results of selected models for each index
mod_stats_df <- data.frame()

# Loop through each vegetative index in the model selections list
for (index_name in names(model_selections_list)) {
  
  # Get the model selections for the current index
  model_selections <- model_selections_list[[index_name]]
  
  # Get the dataframe of models for the current index from all_models_stats
  index_models_df <- all_models_stats[[1]] # [[index_name]]
  
  # Loop through timepoints and model selections
  for (i in seq_along(timepoints)) {
    tp <- timepoints[i]
    model_index <- model_selections[i]
    
    # Check if the timepoint exists in the index_models_df
    if (is.data.frame(index_models_df[[tp]])) {
      # Filter the dataframe to get the model for the selected timepoint
      models_at_tp <- as.data.frame(index_models_df[tp])
      
      # Check if the model index is valid
      if (model_index > 0 && model_index <= nrow(models_at_tp)) {
        sel_model <- models_at_tp[model_index, ]  # Select the entire row
        sel_model$timepoint <- tp  # Add timepoint column to the selected model
        sel_model$VI <- index_name  # Add the index name
        
        # Append the selected model to the data frame
        mod_stats_df <- rbind(mod_stats_df, sel_model)
      } else {
        cat("Model index", model_index, "is out of range for timepoint", tp, "in index", index_name, "\n")
      }
    } else {
      cat("No models available for timepoint", tp, "in index", index_name, "\n")
    }
  }
}

# Display the structure of mod_stats_df
str(mod_stats_df)
table(mod_stats_df$VI, mod_stats_df$timepoint)
# Optionally, you can save the data frame for future use
save(all_env_models_stats, file = "mod_stats_fitNY.Rdata")
save(all_env_models_stats)

mod_stats21$Env<- "KET21"
mod_stats22$Env<- "SNY22"
mod_stats23$Env<- "MCG23"
mod_stats24$Env<- "HELF24"

mod_stats <- rbind(mod_stats21, mod_stats22, mod_stats23, mod_stats24)
save(mod_stats, file= "data/spec_mod_stats_sel.Rdata")
```




## **5. Fit the selected models**
```{r, eval=FALSE }
load("data/wmb_multi_alldata.Rdata")
load("data/spec_mod_stats.Rdata")
multi<- multi %>% filter(Env ==  "HELF24")
multi$GID<- as.factor(multi$GID)
multi<- multi%>% arrange(timepoint)

#################
all_selected_models_df<- mod_stats24
#Needs to be adapted for a loop over several VI-s. 
# Initialize an empty list to store all predictions
all_predictions <- list()


# Loop through each unique vegetative index
for (index in unique(all_selected_models_df$VI)) {
  
  # Filter the dataframe for the selected index
  sel_models_index <- filter(all_selected_models_df, VI == index)
  # Loop through each unique timepoint for the current index
  for (tp in unique(sel_models_index$timepoint)) {
    
    # Filter the selected models for the current timepoint and index
    sel.m <- filter(sel_models_index, timepoint == tp)
    #sel.m<- sel.m[,c(-18, -19)]
    # Filter the original data for the current timepoint
    st.data <- filter(multi, timepoint == tp)
    
    # Fit the model for the selected data and model specifications
    mt.f <- fit.single(st.data, gen = "GID",
                       row = "Row", col = "Column",
                       resp = index, 
                       type.gen = "random",
                       add.row = sel.m$add.row, 
                       add.col = sel.m$add.col,
                       trend.row = sel.m$trend.row,
                       trend.col = sel.m$trend.col,
                       type.residual = sel.m$type.residual, 
                       add.nugget = sel.m$add.nugget,
                       threshold = 3.5)
    
    # Extract predictions and add timepoint and index columns
    pred <- mt.f$predictions
    pred$timepoint <- tp
    pred$VI <- index  # Track the vegetative index in predictions
    
    # Store predictions in the list with a combined key of timepoint and index
    all_predictions[[paste(index, tp, sep = "_")]] <- pred
  }
}
all_predictions1<- all_predictions[c(-2, -18, -26, -42)] 
all_predictions1<- all_predictions1[-2]
# Combine all predictions into a single data frame
final_predictions <- do.call(rbind, all_predictions1)

# Modify the final predictions data frame as needed
preds24<- final_predictions 


preds21$Env<- "KET21"
preds22$Env<- "SNY22"
preds23$Env<- "MCG23"
preds24$Env<- "HELF24"

preds<- rbind(preds21, preds22, preds23, preds24)

table(multi$Env)

#save the predictions

save(preds, file= "data/spec_blup_preds.Rdata")

```



## **6. Adding flight time data to the predicitons and filter for QC**
```{r}
load("data/spec_blup_preds.Rdata")
colnames(multi)
# Create a mapping of timepoint to JD
timepoint_JD <- multi[,c(1, 10,34)] %>%
  distinct()  # Keep only unique combinations
preds <- preds %>%
  left_join(timepoint_JD, by = c("Env","timepoint"))
```


## **7. QC for VI BLUP-s**
```{r}
table(preds$Env , preds$timepoint)

preds$JD <- as.numeric(preds$JDf)

indices <- unique(preds$VI)

# Function to create plots for a given index
create_plots <- function(index) {
  # Filter data for the current index
  preds1 <- preds %>% filter(VI == index)
  
  # Extract unique environment values and collapse them into a string
  env_values <- unique(preds1$Env)
  env_text <- paste(env_values, collapse = ", ")
  
  # Create the histogram plot
  histogram <- ggplot(preds1, aes(x = predicted.value)) +
    geom_histogram(fill = "skyblue", color = "black") +
    facet_wrap(~ Env + timepoint, scales = "free_x") +
    labs(
      x = index,
      y = "Count",
      title = paste("Histogram of", index, "for each timepoint in", env_text)
    ) +
    theme_minimal()
  
  print(histogram)
}

# Loop through all indices
for (index in indices) {
  create_plots(index)
}

#boxplot(mod_stats_df$h2.vc ~  mod_stats_df$timepoint )
#plot(mod_stats_df$h2.vc ~  mod_stats_df$timepoint )

preds1<- preds #%>% filter(!timepoint %in% c("1", "7", "10")) %>% filter(!VI  %in% c("GIPVI"))
# Define a function to create a smooth plot for a given index
create_plot <- function(index) {
  # Filter the data for the current index
  preds1 <- preds %>% filter(VI == index)
  
  # Create the smooth plot
  smooth_plot <- ggplot(preds1, aes(x = JD, 
                                    y = predicted.value, 
                                    group = GID, 
                                    color = as.factor(GID))) +
    geom_smooth(linewidth = 0.5, alpha = 0.01) +
    labs(x = "", 
         y = "VI", 
         title = paste("Vegetative Indices Over Time:", index)) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
      axis.title = element_text(size = 18),
      axis.text = element_text(size = 14),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(color = "black", fill = NA, size = 1),
      legend.position = "none"
    ) +
    facet_wrap(~ Env + VI, scales = "free_x") +
    scale_x_continuous(breaks = c(0, 50, 100, 150, 200), 
                       labels = c("Oct", "Dec", "Feb", "Apr", "Jun")) +
    scale_color_manual(values = rainbow(length(unique(preds1$GID))))
  
  print(smooth_plot)
}

# Get unique indices (assuming the indices are in the 'VI' column)
indices <- unique(preds$VI)

# Loop over each index and create the plot
for (index in indices) {
  create_plot(index)
}
```



## **8. Summary statistics for the BLUP heritabilities **
```{r}
library(patchwork)
load("data/mod_statsNY.Rdata")
mod_stats_df<- mod_stats #%#>% filter(!timepoint %in% c("7"))#, "10", "1"))

summary_stats <- mod_stats_df %>%
  group_by(VI, Env) %>%
  summarize(
    mean_h2.pev = mean(h2.pev, na.rm = TRUE),
    sd_h2.pev = sd(h2.pev, na.rm = TRUE),
    min_h2.pev = min(h2.pev, na.rm = TRUE),
    max_h2.pev = max(h2.pev, na.rm = TRUE),
    n = n()  # Count of observations
  )

# View summary statistics
print(summary_stats)
#boxplot(mod_stats_df$h2.pev ~ mod_stats_df$timepoint, main= "Heritability of VIs at TPs")



summary_stats <- mod_stats_df %>%
  group_by(timepoint, Env) %>%
  summarize(
    mean_h2.pev = mean(h2.pev, na.rm = TRUE),
    sd_h2.pev = sd(h2.pev, na.rm = TRUE),
    min_h2.pev = min(h2.pev, na.rm = TRUE),
    max_h2.pev = max(h2.pev, na.rm = TRUE),
    n = n()  # Count of observations
  )



plot2 <- ggplot(mod_stats_df, aes(x = as.factor(timepoint), y = h2.pev)) +
  geom_boxplot() +  
  scale_fill_brewer() +
  facet_wrap(~ Env ) +
  labs(
    x = "Timepoint",
    y = "h2.pev",  title= "Heritability of TPs"
  ) +
  theme_minimal()
######################################



summary_stats <- mod_stats_df %>%
  group_by(VI, Env) %>%
  summarize(
    mean_h2.pev = mean(h2.pev, na.rm = TRUE),
    sd_h2.pev = sd(h2.pev, na.rm = TRUE),
    min_h2.pev = min(h2.pev, na.rm = TRUE),
    max_h2.pev = max(h2.pev, na.rm = TRUE),
    n = n()  # Count of observations
  )

# View summary statistics
print(summary_stats)


# Create a combined boxplot
# First plot: H2 by Vegetative Index (VI)
plot1 <- ggplot(mod_stats_df, aes(x = VI, y = h2.pev)) +
  geom_boxplot() + 
  facet_wrap(~ Env ) +
  scale_fill_brewer(palette = "Set3") +
  labs(
    title = "Heritability of VIs and Timepoints",
    x = "Vegetative Index (VI)",
    y = "h2.pev"
  ) +
    theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


# Second plot: H2 by Timepoint
plot2 <- ggplot(mod_stats_df, aes(x = as.factor(timepoint), y = h2.pev)) +
  geom_boxplot() +
  facet_wrap(~ Env ) +
  scale_fill_brewer() +
  labs(
    x = "Timepoint",
    y = "h2.pev"
  ) +
  theme_minimal()

# Combine plots side by side
combined_plot <- plot1 + plot2
combined_plot

```

