---
title: "Untitled"
author: "sss322"
date: "2026-02-17"
output: html_document
---


```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning=FALSE, results = "hide")
```


## **Introduction**
This document provides cotains the heritability calculation for multi-spectral drone imagery vegetative indices, maltinq quality and agronomic traits. The goal is to assess the heritability of all traits and choose the ones to proceed with.


## **0. Import packages and data**
```{r, message=FALSE}
pkgs <- c("MASS","dplyr","tidyr","asreml","ASRtriala","ASRgenomics",
          "ggplot2","openxlsx","workflowr","readxl", "forcats", "scales", "corrplot")
suppressPackageStartupMessages(invisible(lapply(pkgs, require, character.only = TRUE)))
Sys.setenv(LANG = "en")
Sys.setlocale("LC_TIME", "C")
```

```{r}
paths <- c("C:/Users/Siim Sepp/NY_WMB/data",
           "~/Documents/GitHub/NY_WMB/data",
           "data")

files <- c("data_ag.Rdata","data_mq.Rdata","data_sp.Rdata","VImat.Rdata", "F_PCA_scores.Rdata")

load_first <- function(fname, dirs = paths) {
  for (d in dirs) {
    p <- file.path(path.expand(d), fname)
    if (file.exists(p)) { load(p, envir = .GlobalEnv); return(invisible(p)) }
  }
  stop("File not found in any known location: ", fname)
}
invisible(lapply(files, load_first))
```


```{r}
data_pc<- FPCA_scores_df %>% filter(VI %in% c("NDVI","PSRI"))

data_pc<- data_pc %>% pivot_wider(values_from = colnames(data_pc)[1:3], names_from = "VI")

data_pcmq <- data_mq[,c(-3)] %>% left_join(data_pc, by= "plot_id")

data_pcag <- data_ag[,c(-3,-5, -6)] %>% left_join(data_pc, by= "plot_id")
```




## **2. Calculating genetic correlation.** 



### **Genetic correlation for malting quality traits**


```{r}
# ROBUST CORRELATION FUNCTION
calculate_pc_trait_correlation_robust <- function(data, pc_trait, target_trait, 
                                                  env = NULL, analysis_type = "Overall") {
  
  failure_reason <- NULL
  
  # Filter data
  data_subset <- data
  if (!is.null(env)) {
    data_subset <- data_subset %>% filter(Env == env)
  }
  
  data_subset <- data_subset %>%
    filter(!is.na(!!sym(pc_trait)), !is.na(!!sym(target_trait))) %>%
    droplevels() %>%
    mutate(
      GID = factor(GID),
      Fam = factor(Fam),
      obs_id = factor(1:n())
    )
  
  n_obs <- nrow(data_subset)
  n_geno <- length(unique(data_subset$GID))
  n_fam <- length(unique(data_subset$Fam))
  
  # More lenient thresholds for Overall
  min_obs <- ifelse(is.null(env), 12, 10)  # Reduced from 15
  min_geno <- ifelse(is.null(env), 6, 5)   # Reduced from 8
  
  if (n_obs < min_obs || n_geno < min_geno) {
    failure_reason <- sprintf("Insufficient data: n_obs=%d (need ≥%d), n_geno=%d (need ≥%d)", 
                             n_obs, min_obs, n_geno, min_geno)
    return(list(failed = TRUE, failure_info = create_failure_info(
      pc_trait, target_trait, analysis_type, env, failure_reason, n_obs, n_geno
    )))
  }
  
  # Check variance BEFORE scaling
  var_pc_raw <- var(data_subset[[pc_trait]], na.rm = TRUE)
  var_target_raw <- var(data_subset[[target_trait]], na.rm = TRUE)
  
  if (var_pc_raw < 1e-10 || var_target_raw < 1e-10) {
    failure_reason <- sprintf("Zero variance before scaling: PC_var=%.2e, Target_var=%.2e", 
                             var_pc_raw, var_target_raw)
    return(list(failed = TRUE, failure_info = create_failure_info(
      pc_trait, target_trait, analysis_type, env, failure_reason, n_obs, n_geno
    )))
  }
  
  # Standardize traits
  data_subset <- data_subset %>%
    mutate(
      !!sym(pc_trait) := as.numeric(scale(!!sym(pc_trait))),
      !!sym(target_trait) := as.numeric(scale(!!sym(target_trait)))
    )
  
  # Reshape to long format
  data_long <- data_subset %>%
    pivot_longer(
      cols = all_of(c(pc_trait, target_trait)), 
      names_to = "trait", 
      values_to = "value"
    ) %>%
    mutate(
      trait = factor(trait, levels = c(pc_trait, target_trait)),
      units = factor(1:n())
    ) %>%
    arrange(trait, units)
  
# ===========================================================================
# MODEL FITTING HIERARCHY
# ===========================================================================
  
  mod <- NULL
  model_attempted <- "None"
  convergence_history <- c()
  
  # MODEL 1: Full model with Family (US covariance)
  if (n_fam >= 3) {  # Need at least 3 families
    mod <- tryCatch({
      model_attempted <- "Model1: us(trait):GID + diag(trait):Fam"
      m <- asreml(
        fixed = value ~ trait,
        random = ~ us(trait):GID + diag(trait):Fam,
        residual = ~ units,
        data = data_long,
        ai.sing = TRUE,
        trace = FALSE,
        maxit = 200,
        workspace = 256e6,
        start.values = TRUE
      )
      convergence_history <- c(convergence_history, 
                              paste0(model_attempted, ": ", m$converge))
      if (m$converge) m else NULL
    }, error = function(e) NULL)
  }
  
  # MODEL 2: US without Family
  if (is.null(mod) || !mod$converge) {
    mod <- tryCatch({
      model_attempted <- "Model2: us(trait):GID"
      m <- asreml(
        fixed = value ~ trait,
        random = ~ us(trait):GID,
        residual = ~ units,
        data = data_long,
        ai.sing = TRUE,
        trace = FALSE,
        maxit = 200,
        workspace = 256e6
      )
      convergence_history <- c(convergence_history, 
                              paste0(model_attempted, ": ", m$converge))
      if (m$converge) m else NULL
    }, error = function(e) NULL)
  }
  
  # MODEL 3: CORGH (heterogeneous correlation) - often more stable
  if (is.null(mod) || !mod$converge) {
    mod <- tryCatch({
      model_attempted <- "Model3: corgh(trait):GID"
      m <- asreml(
        fixed = value ~ trait,
        random = ~ corgh(trait):GID,
        residual = ~ units,
        data = data_long,
        ai.sing = TRUE,
        trace = FALSE,
        maxit = 200,
        workspace = 256e6
      )
      convergence_history <- c(convergence_history, 
                              paste0(model_attempted, ": ", m$converge))
      if (m$converge) m else NULL
    }, error = function(e) NULL)
  }
  
  # MODEL 4: CORV (homogeneous correlation) - most constrained
  if (is.null(mod) || !mod$converge) {
    mod <- tryCatch({
      model_attempted <- "Model4: corv(trait):GID"
      m <- asreml(
        fixed = value ~ trait,
        random = ~ corv(trait):GID,
        residual = ~ units,
        data = data_long,
        ai.sing = TRUE,
        trace = FALSE,
        maxit = 200,
        workspace = 256e6
      )
      convergence_history <- c(convergence_history, 
                              paste0(model_attempted, ": ", m$converge))
      if (m$converge) m else NULL
    }, error = function(e) NULL)
  }
  
  # MODEL 5: Diagonal only (no correlation structure) with IDV residual
  if (is.null(mod) || !mod$converge) {
    mod <- tryCatch({
      model_attempted <- "Model5: diag(trait):GID, idv residual"
      m <- asreml(
        fixed = value ~ trait,
        random = ~ diag(trait):GID,
        residual = ~ idv(units),
        data = data_long,
        ai.sing = TRUE,
        trace = FALSE,
        maxit = 200
      )
      convergence_history <- c(convergence_history, 
                              paste0(model_attempted, ": ", m$converge))
      if (m$converge) m else NULL
    }, error = function(e) NULL)
  }
  
  # MODEL 6: Separate random effects (no covariance structure)
  if (is.null(mod) || !mod$converge) {
    mod <- tryCatch({
      model_attempted <- "Model6: GID + trait:GID (separate)"
      m <- asreml(
        fixed = value ~ trait,
        random = ~ GID + trait:GID,
        residual = ~ units,
        data = data_long,
        ai.sing = TRUE,
        trace = FALSE,
        maxit = 200
      )
      convergence_history <- c(convergence_history, 
                              paste0(model_attempted, ": ", m$converge))
      if (m$converge) m else NULL
    }, error = function(e) NULL)
  }
  
  # MODEL 7: Update previous model if close to convergence
  if (!is.null(mod) && !mod$converge) {
    mod <- tryCatch({
      model_attempted <- paste0(model_attempted, " + update")
      m <- update(mod, maxit = 100)
      convergence_history <- c(convergence_history, 
                              paste0(model_attempted, ": ", m$converge))
      if (m$converge) m else mod  # Keep even if not converged
    }, error = function(e) mod)
  }
  
  # Check if model exists and converged
  if (is.null(mod) || !inherits(mod, "asreml")) {
    failure_reason <- sprintf("All models failed to fit. History: %s", 
                             paste(convergence_history, collapse = "; "))
    return(list(failed = TRUE, failure_info = create_failure_info(
      pc_trait, target_trait, analysis_type, env, failure_reason, n_obs, n_geno
    )))
  }
  
  if (!mod$converge) {
    failure_reason <- sprintf("No model converged. Best attempt: %s. History: %s", 
                             model_attempted, paste(convergence_history, collapse = "; "))
    return(list(failed = TRUE, failure_info = create_failure_info(
      pc_trait, target_trait, analysis_type, env, failure_reason, n_obs, n_geno
    )))
  }
  
  # ===========================================================================
  # EXTRACT VARIANCE COMPONENTS
  # ===========================================================================
  
  vc <- summary(mod)$varcomp
  
  if (is.null(vc) || nrow(vc) == 0) {
    failure_reason <- "No variance components extracted"
    return(list(failed = TRUE, failure_info = create_failure_info(
      pc_trait, target_trait, analysis_type, env, failure_reason, n_obs, n_geno
    )))
  }
  
  # Try different variance component patterns based on model structure
  gid_rows <- grep("trait.*GID|GID.*trait|corgh|corv", rownames(vc), ignore.case = TRUE)
  
  if (length(gid_rows) < 2) {
    # Try alternative extraction for diagonal models
    gid_rows <- grep("GID", rownames(vc))
  }
  
  # Extract variance components based on model structure
  extraction_successful <- FALSE
  
  # Try US/CORGH/CORV structure (3 components)
  if (length(gid_rows) >= 3) {
    V1 <- vc[gid_rows[1], "component"]
    Cov12 <- vc[gid_rows[2], "component"]
    V2 <- vc[gid_rows[3], "component"]
    extraction_successful <- TRUE
  } 
  # Try DIAG structure (2 components - no covariance)
  else if (length(gid_rows) == 2) {
    V1 <- vc[gid_rows[1], "component"]
    V2 <- vc[gid_rows[2], "component"]
    # Estimate correlation from phenotypic correlation as fallback
    pheno_cor <- cor(data_subset[[pc_trait]], data_subset[[target_trait]], 
                     use = "complete.obs")
    Cov12 <- pheno_cor * sqrt(V1 * V2)
    extraction_successful <- TRUE
    model_attempted <- paste0(model_attempted, " (pheno cor fallback)")
  } 
  else {
    failure_reason <- sprintf("Could not extract variance components from %s", 
                             model_attempted)
    return(list(failed = TRUE, failure_info = create_failure_info(
      pc_trait, target_trait, analysis_type, env, failure_reason, n_obs, n_geno
    )))
  }
  
  # Safety checks
  if (is.na(V1) || is.na(V2) || is.na(Cov12)) {
    failure_reason <- sprintf("NA variance: V1=%s, V2=%s, Cov=%s", V1, V2, Cov12)
    return(list(failed = TRUE, failure_info = create_failure_info(
      pc_trait, target_trait, analysis_type, env, failure_reason, n_obs, n_geno
    )))
  }
  
  # Allow negative variances to be set to small positive value
  if (V1 <= 0) V1 <- 1e-6
  if (V2 <= 0) V2 <- 1e-6
  
  sqrt_product <- sqrt(V1 * V2)
  
  if (sqrt_product < 1e-10) {
    failure_reason <- sprintf("Near-zero variances: V1=%.2e, V2=%.2e", V1, V2)
    return(list(failed = TRUE, failure_info = create_failure_info(
      pc_trait, target_trait, analysis_type, env, failure_reason, n_obs, n_geno
    )))
  }
  
  rg <- Cov12 / sqrt_product
  
  # Constrain to valid range
  if (is.na(rg) || is.infinite(rg)) {
    rg <- 0
    model_attempted <- paste0(model_attempted, " (rg set to 0)")
  }
  
  rg <- pmax(-1, pmin(1, rg))
  
  # Calculate SE and p-value
  se_rg <- sqrt((1 - rg^2)^2 / (2 * n_geno))
  z_score <- rg / se_rg
  p_value <- 2 * (1 - pnorm(abs(z_score)))
  sig <- ifelse(p_value < 0.001, "***", 
                ifelse(p_value < 0.01, "**",
                       ifelse(p_value < 0.05, "*", "ns")))
  
  # Parse PC info
  pc_method <- ifelse(grepl("FPC", pc_trait), "FPCA", "PCA")
  pc_index <- ifelse(grepl("NDVI", pc_trait), "NDVI", "PSRI")
  pc_number <- as.numeric(gsub(".*PC(\\d+).*", "\\1", pc_trait))
  
  results <- data.frame(
    Analysis = analysis_type,
    PC_trait = pc_trait,
    PC_Method = pc_method,
    PC_Index = pc_index,
    PC_Number = pc_number,
    Target_trait = target_trait,
    Env = ifelse(is.null(env), "Pooled", as.character(env)),
    V1 = round(V1, 6),
    V2 = round(V2, 6),
    Cov = round(Cov12, 6),
    rg = round(rg, 4),
    se_rg = round(se_rg, 4),
    p_value = round(p_value, 4),
    significance = sig,
    n_obs = n_obs,
    n_geno = n_geno,
    n_fam = n_fam,
    converged = TRUE,
    model_used = model_attempted,
    stringsAsFactors = FALSE
  )
  
  return(list(
    model = mod, 
    results = results,
    failed = FALSE,
    failure_info = NULL
  ))
}

# ==============================================================================
# HELPER FUNCTION: Create failure info dataframe
# ==============================================================================

create_failure_info <- function(pc_trait, target_trait, analysis_type, env, 
                                reason, n_obs, n_geno) {
  data.frame(
    PC_trait = pc_trait,
    Target_trait = target_trait,
    Analysis = analysis_type,
    Env = ifelse(is.null(env), "Pooled", as.character(env)),
    reason = reason,
    n_obs = n_obs,
    n_geno = n_geno,
    stringsAsFactors = FALSE
  )
}

# ==============================================================================
# WRAPPER FUNCTION
# ==============================================================================

run_all_pc_correlations_robust <- function(data, pc_traits, target_traits, 
                                           trait_type = "Agronomic") {
  
  envs <- unique(data$Env[!is.na(data$Env)])
  pc_traits <- pc_traits[pc_traits %in% colnames(data)]
  target_traits <- target_traits[target_traits %in% colnames(data)]
  
  if (length(pc_traits) == 0 || length(target_traits) == 0) {
    cat("No matching traits found in data\n")
    return(list(results = data.frame(), failures = data.frame()))
  }
  
  cat("\n================================================================================\n")
  cat(sprintf("GENETIC CORRELATIONS (ROBUST): PC vs %s TRAITS\n", trait_type))
  cat("================================================================================\n")
  cat("PC traits:", length(pc_traits), "\n")
  cat("Target traits:", length(target_traits), "\n\n")
  
  all_results <- data.frame()
  all_failures <- data.frame()
  
  # OVERALL ANALYSIS
  cat("=== OVERALL ANALYSIS ===\n\n")
  
  overall_success <- 0
  overall_failed <- 0
  
  for (pc in pc_traits) {
    for (target in target_traits) {
      
      result <- calculate_pc_trait_correlation_robust(
        data, pc, target, env = NULL, analysis_type = "Overall"
      )
      
      if (!result$failed) {
        all_results <- rbind(all_results, result$results)
        overall_success <- overall_success + 1
        cat(sprintf("  ✓ %s ~ %-25s: rg=%6.3f %s [%s]\n", 
                   pc, substr(target, 1, 25), 
                   result$results$rg, result$results$significance,
                   result$results$model_used))
      } else {
        all_failures <- rbind(all_failures, result$failure_info)
        overall_failed <- overall_failed + 1
        cat(sprintf("  ✗ %s ~ %-25s: %s\n", 
                   pc, substr(target, 1, 25), 
                   substr(result$failure_info$reason, 1, 60)))
      }
    }
  }
  
  cat(sprintf("\nOverall: %d success, %d failed\n", overall_success, overall_failed))
  
  # BY ENVIRONMENT ANALYSIS
  cat("\n=== BY ENVIRONMENT ANALYSIS ===\n")
  
  for (env_name in envs) {
    
    env_data <- data %>% filter(Env == env_name)
    n_obs_env <- nrow(env_data)
    n_geno_env <- length(unique(env_data$GID))
    
    cat(sprintf("\n--- %s (n=%d, geno=%d) ---\n", env_name, n_obs_env, n_geno_env))
    
    env_success <- 0
    env_failed <- 0
    
    for (pc in pc_traits) {
      for (target in target_traits) {
        
        result <- calculate_pc_trait_correlation_robust(
          data, pc, target, env = env_name, analysis_type = "ByEnv"
        )
        
        if (!result$failed) {
          all_results <- rbind(all_results, result$results)
          env_success <- env_success + 1
        } else {
          all_failures <- rbind(all_failures, result$failure_info)
          env_failed <- env_failed + 1
        }
      }
    }
    
    cat(sprintf("  %s: %d success, %d failed\n", env_name, env_success, env_failed))
  }
  
  cat(sprintf("\n=== FINAL SUMMARY ===\n"))
  cat(sprintf("Total successful: %d\n", nrow(all_results)))
  cat(sprintf("Total failed: %d\n", nrow(all_failures)))
  cat(sprintf("Success rate: %.1f%%\n", 
              100 * nrow(all_results) / (nrow(all_results) + nrow(all_failures))))
  
  return(list(
    results = all_results,
    failures = all_failures
  ))
}
```




```{r}
# ==============================================================================
# USAGE
# ==============================================================================
pc_traits <- c("FPC1_NDVI", "FPC1_PSRI", "FPC2_NDVI", "FPC2_PSRI", "FPC3_NDVI", "FPC3_PSRI")
mq_traits <- colnames(data_mq)[4:15]
ag_traits <- c("yield_kgha", "HD_JD", "winter_survival")

# Run with robust function
output_ag <- run_all_pc_correlations_robust(data_pcag, pc_traits, ag_traits)

output_mq <- run_all_pc_correlations_robust(data_pcmq, pc_traits, mq_traits)

boxplot(output_mq$results$rg~output_mq$results$Target_trait)

output_ag$results$trait_type = "Agronomic"
output_mq$results$trait_type = "Malt quality"

```





#**GENETIC CORRELATIONS: VEGETATION INDICES vs MALT QUALITY TRAITS**

```{r}
colnames(data_pcmq)
colnames(data_sp)
colnames(data_mq)
colnames(data_ag)
#data_sp <- data_sp[,c(1:5,13,14)]
#data_sp <- data_sp %>% pivot_wider(names_from = Timepoint,
 #   values_from = c(PSRI, NDVI))

#######################################
data_spag <-data_ag[,c(-2, -3, -4, -10)] %>% left_join(data_sp, by= "plot_id") 
str(data_spag)
#######################################

data_spmq <-data_mq[,c(-1, -3, -18)] %>% left_join(data_sp, by= "plot_id") 
str(data_spmq)
```

```{r}
# ==============================================================================
# GENETIC CORRELATIONS: VI (by Timepoint) vs MALT QUALITY TRAITS
# Within each environment AND pooled across environments
# ==============================================================================


# ==============================================================================
# CORE FUNCTION: Calculate single VI-MQ correlation (ROBUST)
# ==============================================================================

calculate_vi_mq_correlation_robust <- function(data, vi_trait, mq_trait, 
                                               timepoint = NULL, env = NULL,
                                               analysis_type = "Overall") {
  
  # Filter data
  data_subset <- data
  
  if (!is.null(env)) {
    data_subset <- data_subset %>% filter(Env == env)
  }
  
  if (!is.null(timepoint)) {
    data_subset <- data_subset %>% filter(Timepoint == timepoint)
  }
  
  data_subset <- data_subset %>%
    filter(!is.na(!!sym(vi_trait)), !is.na(!!sym(mq_trait))) %>%
    droplevels() %>%
    mutate(
      GID = factor(GID),
      Fam = factor(Fam),
      obs_id = factor(1:n())
    )
  
  n_obs <- nrow(data_subset)
  n_geno <- length(unique(data_subset$GID))
  
  # Data sufficiency check
  if (n_obs < 15 || n_geno < 8) {
    return(NULL)
  }
  
  # Standardize traits to prevent scale issues
  data_subset <- data_subset %>%
    mutate(
      !!sym(vi_trait) := as.numeric(scale(!!sym(vi_trait))),
      !!sym(mq_trait) := as.numeric(scale(!!sym(mq_trait)))
    )
  
  # Check for zero variance after scaling
  if (sd(data_subset[[vi_trait]], na.rm = TRUE) < 1e-10 || 
      sd(data_subset[[mq_trait]], na.rm = TRUE) < 1e-10) {
    return(NULL)
  }
  
  # Reshape to long format
  data_long <- data_subset %>%
    #select(GID, Fam, obs_id, all_of(c(vi_trait, mq_trait))) %>%
    pivot_longer(
      cols = all_of(c(vi_trait, mq_trait)), 
      names_to = "trait", 
      values_to = "value"
    ) %>%
    mutate(trait = factor(trait, levels = c(vi_trait, mq_trait))) %>%
    arrange(trait, obs_id) %>%
    mutate(units = factor(1:n()))
  
  # Fit bivariate model with fallback options
  mod <- tryCatch({
    asreml(
      fixed = value ~ trait,
      random = ~ us(trait):GID + diag(trait):Fam,
      residual = ~ units,
      data = data_long,
      ai.sing = TRUE,
      trace = FALSE,
      maxit = 100
    )
  }, error = function(e) {
    tryCatch({
      asreml(
        fixed = value ~ trait,
        random = ~ us(trait):GID,
        residual = ~ units,
        data = data_long,
        ai.sing = TRUE,
        trace = FALSE,
        maxit = 100
      )
    }, error = function(e2) NULL)
  })
  
  if (is.null(mod) || !inherits(mod, "asreml") || !mod$converge) {
    return(NULL)
  }
  
  # Extract variance components
  vc <- summary(mod)$varcomp
  if (is.null(vc) || nrow(vc) == 0) return(NULL)
  
  gid_rows <- grep("trait.*GID|GID.*trait", rownames(vc))
  if (length(gid_rows) < 3) return(NULL)
  
  V1 <- vc[gid_rows[1], "component"]
  Cov12 <- vc[gid_rows[2], "component"]
  V2 <- vc[gid_rows[3], "component"]
  
  # Safety checks
  if (is.na(V1) || is.na(V2) || is.na(Cov12) || 
      V1 <= 1e-8 || V2 <= 1e-8) {
    return(NULL)
  }
  
  sqrt_product <- sqrt(V1 * V2)
  if (sqrt_product < 1e-8) return(NULL)
  
  rg <- Cov12 / sqrt_product
  if (is.na(rg) || is.infinite(rg) || abs(rg) > 1.5) return(NULL)
  
  rg <- pmax(-1, pmin(1, rg))
  se_rg <- sqrt((1 - rg^2)^2 / (2 * n_geno))
  
  z_score <- rg / se_rg
  p_value <- 2 * (1 - pnorm(abs(z_score)))
  sig <- ifelse(p_value < 0.001, "***", 
                ifelse(p_value < 0.01, "**",
                       ifelse(p_value < 0.05, "*", "ns")))
  
  results <- data.frame(
    Analysis = analysis_type,
    VI = vi_trait,
    MQ_trait = mq_trait,
    Timepoint = ifelse(is.null(timepoint), "Pooled", as.character(timepoint)),
    Env = ifelse(is.null(env), "Pooled", as.character(env)),
    V1 = round(V1, 6),
    V2 = round(V2, 6),
    Cov = round(Cov12, 6),
    rg = round(rg, 4),
    se_rg = round(se_rg, 4),
    p_value = round(p_value, 4),
    significance = sig,
    n_obs = n_obs,
    n_geno = n_geno,
    converged = TRUE,
    stringsAsFactors = FALSE
  )
  
  return(list(model = mod, results = results))
}

# ==============================================================================
# WRAPPER: Run all VI-MQ correlations
# By timepoint, by environment, and pooled
# ==============================================================================

run_all_vi_mq_correlations <- function(data, vi_traits, mq_traits) {
  
  # Get unique timepoints and environments
  timepoints <- sort(unique(data$Timepoint[!is.na(data$Timepoint)]))
  envs <- unique(data$Env[!is.na(data$Env)])
  
  # Filter to available traits
  vi_traits <- vi_traits[vi_traits %in% colnames(data)]
  mq_traits <- mq_traits[mq_traits %in% colnames(data)]
  
  if (length(vi_traits) == 0 || length(mq_traits) == 0) {
    cat("No matching traits found in data\n")
    return(data.frame())
  }
  
  cat("\n================================================================================\n")
  cat("GENETIC CORRELATIONS: VI (by Timepoint) vs MALT QUALITY TRAITS\n")
  cat("================================================================================\n")
  cat("VI traits:", paste(vi_traits, collapse = ", "), "\n")
  cat("MQ traits:", length(mq_traits), "\n")
  cat("Timepoints:", paste(timepoints, collapse = ", "), "\n")
  cat("Environments:", paste(envs, collapse = ", "), "\n\n")
  
  all_results <- data.frame()
  
  # ===========================================================================
  # ANALYSIS 1: BY TIMEPOINT × ENVIRONMENT (Most granular)
  # ===========================================================================
  
  cat("=== ANALYSIS 1: BY TIMEPOINT × ENVIRONMENT ===\n")
  
  for (env_name in envs) {
    
    cat(sprintf("\n--- Environment: %s ---\n", env_name))
    
    for (tp in timepoints) {
      
      # Check data availability
      tp_env_data <- data %>% 
        filter(Env == env_name, Timepoint == tp)
      
      n_obs <- nrow(tp_env_data)
      n_geno <- length(unique(tp_env_data$GID))
      
      if (n_obs < 15 || n_geno < 8) {
        cat(sprintf("  Timepoint %s: Insufficient data (n=%d, geno=%d)\n", 
                    tp, n_obs, n_geno))
        next
      }
      
      cat(sprintf("  Timepoint %s (n=%d, geno=%d):\n", tp, n_obs, n_geno))
      
      tp_results <- 0
      
      for (vi in vi_traits) {
        for (mq in mq_traits) {
          
          result <- calculate_vi_mq_correlation_robust(
            data, vi, mq, 
            timepoint = tp, 
            env = env_name,
            analysis_type = "Env×Time"
          )
          
          if (!is.null(result)) {
            all_results <- rbind(all_results, result$results)
            tp_results <- tp_results + 1
          }
        }
      }
      
      cat(sprintf("    Completed: %d correlations\n", tp_results))
    }
  }
  
  cat(sprintf("\nCompleted Env×Time: %d correlations\n", 
              nrow(all_results %>% filter(Analysis == "Env×Time"))))
  
  # ===========================================================================
  # ANALYSIS 2: BY TIMEPOINT (POOLED ACROSS ENVIRONMENTS)
  # ===========================================================================
  
  cat("\n\n=== ANALYSIS 2: BY TIMEPOINT (POOLED ACROSS ENVIRONMENTS) ===\n")
  
  for (tp in timepoints) {
    
    # Check data availability
    tp_data <- data %>% filter(Timepoint == tp)
    n_obs <- nrow(tp_data)
    n_geno <- length(unique(tp_data$GID))
    
    if (n_obs < 20 || n_geno < 10) {
      cat(sprintf("\nTimepoint %s: Insufficient data (n=%d, geno=%d)\n", 
                  tp, n_obs, n_geno))
      next
    }
    
    cat(sprintf("\nTimepoint %s (n=%d, geno=%d):\n", tp, n_obs, n_geno))
    
    tp_results <- 0
    
    for (vi in vi_traits) {
      for (mq in mq_traits) {
        
        result <- calculate_vi_mq_correlation_robust(
          data, vi, mq, 
          timepoint = tp, 
          env = NULL,
          analysis_type = "ByTimepoint"
        )
        
        if (!is.null(result)) {
          all_results <- rbind(all_results, result$results)
          tp_results <- tp_results + 1
          
          cat(sprintf("  %s ~ %-25s: rg = %6.3f %s\n", 
                     vi, substr(mq, 1, 25), 
                     result$results$rg, result$results$significance))
        }
      }
    }
    
    cat(sprintf("  Completed: %d correlations\n", tp_results))
  }
  
  cat(sprintf("\nCompleted ByTimepoint: %d correlations\n", 
              nrow(all_results %>% filter(Analysis == "ByTimepoint"))))
  
  # ===========================================================================
  # ANALYSIS 3: BY ENVIRONMENT (POOLED ACROSS TIMEPOINTS)
  # ===========================================================================
  
  cat("\n\n=== ANALYSIS 3: BY ENVIRONMENT (POOLED ACROSS TIMEPOINTS) ===\n")
  
  for (env_name in envs) {
    
    # Check data availability
    env_data <- data %>% filter(Env == env_name)
    n_obs <- nrow(env_data)
    n_geno <- length(unique(env_data$GID))
    
    if (n_obs < 20 || n_geno < 10) {
      cat(sprintf("\n%s: Insufficient data (n=%d, geno=%d)\n", 
                  env_name, n_obs, n_geno))
      next
    }
    
    cat(sprintf("\n%s (n=%d, geno=%d):\n", env_name, n_obs, n_geno))
    
    env_results <- 0
    
    for (vi in vi_traits) {
      for (mq in mq_traits) {
        
        result <- calculate_vi_mq_correlation_robust(
          data, vi, mq, 
          timepoint = NULL, 
          env = env_name,
          analysis_type = "ByEnv"
        )
        
        if (!is.null(result)) {
          all_results <- rbind(all_results, result$results)
          env_results <- env_results + 1
        }
      }
    }
    
    cat(sprintf("  Completed: %d correlations\n", env_results))
  }
  
  cat(sprintf("\nCompleted ByEnv: %d correlations\n", 
              nrow(all_results %>% filter(Analysis == "ByEnv"))))
  
  # ===========================================================================
  # ANALYSIS 4: OVERALL (POOLED ACROSS EVERYTHING)
  # ===========================================================================
  
  cat("\n\n=== ANALYSIS 4: OVERALL (POOLED ACROSS ALL TIMEPOINTS & ENVIRONMENTS) ===\n")
  
  n_total <- nrow(data)
  n_geno_total <- length(unique(data$GID))
  cat(sprintf("Total data: %d obs, %d genotypes\n\n", n_total, n_geno_total))
  
  for (vi in vi_traits) {
    for (mq in mq_traits) {
      
      result <- calculate_vi_mq_correlation_robust(
        data, vi, mq, 
        timepoint = NULL, 
        env = NULL,
        analysis_type = "Overall"
      )
      
      if (!is.null(result)) {
        all_results <- rbind(all_results, result$results)
        cat(sprintf("  %s ~ %-25s: rg = %6.3f %s\n", 
                   vi, substr(mq, 1, 25), 
                   result$results$rg, result$results$significance))
      }
    }
  }
  
  cat(sprintf("\nCompleted Overall: %d correlations\n", 
              nrow(all_results %>% filter(Analysis == "Overall"))))
  
  return(all_results)
}

# ==============================================================================
# EXECUTE ANALYSIS
# ==============================================================================

cat("\n\n################################################################################\n")
cat("STARTING: VI-TIMEPOINT vs MALT QUALITY TRAITS ANALYSIS\n")
cat("################################################################################\n")

# Define VI traits
vi_traits <- c("NDVI", "PSRI")

# Define malt quality traits
mq_traits <- colnames(data_mq)[4:15]

colnames(data_mq)
colnames(data_sp)
data_merged<- data_sp[,c(1:5,13, 14)] %>% left_join(data_mq[,c(-1,-3, -18)], by= "plot_id")
colnames(data_merged)
data_merged<- data_merged %>% filter(Env %in% data_mq$Env)

table(data_merged$Env, data_merged$Timepoint)

# Run comprehensive analysis
results_vi_mq <- run_all_vi_mq_correlations(
  data = data_merged,
  vi_traits = vi_traits,
  mq_traits = mq_traits
)

table(results_vi_mq$converged)


cat("\n\n################################################################################\n")
cat("STARTING: VI-TIMEPOINT vs AGRONOMIC TRAITS ANALYSIS\n")
cat("################################################################################\n")

# Define VI traits
vi_traits <- c("NDVI", "PSRI")

# Define malt quality traits
ag_traits <- colnames(data_ag)[7:9]

colnames(data_ag)
colnames(data_sp)
data_merged<- data_sp[,c(1:5,13, 14)] %>% left_join(data_ag[,c(-2,-3, -4, -10)], by= "plot_id")
colnames(data_merged)
data_merged<- data_merged

table(data_merged$Env, data_merged$Timepoint)

# Run comprehensive analysis
results_vi_ag <- run_all_vi_mq_correlations(
  data = data_merged,
  vi_traits = vi_traits,
  mq_traits = ag_traits
)


results_vi_mq<- rbind(results_vi_mq, results_vi_ag)

















# ==============================================================================
# SUMMARY STATISTICS
# ==============================================================================

cat("\n\n################################################################################\n")
cat("SUMMARY STATISTICS\n")
cat("################################################################################\n\n")

if (nrow(results_vi_mq) > 0) {
  
  # Overall summary by analysis type
  summary_analysis <- results_vi_mq %>%
    group_by(Analysis) %>%
    summarise(
      n_correlations = n(),
      n_significant = sum(p_value < 0.05),
      pct_significant = round(100 * n_significant / n_correlations, 1),
      mean_abs_rg = round(mean(abs(rg)), 3),
      median_rg = round(median(rg), 3),
      .groups = 'drop'
    )
  
  cat("===== SUMMARY BY ANALYSIS TYPE =====\n")
  print(summary_analysis)
  #write.csv(summary_analysis, "VI_MQ_Summary_Analysis.csv", row.names = FALSE)
  
  # Summary by timepoint (pooled across environments)
  summary_timepoint <- results_vi_mq %>%
    filter(Analysis == "ByTimepoint") %>%
    group_by(Timepoint) %>%
    summarise(
      n_correlations = n(),
      n_significant = sum(p_value < 0.05),
      mean_abs_rg = round(mean(abs(rg)), 3),
      .groups = 'drop'
    )
  
  cat("\n===== SUMMARY BY TIMEPOINT =====\n")
  print(summary_timepoint)
  #write.csv(summary_timepoint, "VI_MQ_Summary_Timepoint.csv", row.names = FALSE)
  
  # Summary by environment (pooled across timepoints)
  if (any(results_vi_mq$Analysis == "ByEnv")) {
    summary_env <- results_vi_mq %>%
      filter(Analysis == "ByEnv") %>%
      group_by(Env) %>%
      summarise(
        n_correlations = n(),
        n_significant = sum(p_value < 0.05),
        mean_abs_rg = round(mean(abs(rg)), 3),
        .groups = 'drop'
      )
    
    cat("\n===== SUMMARY BY ENVIRONMENT =====\n")
    print(summary_env)
    #write.csv(summary_env, "VI_MQ_Summary_Environment.csv", row.names = FALSE)
  }
  
  # Best timepoint for each VI-MQ pair
  best_timepoint <- results_vi_mq %>%
    filter(Analysis == "ByTimepoint", p_value < 0.05) %>%
    group_by(VI, MQ_trait) %>%
    slice_max(abs(rg), n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    arrange(VI, desc(abs(rg)))
  
  if (nrow(best_timepoint) > 0) {
    cat("\n===== BEST TIMEPOINT FOR EACH VI-MQ COMBINATION =====\n")
    print(best_timepoint)
    #write.csv(best_timepoint, "VI_MQ_BestTimepoints.csv", row.names = FALSE)
  }
  
  # Top 20 overall correlations
  top_overall <- results_vi_mq %>%
    filter(Analysis == "Overall", p_value < 0.05) %>%
    arrange(desc(abs(rg))) %>%
    head(20)
  
  if (nrow(top_overall) > 0) {
    cat("\n===== TOP 20 CORRELATIONS (OVERALL) =====\n")
    print(top_overall %>% select(VI, MQ_trait, rg, significance))
   # write.csv(top_overall, "VI_MQ_Top20_Overall.csv", row.names = FALSE)
  }
  
  # Top correlations from Env×Time analysis
  top_envtime <- results_vi_mq %>%
    filter(Analysis == "Env×Time", p_value < 0.05) %>%
    arrange(desc(abs(rg))) %>%
    head(20)
  
  if (nrow(top_envtime) > 0) {
    cat("\n===== TOP 20 CORRELATIONS (ENV×TIMEPOINT) =====\n")
    print(top_envtime %>% select(Env, Timepoint, VI, MQ_trait, rg, significance))
    #write.csv(top_envtime, "VI_MQ_Top20_EnvTime.csv", row.names = FALSE)
  }
}

# ==============================================================================
# VISUALIZATIONS
# ==============================================================================

if (nrow(results_vi_mq) > 0) {
  
  # Plot 1: Heatmap - Timepoint × VI vs MQ (key traits)
  if (any(results_vi_mq$Analysis == "ByTimepoint")) {
    
    key_mq <- unique(results_vi_mq$MQ_trait)
    
    plot_data <- results_vi_mq %>%
      filter(Analysis == "ByTimepoint", MQ_trait %in% key_mq) %>%
      mutate(
        VI_Time = paste0(VI, "_T", Timepoint),
        MQ_short = gsub("\\.+$", "", gsub("\\.\\.\\.", "", MQ_trait)),
        rg_label = ifelse(p_value < 0.05, sprintf("%.2f", rg), "")
      )
    
    if (nrow(plot_data) > 0) {
      p1 <- ggplot(plot_data, aes(y = MQ_short, x = VI_Time, fill = rg)) +
        geom_tile(color = "white", linewidth = 1) +
        geom_text(aes(label = rg_label), size = 3) +
        scale_fill_gradient2(
          low = "blue", mid = "white", high = "red",
          midpoint = 0, limits = c(-1, 1),
          name = "rg"
        ) +
        #theme_minimal(base_size = 11) +
        labs(
          title = "Genetic Correlations by Timepoint (Pooled Environments)",
          subtitle = "VI × Timepoint vs Key Malt Quality Traits",
          x = "Malt Quality Trait",
          y = "VI × Timepoint"
        ) +
        theme(
          axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid = element_blank()
        )
      
      #ggsave("VI_MQ_Heatmap_ByTimepoint.png", p1, width = 10, height = 8, dpi = 300)
      print(p1)
    }
  }
  
  # Plot 2: Temporal trends for key MQ traits
  if (any(results_vi_mq$Analysis == "ByTimepoint")) {
    
    #key_mq <- c("MaltExtract...", "BarleyProtein...", "All_MaltQualityScore")
    
    for (mq in key_mq) {
      
      plot_data <- results_vi_mq %>%
        filter(Analysis == "ByTimepoint", MQ_trait == mq) %>%
        mutate(Timepoint = as.numeric(Timepoint))
      
      if (nrow(plot_data) > 0) {
        p <- ggplot(plot_data, aes(x = Timepoint, y = rg, color = VI,
                                    shape = ifelse(p_value < 0.05, "Sig", "NS"))) +
          geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
          geom_line(aes(group = VI), linewidth = 1) +
          geom_point(size = 4) +
          geom_errorbar(aes(ymin = rg - se_rg, ymax = rg + se_rg), width = 0.2) +
          scale_shape_manual(values = c("Sig" = 16, "NS" = 1), name = "p < 0.05") +
          scale_color_manual(values = c("NDVI" = "darkgreen", "PSRI" = "purple")) +
          theme_bw(base_size = 12) +
          labs(
            title = paste("Genetic Correlation with", mq),
            subtitle = "Across growth stages (pooled environments)",
            x = "Timepoint",
            y = "Genetic Correlation (rg)",
            color = "Vegetation Index"
          ) +
          ylim(-1, 1)
        
        print(p)
      }
    }
  }
  
  # Plot 3: Environment comparison for top correlation
  if (any(results_vi_mq$Analysis %in% c("ByEnv", "Env×Time"))) {
    
    top_pair <- results_vi_mq %>%
      filter(Analysis == "Overall", p_value < 0.05) %>%
      arrange(desc(abs(rg))) %>%
      slice(1)
    
    if (nrow(top_pair) > 0) {
      
      env_data <- results_vi_mq %>%
        filter(Analysis == "ByEnv",
               VI == top_pair$VI[1],
               MQ_trait == top_pair$MQ_trait[1])
      
      if (nrow(env_data) > 1) {
        p3 <- ggplot(env_data, aes(x = Env, y = rg)) +
          geom_hline(yintercept = 0, linetype = "dashed") +
          geom_point(size = 4, color = "steelblue") +
          geom_errorbar(aes(ymin = rg - se_rg, ymax = rg + se_rg), width = 0.2) +
          theme_bw(base_size = 12) +
          labs(
            title = "Genetic Correlation Across Environments",
            subtitle = paste(top_pair$VI[1], "vs", top_pair$MQ_trait[1]),
            x = "Environment",
            y = "Genetic Correlation (rg)"
          ) +
          ylim(-1, 1)

        print(p3)
      }
    }
  }
  
  # Plot 4: Heatmap - Env × Timepoint for specific VI-MQ pair
  if (any(results_vi_mq$Analysis == "Env×Time")) {
    
    # Find most significant correlation overall
    top_pair <- results_vi_mq %>%
      filter(Analysis == "Overall", p_value < 0.05) %>%
      arrange(desc(abs(rg))) %>%
      slice(1)
    
    if (nrow(top_pair) > 0) {
      
      envtime_data <- results_vi_mq %>%
        filter(Analysis == "Env×Time",
               VI == top_pair$VI[1],
               MQ_trait == top_pair$MQ_trait[1]) %>%
        mutate(
          Timepoint = factor(Timepoint),
          rg_label = ifelse(p_value < 0.05, sprintf("%.2f", rg), "")
        )
      
      if (nrow(envtime_data) > 0) {
        p4 <- ggplot(envtime_data, aes(x = Timepoint, y = Env, fill = rg)) +
          geom_tile(color = "white", linewidth = 1) +
          geom_text(aes(label = rg_label), size = 3.5) +
          scale_fill_gradient2(
            low = "blue", mid = "white", high = "red",
            midpoint = 0, limits = c(-1, 1),
            name = "rg"
          ) +
         #theme_minimal(base_size = 12) +
          labs(
            title = paste(top_pair$VI[1], "vs", top_pair$MQ_trait[1]),
            subtitle = "Genetic correlations by Environment × Timepoint",
            x = "Timepoint",
            y = "Environment"
          ) +
          theme(panel.grid = element_blank())
        print(p4)
      }
    }
  }
  
  # Plot 5: Which timepoint is best?
  if (nrow(best_timepoint) > 0) {
    
    timepoint_counts <- best_timepoint %>%
      count(Timepoint, VI, name = "n_traits")
    
    p5 <- ggplot(timepoint_counts, aes(x = factor(Timepoint), y = n_traits, fill = VI)) +
      geom_bar(stat = "identity", position = "dodge", color = "black") +
      scale_fill_manual(values = c("NDVI" = "darkgreen", "PSRI" = "purple")) +
      theme_bw(base_size = 12) +
      labs(
        title = "Which Timepoint Best Predicts Malt Quality?",
        subtitle = "Count of MQ traits with strongest correlation at each timepoint",
        x = "Timepoint",
        y = "Number of MQ Traits",
        fill = "Vegetation Index"
      )
    print(p5)
  }
  
  # Plot 6: Overall comparison bar chart
  overall_data <- results_vi_mq %>%
    filter(Analysis == "Overall") %>%
    mutate(MQ_short = gsub("\\.+$", "", gsub("\\.\\.\\.", "", MQ_trait)))
  
  if (nrow(overall_data) > 0) {
    p6 <- ggplot(overall_data, aes(x = reorder(MQ_short, abs(rg)), y = rg, fill = VI)) +
      geom_bar(stat = "identity", position = "dodge", color = "black") +
      geom_text(aes(label = significance), 
                position = position_dodge(width = 0.9), 
                hjust = -0.2, size = 3) +
      coord_flip() +
      scale_fill_manual(values = c("NDVI" = "darkgreen", "PSRI" = "purple")) +
      theme_bw(base_size = 11) +
      labs(
        title = "Overall Genetic Correlations: VI vs Malt Quality",
        subtitle = "Pooled across all timepoints and environments",
        x = "Malt Quality Trait",
        y = "Genetic Correlation (rg)",
        fill = "VI"
      ) +
      ylim(-1, 1)
    print(p6)
  }
}
```





```{r}

results_pc<- rbind(results_pc_ag,results_pc_mq)

hist(results_vi$rg)
hist(results_pc$rg)
g_cor_res<- list(results_vi, results_pc)

save(g_cor_res, file= "gen_cor.Rdata")
```




##**Correlation plots for VI agronomic**
```{r}
colnames(results_vi_ag)
table(corr_summary$Analysis)
corr_summary<- results_vi_ag %>% filter(Analysis == "Env×Time")

env_order<- c( "KET21", "SNY22" ,"MCG23","HELF24" , "MCG25" )
trait_map <- c(
  "KernelWeight.mg." = "KW",
  "on6.64...."       = "on6/64",
  "winter_survival"  = "WS",
  "MaltExtract..."   = "ME",
  "BarleyProtein..." = "BP",
  "WortProtein..."   = "WP",
  "S.T"              = "S.T", 
  "DP..ASBC."       = "DP", 
  "Alpha.amylase.20.DU." = "AA", 
  "Beta.glucan.ppm." = "BG",
  "HD_JD" = "HD", 
  "yield_kgha" = "GY", 
  "FAN.ppm."         = "FAN", 
  "AdjunctQualityScore" = "Adj.Sc",
  "All_MaltQualityScore" = "All.Sc")



pal <- c(
  KET21  = "cornsilk3",
  SNY22  = "palegreen2",
  MCG23  = "wheat2",
  HELF24 = "lightblue3", 
  MCG25 = "yellow3"
)

trait_order<- c( "GY" , "HD", "WS",  "KW", "on6/64" , "FAN" , "DP", "AA", "S.T" , "BG" , "ME",  "BP", "WP", "Adj.Sc", "All.Sc" )

corr_summary <- corr_summary %>%
  mutate(MQ_trait = recode(MQ_trait, !!!trait_map))

corr_summary <- corr_summary %>%  mutate( Env= trimws(as.character(Env)),
    Env = fct_relevel(Env, env_order)  
  ) %>% mutate( MQ_trait= trimws(as.character(MQ_trait)),
    MQ_trait = fct_relevel(MQ_trait, trait_order)  
  )


# ==============================================================================
# HEATMAP VISUALIZATIONS FOR GENETIC CORRELATIONS AGRONOMIC
# ==============================================================================

corr_summary$Timepoint<- as.numeric(corr_summary$Timepoint)
corr_summary$Timepoint[1:42] <- corr_summary$Timepoint[1:42] - 1
corr_summary$Timepoint[79:99] <- corr_summary$Timepoint[79:99]-1
corr_summary$Timepoint[100:145] <- corr_summary$Timepoint[100:145] -11
corr_summary$Timepoint[146:157] <- corr_summary$Timepoint[146:157] -1

JD_map_filtered$Timepoint <- as.numeric(JD_map_filtered$Timepoint) 
JD_map_filtered$Env <- JD_map_filtered$Environment
corr_summary1<- corr_summary %>% left_join(JD_map_filtered[,c(2,3,4,5)], by = c("Env", "Timepoint"))

# ==============================================================================
# OPTION 1: HEATMAP - VI × MQ_trait, FACETED BY ENV
# Timepoint on x-axis, separate panel per environment
# ==============================================================================

heatmap_by_env <- ggplot(
  corr_summary,
  aes(x = factor(Timepoint), y = MQ_trait, fill = rg)
) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.2f%s", rg, significance)), 
              size = 3, color = "black", angle= -90)+
  facet_grid(VI ~ Env, scales = "free") +  # VI in rows, Env in columns
  scale_fill_gradient2(
    low = "lightblue3",
    mid = "white",
    high = "red3",
    midpoint = 0,
    limits = c(-1, 1),
    name = "Genetic\nCorrelation"
  ) +
  labs(
    x = "Timepoint",
    y = "Trait"
  ) +
  #theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold"),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    panel.grid = element_blank(),
    legend.position = "bottom"
  )

print(heatmap_by_env)
```



##**Correlation plots for VI mqlt quality. *
```{r}
results_vi_mq 

corr_summary<- results_vi_mq %>% filter(Analysis == "Env×Time")
table(corr_summary$Analysis)

corr_summary <- corr_summary %>%
  mutate(MQ_trait = recode(MQ_trait, !!!trait_map))

corr_summary <- corr_summary %>%  mutate( Env= trimws(as.character(Env)),
    Env = fct_relevel(Env, env_order)  
  ) %>% mutate( MQ_trait= trimws(as.character(MQ_trait)),
    MQ_trait = fct_relevel(MQ_trait, trait_order)  
  )


# ==============================================================================
# HEATMAP VISUALIZATIONS FOR GENETIC CORRELATIONS AGRONOMIC
# ==============================================================================

corr_summary$Timepoint<- as.numeric(corr_summary$Timepoint)
corr_summary$Timepoint[1:231] <- corr_summary$Timepoint[1:231] - 1
corr_summary$Timepoint[232:390] <- corr_summary$Timepoint[232:390]-11


JD_map_filtered$Timepoint <- as.numeric(JD_map_filtered$Timepoint) 
JD_map_filtered$Env <- JD_map_filtered$Environment
corr_summary1<- corr_summary %>% left_join(JD_map_filtered[,c(2,3,4,5)], by = c("Env", "Timepoint"))

# ==============================================================================
# OPTION 1: HEATMAP - VI × MQ_trait, FACETED BY ENV
# Timepoint on x-axis, separate panel per environment
# ==============================================================================

heatmap_by_env <- ggplot(
  corr_summary,
  aes(x = factor(Timepoint), y = MQ_trait, fill = rg)
) +
  geom_tile(color = "white", linewidth = 0.5) +
  #geom_text(aes(label = sprintf("%.1f%s", rg,"")), 
  #            size = 3, color = "black")+
  facet_grid(VI ~ Env, scales = "free") +  # VI in rows, Env in columns
  scale_fill_gradient2(
    low = "lightblue3",
    mid = "white",
    high = "red3",
    midpoint = 0,
    limits = c(-1, 1),
    name = "Genetic\nCorrelation"
  ) +
  labs(
    x = "Timepoint",
    y = "Trait"
  ) +
  #theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold"),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    panel.grid = element_blank(),
    legend.position = "bottom"
  )

print(heatmap_by_env)

```











##**Correlation plots for PC agronomic**
```{r}
results_pc_ag
results_pc_mq
corr_summary<- results_pc_ag %>% filter(Analysis == "ByEnv")

table(corr_summary$Analysis)

corr_summary <- corr_summary %>%
  mutate(Target_trait = recode(Target_trait, !!!trait_map))

corr_summary <- corr_summary %>%  mutate( Env= trimws(as.character(Env)),
    Env = fct_relevel(Env, env_order)  
  ) %>% mutate( Target_trait= trimws(as.character(Target_trait)),
    Target_trait = fct_relevel(Target_trait, trait_order)  
  )


str(corr_summary)


corr_summary <- corr_summary %>%
  separate(
    col = PC_trait,
    into = c("PC_Index", "PC_Type_Num"),
    sep = "_",
    remove = FALSE  # Keep original column
  ) %>%
  mutate(
    PC_Method = ifelse(grepl("FPC", PC_Type_Num), "FPCA", "PCA"),
    PC_Number = as.numeric(str_extract(PC_Type_Num, "\\d+"))
  ) %>%
  select(-PC_Type_Num)  # Remove temporary column

# View result
head(corr_summary)


heatmap_by_env <- ggplot(
  corr_summary,
  aes(x = factor(PC_Number), y = Target_trait, fill = rg)
) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.1f%s", rg,significance)), 
              size = 3, color = "black")+
  facet_grid(PC_Method ~ PC_Index ~ Env, scales = "free") +  # VI in rows, Env in columns
  scale_fill_gradient2(
    low = "lightblue3",
    mid = "white",
    high = "red3",
    midpoint = 0,
    limits = c(-1, 1),
    name = "Genetic\nCorrelation"
  ) +
  labs(
    x = "Component",
    y = "Trait"
  ) +
  #theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold"),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    panel.grid = element_blank(),
    legend.position = "bottom"
  )

print(heatmap_by_env)
```



##**Correlation plots for PC malt**
```{r}
corr_summary<- results_pc_mq %>% filter(Analysis == "ByEnv")

table(corr_summary$Analysis)

corr_summary <- corr_summary %>%
  mutate(Target_trait = recode(Target_trait, !!!trait_map))

corr_summary <- corr_summary %>%  mutate( Env= trimws(as.character(Env)),
    Env = fct_relevel(Env, env_order)  
  ) %>% mutate( Target_trait= trimws(as.character(Target_trait)),
    Target_trait = fct_relevel(Target_trait, trait_order)  
  )


str(corr_summary)


corr_summary <- corr_summary %>%
  separate(
    col = PC_trait,
    into = c("PC_Index", "PC_Type_Num"),
    sep = "_",
    remove = FALSE  # Keep original column
  ) %>%
  mutate(
    PC_Method = ifelse(grepl("FPC", PC_Type_Num), "FPCA", "PCA"),
    PC_Number = as.numeric(str_extract(PC_Type_Num, "\\d+"))
  ) %>%
  select(-PC_Type_Num)  # Remove temporary column

# View result
head(corr_summary)


heatmap_by_env <- ggplot(
  corr_summary,
  aes(x = factor(PC_Number), y = Target_trait, fill = rg)
) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.1f%s", rg,significance)), 
              size = 2.5, color = "black")+
  facet_grid(PC_Method ~ PC_Index ~ Env, scales = "free") +  # VI in rows, Env in columns
  scale_fill_gradient2(
    low = "lightblue3",
    mid = "white",
    high = "red3",
    midpoint = 0,
    limits = c(-1, 1),
    name = "Genetic\nCorrelation"
  ) +
  labs(
    x = "Component",
    y = "Trait"
  ) +
  #theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold"),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    panel.grid = element_blank(),
    legend.position = "bottom"
  )

print(heatmap_by_env)
```
