---
title: "Phenotypic Correlations"
author: "sss322"
date: "2026-02-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## **Introduction**
Pairwise Pearson correlations between agronomic (ag), malt quality (mq), functional-PCA spectral (pc), and raw spectral VI (sp) traits. Correlations are computed:

- **Pooled** — all environments combined
- **Within environment** — separately per `Env`
- **Within timepoint** — separately per flight timepoint (sp traits only)
- **Within environment × timepoint** — finest stratification for sp traits

sp × pc correlations are **excluded** by design.


## **0. Packages and data**

```{r packages, message=FALSE}
pkgs <- c("dplyr", "tidyr", "ggplot2", "forcats", "scales", "purrr")
suppressPackageStartupMessages(invisible(lapply(pkgs, require, character.only = TRUE)))
Sys.setenv(LANG = "en")
```

```{r load-data}
paths <- c("C:/Users/Siim Sepp/NY_WMB/data",
           "~/Documents/GitHub/NY_WMB/data",
           "data")

load_first <- function(fname, dirs = paths) {
  for (d in dirs) {
    p <- file.path(path.expand(d), fname)
    if (file.exists(p)) { load(p, envir = .GlobalEnv); return(invisible(p)) }
  }
  stop("File not found: ", fname)
}

invisible(lapply(
  c("data_ag.Rdata", "data_mq.Rdata", "data_sp.Rdata", "F_PCA_scores.Rdata"),
  load_first
))
```


## **1. Define traits and prepare merged datasets**

```{r define-traits}
# ---- Agronomic traits (cols 7:9 of data_ag) ----
ag_traits <- colnames(data_ag)[7:9]

# ---- Malt quality traits (cols 4:15 of data_mq) ----
mq_traits <- colnames(data_mq)[4:15]

# ---- Spectral VI traits (cols 5:13 of data_sp) ----
sp_traits <- names(data_sp)[c(5,13)]

# ---- Detect timepoint column in data_sp ----
# (the non-id, non-Env column among cols 1:4)
sp_id_cols <- colnames(data_sp)[1:4]
tp_col <- setdiff(sp_id_cols, c("plot_id", "GID", "Env", "gid", "genotype"))
if (length(tp_col) == 0) {
  # fallback: the 4th column
  tp_col <- colnames(data_sp)[4]
}
tp_col <- tp_col[1]
message("Timepoint column detected: ", tp_col)

# ---- Functional PCA scores (pc traits) ----
data_pc <- FPCA_scores_df %>%
  filter(VI %in% c("NDVI", "PSRI"))

# Identify score columns (not plot_id / VI)
score_cols <- setdiff(colnames(data_pc), c("plot_id", "VI"))
data_pc <- data_pc %>%
  pivot_wider(id_cols = "plot_id",
              names_from  = "VI",
              values_from = all_of(score_cols))

pc_traits <- c("FPC1_NDVI" ,"FPC1_PSRI" ,"FPC2_NDVI" ,"FPC2_PSRI", "FPC3_NDVI", "FPC3_PSRI")

# ---- Convenience: base id columns from each dataset ----
ag_id <- c("plot_id", "GID", "Env")
mq_id <- c("plot_id", "GID", "Env")

# ---- ag base (keep id + ag_traits) ----
ag_base <- data_ag[,c(1:3,7:9)] #%>% select(any_of(ag_id), all_of(ag_traits))

# ---- mq base ----
mq_base <- data_mq[,1:15] #%>% select(any_of(mq_id), all_of(mq_traits))
# ---- sp: one row per plot (average over timepoints) for non-timepoint analyses ----
sp_avg <- data_sp %>%
  group_by(plot_id) %>%
  summarise(across(all_of(sp_traits), \(x) mean(x, na.rm = TRUE)),
            Env = first(Env), .groups = "drop")

# ---- sp: full (per timepoint) dataset for within-timepoint analyses ----
sp_full <- data_sp %>%
  dplyr::select(plot_id, Env, all_of(tp_col), all_of(sp_traits))

# ---- pc base ----
pc_base <- data_pc
```


## **2. Helper functions**

```{r helpers}
# Pearson r with complete-case guard
safe_cor <- function(x, y) {
  ok <- complete.cases(x, y)
  if (sum(ok) < 3) return(NA_real_)
  cor(x[ok], y[ok])
}

# All pairwise correlations between x_traits and y_traits in a data frame
# Returns: data.frame(x_trait, y_trait, r, n)
pair_cors <- function(df, x_traits, y_traits) {
  res <- vector("list", length(x_traits) * length(y_traits))
  k   <- 1L
  for (x in x_traits) {
    for (y in y_traits) {
      ok  <- complete.cases(df[[x]], df[[y]])
      n   <- sum(ok)
      r   <- if (n >= 3) cor(df[[x]][ok], df[[y]][ok]) else NA_real_
      res[[k]] <- data.frame(x_trait = x, y_trait = y, r = r, n = n,
                             stringsAsFactors = FALSE)
      k <- k + 1L
    }
  }
  do.call(rbind, res)
}

# Apply pair_cors within groups defined by group_cols (character vector)
group_cors <- function(df, x_traits, y_traits, group_cols) {
  df %>%
    group_by(across(all_of(group_cols))) %>%
    group_modify(~ pair_cors(.x, x_traits, y_traits)) %>%
    ungroup()
}

# ---- Colour / labelling constants (consistent with project style) ----
env_order <- c("KET21", "SNY22", "MCG23", "HELF24", "MCG25")

trait_map <- c(
  "KernelWeight.mg."    = "KW",
  "on6.64...."          = "on6/64",
  "winter_survival"     = "WS",
  "MaltExtract..."      = "ME",
  "BarleyProtein..."    = "BP",
  "WortProtein..."      = "WP",
  "S.T"                 = "S.T",
  "DP..ASBC."           = "DP",
  "Alpha.amylase.20.DU." = "AA",
  "Beta.glucan.ppm."    = "BG",
  "HD_JD"               = "HD",
  "yield_kgha"          = "GY",
  "FAN.ppm."            = "FAN",
  "AdjunctQualityScore"  = "Adj.Sc",
  "All_MaltQualityScore" = "All.Sc"
)

short_name <- function(x) ifelse(x %in% names(trait_map), trait_map[x], x)

# Summarise a correlation data frame (from cors_all or any tagged df).
# group_vars : character vector of columns to group by
# filter_strat: optional character vector of stratification levels to keep
cor_summary <- function(df, group_vars, filter_strat = NULL) {
  if (!is.null(filter_strat))
    df <- df %>% filter(stratification %in% filter_strat)
  df %>%
    filter(!is.na(r)) %>%
    mutate(.abs_r = abs(r)) %>%
    group_by(across(all_of(group_vars))) %>%
    summarise(
      mean_r     = round(mean(r),      3),
      mean_abs_r = round(mean(.abs_r), 3),
      sd_r       = round(sd(r),        3),
      max_abs_r  = round(max(.abs_r),  3),
      n_pairs    = n(),
      .groups    = "drop"
    ) %>%
    arrange(desc(mean_abs_r))
}

# Heatmap: rows = y_trait, cols = x_trait, tile fill = r
# facet_col: optional grouping column for faceting (e.g. "Env", "Timepoint")
plot_heatmap <- function(cors_df, x_col = "x_trait", y_col = "y_trait",
                         r_col = "r", facet_col = NULL,
                         title = "", base_size = 11) {
  df <- cors_df %>%
    mutate(
      x_label = short_name(.data[[x_col]]),
      y_label = short_name(.data[[y_col]])
    )

  p <- ggplot(df, aes(x = x_label, y = y_label, fill = .data[[r_col]])) +
    geom_tile(colour = "grey90") +
    geom_text(aes(label = ifelse(is.finite(.data[[r_col]]),
                                 sprintf("%.2f", .data[[r_col]]), "")),
              size = 6) +
    scale_fill_gradient2(
      low = "#2c7bb6", mid = "white", high = "#d7191c",
      midpoint = 0, limits = c(-1, 1), name = "r",
      na.value = "grey85"
    ) +
    labs(title = title, x = NULL, y = NULL) +
    theme_minimal(base_size = base_size) +
    theme(
      panel.grid    = element_blank(),
      axis.text   = element_text(size = 18),
      plot.title    = element_text(hjust = 0.5, size = base_size + 1)
    )

  if (!is.null(facet_col)) {
    p <- p + facet_wrap(reformulate(facet_col), scales = "free")
  }
  p
}
```


## **3. ag × mq correlations**

### 3.1 Pooled (over all environments)

```{r agmq-pooled, fig.width=9, fig.height=5}
agmq <- ag_base %>%
  inner_join(mq_base %>% dplyr::select(plot_id, all_of(mq_traits)), by = "plot_id")

cors_agmq_pooled <- pair_cors(agmq, ag_traits, mq_traits)


cors_agmq_pooled <- cors_agmq_pooled %>%
  mutate(x_trait = recode(x_trait, !!!trait_map)) %>%
  mutate(y_trait = recode(y_trait, !!!trait_map)) 

plot_heatmap(cors_agmq_pooled, title = "ag × mq — pooled over environments")


corm<- as.matrix(cor(agmq[,4:18]))

library(tibble)
corm <- corm %>%
  as.data.frame() %>%
  rownames_to_column("Trait") %>%
  mutate(Trait = recode(Trait, !!!trait_map)) %>%
  column_to_rownames("Trait")

colnames(corm) <- recode(colnames(corm), !!!trait_map)
```

### 3.2 Within environment

```{r agmq-within-env, fig.width=14, fig.height=5}
cors_agmq_env <- group_cors(agmq, ag_traits, mq_traits, "Env") %>%
  mutate(Env = factor(Env, levels = env_order))

plot_heatmap(cors_agmq_env, facet_col = "Env",
             title = "ag × mq — within environment")


```

### 3.3 Clustered heatmap (pooled)

Hierarchical clustering (complete linkage, Euclidean distance) applied to both
ag rows and mq columns.

```{r agmq-clustered, fig.width=8, fig.height=5}
library(pheatmap)

# cors_agmq_pooled already has short names from the recode above
cor_mat <- cors_agmq_pooled %>%
  dplyr::select(x_trait, y_trait, r) %>%
  tidyr::pivot_wider(names_from = y_trait, values_from = r) %>%
  tibble::column_to_rownames("x_trait") %>%
  as.matrix()



pheatmap(corm,
         color            = colorRampPalette(c("#2c7bb6", "white", "#d7191c"))(100),
         breaks           = seq(-1, 1, length.out = 101),
         cluster_rows     = TRUE,
         cluster_cols     = TRUE,
         clustering_method = "complete",
         display_numbers  = TRUE,
         number_format    = "%.2f",
         number_color     = "black",
         fontsize_number  = 13,
         fontsize_row     = 14,
         fontsize_col     = 14,
         border_color     = "grey90",
         treeheight_row   = 35,
         treeheight_col   = 35,
         na_col           = "grey85")
```

### 3.4 Clustered heatmap — within environment

```{r agmq-clustered-env, fig.width=7, fig.height=5, results='asis'}
envs_present <- intersect(env_order,
                          unique(as.character(cors_agmq_env$Env)))

for (e in envs_present) {
  cat("\n\n#### ", e, "\n\n")
  mat_e <- cors_agmq_env %>%
    filter(Env == e) %>%
    dplyr::select(x_trait, y_trait, r) %>%
    mutate(x_trait = short_name(x_trait),
           y_trait = short_name(y_trait)) %>%
    tidyr::pivot_wider(names_from = y_trait, values_from = r) %>%
    tibble::column_to_rownames("x_trait") %>%
    as.matrix()

  # Skip if too few non-NA values to cluster
  if (sum(!is.na(mat_e)) < 4) next

  print(
    pheatmap(mat_e,
             color            = colorRampPalette(c("#2c7bb6", "white", "#d7191c"))(100),
             breaks           = seq(-1, 1, length.out = 101),
             cluster_rows     = nrow(mat_e) > 1,
             cluster_cols     = ncol(mat_e) > 1,
             clustering_method = "complete",
             display_numbers  = TRUE,
             number_format    = "%.2f",
             number_color     = "black",
             fontsize_number  = 11,
             fontsize_row     = 13,
             fontsize_col     = 13,
             main             = paste0("ag × mq — ", e, " (clustered)"),
             border_color     = "grey90",
             treeheight_row   = 30,
             treeheight_col   = 30,
             na_col           = "grey85",
             silent           = TRUE)
  )
}
```


## **4. ag × pc correlations**

### 4.1 Pooled

```{r agpc-pooled, fig.width=10, fig.height=5}
agpc <- ag_base %>%
  inner_join(pc_base, by = "plot_id")

table(agpc$Env)

cors_agpc_pooled <- pair_cors(agpc, ag_traits, pc_traits)

plot_heatmap(cors_agpc_pooled, title = "ag × pc — pooled over environments")
```

### 4.2 Within environment

```{r agpc-within-env, fig.width=16, fig.height=5}
cors_agpc_env <- group_cors(agpc, ag_traits, pc_traits, "Env") %>%
  mutate(Env = factor(Env, levels = env_order))

plot_heatmap(cors_agpc_env, facet_col = "Env",
             title = "ag × pc — within environment")
```


## **5. ag × sp correlations**

*sp traits averaged over timepoints for the pooled and within-environment analyses.*

### 5.1 Pooled

```{r agsp-pooled, fig.width=9, fig.height=5}
agsp <- ag_base %>%
  inner_join(sp_avg %>% dplyr::select(plot_id, all_of(sp_traits)), by = "plot_id")

cors_agsp_pooled <- pair_cors(agsp, ag_traits, sp_traits)

plot_heatmap(cors_agsp_pooled, title = "ag × sp — pooled (sp averaged over timepoints)")
```

### 5.2 Within environment

```{r agsp-within-env, fig.width=14, fig.height=5}
cors_agsp_env <- group_cors(agsp, ag_traits, sp_traits, "Env") %>%
  mutate(Env = factor(Env, levels = env_order))

plot_heatmap(cors_agsp_env, facet_col = "Env",
             title = "ag × sp — within environment (sp averaged over timepoints)")
```

### 5.3 Within timepoint (pooled over environments)

```{r agsp-within-tp, fig.width=9, fig.height=10}
agsp_tp <- ag_base %>%
  inner_join(sp_full %>% dplyr::select(plot_id, all_of(tp_col), all_of(sp_traits)),
             by = "plot_id")

cors_agsp_tp <- group_cors(agsp_tp, ag_traits, sp_traits, tp_col) %>%
  rename(Timepoint = all_of(tp_col))

plot_heatmap(cors_agsp_tp, facet_col = "Timepoint",
             title = "ag × sp — within timepoint (pooled over environments)")
```

### 5.4 Within environment × timepoint

```{r agsp-env-tp, fig.width=14, fig.height=14}
cors_agsp_env_tp <- group_cors(agsp_tp, ag_traits, sp_traits, c("Env", tp_col)) %>%
  rename(Timepoint = all_of(tp_col)) %>%
  mutate(Env = factor(Env, levels = env_order),
         EnvTP = paste0(Env, " | ", Timepoint))

plot_heatmap(cors_agsp_env_tp, facet_col = "EnvTP",
             title = "ag × sp — within environment × timepoint",
             base_size = 9)
```


## **6. mq × pc correlations**

### 6.1 Pooled

```{r mqpc-pooled, fig.width=10, fig.height=8}
mqpc <- mq_base %>%
  inner_join(pc_base, by = "plot_id")

cors_mqpc_pooled <- pair_cors(mqpc, mq_traits, pc_traits)

plot_heatmap(cors_mqpc_pooled, title = "mq × pc — pooled over environments")
```

### 6.2 Within environment

```{r mqpc-within-env, fig.width=16, fig.height=8}
cors_mqpc_env <- group_cors(mqpc, mq_traits, pc_traits, "Env") %>%
  mutate(Env = factor(Env, levels = env_order))

plot_heatmap(cors_mqpc_env, facet_col = "Env",
             title = "mq × pc — within environment")
```


## **7. mq × sp correlations**

### 7.1 Pooled

```{r mqsp-pooled, fig.width=10, fig.height=8}
mqsp <- mq_base %>%
  inner_join(sp_avg %>% dplyr::select(plot_id, all_of(sp_traits)), by = "plot_id")

cors_mqsp_pooled <- pair_cors(mqsp, mq_traits, sp_traits)

plot_heatmap(cors_mqsp_pooled, title = "mq × sp — pooled (sp averaged over timepoints)")
```

### 7.2 Within environment

```{r mqsp-within-env, fig.width=14, fig.height=8}
cors_mqsp_env <- group_cors(mqsp, mq_traits, sp_traits, "Env") %>%
  mutate(Env = factor(Env, levels = env_order))

plot_heatmap(cors_mqsp_env, facet_col = "Env",
             title = "mq × sp — within environment (sp averaged over timepoints)")
```

### 7.3 Within timepoint (pooled over environments)

```{r mqsp-within-tp, fig.width=10, fig.height=22}
mqsp_tp <- mq_base %>%
  inner_join(sp_full %>% dplyr::select(plot_id, all_of(tp_col), all_of(sp_traits)),
             by = "plot_id")

cors_mqsp_tp <- group_cors(mqsp_tp, mq_traits, sp_traits, tp_col) %>%
  rename(Timepoint = all_of(tp_col))

plot_heatmap(cors_mqsp_tp, facet_col = "Timepoint",
             title = "mq × sp — within timepoint (pooled over environments)")
```

### 7.4 Within environment × timepoint

```{r mqsp-env-tp, fig.width=14, fig.height=28}
cors_mqsp_env_tp <- group_cors(mqsp_tp, mq_traits, sp_traits, c("Env", tp_col)) %>%
  rename(Timepoint = all_of(tp_col)) %>%
  mutate(Env = factor(Env, levels = env_order),
         EnvTP = paste0(Env, " | ", Timepoint))

plot_heatmap(cors_mqsp_env_tp, facet_col = "EnvTP",
             title = "mq × sp — within environment × timepoint",
             base_size = 9)
```


## **8. sp × sp correlations**

*sp × pc correlations are excluded by design.*

### 8.1 Pooled

```{r spsp-pooled, fig.width=8, fig.height=7}
cors_spsp_pooled <- pair_cors(sp_avg, sp_traits, sp_traits)

plot_heatmap(cors_spsp_pooled, title = "sp × sp — pooled (sp averaged over timepoints)")
```

### 8.2 Within environment

```{r spsp-within-env, fig.width=14, fig.height=7}
sp_avg_env <- sp_avg %>% dplyr::select(plot_id, Env, all_of(sp_traits))

cors_spsp_env <- group_cors(sp_avg_env, sp_traits, sp_traits, "Env") %>%
  mutate(Env = factor(Env, levels = env_order))

plot_heatmap(cors_spsp_env, facet_col = "Env",
             title = "sp × sp — within environment (sp averaged over timepoints)")
```

### 8.3 Within timepoint (pooled over environments)

```{r spsp-within-tp, fig.width=8, fig.height=20}
cors_spsp_tp <- group_cors(sp_full, sp_traits, sp_traits, tp_col) %>%
  rename(Timepoint = all_of(tp_col))

plot_heatmap(cors_spsp_tp, facet_col = "Timepoint",
             title = "sp × sp — within timepoint (pooled over environments)")
```

### 8.4 Within environment × timepoint

```{r spsp-env-tp, fig.width=14, fig.height=28}
cors_spsp_env_tp <- group_cors(sp_full, sp_traits, sp_traits, c("Env", tp_col)) %>%
  rename(Timepoint = all_of(tp_col)) %>%
  mutate(Env = factor(Env, levels = env_order),
         EnvTP = paste0(Env, " | ", Timepoint))

plot_heatmap(cors_spsp_env_tp, facet_col = "EnvTP",
             title = "sp × sp — within environment × timepoint",
             base_size = 9)
```


## **9. Combined correlation dataset**

Stack every pairwise result into one tidy data frame with metadata columns
(`comparison`, `stratification`, `type_x/y`, `VI_x/y`, `PC_x/y`).

```{r combine-all}
# ==============================================================================
# FIX: Convert all Timepoint columns to character before binding
# ==============================================================================

cors_all <- bind_rows(
  # ag × mq
  cors_agmq_pooled %>% mutate(comparison = "ag×mq", stratification = "pooled",
                               Env = NA_character_, Timepoint = NA_character_),
  cors_agmq_env    %>% mutate(comparison = "ag×mq", stratification = "env",
                               Timepoint = NA_character_),
  # ag × pc
  cors_agpc_pooled %>% mutate(comparison = "ag×pc", stratification = "pooled",
                               Env = NA_character_, Timepoint = NA_character_),
  cors_agpc_env    %>% mutate(comparison = "ag×pc", stratification = "env",
                               Timepoint = NA_character_),
  # ag × sp
  cors_agsp_pooled %>% mutate(comparison = "ag×sp", stratification = "pooled",
                               Env = NA_character_, Timepoint = NA_character_),
  cors_agsp_env    %>% mutate(comparison = "ag×sp", stratification = "env",
                               Timepoint = NA_character_),
  cors_agsp_tp     %>% mutate(comparison = "ag×sp", stratification = "timepoint",
                               Env = NA_character_,
                               Timepoint = as.character(Timepoint)),  # Convert to character
  cors_agsp_env_tp %>% mutate(comparison = "ag×sp", stratification = "env×timepoint",
                               Timepoint = as.character(Timepoint)),  # Convert to character
  # mq × pc
  cors_mqpc_pooled %>% mutate(comparison = "mq×pc", stratification = "pooled",
                               Env = NA_character_, Timepoint = NA_character_),
  cors_mqpc_env    %>% mutate(comparison = "mq×pc", stratification = "env",
                               Timepoint = NA_character_),
  # mq × sp
  cors_mqsp_pooled %>% mutate(comparison = "mq×sp", stratification = "pooled",
                               Env = NA_character_, Timepoint = NA_character_),
  cors_mqsp_env    %>% mutate(comparison = "mq×sp", stratification = "env",
                               Timepoint = NA_character_),
  cors_mqsp_tp     %>% mutate(comparison = "mq×sp", stratification = "timepoint",
                               Env = NA_character_,
                               Timepoint = as.character(Timepoint)),  # Convert to character
  cors_mqsp_env_tp %>% mutate(comparison = "mq×sp", stratification = "env×timepoint",
                               Timepoint = as.character(Timepoint)),  # Convert to character
  # sp × sp
  cors_spsp_pooled %>% mutate(comparison = "sp×sp", stratification = "pooled",
                               Env = NA_character_, Timepoint = NA_character_),
  cors_spsp_env    %>% mutate(comparison = "sp×sp", stratification = "env",
                               Timepoint = NA_character_),
  cors_spsp_tp     %>% mutate(comparison = "sp×sp", stratification = "timepoint",
                               Env = NA_character_,
                               Timepoint = as.character(Timepoint)),  # Convert to character
  cors_spsp_env_tp %>% mutate(comparison = "sp×sp", stratification = "env×timepoint",
                               Timepoint = as.character(Timepoint))   # Convert to character
) %>%
  mutate(
    Env            = as.character(Env),
    abs_r          = abs(r),
    type_x = case_when(
      x_trait %in% ag_traits ~ "ag",
      x_trait %in% mq_traits ~ "mq",
      x_trait %in% pc_traits ~ "pc",
      x_trait %in% sp_traits ~ "sp",
      TRUE                   ~ "other"),
    type_y = case_when(
      y_trait %in% ag_traits ~ "ag",
      y_trait %in% mq_traits ~ "mq",
      y_trait %in% pc_traits ~ "pc",
      y_trait %in% sp_traits ~ "sp",
      TRUE                   ~ "other"),
    # VI: for sp traits use trait name directly; for pc traits parse after "_"
    VI_x = case_when(
      type_x == "sp" ~ x_trait,
      type_x == "pc" ~ sub(".*_", "", x_trait),
      TRUE           ~ NA_character_),
    VI_y = case_when(
      type_y == "sp" ~ y_trait,
      type_y == "pc" ~ sub(".*_", "", y_trait),
      TRUE           ~ NA_character_),
    # PC component label (e.g. FPC1, FPC2, FPC3)
    PC_x = ifelse(type_x == "pc", sub("_.*", "", x_trait), NA_character_),
    PC_y = ifelse(type_y == "pc", sub("_.*", "", y_trait), NA_character_),
    comparison     = factor(comparison,
                            levels = c("ag×mq","ag×pc","ag×sp",
                                       "mq×pc","mq×sp","sp×sp")),
    stratification = factor(stratification,
                            levels = c("pooled","env","timepoint","env×timepoint"))
  ) %>%
  select(-any_of("EnvTP"))

cat("Combined dataset:", nrow(cors_all), "rows ×", ncol(cors_all), "cols\n")
cat("Stratification counts:\n")
print(table(cors_all$comparison, cors_all$stratification, useNA = "ifany"))
```


## **10. Summary statistics**

All pooled-stratification summaries use `filter_strat = "pooled"`.
Tables show mean |r|, mean r, SD(r), max |r|, and pair count.

### 10.1 By ag trait (pooled)

```{r summary-ag-trait}
unique(cors_all$comparison)
cor_summary(
  cors_all %>% 
   filter(comparison != "ag×mq") %>%
    filter(type_x == "ag"),
  group_vars   = c("x_trait", "comparison"),
  filter_strat = "pooled"
) %>%
  mutate(x_trait = short_name(x_trait)) %>%
  knitr::kable(caption = "Mean |r| per agronomic trait — pooled")
```

```{r plot-summary-ag, fig.width=8, fig.height=4}
cor_summary(cors_all %>% filter(type_x == "ag"),
            group_vars = c("x_trait", "comparison"), filter_strat = "pooled") %>%
  mutate(x_label = short_name(x_trait)) %>%
  ggplot(aes(x = fct_reorder(x_label, mean_abs_r), y = mean_abs_r, fill = comparison)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Mean |r| by ag trait (pooled)", x = NULL, y = "Mean |r|") +
  theme_minimal()
```

### 10.2 By mq trait (pooled)

```{r summary-mq-trait}
cor_summary(
  cors_all %>% filter(type_x == "mq"),
  group_vars   = c("x_trait", "comparison"),
  filter_strat = "pooled"
) %>%
  mutate(x_trait = short_name(x_trait)) %>%
  knitr::kable(caption = "Mean |r| per malt quality trait — pooled")
```

```{r plot-summary-mq, fig.width=9, fig.height=5}
cor_summary(cors_all %>% filter(type_x == "mq"),
            group_vars = c("x_trait", "comparison"), filter_strat = "pooled") %>%
  mutate(x_label = short_name(x_trait)) %>%
  ggplot(aes(x = fct_reorder(x_label, mean_abs_r), y = mean_abs_r, fill = comparison)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Mean |r| by mq trait (pooled)", x = NULL, y = "Mean |r|") +
  theme_minimal()
```

### 10.3 By VI — raw spectral traits (pooled)

```{r summary-by-vi-sp}
cor_summary(
  cors_all %>% filter(type_x == "sp"),
  group_vars   = c("VI_x", "comparison"),
  filter_strat = "pooled"
) %>%
  rename(VI = VI_x) %>%
  knitr::kable(caption = "Mean |r| per spectral VI (sp) — pooled")
```

```{r plot-summary-vi-sp, fig.width=8, fig.height=4}
cor_summary(cors_all %>% filter(type_x == "sp"),
            group_vars = c("VI_x", "comparison"), filter_strat = "pooled") %>%
  ggplot(aes(x = fct_reorder(VI_x, mean_abs_r), y = mean_abs_r, fill = comparison)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Mean |r| by VI — sp traits (pooled)", x = NULL, y = "Mean |r|") +
  theme_minimal()
```

### 10.4 By VI — functional-PCA traits (pooled)

```{r summary-by-vi-pc}
cor_summary(
  cors_all %>% filter(type_x == "pc"),
  group_vars   = c("VI_x", "comparison"),
  filter_strat = "pooled"
) %>%
  rename(VI = VI_x) %>%
  knitr::kable(caption = "Mean |r| by VI within pc traits — pooled")
```

### 10.5 By PC component (pooled)

```{r summary-by-pc}
cor_summary(
  cors_all %>% filter(type_x == "pc"),
  group_vars   = c("PC_x", "VI_x", "comparison"),
  filter_strat = "pooled"
) %>%
  rename(PC = PC_x, VI = VI_x) %>%
  knitr::kable(caption = "Mean |r| by PC component and VI — pooled")
```

```{r plot-summary-pc, fig.width=8, fig.height=4}
cor_summary(cors_all %>% filter(type_x == "pc"),
            group_vars = c("PC_x", "VI_x", "comparison"), filter_strat = "pooled") %>%
  mutate(pc_label = paste0(PC_x, " (", VI_x, ")")) %>%
  ggplot(aes(x = fct_reorder(pc_label, mean_abs_r), y = mean_abs_r, fill = comparison)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Mean |r| by PC component — pooled", x = NULL, y = "Mean |r|") +
  theme_minimal()
```

### 10.6 By environment

```{r summary-by-env}
cor_summary(
  cors_all %>% filter(!is.na(Env)),
  group_vars   = c("Env", "comparison"),
  filter_strat = "env"
) %>%
  mutate(Env = factor(Env, levels = env_order)) %>%
  arrange(comparison, Env) %>%
  knitr::kable(caption = "Mean |r| by environment (env stratification)")
```

```{r plot-summary-env, fig.width=10, fig.height=5}
cor_summary(cors_all %>% filter(!is.na(Env)),
            group_vars = c("Env", "comparison"), filter_strat = "env") %>%
  mutate(Env = factor(Env, levels = env_order)) %>%
  ggplot(aes(x = Env, y = mean_abs_r, colour = comparison, group = comparison)) +
  geom_line() +
  geom_point(size = 2) +
  scale_colour_brewer(palette = "Dark2") +
  labs(title = "Mean |r| across environments", x = NULL, y = "Mean |r|") +
  theme_minimal() +
  theme(legend.position = "right")
```

### 10.7 Top correlations (pooled, |r| ≥ 0.30, excluding sp×sp)

```{r top-cors}
cors_all %>%
  filter(stratification == "pooled",
         !comparison %in% c("sp×sp"),
         abs_r >= 0.30) %>%
  mutate(x_label = short_name(x_trait),
         y_label = short_name(y_trait)) %>%
  select(comparison, x_label, y_label, r, n) %>%
  arrange(desc(abs(r))) %>%
  head(30) %>%
  knitr::kable(digits = 3,
               caption = "Top 30 pooled correlations |r| ≥ 0.30 (excluding sp×sp)")
```



#VI
```{r}

i<- cor_summary(
  cors_all %>% 
    filter(comparison %in% c( "ag×sp","mq×sp")) %>%
    filter(type_x %in% c("ag", "mq")),
  group_vars   = c("x_trait", "comparison"),
  filter_strat = "pooled"
) %>%
  mutate(x_trait = short_name(x_trait)) %>%
  arrange(comparison, mean_abs_r) %>% 
  knitr::kable(caption = "Mean |r| per agronomic trait — pooled")

mean(i$mean_abs_r)
```


PC
```{r}

i<- cor_summary(
  cors_all %>% 
    filter(comparison %in% c( "ag×pc","mq×pc")) %>%
    filter(type_x %in% c("ag", "mq")),
  group_vars   = c("Env", "VI_y") 
) %>%
  #mutate(x_trait = short_name(x_trait)) %>%
  arrange(VI_y) %>% 
  knitr::kable(caption = "Mean |r| per agronomic trait — pooled")


#Trait Corr
#SNY22  0.61
#MCG23  0.29
#KET21  0.24
#MCG25  0.23
#HELF24 0.12
```

WS
```{r}
data_ag %>%
  group_by(Env) %>%
  summarise(
    n        = sum(!is.na(winter_survival)),
    mean_WS  = mean(winter_survival, na.rm = TRUE),
    sd_WS    = sd(winter_survival, na.rm = TRUE),
    min_WS   = min(winter_survival, na.rm = TRUE),
    max_WS   = max(winter_survival, na.rm = TRUE),
    se_WS    = sd_WS / sqrt(n),
    .groups  = "drop"
  )


```

#WS in different Env-s. 
#MCG23 77%
#SNY22 79.4%
#MCG25 81.9%
#KET21 96.3%
#HELF24 97.8%

#mean_abs_r of VIs with all traits within Envs
#SNY22  NDVI      0.594 
#MCG23  NDVI      0.284 
#KET21  NDVI      0.242 
#MCG25  NDVI      0.233 
#HELF24 NDVI      0.115 

#SNY22  PSRI      0.627 
#MCG23  PSRI      0.292 
#KET21  PSRI      0.235 
#MCG25  PSRI      0.215 
#HELF24 PSRI      0.117 



Table: Mean |r| per agronomic trait — pooled

|Env    |VI_y | | mean_abs_r|  
|:------|:----|------:|----------:|
|SNY22  |NDVI | |      0.426|
|MCG23  |NDVI | |      0.169| 
|SNY22  |PSRI | |      0.393| 
|MCG23  |PSRI | |      0.206| 


|KET21  |NDVI | |      0.206| 
|MCG25  |NDVI | |      0.194| 

|HELF24 |NDVI | |      0.125| 
|KET21  |PSRI | |      0.202| 
|MCG25  |PSRI | |      0.196|
|NA     |PSRI | |      0.146| 
|HELF24 |PSRI | |      0.144| 



