---
title: "Phenotypic Correlations"
author: "sss322"
date: "2026-02-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## **Introduction**
Pairwise Pearson correlations between agronomic (ag), malt quality (mq), functional-PCA spectral (pc), and raw spectral VI (sp) traits. Correlations are computed:

- **Pooled** — all environments combined
- **Within environment** — separately per `Env`
- **Within timepoint** — separately per flight timepoint (sp traits only)
- **Within environment × timepoint** — finest stratification for sp traits

sp × pc correlations are **excluded** by design.


## **0. Packages and data**

```{r packages, message=FALSE}
pkgs <- c("dplyr", "tidyr", "ggplot2", "forcats", "scales", "purrr")
suppressPackageStartupMessages(invisible(lapply(pkgs, require, character.only = TRUE)))
Sys.setenv(LANG = "en")
```

```{r load-data}
paths <- c("C:/Users/Siim Sepp/NY_WMBCL/data",
           "~/Documents/GitHub/NY_WMBCL/data",
           "data")

load_first <- function(fname, dirs = paths) {
  for (d in dirs) {
    p <- file.path(path.expand(d), fname)
    if (file.exists(p)) { load(p, envir = .GlobalEnv); return(invisible(p)) }
  }
  stop("File not found: ", fname)
}

invisible(lapply(
  c("data_ag.Rdata", "data_mq.Rdata", "data_sp.Rdata", "F_PCA_scores.Rdata"),
  load_first
))
```


## **1. Define traits and prepare merged datasets**

```{r define-traits}
# ---- Agronomic traits (cols 7:9 of data_ag) ----
ag_traits <- colnames(data_ag)[7:9]

# ---- Malt quality traits (cols 4:15 of data_mq) ----
mq_traits <- colnames(data_mq)[4:15]

# ---- Spectral VI traits (cols 5:13 of data_sp) ----
sp_traits <- names(data_sp)[c(5,13)]

# ---- Detect timepoint column in data_sp ----
# (the non-id, non-Env column among cols 1:4)
sp_id_cols <- colnames(data_sp)[1:4]
tp_col <- setdiff(sp_id_cols, c("plot_id", "GID", "Env", "gid", "genotype"))
if (length(tp_col) == 0) {
  # fallback: the 4th column
  tp_col <- colnames(data_sp)[4]
}
tp_col <- tp_col[1]
message("Timepoint column detected: ", tp_col)

# ---- Functional PCA scores (pc traits) ----
data_pc <- FPCA_scores_df %>%
  filter(VI %in% c("NDVI", "PSRI"))

# Identify score columns (not plot_id / VI)
score_cols <- setdiff(colnames(data_pc), c("plot_id", "VI"))
data_pc <- data_pc %>%
  pivot_wider(id_cols = "plot_id",
              names_from  = "VI",
              values_from = all_of(score_cols))

pc_traits <- c("FPC1_NDVI" ,"FPC1_PSRI" ,"FPC2_NDVI" ,"FPC2_PSRI", "FPC3_NDVI", "FPC3_PSRI")

# ---- Convenience: base id columns from each dataset ----
ag_id <- c("plot_id", "GID", "Env")
mq_id <- c("plot_id", "GID", "Env")

# ---- ag base (keep id + ag_traits) ----
ag_base <- data_ag[,c(1:3,7:9)] #%>% select(any_of(ag_id), all_of(ag_traits))

# ---- mq base ----
mq_base <- data_mq[,1:15] #%>% select(any_of(mq_id), all_of(mq_traits))
# ---- sp: one row per plot (average over timepoints) for non-timepoint analyses ----
sp_avg <- data_sp %>%
  group_by(plot_id) %>%
  summarise(across(all_of(sp_traits), \(x) mean(x, na.rm = TRUE)),
            Env = first(Env), .groups = "drop")

# ---- sp: full (per timepoint) dataset for within-timepoint analyses ----
sp_full <- data_sp %>%
  dplyr::select(plot_id, Env, all_of(tp_col), all_of(sp_traits))

# ---- pc base ----
pc_base <- data_pc
```


## **2. Helper functions**

```{r helpers}
# Pearson r with complete-case guard
safe_cor <- function(x, y) {
  ok <- complete.cases(x, y)
  if (sum(ok) < 3) return(NA_real_)
  cor(x[ok], y[ok])
}

# All pairwise correlations between x_traits and y_traits in a data frame
# Returns: data.frame(x_trait, y_trait, r, n)
pair_cors <- function(df, x_traits, y_traits) {
  res <- vector("list", length(x_traits) * length(y_traits))
  k   <- 1L
  for (x in x_traits) {
    for (y in y_traits) {
      ok  <- complete.cases(df[[x]], df[[y]])
      n   <- sum(ok)
      r   <- if (n >= 3) cor(df[[x]][ok], df[[y]][ok]) else NA_real_
      res[[k]] <- data.frame(x_trait = x, y_trait = y, r = r, n = n,
                             stringsAsFactors = FALSE)
      k <- k + 1L
    }
  }
  do.call(rbind, res)
}

# Apply pair_cors within groups defined by group_cols (character vector)
group_cors <- function(df, x_traits, y_traits, group_cols) {
  df %>%
    group_by(across(all_of(group_cols))) %>%
    group_modify(~ pair_cors(.x, x_traits, y_traits)) %>%
    ungroup()
}

# ---- Colour / labelling constants (consistent with project style) ----
env_order <- c("KET21", "SNY22", "MCG23", "HELF24", "MCG25")

trait_map <- c(
  "KernelWeight.mg."    = "KW",
  "on6.64...."          = "on6/64",
  "winter_survival"     = "WS",
  "MaltExtract..."      = "ME",
  "BarleyProtein..."    = "BP",
  "WortProtein..."      = "WP",
  "S.T"                 = "S.T",
  "DP..ASBC."           = "DP",
  "Alpha.amylase.20.DU." = "AA",
  "Beta.glucan.ppm."    = "BG",
  "HD_JD"               = "HD",
  "yield_kgha"          = "GY",
  "FAN.ppm."            = "FAN",
  "AdjunctQualityScore"  = "Adj.Sc",
  "All_MaltQualityScore" = "All.Sc"
)

short_name <- function(x) ifelse(x %in% names(trait_map), trait_map[x], x)

# Summarise a correlation data frame (from cors_all or any tagged df).
# group_vars : character vector of columns to group by
# filter_strat: optional character vector of stratification levels to keep
cor_summary <- function(df, group_vars, filter_strat = NULL) {
  if (!is.null(filter_strat))
    df <- df %>% filter(stratification %in% filter_strat)
  df %>%
    filter(!is.na(r)) %>%
    mutate(.abs_r = abs(r)) %>%
    group_by(across(all_of(group_vars))) %>%
    summarise(
      mean_r     = round(mean(r),      3),
      mean_abs_r = round(mean(.abs_r), 3),
      sd_r       = round(sd(r),        3),
      max_abs_r  = round(max(.abs_r),  3),
      min_abs_r  = round(min(.abs_r),  3),
      range_abs_r  = max_abs_r - min_abs_r,
      n_pairs    = n(),
      .groups    = "drop"
    ) %>%
    arrange(desc(mean_abs_r))
}

# Heatmap: rows = y_trait, cols = x_trait, tile fill = r
# facet_col: optional grouping column for faceting (e.g. "Env", "Timepoint")
plot_heatmap <- function(cors_df, x_col = "x_trait", y_col = "y_trait",
                         r_col = "r", facet_col = NULL,
                         title = "", base_size = 11) {
  df <- cors_df %>%
    mutate(
      x_label = short_name(.data[[x_col]]),
      y_label = short_name(.data[[y_col]])
    )

  p <- ggplot(df, aes(x = x_label, y = y_label, fill = .data[[r_col]])) +
    geom_tile(colour = "grey90") +
    geom_text(aes(label = ifelse(is.finite(.data[[r_col]]),
                                 sprintf("%.2f", .data[[r_col]]), "")),
              size = 6) +
    scale_fill_gradient2(
      low = "#2c7bb6", mid = "white", high = "#d7191c",
      midpoint = 0, limits = c(-1, 1), name = "r",
      na.value = "grey85"
    ) +
    labs(title = title, x = NULL, y = NULL) +
    theme_minimal(base_size = base_size) +
    theme(
      panel.grid    = element_blank(),
      axis.text   = element_text(size = 18),
      plot.title    = element_text(hjust = 0.5, size = base_size + 1)
    )

  if (!is.null(facet_col)) {
    p <- p + facet_wrap(reformulate(facet_col), scales = "free")
  }
  p
}
```


## **3. ag × mq correlations**

### 3.1 Pooled (over all environments)

```{r agmq-pooled, fig.width=9, fig.height=5}
agmq <- ag_base %>%
  inner_join(mq_base %>% dplyr::select(plot_id, all_of(mq_traits)), by = "plot_id")

cors_agmq_pooled <- pair_cors(agmq, ag_traits, mq_traits)


cors_agmq_pooled <- cors_agmq_pooled %>%
  mutate(x_trait = recode(x_trait, !!!trait_map)) %>%
  mutate(y_trait = recode(y_trait, !!!trait_map)) 

plot_heatmap(cors_agmq_pooled, title = "ag × mq — pooled over environments")


corm<- as.matrix(cor(agmq[,4:18]))

library(tibble)
corm <- corm %>%
  as.data.frame() %>%
  rownames_to_column("Trait") %>%
  mutate(Trait = recode(Trait, !!!trait_map)) %>%
  column_to_rownames("Trait")

colnames(corm) <- recode(colnames(corm), !!!trait_map)
```

### 3.2 Within environment

```{r agmq-within-env, fig.width=14, fig.height=5}
cors_agmq_env <- group_cors(agmq, ag_traits, mq_traits, "Env") %>%
  mutate(Env = factor(Env, levels = env_order))

plot_heatmap(cors_agmq_env, facet_col = "Env",
             title = "ag × mq — within environment")


```

### 3.3 Clustered heatmap (pooled)

Hierarchical clustering (complete linkage, Euclidean distance) applied to both
ag rows and mq columns.

```{r agmq-clustered, fig.width=8, fig.height=5}
library(pheatmap)

# cors_agmq_pooled already has short names from the recode above
cor_mat <- cors_agmq_pooled %>%
  dplyr::select(x_trait, y_trait, r) %>%
  tidyr::pivot_wider(names_from = y_trait, values_from = r) %>%
  tibble::column_to_rownames("x_trait") %>%
  as.matrix()



pheatmap(corm,
         color            = colorRampPalette(c("#2c7bb6", "white", "#d7191c"))(100),
         breaks           = seq(-1, 1, length.out = 101),
         cluster_rows     = TRUE,
         cluster_cols     = TRUE,
         clustering_method = "complete",
         display_numbers  = TRUE,
         number_format    = "%.2f",
         number_color     = "black",
         fontsize_number  = 13,
         fontsize_row     = 14,
         fontsize_col     = 14,
         border_color     = "grey90",
         treeheight_row   = 35,
         treeheight_col   = 35,
         na_col           = "grey85")


```



## **4. pc correlations**

### 4.1 Pooled
```{r agpc-pooled, fig.width=10, fig.height=6}
agpc <- ag_base %>%
  inner_join(pc_base, by = "plot_id")

pcs<- agpc %>% left_join(data_mq[,c(-1,-3)], by= "plot_id")
mqagtraits<- c(ag_traits, mq_traits)

cors_pc_pooled <- pair_cors(pcs, mqagtraits, pc_traits)

plot_heatmap(
  cors_pc_pooled %>% mutate(x_trait = short_name(x_trait)),
  title = "(ag+mq) \u00d7 pc \u2014 pooled"
)

cors_pc_pooled <- cors_pc_pooled %>%
  separate(y_trait, into = c("FPC", "VI"), sep = "_", remove = FALSE) %>% arrange(FPC) %>%   mutate(x_trait = recode(x_trait, !!!trait_map)) 





p <- ggplot(cors_pc_pooled, aes(y = x_trait, x = FPC, fill = r)) +
  geom_tile(color = "grey90", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.2f", r)), 
            size = 4, 
            fontface = "bold") +
  scale_color_manual(values = c("black", "white"), guide = "none") +
  facet_wrap(~ VI, nrow = 1) +
  scale_fill_gradientn(
    colors = colorRampPalette(c("#2c7bb6", "white", "#d7191c"))(100),
    limits = c(-1, 1), 
    name = "r",
    na.value = "grey85"
  ) +
  scale_x_discrete(labels = function(x) str_wrap(short_name(x), width = 10)) +
  labs(
      x = "Trait",
    y = "FPC"
  ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1, size= 14),
    axis.text.y =element_text(size= 14), 
    strip.text = element_text(face = "bold", size = 16),
    panel.grid = element_blank(),
    axis.title = element_text(size= 14),
    panel.border = element_rect(fill = NA),
    legend.position = "bottom",
    legend.text = element_text(size = 14, face="bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

p
```


### 4.2 Within environment
```{r agpc-within-env, fig.width=16, fig.height=6}
cors_pc_env <- group_cors(pcs, mqagtraits, pc_traits, "Env") %>%
  mutate(Env = factor(Env, levels = env_order))

plot_heatmap(
  cors_pc_env %>% mutate(x_trait = short_name(x_trait)),
  facet_col = "Env",
  title = "(ag+mq) \u00d7 pc \u2014 within environment"
)



cors_pc_env <- cors_pc_env %>%
  separate(y_trait, into = c("FPC", "VI"), sep = "_", remove = FALSE) %>% arrange(FPC) #%>%   mutate(x_trait = recode(x_trait, !!!trait_map)) 



crpc1 <- cors_pc_env %>% filter(x_trait %in% ag_traits) %>%   mutate(x_trait = recode(x_trait, !!!trait_map)) 

p <- ggplot(crpc1, aes(y = x_trait, x = FPC, fill = r)) +
  geom_tile(color = "grey90", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.2f", r)), 
            size = 4, 
            fontface = "bold") +
  scale_color_manual(values = c("black", "white"), guide = "none") +
  facet_grid(VI ~ Env) +
  scale_fill_gradientn(
    colors = colorRampPalette(c("#2c7bb6", "white", "#d7191c"))(100),
    limits = c(-1, 1), 
    name = "r",
    na.value = "grey85"
  ) +
  scale_x_discrete(labels = function(x) str_wrap(short_name(x), width = 10)) +
  labs(
      x = "Trait",
    y = "FPC"
  ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1, size= 14),
    axis.text.y =element_text(size= 14), 
    strip.text = element_text(face = "bold", size = 16),
    panel.grid = element_blank(),
    axis.title = element_text(size= 14),
    panel.border = element_rect(fill = NA),
    legend.position = "bottom",
    legend.text = element_text(size = 14, face="bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

p

crpc1 <- cors_pc_env %>% filter(x_trait %in% mq_traits) %>%   mutate(x_trait = recode(x_trait, !!!trait_map))  %>% filter(Env %in% data_mq$Env)

p <- ggplot(crpc1, aes(y = x_trait, x = FPC, fill = r)) +
  geom_tile(color = "grey90", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.2f", r)), 
            size = 4, 
            fontface = "bold") +
  scale_color_manual(values = c("black", "white"), guide = "none") +
  facet_grid(VI ~ Env) +
  scale_fill_gradientn(
    colors = colorRampPalette(c("#2c7bb6", "white", "#d7191c"))(100),
    limits = c(-1, 1), 
    name = "r",
    na.value = "grey85"
  ) +
  scale_x_discrete(labels = function(x) str_wrap(short_name(x), width = 10)) +
  labs(
      x = "Trait",
    y = "FPC"
  ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1, size= 14),
    axis.text.y =element_text(size= 14), 
    strip.text = element_text(face = "bold", size = 16),
    panel.grid = element_blank(),
    axis.title = element_text(size= 14),
    panel.border = element_rect(fill = NA),
    legend.position = "bottom",
    legend.text = element_text(size = 14, face="bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

p
```



## **5. sp correlations**

### 5.1 Pooled

```{r agsp-pooled, fig.width=7, fig.height=6}
agsp <- ag_base %>%
  inner_join(sp_avg %>% dplyr::select(plot_id, all_of(sp_traits)), by = "plot_id")

sps<- agsp %>% left_join(data_mq[,c(-1,-3)], by= "plot_id")

cors_sp_pooled <- pair_cors(sps, mqagtraits, sp_traits)

plot_heatmap(
  cors_sp_pooled %>% mutate(x_trait = short_name(x_trait)),
  title = "(ag+mq) \u00d7 sp \u2014 pooled"
)
```

### 5.2 Within environment

```{r agsp-within-env, fig.width=16, fig.height=6}
cors_sp_env <- group_cors(sps, mqagtraits, sp_traits, "Env") %>%
  mutate(Env = factor(Env, levels = env_order))

plot_heatmap(
  cors_sp_env %>% mutate(x_trait = short_name(x_trait)),
  facet_col = "Env",
  title = "(ag+mq) \u00d7 sp \u2014 within environment"
)
```


### 5.3 Within environment × timepoint

```{r agsp-env-tp, fig.width=16, fig.height=14}
agsp_tp<- sps[,-7:-8] %>% inner_join(sp_full %>% dplyr::select(plot_id, all_of(tp_col), all_of(sp_traits)),
             by = "plot_id")

cors_sp_env_tp <- group_cors(agsp_tp, mqagtraits, sp_traits, c("Env", tp_col)) %>%
  rename(Timepoint = all_of(tp_col)) %>%
  mutate(Env = factor(Env, levels = env_order),
         EnvTP = paste0(Env, " | ", Timepoint))

plot_heatmap(
  cors_sp_env_tp %>% mutate(x_trait = short_name(x_trait)),
  facet_col = "EnvTP",
  title = "(ag+mq) \u00d7 sp \u2014 within environment \u00d7 timepoint"
)


crsp1 <- cors_sp_env_tp %>% filter(x_trait %in% mq_traits) %>%   mutate(x_trait = recode(x_trait, !!!trait_map))  %>% filter(Env %in% data_mq$Env)

p <- ggplot(crsp1, aes(y = x_trait, x = Timepoint, fill = r)) +
  geom_tile(color = "grey90", linewidth = 0.5) +
  #geom_text(aes(label = sprintf("%.2f", r)), 
   #         size = 4, 
    #        fontface = "bold") +
  scale_color_manual(values = c("black", "white"), guide = "none") +
  facet_grid(y_trait ~ Env, scales= "free") +
  scale_fill_gradientn(
    colors = colorRampPalette(c("#2c7bb6", "white", "#d7191c"))(100),
    limits = c(-1, 1), 
    name = "r",
    na.value = "grey85"
  ) +
  scale_x_discrete(labels = function(x) str_wrap(short_name(x), width = 10)) +
  labs(
      x = "Time",
    y = "Trait"
  ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1, size= 14),
    axis.text.y =element_text(size= 14), 
    strip.text = element_text(face = "bold", size = 16),
    panel.grid = element_blank(),
    axis.title = element_text(size= 14),
    panel.border = element_rect(fill = NA),
    legend.position = "bottom",
    legend.text = element_text(size = 14, face="bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

p

load("JD_map.Rdata")


colnames(JD_map)[6] <- "Env"
crsp1 <- cors_sp_env_tp %>% filter(x_trait %in% ag_traits) %>%   mutate(x_trait = recode(x_trait, !!!trait_map))%>% left_join(JD_map[,-4], by= c("Env", "Timepoint")) %>% mutate(Env = recode(Env, !!!env_order))

library(dplyr)

crsp1$Month <- format(as.Date(crsp1$Date), "%b")  # "Mar", "Apr", etc.


crsp1$Month <- cut(
  crsp1$DAS,
  breaks = c(0, 120, 150, 180, 210, 240),  # adjust to your season
  labels = c("Mar", "Apr", "May", "Jun", "Jul"),
  include.lowest = TRUE
)


library(dplyr)

crsp1 <- crsp1 %>%
  group_by(Env, y_trait, x_trait, Month) %>%
  arrange(DAS) %>%
  mutate(Month_id = paste0(Month, "_", row_number())) %>%
  ungroup()


p <- ggplot(crsp1, aes(y = x_trait, x = Month_id, fill = r)) +
  geom_tile(color = "grey85", linewidth = 0.3) +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3) +
  
  facet_grid(y_trait ~ Env, scales = "free_x", space = "free_x") +
  
  scale_fill_gradient2(
    low = "#2c7bb6", mid = "white", high = "#d7191c",
    midpoint = 0, limits = c(-1, 1), name = "r"
  ) +
  
  scale_x_discrete(
    labels = function(x) sub("_.*", "", x)  # show only month
  ) +
  theme(
    axis.text.x = element_text(angle = 0, size = 12),
    strip.text = element_text(face = "bold"),
    panel.grid = element_blank(),
    legend.position = "bottom"
  )

p
```


## **9. Combined correlation dataset**

Stack every pairwise result into one tidy data frame with metadata columns
(`comparison`, `stratification`, `type_x/y`, `VI_x/y`, `PC_x/y`).

```{r combine-all}
# ── classify x_trait as "ag" or "mq" (uses original trait names) ─────────────
classify_agmq <- function(v) {
  dplyr::case_when(
    v %in% ag_traits ~ "ag",
    v %in% mq_traits ~ "mq",
    TRUE             ~ NA_character_
  )
}

extract_VI_name <- function(v) {
  dplyr::case_when(
    grepl("NDVI", v, ignore.case = TRUE) ~ "NDVI",
    grepl("NDRE", v, ignore.case = TRUE) ~ "NDRE",
    grepl("PSRI", v, ignore.case = TRUE) ~ "PSRI",
    TRUE ~ NA_character_
  )
}

# Ensure Env / Timepoint columns exist before binding
add_std_cols <- function(df) {
  if (!"Env"       %in% colnames(df)) df$Env       <- NA_character_
  if (!"Timepoint" %in% colnames(df)) df$Timepoint <- NA_character_
  df %>% mutate(abs_r     = abs(r),
                Env       = as.character(Env),
                Timepoint = as.character(Timepoint))
}

cors_all <- bind_rows(

  # ---- ag × mq (traits already recoded to short names in section 3.1) -------
  cors_agmq_pooled %>%
    add_std_cols() %>%
    mutate(stratification = "pooled", type_x = "ag", type_y = "mq",
           comparison = "ag\u00d7mq",
           VI_x = NA_character_, VI_y = NA_character_,
           PC_x = NA_character_, PC_y = NA_character_),

  cors_agmq_env %>%
    add_std_cols() %>%
    mutate(stratification = "env", type_x = "ag", type_y = "mq",
           comparison = "ag\u00d7mq",
           VI_x = NA_character_, VI_y = NA_character_,
           PC_x = NA_character_, PC_y = NA_character_),

  # ---- (ag+mq) × pc ----------------------------------------------------------
  cors_pc_pooled %>%
    add_std_cols() %>%
    mutate(stratification = "pooled",
           type_x     = classify_agmq(x_trait),
           type_y     = "pc",
           comparison = paste0(type_x, "\u00d7pc"),
           VI_x       = NA_character_,
           VI_y       = extract_VI_name(y_trait),
           PC_x       = NA_character_,
           PC_y       = sub("_(NDVI|NDRE|PSRI)$", "", y_trait)),

  cors_pc_env %>%
    add_std_cols() %>%
    mutate(stratification = "env",
           type_x     = classify_agmq(x_trait),
           type_y     = "pc",
           comparison = paste0(type_x, "\u00d7pc"),
           VI_x       = NA_character_,
           VI_y       = extract_VI_name(y_trait),
           PC_x       = NA_character_,
           PC_y       = sub("_(NDVI|NDRE|PSRI)$", "", y_trait)),

  # ---- (ag+mq) × sp ----------------------------------------------------------
  cors_sp_pooled %>%
    add_std_cols() %>%
    mutate(stratification = "pooled",
           type_x     = classify_agmq(x_trait),
           type_y     = "sp",
           comparison = paste0(type_x, "\u00d7sp"),
           VI_x       = NA_character_,
           VI_y       = extract_VI_name(y_trait),
           PC_x       = NA_character_,
           PC_y       = NA_character_),

  cors_sp_env %>%
    add_std_cols() %>%
    mutate(stratification = "env",
           type_x     = classify_agmq(x_trait),
           type_y     = "sp",
           comparison = paste0(type_x, "\u00d7sp"),
           VI_x       = NA_character_,
           VI_y       = extract_VI_name(y_trait),
           PC_x       = NA_character_,
           PC_y       = NA_character_),

  cors_sp_env_tp %>%
    add_std_cols() %>%
    mutate(stratification = "env_tp",
           type_x     = classify_agmq(x_trait),
           type_y     = "sp",
           comparison = paste0(type_x, "\u00d7sp"),
           VI_x       = NA_character_,
           VI_y       = extract_VI_name(y_trait),
           PC_x       = NA_character_,
           PC_y       = NA_character_)

) %>%
  mutate(Env = factor(Env, levels = env_order))

cat("cors_all:", nrow(cors_all), "rows\n")
cat("comparisons:", paste(sort(unique(cors_all$comparison)), collapse = ", "), "\n")
cat("stratifications:", paste(sort(unique(cors_all$stratification)), collapse = ", "), "\n")
```


## **10. Summary statistics**

All pooled-stratification summaries use `filter_strat = "pooled"`.
Tables show mean |r|, mean r, SD(r), max |r|, and pair count.




### 10.1 ag and mq  with VI

```{r summary-ag-trait}

cor_summary(
  cors_all %>% 
   filter(comparison %in% c("ag×sp", "mq×sp")),
  group_vars   = c("x_trait","y_trait", "comparison"),
  filter_strat = "pooled") %>%
  mutate(x_trait = short_name(x_trait)) %>% 
  arrange(mean_abs_r)%>%
  knitr::kable(caption = "Mean |r| per agronomic trait — pooled") 


cor_summary(
  cors_all %>% 
   filter(comparison %in% c("ag×sp", "mq×sp")),
  group_vars   = c("x_trait"),
  filter_strat = "env") %>%
  mutate(x_trait = short_name(x_trait)) %>% 
  arrange(mean_abs_r)%>%
  knitr::kable(caption = "Mean |r| per agronomic trait — pooled") 
```


```{r plot-summary-vi-sp, fig.width=8, fig.height=4}

#WS
ag_r<- cors_all %>% 
  filter(x_trait %in% c("winter_survival")) %>% 
  filter(comparison %in% c("ag×sp")) %>% 
  filter(stratification != "env_tp") %>% 
  group_by(Env) %>%
  summarise(
    n = n(),
    mean_abs_r = mean(abs_r, na.rm = TRUE),
    median_abs_r = median(abs_r, na.rm = TRUE),
    min_abs_r = min(abs_r, na.rm = TRUE),
    max_abs_r = max(abs_r, na.rm = TRUE),
    sd_abs_r = sd(abs_r, na.rm = TRUE),
    .groups = 'drop'
  ) 

ws<- data_ag %>%
  group_by(Env) %>%
  summarise(
    n        = sum(!is.na(winter_survival)),
    mean_WS  = mean(winter_survival, na.rm = TRUE),
    sd_WS    = sd(winter_survival, na.rm = TRUE),
    min_WS   = min(winter_survival, na.rm = TRUE),
    max_WS   = max(winter_survival, na.rm = TRUE),
    se_WS    = sd_WS / sqrt(n),
    .groups  = "drop"
  )

ws_r<- ws[1:3] %>% left_join(ag_r[,c(1,3)], by= "Env")

cor(ws_r$mean_WS, ws_r$mean_abs_r)



# Summary statistics by EnvTP
p<- cors_all %>% 
  filter(comparison %in% c("ag×sp", "mq×sp")) %>% 
  filter(stratification == "env_tp") %>% 
  group_by(EnvTP) %>%
  summarise(
    n = n(),
    mean_abs_r = mean(abs_r, na.rm = TRUE),
    median_abs_r = median(abs_r, na.rm = TRUE),
    min_abs_r = min(abs_r, na.rm = TRUE),
    max_abs_r = max(abs_r, na.rm = TRUE),
    sd_abs_r = sd(abs_r, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  arrange(EnvTP)

print(p, n= 41)

#ag/mq corr with sp. 
cor_summary(
  cors_all %>% filter(comparison %in% c("ag×sp", "mq×sp")),
  group_vars   = c( "comparison"),
  filter_strat = "pooled"
) %>%
  knitr::kable(caption = "Mean |r| by VI within pc traits — pooled")


#ag/mq corr with sp over env.  
cor_summary(
  cors_all %>% filter(comparison %in% c("ag×sp", "mq×sp")),
  group_vars   = c( "comparison"),
  filter_strat = "env"
) %>%
  knitr::kable(caption = "Mean |r| by VI within pc traits — pooled")

#ag/mq corr with pc. 
cor_summary(
  cors_all %>% filter(comparison %in% c("ag×sp", "mq×sp")),
  group_vars   = c("VI_x", "comparison"),
  filter_strat = "pooled"
) %>%
  rename(VI = VI_x) %>%
  knitr::kable(caption = "Mean |r| by VI within pc traits — pooled")

```




### 10.4 By VI — functional-PCA traits (pooled)

```{r summary-by-vi-pc}

#fpc1
cor_summary(
  cors_all %>% filter(comparison %in% c("ag×pc", "mq×pc")) %>% 
  filter(y_trait %in% c("FPC1_NDVI", "FPC1_PSRI")),
  group_vars   = c("x_trait"),
  filter_strat = "env") %>% 
  arrange(mean_abs_r) %>% 
  knitr::kable(caption = "Mean |r| by VI within pc traits — pooled")
#fpc2
cor_summary(
  cors_all %>% filter(comparison %in% c("ag×pc", "mq×pc")) %>% 
  filter(y_trait %in% c("FPC2_NDVI", "FPC2_PSRI")),
  group_vars   = c("x_trait"),
  filter_strat = "env") %>% 
  arrange(mean_abs_r) %>% 
  knitr::kable(caption = "Mean |r| by VI within pc traits — pooled")

#fpc3
cor_summary(
  cors_all %>% filter(comparison %in% c("ag×pc", "mq×pc")) %>% 
  filter(y_trait %in% c("FPC3_NDVI", "FPC3_PSRI")),
  group_vars   = c("x_trait"),
  filter_strat = "env") %>% 
  arrange(mean_abs_r) %>% 
  knitr::kable(caption = "Mean |r| by VI within pc traits — pooled")


#compare ag vs mq. 
cor_summary(
  cors_all %>% filter(comparison %in% c("ag×pc", "mq×pc")),
  group_vars   = c("comparison"),
  filter_strat = "env") %>% 
  arrange(mean_abs_r) %>% 
  knitr::kable(caption = "Mean |r| by VI within pc traits — pooled")


#compare env-s
cor_summary(
  cors_all %>% filter(comparison %in% c("ag×pc", "mq×pc")) %>% 
  filter(y_trait %in% c("FPC1_NDVI", "FPC1_PSRI")),
  group_vars   = c("Env"),
  filter_strat = "env") %>% 
  arrange(mean_abs_r) %>% 
  knitr::kable(caption = "Mean |r| by VI within pc traits — pooled")

```

